### 2014-12-12 ebayes_rnaseq_heterosis.tex
copied style files

%  template.tex for Biometrics papers
%
%  This file provides a template for Biometrics authors.  Use this
%  template as the starting point for creating your manuscript document.
%  See the file biomsample.tex for an example of a full-blown manuscript.

%  ALWAYS USE THE referee OPTION WITH PAPERS SUBMITTED TO BIOMETRICS!!!
%  You can see what your paper would look like typeset by removing
%  the referee option.  Because the typeset version will be in two
%  columns, however, some of your equations may be too long. DO NOT
%  use the \longequation option discussed in the user guide!!!  This option
%  is reserved ONLY for equations that are impossible to split across 
%  multiple lines; e.g., a very wide matrix.  Instead, type your equations 
%  so that they stay in one column and are split across several lines, 
%  as are almost all equations in the journal.  Use a recent version of the
%  journal as a guide. 
%  
\documentclass[useAMS,referee]{biom}
%documentclass[useAMS]{biom}
%
%  If your system does not have the AMS fonts version 2.0 installed, then
%  remove the useAMS option.
%
%  useAMS allows you to obtain upright Greek characters.
%  e.g. \umu, \upi etc.  See the section on "Upright Greek characters" in
%  this guide for further information.
%
%  If you are using AMS 2.0 fonts, bold math letters/symbols are available
%  at a larger range of sizes for NFSS release 1 and 2 (using \boldmath or
%  preferably \bmath).
% 
%  Other options are described in the user guide. Here are a few:
% 
%  -  If you use Patrick Daly's natbib  to cross-reference your 
%     bibliography entries, use the usenatbib option
%
%  -  If you use \includegraphics (graphicx package) for importing graphics
%     into your figures, use the usegraphicx option
% 
%  If you wish to typeset the paper in Times font (if you do not have the
%  PostScript Type 1 Computer Modern fonts you will need to do this to get
%  smoother fonts in a PDF file) then uncomment the next line
%  \usepackage{Times}

%%%%% PLACE YOUR OWN MACROS HERE %%%%%

\def\bSig\mathbf{\Sigma}
\newcommand{\VS}{V\&S}
\newcommand{\tr}{\mbox{tr}}

%  The rotating package allows you to have tables displayed in landscape
%  mode.  The rotating package is NOT included in this distribution, but
%  can be obtained from the CTAN archive.  USE OF LANDSCAPE TABLES IS
%  STRONGLY DISCOURAGED -- create landscape tables only as a last resort if
%  you see no other way to display the information.  If you do do this,
%  then you need the following command.

%\usepackage[figuresright]{rotating}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%  Here, place your title and author information.  Note that in 
%  use of the \author command, you create your own footnotes.  Follow
%  the examples below in creating your author and affiliation information.
%  Also consult a recent issue of the journal for examples of formatting.

\title[This is an Example of Recto Running Head]{This is an Example of
an Article Title This is an Example of an Article Title}

%  Here are examples of different configurations of author/affiliation
%  displays.  According to the Biometrics style, in some instances,
%  the convention is to have superscript *, **, etc footnotes to indicate 
%  which of multiple email addresses belong to which author.  In this case,
%  use the \email{ } command to produce the emails in the display.

%  In other cases, such as a single author or two authors from 
%  different institutions, there should be no footnoting.  Here, use
%  the \emailx{ } command instead.

%  The examples below corrspond to almost every possible configuration
%  of authors and may be used as a guide.  For other configurations, consult
%  a recent issue of the the journal.

%  Single author -- USE \emailx{ } here so that no asterisk footnoting
%  for the email address will be produced.

%\author{John Author\emailx{email@address.edu} \\
%Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K.}

%  Two authors from the same institution, with both emails -- use
%  \email{ } here to produce the asterisk footnoting for each email address

%\author{John Author$^{*}$\email{author@address.edu} and
%Kathy Authoress$^{**}$\email{email2@address.edu} \\
%Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K.}

%  Exactly two authors from different institutions, with both emails  
%  USE \emailx{ } here so that no asterisk footnoting for the email address
%  is produced.

\author
{John Author\emailx{author@address.edu} \\
Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K. 
\and
Kathy Author\emailx{anotherauthor@address.edu} \\
Department of Biostatistics, University of North Carolina at Chapel Hill, 
Chapel Hill, North Carolina, U.S.A.}

%  Three or more authors from same institution with all emails displayed
%  and footnoted using asterisks -- use \email{ }

%\author{John Author$^*$\email{author@address.edu}, 
%Jane Author$^{**}$\email{jane@address.edu}, and 
%Dick Author$^{***}$\email{dick@address.edu} \\
%Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K}

%  Three or more authors from same institution with one corresponding email
%  displayed

%\author{John Author$^*$\email{author@address.edu}, 
%Jane Author, and Dick Author \\
%Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K}

%  Three or more authors, with at least two different institutions,
%  more than one email displayed

%\author{John Author$^{1,*}$\email{author@address.edu}, 
%Kathy Author$^{2,**}$\email{anotherauthor@address.edu}, and 
%Wilma Flinstone$^{3,***}$\email{wilma@bedrock.edu} \\
%$^{1}$Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K \\
%$^{2}$Department of Biostatistics, University of North Carolina at 
%Chapel Hill, Chapel Hill, North Carolina, U.S.A. \\
%$^{3}$Department of Geology, University of Bedrock, Bedrock, Kansas, U.S.A.}

%  Three or more authors with at least two different institutions and only
%  one email displayed

%\author{John Author$^{1,*}$\email{author@address.edu}, 
%Wilma Flinstone$^{2}$, and Barney Rubble$^{2}$ \\
%$^{1}$Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K \\
%$^{2}$Department of Geology, University of Bedrock, Bedrock, Kansas, U.S.A.}

\begin{document}

%  This will produce the submission and review information that appears
%  right after the reference section.  Of course, it will be unknown when
%  you submit your paper, so you can either leave this out or put in 
%  sample dates (these will have no effect on the fate of your paper in the
%  review process!)

\date{{\it Received October} 2007. {\it Revised February} 2008.  {\it
Accepted March} 2008.}

%  These options will count the number of pages and provide volume
%  and date information in the upper left hand corner of the top of the 
%  first page as in published papers.  The \pagerange command will only
%  work if you place the command \label{firstpage} near the beginning
%  of the document and \label{lastpage} at the end of the document, as we
%  have done in this template.

%  Again, putting a volume number and date is for your own amusement and
%  has no bearing on what actually happens to your paper!

\pagerange{\pageref{firstpage}--\pageref{lastpage}} 
\volume{64}
\pubyear{2008}
\artmonth{December}

%  The \doi command is where the DOI for your paper would be placed should it
%  be published.  Again, if you make one up and stick it here, it means 
%  nothing!

\doi{10.1111/j.1541-0420.2005.00454.x}

%  This label and the label ``lastpage'' are used by the \pagerange
%  command above to give the page range for the article.  You may have 
%  to process the document twice to get this to match up with what you 
%  expect.  When using the referee option, this will not count the pages
%  with tables and figures.

\label{firstpage}

%  put the summary for your paper here

\begin{abstract}
This is the summary for this paper.
\end{abstract}

%  Please place your key words in alphabetical order, separated
%  by semicolons, with the first letter of the first word capitalized,
%  and a period at the end of the list.
%

\begin{keywords}
A key word; But another key word; Still another key word; Yet another key word.
\end{keywords}

%  As usual, the \maketitle command creates the title and author/affiliations
%  display

\maketitle

%  If you are using the referee option, a new page, numbered page 1, will
%  start after the summary and keywords.  The page numbers thus count the
%  number of pages of your manuscript in the preferred submission style.
%  Remember, ``Normally, regular papers exceeding 25 pages and Reader Reaction 
%  papers exceeding 12 pages in (the preferred style) will be returned to 
%  the authors without review. The page limit includes acknowledgements, 
%  references, and appendices, but not tables and figures. The page count does 
%  not include the title page and abstract. A maximum of six (6) tables or 
%  figures combined is often required.''

%  You may now place the substance of your manuscript here.  Please use
%  the \section, \subsection, etc commands as described in the user guide.
%  Please use \label and \ref commands to cross-reference sections, equations,
%  tables, figures, etc.
%
%  Please DO NOT attempt to reformat the style of equation numbering!
%  For that matter, please do not attempt to redefine anything!

\section{Introduction}
\label{s:intro}

Here is the introduction. The file \texttt{biomsample.tex} gives an
example of using the \texttt{natbib} package to cross-reference
citations from the bibliography.  Here, we just do this manually to
demonstrate that the old-fashioned way also is acceptable.  See the
comment right before the bibliography section below for information on
using BiBTeX.

Please note that, although this document class produces a final
product that is very close to the format and appearance of a typeset
\textit{Biometrics} article, there may be some idiosyncracies that
cause the format to deviate slightly from that in the journal.  These
will be corrected at the typesetting phase should your paper be accepted
and forwarded for publication.  So do not worry if such things occur!

\section{Model}
\label{s:model}

\subsection{First Model Subsection}

The Cox model (Cox, 1972) is one of the most widely used statistical
models.  Hastie, Tibshirani, and Friedman (2001) is an example of a
citation to a work with three authors.  The first time you reference
one of these in the text, use all the authors names. However, in all
subsequent references, just use Hastie et al. (2001).  Works with four
or more authors are always referenced in the text using ``et al.''
All authors names should appear in the bibliography for all entries.

\subsection{Second Model Subsection}

Please use a recent issue of \textit{Biometrics} as a guide to the style
for citations and bibliography entries, and follow that style exactly!!

\section{Inference}
\label{s:inf}

Please see the file \texttt{biomsample.tex} for fancy examples of making
tables.  Here is a very simple one.  Use \texttt{table} for tables
that are narrow enough to fit in one column of the typeset journl; use
\texttt{table*} for tables that need to span two columns.  For
figures, use of \texttt{figure} and \texttt{figure*} is analogous.

\begin{table}
\caption{This is a simple table.}
\label{t:one}
\begin{center}
\begin{tabular}{lrrr}
\Hline
Estimator & \multicolumn{1}{c}{$\beta_1$} &  \multicolumn{1}{c}{$\beta_2$} & 
\multicolumn{1}{c}{$\beta_3$} \\ \hline
MLE & 10.18 & $-$3.26 & 0.13 \\
OLS & 9.92 & $-$3.19 & 0.11 \\
WLS & 9.88 & $-$3.33 & 0.12 \\
\hline
\end{tabular}
\end{center}
\end{table}

You can experiment with fancier tables than Table~\ref{t:one}.

We can get bold symbols using \verb+\bmath+, for example, $\bmath{\alpha}_i$.

\section{Discussion}
\label{s:discuss}

Put your final comments here.

%  The \backmatter command formats the subsequent headings so that they
%  are in the journal style.  Please keep this command in your document
%  in this position, right after the final section of the main part of 
%  the paper and right before the Acknowledgements, Supplementary Materials,
%  and References sections.

\backmatter

%  This section is optional.  Here is where you will want to cite
%  grants, people who helped with the paper, etc.  But keep it short!

\section*{Acknowledgements}

The authors thank Professor A. Sen for some helpful suggestions,
Dr C. R. Rangarajan for a critical reading of the original version of the
paper, and an anonymous referee for very useful comments that improved
the presentation of the paper.\vspace*{-8pt}

%  If your paper refers to supplementary web material, then you MUST
%  include this section!!  See Instructions for Authors at the journal
%  website http://www.biometrics.tibs.org

\section*{Supplementary Materials}

Web Appendix A, referenced in Section~\ref{s:model}, is available with
this paper at the Biometrics website on Wiley Online
Library.\vspace*{-8pt}

%  Here, we create the bibliographic entries manually, following the
%  journal style.  If you use this method or use natbib, PLEASE PAY
%  CAREFUL ATTENTION TO THE BIBLIOGRAPHIC STYLE IN A RECENT ISSUE OF
%  THE JOURNAL AND FOLLOW IT!  Failure to follow stylistic conventions
%  just lengthens the time spend copyediting your paper and hence its
%  position in the publication queue should it be accepted.

%  We greatly prefer that you incorporate the references for your
%  article into the body of the article as we have done here 
%  (you can use natbib or not as you choose) than use BiBTeX,
%  so that your article is self-contained in one file.
%  If you do use BiBTeX, please use the .bst file that comes with 
%  the distribution.

\begin{thebibliography}{}

\bibitem{ } Cox, D. R. (1972). Regression models and life tables (with
discussion).  \textit{Journal of the Royal Statistical Society, Series B}
\textbf{34,} 187--200.

\bibitem{ }  Hastie, T., Tibshirani, R., and Friedman, J. (2001). \textit{The 
Elements of Statistical Learning: Data Mining, Inference, and Prediction}.
New York: Springer.

\end{thebibliography}

\appendix

%  To get the journal style of heading for an appendix, mimic the following.

\section{}
\subsection{Title of appendix}

Put your short appendix here.  Remember, longer appendices are
possible when presented as Supplementary Web Material.  Please 
review and follow the journal policy for this material, available
under Instructions for Authors at \texttt{http://www.biometrics.tibs.org}.

\label{lastpage}

\end{document}

### 2014-12-12 ebayes_rnaseq_heterosis.tex
updated template with our info

@@
 
-%  The rotating package allows you to have tables displayed in landscape
-%  mode.  The rotating package is NOT included in this distribution, but
-%  can be obtained from the CTAN archive.  USE OF LANDSCAPE TABLES IS
-%  STRONGLY DISCOURAGED -- create landscape tables only as a last resort if
-%  you see no other way to display the information.  If you do do this,
-%  then you need the following command.
+\title[Empirical Bayes analysis for detection of gene heterosis in RNAseq data]{Empirical Bayes analysis for detection of gene heterosis in RNAseq data}
 
-%\usepackage[figuresright]{rotating}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%  Here, place your title and author information.  Note that in 
-%  use of the \author command, you create your own footnotes.  Follow
-%  the examples below in creating your author and affiliation information.
-%  Also consult a recent issue of the journal for examples of formatting.
-
-\title[This is an Example of Recto Running Head]{This is an Example of
-an Article Title This is an Example of an Article Title}
-
-%  Here are examples of different configurations of author/affiliation
-%  displays.  According to the Biometrics style, in some instances,
-%  the convention is to have superscript *, **, etc footnotes to indicate 
-%  which of multiple email addresses belong to which author.  In this case,
-%  use the \email{ } command to produce the emails in the display.
-
-%  In other cases, such as a single author or two authors from 
-%  different institutions, there should be no footnoting.  Here, use
-%  the \emailx{ } command instead.
-
-%  The examples below corrspond to almost every possible configuration
-%  of authors and may be used as a guide.  For other configurations, consult
-%  a recent issue of the the journal.
-
-%  Single author -- USE \emailx{ } here so that no asterisk footnoting
-%  for the email address will be produced.
-
-%\author{John Author\emailx{email@address.edu} \\
-%Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K.}
-
-%  Two authors from the same institution, with both emails -- use
-%  \email{ } here to produce the asterisk footnoting for each email address
-
-%\author{John Author$^{*}$\email{author@address.edu} and
-%Kathy Authoress$^{**}$\email{email2@address.edu} \\
-%Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K.}
-
-%  Exactly two authors from different institutions, with both emails  
-%  USE \emailx{ } here so that no asterisk footnoting for the email address
-%  is produced.
-
-\author
-{John Author\emailx{author@address.edu} \\
-Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K. 
-\and
-Kathy Author\emailx{anotherauthor@address.edu} \\
-Department of Biostatistics, University of North Carolina at Chapel Hill, 
-Chapel Hill, North Carolina, U.S.A.}
-
-%  Three or more authors from same institution with all emails displayed
-%  and footnoted using asterisks -- use \email{ }
-
-%\author{John Author$^*$\email{author@address.edu}, 
-%Jane Author$^{**}$\email{jane@address.edu}, and 
-%Dick Author$^{***}$\email{dick@address.edu} \\
-%Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K}
-
-%  Three or more authors from same institution with one corresponding email
-%  displayed
-
-%\author{John Author$^*$\email{author@address.edu}, 
-%Jane Author, and Dick Author \\
-%Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K}
-
-%  Three or more authors, with at least two different institutions,
-%  more than one email displayed
-
-%\author{John Author$^{1,*}$\email{author@address.edu}, 
-%Kathy Author$^{2,**}$\email{anotherauthor@address.edu}, and 
-%Wilma Flinstone$^{3,***}$\email{wilma@bedrock.edu} \\
-%$^{1}$Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K \\
-%$^{2}$Department of Biostatistics, University of North Carolina at 
-%Chapel Hill, Chapel Hill, North Carolina, U.S.A. \\
-%$^{3}$Department of Geology, University of Bedrock, Bedrock, Kansas, U.S.A.}
-
-%  Three or more authors with at least two different institutions and only
-%  one email displayed
-
-%\author{John Author$^{1,*}$\email{author@address.edu}, 
-%Wilma Flinstone$^{2}$, and Barney Rubble$^{2}$ \\
-%$^{1}$Department of Statistics, University of Warwick, Coventry CV4 7AL, U.K \\
-%$^{2}$Department of Geology, University of Bedrock, Bedrock, Kansas, U.S.A.}
+\author{Jarad Niemi $^*$\email{niemi@iastate.edu}, 
+Eric Mittman, 
+Will Landau, and 
+Dan Nettleton \\
+Department of Statistics, Iowa State University, Ames, Iowa, U.S.A.}
 
@@
 
-%  This will produce the submission and review information that appears
-%  right after the reference section.  Of course, it will be unknown when
-%  you submit your paper, so you can either leave this out or put in 
-%  sample dates (these will have no effect on the fate of your paper in the
-%  review process!)
-
-\date{{\it Received October} 2007. {\it Revised February} 2008.  {\it
-Accepted March} 2008.}
-
-%  These options will count the number of pages and provide volume
-%  and date information in the upper left hand corner of the top of the 
-%  first page as in published papers.  The \pagerange command will only
-%  work if you place the command \label{firstpage} near the beginning
-%  of the document and \label{lastpage} at the end of the document, as we
-%  have done in this template.
-
-%  Again, putting a volume number and date is for your own amusement and
-%  has no bearing on what actually happens to your paper!
+\date{{\it Received December} 2014}
 
 \pagerange{\pageref{firstpage}--\pageref{lastpage}} 
-\volume{64}
-\pubyear{2008}
-\artmonth{December}
-
-%  The \doi command is where the DOI for your paper would be placed should it
-%  be published.  Again, if you make one up and stick it here, it means 
-%  nothing!
-
-\doi{10.1111/j.1541-0420.2005.00454.x}
-
-%  This label and the label ``lastpage'' are used by the \pagerange
-%  command above to give the page range for the article.  You may have 
-%  to process the document twice to get this to match up with what you 
-%  expect.  When using the referee option, this will not count the pages
-%  with tables and figures.
+\volume{VV}
+\pubyear{YYYY}
+\artmonth{Month}
+\doi{}
 
@@
 
-%  Please place your key words in alphabetical order, separated
-%  by semicolons, with the first letter of the first word capitalized,
-%  and a period at the end of the list.
-%
-
 \begin{keywords}
-A key word; But another key word; Still another key word; Yet another key word.
+Empirical Bayes; Heterosis; Hierarchical Model; Negative binomial; RNAseq.
 \end{keywords}
-
-%  As usual, the \maketitle command creates the title and author/affiliations
-%  display
 
@@
 
-%  If you are using the referee option, a new page, numbered page 1, will
-%  start after the summary and keywords.  The page numbers thus count the
-%  number of pages of your manuscript in the preferred submission style.
-%  Remember, ``Normally, regular papers exceeding 25 pages and Reader Reaction 
-%  papers exceeding 12 pages in (the preferred style) will be returned to 
-%  the authors without review. The page limit includes acknowledgements, 
-%  references, and appendices, but not tables and figures. The page count does 
-%  not include the title page and abstract. A maximum of six (6) tables or 
-%  figures combined is often required.''
-
-%  You may now place the substance of your manuscript here.  Please use
-%  the \section, \subsection, etc commands as described in the user guide.
-%  Please use \label and \ref commands to cross-reference sections, equations,
-%  tables, figures, etc.
-%
-%  Please DO NOT attempt to reformat the style of equation numbering!
-%  For that matter, please do not attempt to redefine anything!
+%  A maximum of six (6) tables or figures combined is often required.''
 
@@
 
-Here is the introduction. The file \texttt{biomsample.tex} gives an
-example of using the \texttt{natbib} package to cross-reference
-citations from the bibliography.  Here, we just do this manually to
-demonstrate that the old-fashioned way also is acceptable.  See the
-comment right before the bibliography section below for information on
-using BiBTeX.
+\section{Heterosis}
+\label{s:heterosis}
 
-Please note that, although this document class produces a final
-product that is very close to the format and appearance of a typeset
-\textit{Biometrics} article, there may be some idiosyncracies that
-cause the format to deviate slightly from that in the journal.  These
-will be corrected at the typesetting phase should your paper be accepted
-and forwarded for publication.  So do not worry if such things occur!
-
-\section{Model}
+\section{Hierarchical model}
 \label{s:model}
 
-\subsection{First Model Subsection}
-
-The Cox model (Cox, 1972) is one of the most widely used statistical
-models.  Hastie, Tibshirani, and Friedman (2001) is an example of a
-citation to a work with three authors.  The first time you reference
-one of these in the text, use all the authors names. However, in all
-subsequent references, just use Hastie et al. (2001).  Works with four
-or more authors are always referenced in the text using ``et al.''
-All authors names should appear in the bibliography for all entries.
-
-\subsection{Second Model Subsection}
-
-Please use a recent issue of \textit{Biometrics} as a guide to the style
-for citations and bibliography entries, and follow that style exactly!!
-
-\section{Inference}
+\section{Empirical Bayes}
 \label{s:inf}
 
-Please see the file \texttt{biomsample.tex} for fancy examples of making
-tables.  Here is a very simple one.  Use \texttt{table} for tables
-that are narrow enough to fit in one column of the typeset journl; use
-\texttt{table*} for tables that need to span two columns.  For
-figures, use of \texttt{figure} and \texttt{figure*} is analogous.
+\section{Simulation study}
+\label{s:simulation}
 
-\begin{table}
-\caption{This is a simple table.}
-\label{t:one}
-\begin{center}
-\begin{tabular}{lrrr}
-\Hline
-Estimator & \multicolumn{1}{c}{$\beta_1$} &  \multicolumn{1}{c}{$\beta_2$} & 
-\multicolumn{1}{c}{$\beta_3$} \\ \hline
-MLE & 10.18 & $-$3.26 & 0.13 \\
-OLS & 9.92 & $-$3.19 & 0.11 \\
-WLS & 9.88 & $-$3.33 & 0.12 \\
-\hline
-\end{tabular}
-\end{center}
-\end{table}
-
-You can experiment with fancier tables than Table~\ref{t:one}.
-
-We can get bold symbols using \verb+\bmath+, for example, $\bmath{\alpha}_i$.
+\section{Heterosis in corn ??}
+\label{s:corn}
 
@@
 
-%  The \backmatter command formats the subsequent headings so that they
-%  are in the journal style.  Please keep this command in your document
-%  in this position, right after the final section of the main part of 
-%  the paper and right before the Acknowledgements, Supplementary Materials,
-%  and References sections.
-
-\backmatter
-
-%  This section is optional.  Here is where you will want to cite
-%  grants, people who helped with the paper, etc.  But keep it short!
+\backmatter %  Please keep this command in your document in this position
 
@@
 
-The authors thank Professor A. Sen for some helpful suggestions,
-Dr C. R. Rangarajan for a critical reading of the original version of the
-paper, and an anonymous referee for very useful comments that improved
-the presentation of the paper.\vspace*{-8pt}
+The authors thank Andrew Lithio for help implementing our model in {\tt ShrinkBayes}.
 
@@
 
-%  Here, we create the bibliographic entries manually, following the
-%  journal style.  If you use this method or use natbib, PLEASE PAY
-%  CAREFUL ATTENTION TO THE BIBLIOGRAPHIC STYLE IN A RECENT ISSUE OF
-%  THE JOURNAL AND FOLLOW IT!  Failure to follow stylistic conventions
-%  just lengthens the time spend copyediting your paper and hence its
-%  position in the publication queue should it be accepted.
-
-%  We greatly prefer that you incorporate the references for your
-%  article into the body of the article as we have done here 
-%  (you can use natbib or not as you choose) than use BiBTeX,
-%  so that your article is self-contained in one file.
-%  If you do use BiBTeX, please use the .bst file that comes with 
-%  the distribution.
-
-\begin{thebibliography}{}
-
-\bibitem{ } Cox, D. R. (1972). Regression models and life tables (with
-discussion).  \textit{Journal of the Royal Statistical Society, Series B}
-\textbf{34,} 187--200.
-
-\bibitem{ }  Hastie, T., Tibshirani, R., and Friedman, J. (2001). \textit{The 
-Elements of Statistical Learning: Data Mining, Inference, and Prediction}.
-New York: Springer.
-
-\end{thebibliography}
+\bibliography{jarad}
+\bibliographystyle{biom}
 
@@
 \section{}
-\subsection{Title of appendix}
-
-Put your short appendix here.  Remember, longer appendices are
-possible when presented as Supplementary Web Material.  Please 
-review and follow the journal policy for this material, available
-under Instructions for Authors at \texttt{http://www.biometrics.tibs.org}.
+\subsection{Stan model for heterosis}
 

### 2014-12-12 ebayes_rnaseq_heterosis.tex
added figures
start of manuscript

@@
-%  template.tex for Biometrics papers
-%
-%  This file provides a template for Biometrics authors.  Use this
-%  template as the starting point for creating your manuscript document.
-%  See the file biomsample.tex for an example of a full-blown manuscript.
+\documentclass[useAMS,usenatbib,referee]{biom}
 
-%  ALWAYS USE THE referee OPTION WITH PAPERS SUBMITTED TO BIOMETRICS!!!
-%  You can see what your paper would look like typeset by removing
-%  the referee option.  Because the typeset version will be in two
-%  columns, however, some of your equations may be too long. DO NOT
-%  use the \longequation option discussed in the user guide!!!  This option
-%  is reserved ONLY for equations that are impossible to split across 
-%  multiple lines; e.g., a very wide matrix.  Instead, type your equations 
-%  so that they stay in one column and are split across several lines, 
-%  as are almost all equations in the journal.  Use a recent version of the
-%  journal as a guide. 
-%  
-\documentclass[useAMS,referee]{biom}
-%documentclass[useAMS]{biom}
-%
-%  If your system does not have the AMS fonts version 2.0 installed, then
-%  remove the useAMS option.
-%
-%  useAMS allows you to obtain upright Greek characters.
-%  e.g. \umu, \upi etc.  See the section on "Upright Greek characters" in
-%  this guide for further information.
-%
-%  If you are using AMS 2.0 fonts, bold math letters/symbols are available
-%  at a larger range of sizes for NFSS release 1 and 2 (using \boldmath or
-%  preferably \bmath).
-% 
-%  Other options are described in the user guide. Here are a few:
-% 
-%  -  If you use Patrick Daly's natbib  to cross-reference your 
-%     bibliography entries, use the usenatbib option
-%
-%  -  If you use \includegraphics (graphicx package) for importing graphics
-%     into your figures, use the usegraphicx option
-% 
-%  If you wish to typeset the paper in Times font (if you do not have the
-%  PostScript Type 1 Computer Modern fonts you will need to do this to get
-%  smoother fonts in a PDF file) then uncomment the next line
-%  \usepackage{Times}
+\usepackage{amsmath}
+\usepackage[figuresright]{rotating}
+
+\graphicspath{{../temp/rna-seq-stan/fig-rm-stan_laplace/}}
 
@@
 
-\title[Empirical Bayes analysis for detection of gene heterosis in RNAseq data]{Empirical Bayes analysis for detection of gene heterosis in RNAseq data}
+\title[eBayes for heterosis in RNAseq]{Empirical Bayes analysis for detection of gene heterosis in RNAseq data}
 
@@
 \artmonth{Month}
-\doi{}
+\doi{x}
 
@@
 
-\section{Hierarchical model}
+\section{Empirical Bayes}
+\subsection{Hierarchical model for RNAseq counts}
 \label{s:model}
 
-\section{Empirical Bayes}
-\label{s:inf}
+Let $Y_{gvi}$ be the count for gene $g=1,\ldots,F$, variety $v=1,\ldots,V$, and replicate $i=1,\ldots,n_v$.
+As shown in equation \eqref{e:data}, we assume a negative binomial for the counts with a mean that depends on the gene-variety combination through $\mu_{gv}$ and the sequencing depth $c_{vi}$ and a gene-specific over-dispersion $\psi_g$.
+Here $NB(\eta,\psi)$ indicates a negative-binomial with expectation $\eta$ and variance $\eta+\psi\eta^2$.
+
+\begin{equation} 
+Y_{gvi} \stackrel{ind}{\sim} NB(e^{\mu_{gv}+\delta_{vi}},\psi_g) 
+\label{e:data}
+\end{equation}
+
+A hierarchical model is constructed for the sequencing depth, overdispersion, and gene-variety parameters.
+The sample-specific sequencing depth terms are assumed to follow a normal distribution, i.e.
+$\delta_{vi} \stackrel{ind}{\sim} N(0,\tau_\delta^2)$.
+The logarithm of the gene-specific over-dispersion parameters also follow a normal distribution $\log(\psi_g) \stackrel{ind}{\sim} N(\zeta_\psi,\tau_\psi^2)$.
+The variety-specific effects for each gene, $\mu_g = (\mu_{g1},\ldots,\mu_{gV})'$ are assumed to have a joint normal distribution $\mu_g \stackrel{ind}{\sim} N(\eta, \Sigma)$.
+We consider an \emph{independent} model where $\Sigma$ is diagonal with elements $\sigma_1^2,\ldots,\sigma_V^2$ and a \emph{covariance} model where we estimate all parameters in $\Sigma$.
+
+We assume improper uniform priors for all hierarchical means and half-Cauchy priors for standard deviations, e.g.
+$\tau_\delta, \tau_\psi \stackrel{ind}{\sim} Ca^+(0,3)$ where 0 and 3 are the location and scale parameters on the untruncated distribution.
+In the independent model, we also assume $Ca^+(0,3)$ on the diagonal elements while in the covariance model, we assume a conjugate, proper inverse Wishart with degrees of freedom $V+1$ and an identity scale matrix.
+
+\subsection{Empirical Bayes}
+\label{s:ebayes}
+
+The parameters can be group by gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ and $\theta_g = (\mu_g,\psi_g)$ and non-gene-specific parameters $\phi = (\delta_{11},\ldots,\delta_{Vn_v}, \zeta_\psi, \tau_\psi^2, \Sigma)$.
+We employ an optimization procedure to obtain the \emph{maximum a posterior} (MAP) estimate for $\phi$, i.e.
+$\hat{\phi}_{MAP}$ and then base gene-specific inference on the posterior conditional on this estimated hyperparameter.
+Fortunately, conditional on the hyperparameters the gene specific parameters are conditionally independent as in equation \eqref{e:condind}.
+
+\begin{equation}
+p(\theta|y,\hat{\phi}_{MAP}) = \prod_{g=1}^G p(\theta_g|y_g,\hat{\phi}_{MAP}) 
+\label{e:condind}
+\end{equation}
 
@@
 \label{s:simulation}
+
+\subsection{Coverage for our model}
+
+\input{../temp/analyze-heterosis/FIGURES/coverage}
+
+\subsection{Heterosis detection}
+
+\begin{figure}[htbp]
+\centerline{\includegraphics[width=0.8\textwidth]{exampleROC0_1}}
+\caption{Example ROC curves for the first simulation}
+\label{f:roc}
+\end{figure}
+
+\begin{figure}
+\centerline{\includegraphics[width=0.8\textwidth]{auc-facet-TRUE}}
+\caption{Area under the ROC curves (AUC) for 3 different sample sizes (facets) for the methods under comparison for false positive rates below 0.1.
+Each line is a different simulation while the blue box indicates mean AUC and its standard error across the simulations.}
+\end{figure}
 
@@
 
-Web Appendix A, referenced in Section~\ref{s:model}, is available with
-this paper at the Biometrics website on Wiley Online
-Library.\vspace*{-8pt}
-
 \bibliography{jarad}

### 2014-12-15 ebayes_rnaseq_heterosis.tex
some work

@@
 
-\title[eBayes for heterosis in RNAseq]{Empirical Bayes analysis for detection of gene heterosis in RNAseq data}
+\title{Empirical Bayes analysis for detection of gene heterosis in RNAseq data}
 
-\author{Jarad Niemi $^*$\email{niemi@iastate.edu}, 
+\author{Jarad Niemi$^*$\email{niemi@iastate.edu}, 
 Eric Mittman, 
@@
 
-\section{Empirical Bayes}
+Well before heterosis was scientifically described by \cite{darwin1876effects}, humans had been using heterosis for various practical purposes.
+Within the last century, heterosis has been used to improve many species for food, feed, and fuel industries.
+A few relatively recent examples include rice \cite{yu1997importance}, alfalfa \cite{riday2002heterosis}, tomatoes \cite{krieger2010flowering}, and fish \cite{wohlfarth1993heterosis}.
+Despite the intensive study and successful utilization of heterosis, the basic genomic mechanisms responsible for heterosis remain unclear \cite{lippman2007heterosis}.
+
+\cite{shull1952beginnings} coined the term heterosis and defined it as 
+\begin{quote}
+the interpretation of increased vigor, size, fruitfulness, speed of development, resistance to disease and to insect pests, or to climatic rigors of any kind, manifested by crossbred organisms as compared with corresponding inbreds, as the specific results of unlikeness in the constitutions of the uniting parental gametes.
+\end{quote}
+
+For example, the maize hybrid produced from a cross of parental inbred lines B73 and Mo17 is larger in size, has lower maturation time, and has higher grain yield than both of its parents \cite{hallauer2010quantitative}.
+
+\section{Empirical Bayes identification of gene heterosis from RNAseq counts}
+
+To identify genes displaying heterosis using RNAseq data, we build a hierarchical model to borrow information on gene-variety means as well as overdispersion parameters, estimate the parameters using a empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate posterior probabilities for heterosis.
+
 \subsection{Hierarchical model for RNAseq counts}
@@
 As shown in equation \eqref{e:data}, we assume a negative binomial for the counts with a mean that depends on the gene-variety combination through $\mu_{gv}$ and the sequencing depth $c_{vi}$ and a gene-specific over-dispersion $\psi_g$.
-Here $NB(\eta,\psi)$ indicates a negative-binomial with expectation $\eta$ and variance $\eta+\psi\eta^2$.
+Here $NB(\eta,\psi)$ indicates a negative-binomial with expectation $\eta$ and variance $\eta+\psi\eta^2$ and $ind$ indicates the observations are conditionally independent.
 
@@
 We assume improper uniform priors for all hierarchical means and half-Cauchy priors for standard deviations, e.g.
-$\tau_\delta, \tau_\psi \stackrel{ind}{\sim} Ca^+(0,3)$ where 0 and 3 are the location and scale parameters on the untruncated distribution.
+$\tau_\delta, \tau_\psi \stackrel{ind}{\sim} Ca^+(0,3)$ where 0 and 3 are the location and scale parameters of the untruncated distribution.
 In the independent model, we also assume $Ca^+(0,3)$ on the diagonal elements while in the covariance model, we assume a conjugate, proper inverse Wishart with degrees of freedom $V+1$ and an identity scale matrix.
@@
 
-The parameters can be group by gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ and $\theta_g = (\mu_g,\psi_g)$ and non-gene-specific parameters $\phi = (\delta_{11},\ldots,\delta_{Vn_v}, \zeta_\psi, \tau_\psi^2, \Sigma)$.
-We employ an optimization procedure to obtain the \emph{maximum a posterior} (MAP) estimate for $\phi$, i.e.
-$\hat{\phi}_{MAP}$ and then base gene-specific inference on the posterior conditional on this estimated hyperparameter.
-Fortunately, conditional on the hyperparameters the gene specific parameters are conditionally independent as in equation \eqref{e:condind}.
-
+The parameters from the model in Section \ref{s:model} can be group by gene-specific parameters $\theta_g = (\mu_g,\psi_g)$, collectively $\theta = (\theta_1,\ldots,\theta_G)$, and hyperparameter $\phi = (\delta, \zeta_\psi, \tau_\psi^2, \Sigma)$ where $\delta = (\delta_{11},\ldots,\delta_{Vn_V})$.
+We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\phi$ and then base gene-specific inference on the posterior conditional on this estimated hyperparameter.
+Conditional on the hyperparameters, the gene specific parameters are conditionally independent as in equation \eqref{e:condind}.
 \begin{equation}
-p(\theta|y,\hat{\phi}_{MAP}) = \prod_{g=1}^G p(\theta_g|y_g,\hat{\phi}_{MAP}) 
+p(\theta|y,\hat{\phi}_{MAP}) = \prod_{g=1}^G p(\theta_g|y_g,\hat{\phi}_{MAP}) \propto \prod_{g=1}^G NB(y_g|\theta_g,\hat{\delta}_{MAP})p(\theta_g|\hat{\phi}_{MAP}) 
 \label{e:condind}
 \end{equation}
+Thus, conditional posterior inference for the gene specific parameters can be obtained independently and in parallel.
+
+\subsection{Gene heterosis}
+\label{s:gene_heterosis}
+
+As defined in Section \ref{s:heterosis}, heterosis is increased (or decreased) phenotypic expression in a hybrid relative to its inbred parents.
+Thus, when attempting to identify heterosis, there are three varieties of interest, i.e.
+$V=3$.
+We let $v=1,2$ be the two parental varieties and $v=3$ be the hybrid.
+We define gene heterosis to be the differential expression of the hybrid relative to its parents.
+For a specific gene $g$, low-parent gene heteorisis (LPH) occurs when $\mu_{g3}< \min(\mu_{g1},\mu_{g2})$, mid-parent gene heterosis occurs when $\mu_{g1}\le \mu_{g3}\le \mu_{g2}$ if $\mu_{g1}\le \mu_{g2}$ (or $\mu_{g1}\ge \mu_{g3}\ge \mu_{g2}$ if $\mu_{g1}\ge \mu_{g2}$), and high-parent gene heterosis (HPH) occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$.
+For simplicity, we define \emph{extreme gene heterosis} (EH) which is the event LPH or HPH.
+Of interest is to evaluate the hypotheses shown in equation \eqref{e:hypotheses} where $H_0$ indicates no EH and $H_1$ indicates EH.
+\begin{align}
+\label{e:hypotheses}
+H_0:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
+H_1:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}.
+\end{align}
 

### 2014-12-15 ebayes_rnaseq_heterosis.tex
through edgeR

@@
 For example, the maize hybrid produced from a cross of parental inbred lines B73 and Mo17 is larger in size, has lower maturation time, and has higher grain yield than both of its parents \cite{hallauer2010quantitative}.
+Data from \citep{paschold2012complementation}
+
+\citep{ji2014estimation}
 
@@
 Here $NB(\eta,\psi)$ indicates a negative-binomial with expectation $\eta$ and variance $\eta+\psi\eta^2$ and $ind$ indicates the observations are conditionally independent.
-
 \begin{equation} 
@@
 
-We assume improper uniform priors for all hierarchical means and half-Cauchy priors for standard deviations, e.g.
+We assume improper uniform priors for all hierarchical means and half-Cauchy priors for standard deviations \citep{gelman2006prior}, e.g.
 $\tau_\delta, \tau_\psi \stackrel{ind}{\sim} Ca^+(0,3)$ where 0 and 3 are the location and scale parameters of the untruncated distribution.
@@
 \begin{equation}
-p(\theta|y,\hat{\phi}_{MAP}) = \prod_{g=1}^G p(\theta_g|y_g,\hat{\phi}_{MAP}) \propto \prod_{g=1}^G NB(y_g|\theta_g,\hat{\delta}_{MAP})p(\theta_g|\hat{\phi}_{MAP}) 
+p(\theta|y,\hat{\phi}_{MAP}) = \prod_{g=1}^G p(\theta_g|y_g,\hat{\phi}_{MAP}) \propto \prod_{g=1}^G NB(y_g|\theta_g,\hat{\delta}_{MAP})N(\theta_g|\hat{\mu}_{MAP}, \hat{Sigma}_{MAP}) 
 \label{e:condind}
@@
 Thus, conditional posterior inference for the gene specific parameters can be obtained independently and in parallel.
+
+To perform the optimization, we use the statistical software {\tt Stan} \citep{stan-software:2014} run through the RStan interface \citep{rstan-software:2014} in R \citep{R2014}.
+The Stan software allows construction of general Bayesian models with the primary restriction that all unknown parameters must be continuous.
+Once a model is constructed, the software can be used to obtain MAP estimates \cite[see Section 50.3]{stan-manual:2014} using the {\tt optimizing} function in RStan.
+We found it necessary to use the L-BFGS algorithm \cite[see Section 55]{stan-manual:2014}.
+Alternatively, the software can be used to run a Markov chain Monte Carlo (MCMC) algorithm to obtain samples from the posterior using the {\tt sampling} function in RStan.
+By default, Stan uses a variant of Hamiltonion Monte Carlo \citep{neal2011mcmc} called the No-U-turn sampler \citep{hoffman2013no}.
+
+As we have not integrated out the gene-specific parameters when obtaining the MAP estimate, we actually receive a joint MAP estimate for $\theta$ and $\phi$.
+We explicitly use the MAP estimate for $\phi$ when obtaining empirical Bayes posterior samples for the gene specific parameters.
+The MAP estimate for $\theta$ can also be used in this step as initial values for the MCMC.
 
@@
 \label{e:hypotheses}
-H_0:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
-H_1:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}.
+H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
+H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}.
 \end{align}
+We evaluate these hypotheses based on empirical Bayes estimates of their posterior probability, e.g. 
+\begin{align}
+P\left(H_{g1}|y, \hat{\phi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\phi}_{MAP}\right) \nonumber \\
+&\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right) \label{e:probs}
+\end{align}
+where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise.
+The set of probabilities $P\left(H_{g1}|y, \hat{\phi}_{MAP}\right)$ for $g=1,\ldots,G$ can be used to provide a sorted list of genes potentially involved in phenotypic heterosis.
 
-\section{Simulation study}
+\section{Simulation study based on maize experiment}
 \label{s:simulation}
 
-\subsection{Coverage for our model}
+% \subsection{Coverage for our model}
+% \input{../temp/analyze-heterosis/FIGURES/coverage}
 
-\input{../temp/analyze-heterosis/FIGURES/coverage}
+To assess the efficacy of our method to identify genes demonstrating heterosis, we used a maize data set with parental varieties B73 and Mo17 and the hybrid variety (B73 $\times$ Mo17) \citep{paschold2012complementation}.
+We compared our approach to an approach using {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes}.
 
-\subsection{Heterosis detection}
+\subsection{Construct simulated data}
+
+Initially, the maize experiment had RNAseq counts for 39,656 genes and 4 replicates per variety.
+We eliminated low count genes by removing any gene with three or more zeros for any variety and genes whose mean count across the varieties was less than one which resulted in 27,888 genes remaining.
+
+To obtain gene-variety-specific location parameters and gene-specific dispersion parameters, we used {\tt edgeR} in order to maintain existing covariance structure amongst the gene-variety means.
+We estimated sample-specific normalization factors to account for sequencing depth.
+Then, we estimated common a common dispersion for each gene, but shared across the samples.
+Finally, we obtained independent estimates for the gene-variety specific location parameters.
+The location parameters, dispersion parameters, and normalization factors were then treated as the truth for the simulation study.
+A gene with $\mu_{g3} < \min(\mu_{g1}, \mu_{g2})$ was considered to have low parent heterosis while a gene with $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$ was considered to have high parent heterosis.
+
+Using these parameters and normalization factors, we simulated data according to the negative-binomial model in equation \eqref{e:model} independently for each gene.
+For each simulation, low count genes, i.e.
+those with three or more zeros in any variety or whose mean count across the varieties was less than one, were removed.
+A random subset of 25,000 genes was retained from those with high enough counts.
+This simulation process was repeated 10 times for each of 4, 8, and 16 replicates per variety (reusing normalization factors for the higher number of replicates).
+
+\subsection{Alternative methods}
+
+Existing methods are not designed to identify gene heterosis, but we adapted the approaches of used in {\tt edgeR}, {\tt baySeq}.
+and {\tt ShrinkBayes} to provide a list of genes ranked by plausibility of heterosis.
+
+The approach used by {\tt edgeR} naturally provides testing of hypotheses for differential expression amongst two varieties.
+If the hybrid's mean expression does not differ from a parent, then the hybrid does not exhibit heterosis.
+We can calculate two pvalues: one for the null hypothesis that the hybrid's mean expression is the same as parent 1 and the second for the null hypothesis that the hybrid's mean expression is the same as parent 2.
+But, even if both pvalues are small, the hybrid's mean expression could be between the two parents, i.e.
+mid-parent heterosis, and therefore not be exhibiting extreme heterosis.
+We construct a statistic that is 1 if the estimate of mean expression
+for the hybrid is between the estimates for the parents and, otherwise, is selected from the pvalues from above by whichever parent's mean estimate is closer to that of the hybrid.
 
@@
 \section{}
-\subsection{Stan model for heterosis}
+
+The following sections provide Stan model code for performing the optimization and MCMC described in Section \ref{s:ebayes}.
+In these models, it is easier to construct the model using a single index, i.e.
+$y_i$, and then provide vectors that provide relevant quantities for this observation.
+These vectors include $g$ which indicates the gene, $v$ which indicates the variety, and $s$ which gives the sample index, i.e.
+a distinct identifier for the variety-replicate combation.
+
+\subsection{Optimization models}
+
+The following two sections provide Stan model code for the optimization procedure, using the {\tt optimizing} function in RStan, either assuming independence or estimating a covariance amongst the location and (log) dispersion parameters.
+
+\subsubsection{All genes model}
+\label{s:all_genes_model}
+
+This section provides Stan model code for the model that assumes prior independence amongst the location and dispersion parameters.
+
+\begin{verbatim}
+data{
+  int N; //total # obs
+  int G; // # genes
+  int S; // # samples
+  int<lower=0> y[N];
+  int<lower=1,upper=G> g[N]; //obs -> gene
+  int<lower=1,upper=3> t[N]; //obs -> genotype
+  int<lower=1,upper=S> s[N]; //obs -> sample (genotype*rep)
+  row_vector[3] X[3]; // mean structure (genotype)
+}
+parameters{
+  matrix[3,G] B;         //phi, alpha, gamma by gene
+  vector[G] lpsi;        //'dispersion'
+  vector[S] c;           //lane sequencing depth
+  real<lower=-20,upper=20> mu_phi;
+  real<lower=-20,upper=20> mu_alpha;
+  real<lower=-20,upper=20> mu_delta;
+  real<lower=-20,upper=20> mu_psi;
+  real<lower=0,upper = 20> sigma_phi;
+  real<lower=0,upper = 20> sigma_alpha;
+  real<lower=0,upper = 20> sigma_delta;
+  real<lower=0,upper = 20> sigma_psi;
+  real<lower=0,upper=5> sigma_c;
+}
+
+model{
+  lpsi ~ normal(mu_psi,sigma_psi);
+  B[1] ~ normal(mu_phi,sigma_phi);
+  B[2] ~ normal(mu_alpha,sigma_alpha);
+  B[3] ~ normal(mu_delta,sigma_delta);
+  c ~ normal(0,sigma_c);
+  for(n in 1:N){
+    y[n] ~ neg_binomial_2_log(X[t[n]]*col(B,g[n]) + c[s[n]], 1 / exp(lpsi[g[n]]));
+  }
+  sigma_psi ~ cauchy(0,3);
+  sigma_phi ~ cauchy(0,3);
+  sigma_alpha ~ cauchy(0,3);
+  sigma_delta ~ cauchy(0,3);
+}
+\end{verbatim}
+
+\subsubsection{All genes model with covariance}
+\label{s:all_genes_model_with_covariance}
+
+This section provides Stan model code for the model that estimates the covariance structure amongst the location and (log) dispersion parameters.
+
+\begin{verbatim}
+data{
+  int N; //total # obs
+  int G; // # genes
+  int S; // # samples
+  int<lower=0> y[N];
+  int<lower=1,upper=G> g[N]; //obs -> gene
+  int<lower=1,upper=3> t[N]; //obs -> genotype
+  int<lower=1,upper=S> s[N]; //obs -> sample (genotype*rep)
+  row_vector[4] X[3]; // mean structure (genotype)
+}
+parameters{
+  vector[4] B[G];         //mu1, mu2, mu3, lpsi by gene
+  vector[S] c;            //lane sequencing depth
+  vector<lower=-20,upper=20>[4] mu;
+  cov_matrix[4] Sigma;
+  real<lower=0,upper=5> sigma_c;
+}
+
+model{
+  vector[4] x;
+  for(i in 1:G){
+    B[i] ~ multi_normal(mu, Sigma);
+  }
+  c ~ normal(0,sigma_c);
+  for(n in 1:N){
+    y[n] ~ neg_binomial_2_log(X[t[n]]*B[g[n]] + c[s[n]], 1 / exp(B[g[n],4]));
+  }
+  x <-rep_vector(1,4);
+  Sigma ~ inv_wishart(5.0,diag_matrix(x));
+}
+\end{verbatim}
+
+\subsection{Models for MCMC}
+
+As described in Section \ref{s:ebayes}, conditional on the hyperparameters inference on the gene-specific parameters can be performed independently and in parallel.
+This section provides Stan model code for single-gene models either assuming independence or with a covariance structure amongst the location and (log) dispersion parameter.
+These model files are used in conjunction with the {\tt sampling} function in RStan.
+
+\subsubsection{Single-gene model assuming independence}
+\label{s:single_gene_model}
+
+This section provides Stan model code for the single-gene model where independence is assumed amongst the location and (log) dispersion parameters.
+
+\begin{verbatim}
+data{
+  int N; //total # obs
+  int S; // # samples
+  int<lower=0> y[N];
+  int<lower=1,upper=3> t[N]; //obs -> genotype
+  int<lower=1,upper=S> s[N]; //obs -> sample (genotype*rep)
+  row_vector[3] X[3]; // mean structure (genotype)
+  real<lower=-20,upper=20> mu_phi;
+  real<lower=-20,upper=20> mu_alpha;
+  real<lower=-20,upper=20> mu_delta;
+  real<lower=-20,upper=20> mu_psi;
+  real<lower=0,upper = 20> sigma_phi;
+  real<lower=0,upper = 20> sigma_alpha;
+  real<lower=0,upper = 20> sigma_delta;
+  real<lower=0,upper = 20> sigma_psi;
+  vector[S] c;            //lane sequencing depth
+}
+transformed data{
+  vector[3] Mu;
+  cov_matrix[3] Sigma;
+  Mu[1] <- mu_phi;
+  Mu[2] <- mu_alpha;
+  Mu[3] <- mu_delta;
+  Sigma[1,2] <- 0;
+  Sigma[2,1] <- 0;
+  Sigma[1,3] <- 0;
+  Sigma[3,1] <- 0;
+  Sigma[2,3] <- 0;
+  Sigma[3,2] <- 0;
+  Sigma[1,1] <- square(sigma_phi);
+  Sigma[2,2] <- square(sigma_alpha);
+  Sigma[3,3] <- square(sigma_delta);
+}
+parameters{
+  vector[3] B;         //phi, alpha, gamma by gene
+  real lpsi;       //'dispersion'
+}
+
+model{
+  lpsi ~ normal(mu_psi,sigma_psi);
+  B    ~ multi_normal(Mu,Sigma);
+  for(n in 1:N){
+    y[n] ~ neg_binomial_2_log(X[t[n]]*B + c[s[n]], 1 / exp(lpsi));
+  }
+}
+\end{verbatim}
+
+\subsection{Single-gene model with covariance}
+\label{s:single_gene_model_with_covariance}
+
+This section provides Stan model code for the single-gene model where a covariance amongst the location and (log) dispersion parameters is used.
+
+\begin{verbatim}
+data{
+  int N; //total # obs
+  int S; // # samples
+  int<lower=0> y[N];
+  int<lower=1,upper=3> t[N]; //obs -> genotype
+  int<lower=1,upper=S> s[N]; //obs -> sample (genotype*rep)
+  row_vector[4] X[3]; // mean structure (genotype)
+  vector[S] c;            //lane sequencing depth
+  vector<lower=-20,upper=20>[4] mu;
+  real<lower=0> sigma_phi;
+  real<lower=0> sigma_alpha;
+  real<lower=0> sigma_delta;
+  real<lower=0> sigma_psi;
+  real cor_12;
+  real cor_13;
+  real cor_14;
+  real cor_23;
+  real cor_24;
+  real cor_34;
+  real<lower=0,upper=5> sigma_c;
+}
+transformed data{
+  cov_matrix[4] Sigma;
+  Sigma[1,2] <- cor_12;
+  Sigma[2,1] <- cor_12;
+  Sigma[1,3] <- cor_13;
+  Sigma[3,1] <- cor_13;
+  Sigma[1,4] <- cor_14;
+  Sigma[4,1] <- cor_14;
+  Sigma[2,3] <- cor_23;
+  Sigma[3,2] <- cor_23;
+  Sigma[2,4] <- cor_24;
+  Sigma[4,2] <- cor_24;
+  Sigma[3,4] <- cor_34;
+  Sigma[4,3] <- cor_34;
+  Sigma[1,1] <- sigma_phi;
+  Sigma[2,2] <- sigma_alpha;
+  Sigma[3,3] <- sigma_alpha;
+  Sigma[4,4] <- sigma_psi;
+}
+parameters{
+  vector[4] B;         //phi, alpha, gamma by gene
+}
+
+model{
+  B    ~ multi_normal(mu,Sigma);
+  for(n in 1:N){
+    y[n] ~ neg_binomial_2_log(X[t[n]]*B + c[s[n]], 1 / exp(B[4]));
+  }
+}
+\end{verbatim}
 

### 2014-12-16 ebayes_rnaseq_heterosis.tex
self-contained
through other methods

--- 2014-12-15 ebayes_rnaseq_heterosis.tex
@@
 
-\graphicspath{{../temp/rna-seq-stan/fig-rm-stan_laplace/}}
+\graphicspath{{include/}}
 
@@
 \label{e:hypotheses}
-H_0:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
-H_1:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}.
+H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
+H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}.
 \end{align}
+We evaluate these hypotheses based on empirical Bayes estimates of their posterior probability, e.g. 
+\begin{align}
+P\left(H_{g1}|y, \hat{\phi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\phi}_{MAP}\right) \nonumber \\
+&\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right) \label{e:probs}
+\end{align}
+where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise.
+The set of probabilities $P\left(H_{g1}|y, \hat{\phi}_{MAP}\right)$ for $g=1,\ldots,G$ can be used to provide a sorted list of genes potentially involved in phenotypic heterosis.
 
-\section{Simulation study}
+\section{Simulation study based on maize experiment}
 \label{s:simulation}
 
-\subsection{Coverage for our model}
+% \subsection{Coverage for our model}
+% \input{../temp/analyze-heterosis/FIGURES/coverage}
 
-\input{../temp/analyze-heterosis/FIGURES/coverage}
+To assess the efficacy of our method to identify genes demonstrating heterosis, we used a maize data set with parental varieties B73 and Mo17 and the hybrid variety (B73 $\times$ Mo17) \citep{paschold2012complementation}.
+We compared our approach to approaches using {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes}.
 
-\subsection{Heterosis detection}
+\subsection{Construct simulated data}
+
+Initially, the maize experiment had RNAseq counts for 39,656 genes and 4 replicates per variety.
+We eliminated low count genes by removing any gene with three or more zeros for any variety and genes whose mean count across the varieties was less than one which resulted in 27,888 genes remaining.
+
+To obtain gene-variety-specific location parameters and gene-specific dispersion parameters, we used {\tt edgeR} in order to maintain existing covariance structure amongst the gene-variety means.
+We estimated sample-specific normalization factors to account for sequencing depth.
+Then, we estimated common a common dispersion for each gene, but shared across the samples.
+Finally, we obtained independent estimates for the gene-variety specific location parameters.
+The location parameters, dispersion parameters, and normalization factors were then treated as the truth for the simulation study.
+A gene with $\mu_{g3} < \min(\mu_{g1}, \mu_{g2})$ was considered to have low parent heterosis while a gene with $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$ was considered to have high parent heterosis.
+
+Using these parameters and normalization factors, we simulated data according to the negative-binomial model in equation \eqref{e:data} independently for each gene.
+For each simulation, low count genes, i.e.
+those with three or more zeros in any variety or whose mean count across the varieties was less than one, were removed.
+A random subset of 25,000 genes was retained from those with high enough counts.
+This simulation process was repeated 10 times for each of 4, 8, and 16 replicates per variety (reusing normalization factors for the higher number of replicates).
+
+\subsection{Alternative methods}
+
+Existing methods are not designed to identify gene heterosis, but we adapted the approaches in {\tt edgeR}, {\tt baySeq}.
+and {\tt ShrinkBayes} to provide a statistic representing the plausibility of heterosis.
+
+The approach used by {\tt edgeR} naturally provides testing of hypotheses for differential expression amongst two varieties \citep{robinson2007moderated, robinson2010edgeR}.
+If the hybrid's mean expression does not differ from a parent, then the hybrid does not exhibit heterosis.
+We can calculate two pvalues: one for the null hypothesis that the hybrid's mean expression is the same as parent 1 and the second for the null hypothesis that the hybrid's mean expression is the same as parent 2.
+But, even if both pvalues are small, the hybrid's mean expression could be between the two parents, i.e.
+mid-parent heterosis, and therefore not be exhibiting extreme heterosis.
+We construct a statistic that is 0 if the estimate of mean expression
+for the hybrid is between the estimates for the parents. Otherwise, the statistic is one minus the pvalue above corresponding to whichever parent's mean estimate is closer to that of the hybrid.
+
+The {\tt baySeq} approach allows for a wider range of hypotheses including $H_1: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H_2: \mu_{g1}=\mu_{g2}$, $H_3: \mu_{g1}=\mu_{g3}$, $H_4: \mu_{g2}=\mu_{g3}$, and $H_5:$ all $\mu_g$'s are distinct \citep{hardcastle2010bayseq, hardcastle2012baySeq}.
+Hypotheses $H_2$ and $H_5$ are the only two that indicate the possibility of heterosis.
+Similar to the {\tt edgeR} approach, we construct a statistic that is 0 if the estimated hybrid mean is between the parental means and otherwise is the sum of the posterior probabilities for hypotheses $H_2$ and $H_5$.
+
+The {\tt ShrinkBayes} package implements fully Bayesian inference through integrated nested Laplace approximation (INLA) \citep{rue2009approximate,van2014shrinkbayes}.
+A limitation with INLA is that it only provides approximations to marginal posteriors rather than the full posterior.
+Thus we cannot directly estimate nonlinear quantities such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2})$.
+To circumvent this limitation, we reparameterized following \cite{ji2014estimation}.
+Let $\phi_g = (\mu_{g1}+\mu_{g2})/2$ be the parental log-mean, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental difference, and $\delta_g = \mu_{g3} - \phi_g$ be the hybrid effect.
+We implemented this model using {\tt ShrinkBayes} and constructed a statistic based on posterior median of these parameters.
+Specifically, the statistic is 0 if $\left|\hat{\delta}_f  \right| \le \left| \hat{\alpha}_f \right|$, $P(\delta_f + \alpha_f > 0 |y)$ if $\hat{\delta}_f > -\hat{\alpha}_f \ge 0$, $P(\delta_f - \alpha_f > 0 |y)$ if $\hat{\delta}_f > \hat{\alpha}_f \ge 0$, $P(\delta_f - \alpha_f < 0 |y)$ if $\hat{\delta}_f < \hat{\alpha}_f \le 0$, and $P(\delta_f + \alpha_f < 0 |y)$ if $\hat{\delta}_f < -\hat{\alpha}_f \le 0$.
+
+\subsection{Results}
 
@@
 \section{}
-\subsection{Stan model for heterosis}
+
+%\input{stan_models}
 

### 2014-12-16 stan_models.tex
moved stan_models to another file

The following sections provide Stan model code for performing the optimization and MCMC described in Section \ref{s:ebayes}. In these models, it is easier to construct the model using a single index, i.e. $y_i$, and then provide vectors that provide relevant quantities for this observation. These vectors include $g$ which indicates the gene, $v$ which indicates the variety, and $s$ which gives the sample index, i.e. a distinct identifier for the variety-replicate combation.

\subsection{Optimization models}

The following two sections provide Stan model code for the optimization procedure, using the {\tt optimizing} function in RStan, either assuming independence or estimating a covariance amongst the location and (log) dispersion parameters.

\subsubsection{All genes model}
\label{s:all_genes_model}

This section provides Stan model code for the model that assumes prior independence amongst the location and dispersion parameters.

\begin{verbatim}
data{
  int N; //total # obs
  int G; // # genes
  int S; // # samples
  int<lower=0> y[N];
  int<lower=1,upper=G> g[N]; //obs -> gene
  int<lower=1,upper=3> t[N]; //obs -> genotype
  int<lower=1,upper=S> s[N]; //obs -> sample (genotype*rep)
  row_vector[3] X[3]; // mean structure (genotype)
}
parameters{
  matrix[3,G] B;         //phi, alpha, gamma by gene
  vector[G] lpsi;        //'dispersion'
  vector[S] c;           //lane sequencing depth
  real<lower=-20,upper=20> mu_phi;
  real<lower=-20,upper=20> mu_alpha;
  real<lower=-20,upper=20> mu_delta;
  real<lower=-20,upper=20> mu_psi;
  real<lower=0,upper = 20> sigma_phi;
  real<lower=0,upper = 20> sigma_alpha;
  real<lower=0,upper = 20> sigma_delta;
  real<lower=0,upper = 20> sigma_psi;
  real<lower=0,upper=5> sigma_c;
}

model{
  lpsi ~ normal(mu_psi,sigma_psi);
  B[1] ~ normal(mu_phi,sigma_phi);
  B[2] ~ normal(mu_alpha,sigma_alpha);
  B[3] ~ normal(mu_delta,sigma_delta);
  c ~ normal(0,sigma_c);
  for(n in 1:N){
    y[n] ~ neg_binomial_2_log(X[t[n]]*col(B,g[n]) + c[s[n]], 1 / exp(lpsi[g[n]]));
  }
  sigma_psi ~ cauchy(0,3);
  sigma_phi ~ cauchy(0,3);
  sigma_alpha ~ cauchy(0,3);
  sigma_delta ~ cauchy(0,3);
}
\end{verbatim}

\subsubsection{All genes model with covariance}
\label{s:all_genes_model_with_covariance}

This section provides Stan model code for the model that estimates the covariance structure amongst the location and (log) dispersion parameters.

\begin{verbatim}
data{
  int N; //total # obs
  int G; // # genes
  int S; // # samples
  int<lower=0> y[N];
  int<lower=1,upper=G> g[N]; //obs -> gene
  int<lower=1,upper=3> t[N]; //obs -> genotype
  int<lower=1,upper=S> s[N]; //obs -> sample (genotype*rep)
  row_vector[4] X[3]; // mean structure (genotype)
}
parameters{
  vector[4] B[G];         //mu1, mu2, mu3, lpsi by gene
  vector[S] c;            //lane sequencing depth
  vector<lower=-20,upper=20>[4] mu;
  cov_matrix[4] Sigma;
  real<lower=0,upper=5> sigma_c;
}

model{
  vector[4] x;
  for(i in 1:G){
    B[i] ~ multi_normal(mu, Sigma);
  }
  c ~ normal(0,sigma_c);
  for(n in 1:N){
    y[n] ~ neg_binomial_2_log(X[t[n]]*B[g[n]] + c[s[n]], 1 / exp(B[g[n],4]));
  }
  x <-rep_vector(1,4);
  Sigma ~ inv_wishart(5.0,diag_matrix(x));
}
\end{verbatim}

\subsection{Models for MCMC}

As described in Section \ref{s:ebayes}, conditional on the hyperparameters inference on the gene-specific parameters can be performed independently and in parallel. This section provides Stan model code for single-gene models either assuming independence or with a covariance structure amongst the location and (log) dispersion parameter. These model files are used in conjunction with the {\tt sampling} function in RStan.

\subsubsection{Single-gene model assuming independence}
\label{s:single_gene_model}

This section provides Stan model code for the single-gene model where independence is assumed amongst the location and (log) dispersion parameters.

\begin{verbatim}
data{
  int N; //total # obs
  int S; // # samples
  int<lower=0> y[N];
  int<lower=1,upper=3> t[N]; //obs -> genotype
  int<lower=1,upper=S> s[N]; //obs -> sample (genotype*rep)
  row_vector[3] X[3]; // mean structure (genotype)
  real<lower=-20,upper=20> mu_phi;
  real<lower=-20,upper=20> mu_alpha;
  real<lower=-20,upper=20> mu_delta;
  real<lower=-20,upper=20> mu_psi;
  real<lower=0,upper = 20> sigma_phi;
  real<lower=0,upper = 20> sigma_alpha;
  real<lower=0,upper = 20> sigma_delta;
  real<lower=0,upper = 20> sigma_psi;
  vector[S] c;            //lane sequencing depth
}
transformed data{
  vector[3] Mu;
  cov_matrix[3] Sigma;
  Mu[1] <- mu_phi;
  Mu[2] <- mu_alpha;
  Mu[3] <- mu_delta;
  Sigma[1,2] <- 0;
  Sigma[2,1] <- 0;
  Sigma[1,3] <- 0;
  Sigma[3,1] <- 0;
  Sigma[2,3] <- 0;
  Sigma[3,2] <- 0;
  Sigma[1,1] <- square(sigma_phi);
  Sigma[2,2] <- square(sigma_alpha);
  Sigma[3,3] <- square(sigma_delta);
}
parameters{
  vector[3] B;         //phi, alpha, gamma by gene
  real lpsi;       //'dispersion'
}

model{
  lpsi ~ normal(mu_psi,sigma_psi);
  B    ~ multi_normal(Mu,Sigma);
  for(n in 1:N){
    y[n] ~ neg_binomial_2_log(X[t[n]]*B + c[s[n]], 1 / exp(lpsi));
  }
}
\end{verbatim}

\subsection{Single-gene model with covariance}
\label{s:single_gene_model_with_covariance}

This section provides Stan model code for the single-gene model where a covariance amongst the location and (log) dispersion parameters is used.

\begin{verbatim}
data{
  int N; //total # obs
  int S; // # samples
  int<lower=0> y[N];
  int<lower=1,upper=3> t[N]; //obs -> genotype
  int<lower=1,upper=S> s[N]; //obs -> sample (genotype*rep)
  row_vector[4] X[3]; // mean structure (genotype)
  vector[S] c;            //lane sequencing depth
  vector<lower=-20,upper=20>[4] mu;
  real<lower=0> sigma_phi;
  real<lower=0> sigma_alpha;
  real<lower=0> sigma_delta;
  real<lower=0> sigma_psi;
  real cor_12;
  real cor_13;
  real cor_14;
  real cor_23;
  real cor_24;
  real cor_34;
  real<lower=0,upper=5> sigma_c;
}
transformed data{
  cov_matrix[4] Sigma;
  Sigma[1,2] <- cor_12;
  Sigma[2,1] <- cor_12;
  Sigma[1,3] <- cor_13;
  Sigma[3,1] <- cor_13;
  Sigma[1,4] <- cor_14;
  Sigma[4,1] <- cor_14;
  Sigma[2,3] <- cor_23;
  Sigma[3,2] <- cor_23;
  Sigma[2,4] <- cor_24;
  Sigma[4,2] <- cor_24;
  Sigma[3,4] <- cor_34;
  Sigma[4,3] <- cor_34;
  Sigma[1,1] <- sigma_phi;
  Sigma[2,2] <- sigma_alpha;
  Sigma[3,3] <- sigma_alpha;
  Sigma[4,4] <- sigma_psi;
}
parameters{
  vector[4] B;         //phi, alpha, gamma by gene
}

model{
  B    ~ multi_normal(mu,Sigma);
  for(n in 1:N){
    y[n] ~ neg_binomial_2_log(X[t[n]]*B + c[s[n]], 1 / exp(B[4]));
  }
}
\end{verbatim}

### 2014-12-16 ebayes_rnaseq_heterosis.tex
started sim results

@@
 We implemented this model using {\tt ShrinkBayes} and constructed a statistic based on posterior median of these parameters.
-Specifically, the statistic is 0 if $\left|\hat{\delta}_f  \right| \le \left| \hat{\alpha}_f \right|$, $P(\delta_f + \alpha_f > 0 |y)$ if $\hat{\delta}_f > -\hat{\alpha}_f \ge 0$, $P(\delta_f - \alpha_f > 0 |y)$ if $\hat{\delta}_f > \hat{\alpha}_f \ge 0$, $P(\delta_f - \alpha_f < 0 |y)$ if $\hat{\delta}_f < \hat{\alpha}_f \le 0$, and $P(\delta_f + \alpha_f < 0 |y)$ if $\hat{\delta}_f < -\hat{\alpha}_f \le 0$.
+Specifically, the statistic is 0 if $|\hat{\delta}_f| \le |\hat{\alpha}_f|$, $P(\delta_f + \alpha_f > 0 |y)$ if $\hat{\delta}_f > -\hat{\alpha}_f \ge 0$, $P(\delta_f - \alpha_f > 0 |y)$ if $\hat{\delta}_f > \hat{\alpha}_f \ge 0$, $P(\delta_f - \alpha_f < 0 |y)$ if $\hat{\delta}_f < \hat{\alpha}_f \le 0$, and $P(\delta_f + \alpha_f < 0 |y)$ if $\hat{\delta}_f < -\hat{\alpha}_f \le 0$.
 
@@
 
+For each of the methods above as well as the method described in Section \label{s:gene_heterosis}, we sorted genes according to the plausibility of heterosis.
+From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to identify genes with heterosis.
+A representative set of ROC curves is shown in Figure \ref{f:roc}.
 \begin{figure}[htbp]
-\centerline{\includegraphics[width=0.8\textwidth]{exampleROC0_1}}
-\caption{Example ROC curves for the first simulation}
+\centerline{\includegraphics[width=\textwidth]{exampleROC0_1}}
+\caption{Example ROC curves for the first simulation for false positive rates below 0.1.}
 \label{f:roc}
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.8\textwidth]{auc-facet-TRUE}}
+\centerline{\includegraphics[width=\textwidth]{auc-facet-TRUE}}
 \caption{Area under the ROC curves (AUC) for 3 different sample sizes (facets) for the methods under comparison for false positive rates below 0.1.

### 2014-12-16 ebayes_rnaseq_heterosis.tex
through simulation study

@@
 $\delta_{vi} \stackrel{ind}{\sim} N(0,\tau_\delta^2)$.
-The logarithm of the gene-specific over-dispersion parameters also follow a normal distribution $\log(\psi_g) \stackrel{ind}{\sim} N(\zeta_\psi,\tau_\psi^2)$.
-The variety-specific effects for each gene, $\mu_g = (\mu_{g1},\ldots,\mu_{gV})'$ are assumed to have a joint normal distribution $\mu_g \stackrel{ind}{\sim} N(\eta, \Sigma)$.
-We consider an \emph{independent} model where $\Sigma$ is diagonal with elements $\sigma_1^2,\ldots,\sigma_V^2$ and a \emph{covariance} model where we estimate all parameters in $\Sigma$.
+The logarithm of the gene-specific over-dispersion parameters and the gene-specific variety effects are assumed to have a joint normal distribution, i.e.
+$\theta_g = [\mu_g, \log(\psi_g)]^\top \stackrel{ind}{\sim} N\left(\eta, \Sigma\right)$
+where $\mu_g = (\mu_{g1},\ldots,\mu_{gV})'$.
+We consider an \emph{independent} model where $\Sigma$ is diagonal with elements $\sigma_1^2,\ldots,\sigma_{V+1}^2$ and a \emph{covariance} model where we estimate all parameters in $\Sigma$.
 
 We assume improper uniform priors for all hierarchical means and half-Cauchy priors for standard deviations \citep{gelman2006prior}, e.g.
-$\tau_\delta, \tau_\psi \stackrel{ind}{\sim} Ca^+(0,3)$ where 0 and 3 are the location and scale parameters of the untruncated distribution.
-In the independent model, we also assume $Ca^+(0,3)$ on the diagonal elements while in the covariance model, we assume a conjugate, proper inverse Wishart with degrees of freedom $V+1$ and an identity scale matrix.
+$\tau_\delta\stackrel{ind}{\sim} Ca^+(0,3)$ where 0 and 3 are the location and scale parameters of the untruncated distribution.
+In the independent model, we also assume $Ca^+(0,3)$ on the diagonal elements of $\Sigma$ while in the covariance model, we assume a conjugate, proper inverse Wishart with degrees of freedom $V+2$ and an identity scale matrix which results in marginally uniform correlations.
 
@@
 
-The parameters from the model in Section \ref{s:model} can be group by gene-specific parameters $\theta_g = (\mu_g,\psi_g)$, collectively $\theta = (\theta_1,\ldots,\theta_G)$, and hyperparameter $\phi = (\delta, \zeta_\psi, \tau_\psi^2, \Sigma)$ where $\delta = (\delta_{11},\ldots,\delta_{Vn_V})$.
-We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\phi$ and then base gene-specific inference on the posterior conditional on this estimated hyperparameter.
-Conditional on the hyperparameters, the gene specific parameters are conditionally independent as in equation \eqref{e:condind}.
-\begin{equation}
-p(\theta|y,\hat{\phi}_{MAP}) = \prod_{g=1}^G p(\theta_g|y_g,\hat{\phi}_{MAP}) \propto \prod_{g=1}^G NB(y_g|\theta_g,\hat{\delta}_{MAP})N(\theta_g|\hat{\mu}_{MAP}, \hat{Sigma}_{MAP}) 
+The parameters from the model in Section \ref{s:model} can be group by gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$, and hyperparameter $\phi = (\delta,\eta, \Sigma)$ where $\delta = (\delta_{11},\ldots,\delta_{Vn_V})$.
+We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\phi$, denoted $\hat{\phi}$, and then base gene-specific inference on the posterior conditional on this estimated hyperparameter.
+Conditional on the hyperparameters, the gene-specific parameters are conditionally independent as in equation \eqref{e:condind}.
+Thus, conditional posterior inference for the gene-specific parameters can be obtained independently and in parallel.
+\begin{align}
+p\left(\theta\left|y,\hat{\phi}\right.\right) 
+%&= \prod_{g=1}^G p\left(\theta_g\left|y_g,\hat{\phi}\right.\right) = \prod_{g=1}^G p\left(y_g\left|\theta_g,\hat{\delta}\right.\right) p\left(\theta_g\left|\hat{\phi}\right.\right) \nonumber \\
+&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\delta}_{vi}},\psi_g\right.\right) \right] N\left(\left.\left[\mu_g, \log(\psi_g) \right]^\top\right|\hat{\eta}, \hat{\Sigma} \right)
 \label{e:condind}
-\end{equation}
-Thus, conditional posterior inference for the gene specific parameters can be obtained independently and in parallel.
+\end{align}
 
@@
 The Stan software allows construction of general Bayesian models with the primary restriction that all unknown parameters must be continuous.
-Once a model is constructed, the software can be used to obtain MAP estimates \cite[see Section 50.3]{stan-manual:2014} using the {\tt optimizing} function in RStan.
-We found it necessary to use the L-BFGS algorithm \cite[see Section 55]{stan-manual:2014}.
-Alternatively, the software can be used to run a Markov chain Monte Carlo (MCMC) algorithm to obtain samples from the posterior using the {\tt sampling} function in RStan.
-By default, Stan uses a variant of Hamiltonion Monte Carlo \citep{neal2011mcmc} called the No-U-turn sampler \citep{hoffman2013no}.
+Once a model is constructed, the software can be used to obtain MAP estimates \cite[Section 50.3,][]{stan-manual:2014} using the {\tt optimizing} function in RStan.
+We found it necessary to use the L-BFGS algorithm \cite[Section 55,][]{stan-manual:2014} presumably due to the large dimension of the optimization.
+The {\tt Stan} software can also be used to run a Markov chain Monte Carlo (MCMC) algorithm to obtain samples from the posterior using the {\tt sampling} function in RStan.
+By default, Stan uses a variant of Hamiltonion Monte Carlo \citep{neal2011mcmc} called the no-U-turn sampler \citep{hoffman2013no}.
 
 As we have not integrated out the gene-specific parameters when obtaining the MAP estimate, we actually receive a joint MAP estimate for $\theta$ and $\phi$.
-We explicitly use the MAP estimate for $\phi$ when obtaining empirical Bayes posterior samples for the gene specific parameters.
+We explicitly use the MAP estimate for $\phi$ when obtaining empirical Bayes posterior samples for the gene-specific parameters as in equation \eqref{e:condind}.
 The MAP estimate for $\theta$ can also be used in this step as initial values for the MCMC.
@@
 
-\subsection{Construct simulated data}
+\subsection{Constructing simulated data}
+\label{s:sim_data}
 
@@
 Finally, we obtained independent estimates for the gene-variety specific location parameters.
-The location parameters, dispersion parameters, and normalization factors were then treated as the truth for the simulation study.
-A gene with $\mu_{g3} < \min(\mu_{g1}, \mu_{g2})$ was considered to have low parent heterosis while a gene with $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$ was considered to have high parent heterosis.
+The location parameters, dispersion parameters, and normalization factors were then treated as the truth for the simulation study, i.e.
+a gene with $\mu_{g3} < \min(\mu_{g1}, \mu_{g2})$ was considered to have LPH while a gene with $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$ was considered to have HPH.
 
@@
 those with three or more zeros in any variety or whose mean count across the varieties was less than one, were removed.
-A random subset of 25,000 genes was retained from those with high enough counts.
-This simulation process was repeated 10 times for each of 4, 8, and 16 replicates per variety (reusing normalization factors for the higher number of replicates).
+A random set of 25,000 genes was retained from those with high enough counts.
+This simulation process was repeated 10 times for each of 4, 8, and 16 replicates per variety (reusing normalization factors when necessary).
 
 \subsection{Alternative methods}
+\label{s:alternative}
 
-Existing methods are not designed to identify gene heterosis, but we adapted the approaches in {\tt edgeR}, {\tt baySeq}.
-and {\tt ShrinkBayes} to provide a statistic representing the plausibility of heterosis.
+Existing RNAseq methods are not designed to identify gene heterosis, but we adapted the approaches in {\tt edgeR}, {\tt baySeq}.
+and {\tt ShrinkBayes} to provide a statistic representing the plausibility of heterosis, i.e.
+higher values indicating higher plausibility.
 
@@
 for the hybrid is between the estimates for the parents. Otherwise, the statistic is one minus the pvalue above corresponding to whichever parent's mean estimate is closer to that of the hybrid. 
+% NEED MLE
 
@@
 Hypotheses $H_2$ and $H_5$ are the only two that indicate the possibility of heterosis.
-Similar to the {\tt edgeR} approach, we construct a statistic that is 0 if the estimated hybrid mean is between the parental means and otherwise is the sum of the posterior probabilities for hypotheses $H_2$ and $H_5$.
+Similar to the {\tt edgeR} approach, we construct a statistic that is 0 if the estimated hybrid mean, based on the MLE's from {\tt edgeR}, is between the parental means and otherwise is the sum of the posterior probabilities for hypotheses $H_2$ and $H_5$.
 
@@
 Thus we cannot directly estimate nonlinear quantities such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2})$.
-To circumvent this limitation, we reparameterized following \cite{ji2014estimation}.
+To circumvent this limitation, we reparameterized the independence model in Section \ref{s:model} following \cite{ji2014estimation}.
 Let $\phi_g = (\mu_{g1}+\mu_{g2})/2$ be the parental log-mean, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental difference, and $\delta_g = \mu_{g3} - \phi_g$ be the hybrid effect.
@@
 
-For each of the methods above as well as the method described in Section \label{s:gene_heterosis}, we sorted genes according to the plausibility of heterosis.
-From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to identify genes with heterosis.
+For each of the methods above as well as the method described in Section \ref{s:gene_heterosis}, we sorted genes according to the plausibility of heterosis.
+From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to identify genes with heterosis, as defined in Section \ref{s:sim_data}.
 A representative set of ROC curves is shown in Figure \ref{f:roc}.
@@
 \centerline{\includegraphics[width=\textwidth]{exampleROC0_1}}
-\caption{Example ROC curves for the first simulation for false positive rates below 0.1.}
+\caption{Repersentative ROC curves for false positive rates below 0.1 for the approaches using {\tt edgeR}, {\tt baySeq},  and {\tt ShrinkBayes} described in Section \ref{s:alternative} and the method using {\tt RStan} assuming independence and with a covariance described in Section \ref{s:gene_heterosis}.}
 \label{f:roc}
 \end{figure}
+For this simulation, we can see the approaches based on the model in Section \ref{s:model}, i.e.
+both RStan approaches and the ShrinkBayes approach, provide the best true positive rate for a given false positive rate.
+Also, as expected, as the sample size increases our ability to distinguish genes with heterosis from genes without improves.
 
+Figure \ref{f:auc} provides the area under the ROC below a false positive rate of 0.1 across the 10 simulations for each of the 3 different sample sizes. 
 \begin{figure}
 \centerline{\includegraphics[width=\textwidth]{auc-facet-TRUE}}
-\caption{Area under the ROC curves (AUC) for 3 different sample sizes (facets) for the methods under comparison for false positive rates below 0.1.
+\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different sample sizes for the approaches using {\tt edgeR}, {\tt baySeq},  and {\tt ShrinkBayes} described in Section \ref{s:alternative} and the method using {\tt RStan} assuming independence and with a covariance described in Section \ref{s:gene_heterosis}.
 Each line is a different simulation while the blue box indicates mean AUC and its standard error across the simulations.}
+\label{f:auc}
 \end{figure}
+The pattern appears to hold with the {\tt RStan} and {\tt ShrinkBayes} approaches outperforming the other two.
+While there does not appear to be a significant difference between the independence model using RStan or ShrinkBayes, there does appear to be a significant decreased performance, at least for sample sizes, when using the model where a covariance is estimated.
 
-\section{Heterosis in corn ??}
+\section{Heterosis in maize ??}
 \label{s:corn}

### 2014-12-16 ebayes_rnaseq_heterosis.tex
start of introduction

@@
 \begin{keywords}
-Empirical Bayes; Heterosis; Hierarchical Model; Negative binomial; RNAseq.
+Empirical Bayes; Gene expression; Heterosis; Hierarchical model; Negative binomial; RNAseq.
 \end{keywords}
@@
 
-\section{Heterosis}
-\label{s:heterosis}
+% Need to udpate as this is directly from Tieming's paper
+Heterosis, or hybrid vigor, refers to the enhances phenotype of hybrid progeny relative to their inbred parents.
+Taking maize as an example, the offspring from crossing the inbred lines B73 and Mo17 are taller, mature faster, and produce greater yields than their parent lines \citep{hallauer1981quantitative, hallauer2010quantitative}.
+Since heterosis was scientifically documented by \cite{darwin1876effects}, it has been successfully manipulated to improve many species for food, feed, and fuel industries, such as rice \citep{yu1997importance}, alfalfa \citep{riday2002heterosis}, tomatoes \citep{krieger2010flowering}, and fish \citep{wohlfarth1993heterosis}.
+Despite the intensive study and successful utilization of heterosis, the basic genomic mechanisms remain unclear \citep{coors1999genetics, lippman2007heterosis}.
+Researchers speculate that gene-expression heterosis could be among the mechanisms responsible for the phenotypic heterosis \citep{swanson2006all, springer2007allelic}.
 
-Well before heterosis was scientifically described by \cite{darwin1876effects}, humans had been using heterosis for various practical purposes.
+Recently \citep{ji2014estimation} provided an approach to assess heterosis using micro-array data using a hierarchical model for the ...
+
+%\section{Heterosis}
+%\label{s:heterosis}
+
+% Well before heterosis was scientifically described by , humans had been using heterosis for various practical purposes.
 Within the last century, heterosis has been used to improve many species for food, feed, and fuel industries.
-A few relatively recent examples include rice \cite{yu1997importance}, alfalfa \cite{riday2002heterosis}, tomatoes \cite{krieger2010flowering}, and fish \cite{wohlfarth1993heterosis}.
-Despite the intensive study and successful utilization of heterosis, the basic genomic mechanisms responsible for heterosis remain unclear \cite{lippman2007heterosis}.
-
-\cite{shull1952beginnings} coined the term heterosis and defined it as 
-\begin{quote}
-the interpretation of increased vigor, size, fruitfulness, speed of development, resistance to disease and to insect pests, or to climatic rigors of any kind, manifested by crossbred organisms as compared with corresponding inbreds, as the specific results of unlikeness in the constitutions of the uniting parental gametes.
-\end{quote}
-
-For example, the maize hybrid produced from a cross of parental inbred lines B73 and Mo17 is larger in size, has lower maturation time, and has higher grain yield than both of its parents \cite{hallauer2010quantitative}.
+A few relatively recent examples include rice , alfalfa , tomatoes .
+Despite the intensive study and successful utilization of heterosis, the basic genomic mechanisms responsible for heterosis remain unclear .
+% 
+% \cite{shull1952beginnings} coined the term heterosis and defined it as 
+% \begin{quote}
+% the interpretation of increased vigor, size, fruitfulness, speed of development, resistance to disease and to insect pests, or to climatic rigors of any kind, manifested by crossbred organisms as compared with corresponding inbreds, as the specific results of unlikeness in the constitutions of the uniting parental gametes.
+% \end{quote}
+% 
+% For example, the maize hybrid produced from a cross of parental inbred lines B73 and Mo17 is larger in size, has lower maturation time, and has higher grain yield than both of its parents \cite{hallauer2010quantitative}.
 Data from \citep{paschold2012complementation}
-
-\citep{ji2014estimation}
+% 
+% \citep{ji2014estimation}
 
@@
 The {\tt ShrinkBayes} package implements fully Bayesian inference through integrated nested Laplace approximation (INLA) \citep{rue2009approximate,van2014shrinkbayes}.
-A limitation with INLA is that it only provides approximations to marginal posteriors rather than the full posterior.
-Thus we cannot directly estimate nonlinear quantities such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2})$.
+A limitation with INLA is that it only provides approximations to marginal posteriors for parameters or linear combinations of parameters rather than an approximation to the full posterior.
+Thus, we cannot directly estimate nonlinear quantities such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2}))$.
 To circumvent this limitation, we reparameterized the independence model in Section \ref{s:model} following \cite{ji2014estimation}.
 Let $\phi_g = (\mu_{g1}+\mu_{g2})/2$ be the parental log-mean, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental difference, and $\delta_g = \mu_{g3} - \phi_g$ be the hybrid effect.
-We implemented this model using {\tt ShrinkBayes} and constructed a statistic based on posterior median of these parameters.
+We estimated this model using {\tt ShrinkBayes} and constructed a statistic based on the posterior mean of these parameters.
 Specifically, the statistic is 0 if $|\hat{\delta}_f| \le |\hat{\alpha}_f|$, $P(\delta_f + \alpha_f > 0 |y)$ if $\hat{\delta}_f > -\hat{\alpha}_f \ge 0$, $P(\delta_f - \alpha_f > 0 |y)$ if $\hat{\delta}_f > \hat{\alpha}_f \ge 0$, $P(\delta_f - \alpha_f < 0 |y)$ if $\hat{\delta}_f < \hat{\alpha}_f \le 0$, and $P(\delta_f + \alpha_f < 0 |y)$ if $\hat{\delta}_f < -\hat{\alpha}_f \le 0$.
@@
 The pattern appears to hold with the {\tt RStan} and {\tt ShrinkBayes} approaches outperforming the other two.
-While there does not appear to be a significant difference between the independence model using RStan or ShrinkBayes, there does appear to be a significant decreased performance, at least for sample sizes, when using the model where a covariance is estimated.
+While there does not appear to be a significant difference between the independence model using RStan or ShrinkBayes, there does appear to be a significant decreased performance, at least for smaller sample sizes, when using the model where a covariance is estimated.
 
@@
 The authors thank Andrew Lithio for help implementing our model in {\tt ShrinkBayes}.
+Research reported in this publication was supported by National Institute of General Medical Sciences of the National Institutes of Health under award number R01GM109458.
+The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health.
 

### 2014-12-17 ebayes_rnaseq_heterosis.tex
comment and little edit to readme

@@
 Despite the intensive study and successful utilization of heterosis, the basic genomic mechanisms remain unclear \citep{coors1999genetics, lippman2007heterosis}.
-Researchers speculate that gene-expression heterosis could be among the mechanisms responsible for the phenotypic heterosis \citep{swanson2006all, springer2007allelic}.
+Researchers speculate that gene-expression heterosis $-$ i.e, enhanced expression of one or more genes in the hybrid relative to the inbred parents $-$ could be among the mechanisms responsible for the phenotypic heterosis \citep{swanson2006all, springer2007allelic}.
+
+% WL: Since you mainly talk about phenotypic heterosis, I added a little clarification about gene-expression heterosis.
+However, the reader will still be mostly focused on phenotypic heterosis, and more may need to be done to refocus on the transcriptome.
+Do we have good real-world examples of heterosis in the expression important or interesting genes?
+This may be a good question for Dan.
 

### 2014-12-17 ebayes_rnaseq_heterosis.tex
more comments

@@
 
-% WL: Since you mainly talk about phenotypic heterosis, I added a little clarification about gene-expression heterosis.
+% WML: Since you mainly talk about phenotypic heterosis, I added a little clarification about gene-expression heterosis.
 However, the reader will still be mostly focused on phenotypic heterosis, and more may need to be done to refocus on the transcriptome.
@@
 
+% WML: Maybe it would be good to quickly clarify what a formal heterosis RNA-seq experiment is.
+
 To identify genes displaying heterosis using RNAseq data, we build a hierarchical model to borrow information on gene-variety means as well as overdispersion parameters, estimate the parameters using a empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate posterior probabilities for heterosis.
+
+% WML: "borrowing information" is intuitive and informal. I agree that the term should be used, but there might be a slightly more appropriate use. Not a big deal.
 
@@
 \label{s:model}
+
+% WML: We were debating whether to use g for gene or f for feature, and either choice would work. But we use g, we should probably also use G to denote the number of genes, rather than F.
 
@@
 
+% WML: Are the sequencing depths c_{vi} or \delta_{vi}? I think c_{vi} is better because \delta is used in the phi-alpha-delta parameterization of gene expression effects.
+
 A hierarchical model is constructed for the sequencing depth, overdispersion, and gene-variety parameters.
@@
 
+% WML: Is \mu_g the vector of \mu_{gv} values? If so, please clarify.
+
 We assume improper uniform priors for all hierarchical means and half-Cauchy priors for standard deviations \citep{gelman2006prior}, e.g.
@@
 
+% WML: I assume the "hierarchical means" are the hyperparameters in the \eta vector and the "marginally uniform correlations" are the ones derived from the off-diagonal covariance elements of \Sigma.
+Small clarifications might be helpful in passing.
+
 \subsection{Empirical Bayes}
 \label{s:ebayes}
+
+% WML: The use of \phi in this section conflicts with the phi-alpha-delta parameterization in other heterosis models.
+Maybe you could replace phi, alpha, and delta by other letters in the ShrinkBayes section to avoid this conflict.
 
@@
 
+% WML: \ref{s:intro} or \ref{s:heterosis}?
+
+% WML: the first two sentences are unclear. I would delete the first sentence altogether and compare "varieties" to treatment groups or blocks for the sake of clarity.
+
 As defined in Section \ref{s:heterosis}, heterosis is increased (or decreased) phenotypic expression in a hybrid relative to its inbred parents.
@@
 We define gene heterosis to be the differential expression of the hybrid relative to its parents.
-For a specific gene $g$, low-parent gene heteorisis (LPH) occurs when $\mu_{g3}< \min(\mu_{g1},\mu_{g2})$, mid-parent gene heterosis occurs when $\mu_{g1}\le \mu_{g3}\le \mu_{g2}$ if $\mu_{g1}\le \mu_{g2}$ (or $\mu_{g1}\ge \mu_{g3}\ge \mu_{g2}$ if $\mu_{g1}\ge \mu_{g2}$), and high-parent gene heterosis (HPH) occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$.
-For simplicity, we define \emph{extreme gene heterosis} (EH) which is the event LPH or HPH.
+For a specific gene $g$, low-parent gene heteorisis (LPH) occurs when $\mu_{g3}< \min(\mu_{g1},\mu_{g2})$,
+
+% WML: For mid-parent heterosis, you need \mu_{g3} \ne \frac{\mu_{g1} + \mu_{g2}}{2}.
+I don't know whether you need \mu_{g1}\le \mu_{g3}\le \mu_{g2} or \mu_{g2}\le \mu_{g3}\le \mu_{g1}, and I have doubts.
+It might be good to check with Dan on this.
+
+mid-parent gene heterosis occurs when $\mu_{g1}\le \mu_{g3}\le \mu_{g2}$ if $\mu_{g1}\le \mu_{g2}$ (or $\mu_{g1}\ge \mu_{g3}\ge \mu_{g2}$ if $\mu_{g1}\ge \mu_{g2}$), and high-parent gene heterosis (HPH) occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$.
+For simplicity, we define \emph{extreme gene heterosis} (EH) which is the event of LPH or HPH.
 Of interest is to evaluate the hypotheses shown in equation \eqref{e:hypotheses} where $H_0$ indicates no EH and $H_1$ indicates EH.
@@
 The set of probabilities $P\left(H_{g1}|y, \hat{\phi}_{MAP}\right)$ for $g=1,\ldots,G$ can be used to provide a sorted list of genes potentially involved in phenotypic heterosis.
+
+% WML: You may want to clarify that the genes in the list are ranked by the strength of EH and then justify the lists's utility for geneticists and in the comparison study to follow.
+This could be a good segue.
 

### 2014-12-17 ebayes_rnaseq_heterosis.tex
more comments and some changes

@@
 
+% WML: Please clarify that {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes} are R packages. You may even want to cite R somewhere.
+
 \subsection{Constructing simulated data}
@@
 Initially, the maize experiment had RNAseq counts for 39,656 genes and 4 replicates per variety.
-We eliminated low count genes by removing any gene with three or more zeros for any variety and genes whose mean count across the varieties was less than one which resulted in 27,888 genes remaining.
+We eliminated low count genes by removing any gene with three or more zeros for any variety and genes whose mean count across the varieties was less than one, which resulted in 27,888 genes remaining.
 
-To obtain gene-variety-specific location parameters and gene-specific dispersion parameters, we used {\tt edgeR} in order to maintain existing covariance structure amongst the gene-variety means.
+% WML: Sentences 2-4 in the paragraph below are redundant with sentence 1. I commented out the paragraph below and inserted a modified one afterwards
+%To obtain gene-variety-specific location parameters and gene-specific dispersion parameters, we used {\tt edgeR} in order to maintain existing covariance structure amongst the gene-variety means.
 We estimated sample-specific normalization factors to account for sequencing depth.
@@
 a gene with $\mu_{g3} < \min(\mu_{g1}, \mu_{g2})$ was considered to have LPH while a gene with $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$ was considered to have HPH.
+
+To begin, we applied the {\tt edgeR} package from Bioconductor (citation needed) to the maize data to obtain sample-specific normalization factors, then gene-specific dispersions, and finally gene-variety-specific location parameters.
+We treated all parameters as truth for the simulation study:  i.e.
+a gene with $\mu_{g3} < \min(\mu_{g1}, \mu_{g2})$ was considered to have LPH, while a gene with $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$ was considered to have HPH.
+This technique ensured that our simulated datasets maintained the existing covariance structure among the gene-variety means of the original maize data.
 
@@
 
-Existing RNAseq methods are not designed to identify gene heterosis, but we adapted the approaches in {\tt edgeR}, {\tt baySeq}.
-and {\tt ShrinkBayes} to provide a statistic representing the plausibility of heterosis, i.e.
-higher values indicating higher plausibility.
+Existing RNAseq data analysis methods are not specifically designed to identify gene heterosis, but we adapted the approaches in {\tt edgeR}, {\tt baySeq}.
+and {\tt ShrinkBayes} to provide statistics to measure the strength of heterosis of each gene.
 
-The approach used by {\tt edgeR} naturally provides testing of hypotheses for differential expression amongst two varieties \citep{robinson2007moderated, robinson2010edgeR}.
+% WML: "...
+provide a statistic representing the plausibility of heterosis, i.e.
+higher values indicating higher plausibility." => "...provide statistics to measure the strength of heterosis of each gene."
+
+% WML: change: "edgeR was originally designed to..."
+
+{\tt edgeR} was originally designed to test for differential expression between two varieties \citep{robinson2007moderated, robinson2010edgeR}, and we used this capability to construct a statistic for each gene.
 If the hybrid's mean expression does not differ from a parent, then the hybrid does not exhibit heterosis.
@@
 for the hybrid is between the estimates for the parents. Otherwise, the statistic is one minus the pvalue above corresponding to whichever parent's mean estimate is closer to that of the hybrid.
+
 % NEED MLE

### 2014-12-17 ebayes_rnaseq_heterosis.tex
I rewrote the paragraphs on edgeR and baySeq.

@@
 
-% WML: change: "edgeR was originally designed to..."
+% WML: I rewrote the paragraph on edgeR.
 
-{\tt edgeR} was originally designed to test for differential expression between two varieties \citep{robinson2007moderated, robinson2010edgeR}, and we used this capability to construct a statistic for each gene.
-If the hybrid's mean expression does not differ from a parent, then the hybrid does not exhibit heterosis.
-We can calculate two pvalues: one for the null hypothesis that the hybrid's mean expression is the same as parent 1 and the second for the null hypothesis that the hybrid's mean expression is the same as parent 2.
-But, even if both pvalues are small, the hybrid's mean expression could be between the two parents, i.e.
-mid-parent heterosis, and therefore not be exhibiting extreme heterosis.
-We construct a statistic that is 0 if the estimate of mean expression
-for the hybrid is between the estimates for the parents. Otherwise, the statistic is one minus the pvalue above corresponding to whichever parent's mean estimate is closer to that of the hybrid.
+{\tt edgeR} is designed to apply a negative binomial loglinear model to test for differential expression between any two varieties \citep{robinson2007moderated, robinson2010edgeR}, and we used this capability to construct a pseudo-probability of heterosis for each gene.
+We first computed the maximum likelihood estimates of the mean expression effects of all genes using {\tt edgeR}'s built-in Fisher scoring algorithm, and then we used the likelihood ratio test functionality to calculate two legitimate p-values for each gene: one for the null hypothesis that the hybrid's mean expression is the same as that of parent 1 and the second for the analogous test involving parent 2.
+For each gene, if the estimated mean expression level of the hybrid is not between the estimated mean expression levels of the parents, we set the pseudo-probability of heterosis to be one minus the p-value corresponding to the parent with estimated expression level closer to that of the hybrid.
+Otherwise, the pseudo-probability is zero.
 
@@
 
-The {\tt baySeq} approach allows for a wider range of hypotheses including $H_1: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H_2: \mu_{g1}=\mu_{g2}$, $H_3: \mu_{g1}=\mu_{g3}$, $H_4: \mu_{g2}=\mu_{g3}$, and $H_5:$ all $\mu_g$'s are distinct \citep{hardcastle2010bayseq, hardcastle2012baySeq}.
-Hypotheses $H_2$ and $H_5$ are the only two that indicate the possibility of heterosis.
-Similar to the {\tt edgeR} approach, we construct a statistic that is 0 if the estimated hybrid mean, based on the MLE's from {\tt edgeR}, is between the parental means and otherwise is the sum of the posterior probabilities for hypotheses $H_2$ and $H_5$.
+% WML: I rewrote the baySeq paragraph.
+
+The {\tt baySeq} approach allows for a wider range of hypotheses for each gene, including $H_1: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H_2: \mu_{g1}=\mu_{g2}$, $H_3: \mu_{g1}=\mu_{g3}$, $H_4: \mu_{g2}=\mu_{g3}$, and $H_5:$ all $\mu_g$'s are distinct \citep{hardcastle2010bayseq, hardcastle2012baySeq}.
+In a technique similar to our application of {\tt edgeR}, we used used the posterior probabilities of these hypotheses to construct a pseudo-probability of heterosis for each gene.
+We set this pseudo-probability to zero if the hybrid's maximum likelihood estimate of the mean expression level calculated with {\tt edgeR} is between those of the parents.
+Otherwise, the pseudo-probability is the sum of the posterior probabilities of $H_2$ and $H_5$.
+Note that $H_2$ and $H_5$ are the only hypotheses that allow for heterosis.
 

### 2014-12-17 ebayes_rnaseq_heterosis.tex
rewrote part of the ShrinkBayes paragraph and added comments

@@
 The {\tt ShrinkBayes} package implements fully Bayesian inference through integrated nested Laplace approximation (INLA) \citep{rue2009approximate,van2014shrinkbayes}.
-A limitation with INLA is that it only provides approximations to marginal posteriors for parameters or linear combinations of parameters rather than an approximation to the full posterior.
+A limitation of INLA is that it only provides approximations to marginal posteriors for parameters or linear combinations of parameters rather than an approximation to the full posterior.
 Thus, we cannot directly estimate nonlinear quantities such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2}))$.
 To circumvent this limitation, we reparameterized the independence model in Section \ref{s:model} following \cite{ji2014estimation}.
+
+% WML: your use of "following" in this last sentence is not clear to me.
+
+% WML: Here, you use the phi-alpha-delta parameterization.
+You already defined phi and delta in previous sections.
+To avoid a conflict, you could easily change the notation here.
+It doesn't even have to be super intuitive because you only use it in this one paragraph.
+
  Let $\phi_g = (\mu_{g1}+\mu_{g2})/2$ be the parental log-mean, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental difference, and $\delta_g = \mu_{g3} - \phi_g$ be the hybrid effect.
-We estimated this model using {\tt ShrinkBayes} and constructed a statistic based on the posterior mean of these parameters.
-Specifically, the statistic is 0 if $|\hat{\delta}_f| \le |\hat{\alpha}_f|$, $P(\delta_f + \alpha_f > 0 |y)$ if $\hat{\delta}_f > -\hat{\alpha}_f \ge 0$, $P(\delta_f - \alpha_f > 0 |y)$ if $\hat{\delta}_f > \hat{\alpha}_f \ge 0$, $P(\delta_f - \alpha_f < 0 |y)$ if $\hat{\delta}_f < \hat{\alpha}_f \le 0$, and $P(\delta_f + \alpha_f < 0 |y)$ if $\hat{\delta}_f < -\hat{\alpha}_f \le 0$.
+After fitting the model with {\tt ShrinkBayes}, we constructed a pseudo-probability of heterosis for each gene.
+For gene $g$, we used the estimated posterior means $\hat{\delta}_g$ and $\hat{\alpha}_g$ of $\delta_g$ and $\alpha_g$, respectively, along with appropriate posterior probabilities $P(\cdot \mid y)$.
+Specifically, the pseudo-probability of heterosis for gene $g$ is zero if $|\hat{\delta}_f| \le |\hat{\alpha}_f|$, $P(\delta_f + \alpha_f > 0 |y)$ if $\hat{\delta}_f > -\hat{\alpha}_f \ge 0$, $P(\delta_f - \alpha_f > 0 |y)$ if $\hat{\delta}_f > \hat{\alpha}_f \ge 0$, $P(\delta_f - \alpha_f < 0 |y)$ if $\hat{\delta}_f < \hat{\alpha}_f \le 0$, and $P(\delta_f + \alpha_f < 0 |y)$ if $\hat{\delta}_f < -\hat{\alpha}_f \le 0$.
 

### 2014-12-17 data/for_supp.R
created (most of) the supplementary materials

@@ -1,0 +1,17 @@
+d = read.table("rna_seq.txt", header=TRUE)
+
+d = d[,c(1, grep("total", names(d)))]
+
+names(d)[-1] = paste(c("B73","Mo17","B73xMo17","Mo17xB73"), 
+                     rep(1:4, each=4), 
+                     sep="_")
+
+# Order columns
+B_cols = c(2,6,10,14)
+d = d[,c(1, B_cols, B_cols+1, B_cols+2, B_cols+3)]
+
+# Assuming we are using the BM cross
+d = d[,1:13]
+
+# Write file
+write.csv(d, file="../supp/data.csv", row.names=FALSE)

### 2014-12-17 supp/script.R

@@ -1,0 +1,2 @@
+# We need an R script to perform the analysis
+

### 2014-12-17 ebayes_rnaseq_heterosis.tex
finished a round of specific comments
WML: first round of edits complete.
created (most of) the supplementary materials
post will's comments

@@
 
-% WML: "borrowing information" is intuitive and informal. I agree that the term should be used, but there might be a slightly more appropriate use. Not a big deal.
-
 \subsection{Hierarchical model for RNAseq counts}
@@
 
-% WML: We were debating whether to use g for gene or f for feature, and either choice would work. But we use g, we should probably also use G to denote the number of genes, rather than F.
-
-Let $Y_{gvi}$ be the count for gene $g=1,\ldots,F$, variety $v=1,\ldots,V$, and replicate $i=1,\ldots,n_v$.
+Let $Y_{gvi}$ be the count for gene $g=1,\ldots,G$, variety $v=1,\ldots,V$, and replicate $i=1,\ldots,n_v$.
 As shown in equation \eqref{e:data}, we assume a negative binomial for the counts with a mean that depends on the gene-variety combination through $\mu_{gv}$ and the sequencing depth $c_{vi}$ and a gene-specific over-dispersion $\psi_g$.
@@
 \end{equation}
-
-% WML: Are the sequencing depths c_{vi} or \delta_{vi}? I think c_{vi} is better because \delta is used in the phi-alpha-delta parameterization of gene expression effects.
 
@@
 
-% WML: Is \mu_g the vector of \mu_{gv} values? If so, please clarify.
-
-We assume improper uniform priors for all hierarchical means and half-Cauchy priors for standard deviations \citep{gelman2006prior}, e.g.
-$\tau_\delta\stackrel{ind}{\sim} Ca^+(0,3)$ where 0 and 3 are the location and scale parameters of the untruncated distribution.
-In the independent model, we also assume $Ca^+(0,3)$ on the diagonal elements of $\Sigma$ while in the covariance model, we assume a conjugate, proper inverse Wishart with degrees of freedom $V+2$ and an identity scale matrix which results in marginally uniform correlations.
-
-% WML: I assume the "hierarchical means" are the hyperparameters in the \eta vector and the "marginally uniform correlations" are the ones derived from the off-diagonal covariance elements of \Sigma.
-Small clarifications might be helpful in passing.
+We assume improper uniform priors for $\eta$ and $\tau_\delta\stackrel{ind}{\sim} Ca^+(0,3)$ where 0 and 3 are the location and scale parameters of the untruncated distribution.
+In the independent model, we also $Ca^+(0,3)$ on the diagonal elements of $\Sigma$ while in the covariance model, we assume a conjugate, proper inverse Wishart with degrees of freedom $V+2$ and an identity scale matrix which results in marginally uniform correlations.
 
@@
 \label{s:ebayes}
-
-% WML: The use of \phi in this section conflicts with the phi-alpha-delta parameterization in other heterosis models.
-Maybe you could replace phi, alpha, and delta by other letters in the ShrinkBayes section to avoid this conflict.
 
@@
 
-% WML: \ref{s:intro} or \ref{s:heterosis}?
-
-% WML: the first two sentences are unclear. I would delete the first sentence altogether and compare "varieties" to treatment groups or blocks for the sake of clarity.
-
-As defined in Section \ref{s:heterosis}, heterosis is increased (or decreased) phenotypic expression in a hybrid relative to its inbred parents.
+As defined in Section \ref{s:intro}, heterosis is increased (or decreased) phenotypic expression in a hybrid relative to its inbred parents.
 Thus, when attempting to identify heterosis, there are three varieties of interest, i.e.
@@
 We define gene heterosis to be the differential expression of the hybrid relative to its parents.
-For a specific gene $g$, low-parent gene heteorisis (LPH) occurs when $\mu_{g3}< \min(\mu_{g1},\mu_{g2})$,
-
-% WML: For mid-parent heterosis, you need \mu_{g3} \ne \frac{\mu_{g1} + \mu_{g2}}{2}.
-I don't know whether you need \mu_{g1}\le \mu_{g3}\le \mu_{g2} or \mu_{g2}\le \mu_{g3}\le \mu_{g1}, and I have doubts.
-It might be good to check with Dan on this.
-
-mid-parent gene heterosis occurs when $\mu_{g1}\le \mu_{g3}\le \mu_{g2}$ if $\mu_{g1}\le \mu_{g2}$ (or $\mu_{g1}\ge \mu_{g3}\ge \mu_{g2}$ if $\mu_{g1}\ge \mu_{g2}$), and high-parent gene heterosis (HPH) occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$.
+For a specific gene $g$, low-parent gene heteorisis (LPH) occurs when $\mu_{g3}< \min(\mu_{g1},\mu_{g2})$, mid-parent gene heterosis occurs when $\mu_{g1}\le \mu_{g3}\le \mu_{g2}$ if $\mu_{g1}\le \mu_{g2}$ (or $\mu_{g1}\ge \mu_{g3}\ge \mu_{g2}$ if $\mu_{g1}\ge \mu_{g2}$), and high-parent gene heterosis (HPH) occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$.
 For simplicity, we define \emph{extreme gene heterosis} (EH) which is the event of LPH or HPH.
@@
 where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise.
-The set of probabilities $P\left(H_{g1}|y, \hat{\phi}_{MAP}\right)$ for $g=1,\ldots,G$ can be used to provide a sorted list of genes potentially involved in phenotypic heterosis.
-
-% WML: You may want to clarify that the genes in the list are ranked by the strength of EH and then justify the lists's utility for geneticists and in the comparison study to follow.
-This could be a good segue.
+The set of probabilities $P\left(H_{g1}|y, \hat{\phi}_{MAP}\right)$ for $g=1,\ldots,G$ can be used to provide a list of genes sorted by the relative strength of evidence for extreme heterosis.
+This list can be used by geneticists to prioritize future experiments to understand the genetic mechanism for heterosis.
 
@@
 To assess the efficacy of our method to identify genes demonstrating heterosis, we used a maize data set with parental varieties B73 and Mo17 and the hybrid variety (B73 $\times$ Mo17) \citep{paschold2012complementation}.
-We compared our approach to approaches using {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes}.
-
-% WML: Please clarify that {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes} are R packages. You may even want to cite R somewhere.
+We compared our method to approaches using the R packages {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes}.
 
@@
 
-% WML: Sentences 2-4 in the paragraph below are redundant with sentence 1. I commented out the paragraph below and inserted a modified one afterwards
-%To obtain gene-variety-specific location parameters and gene-specific dispersion parameters, we used {\tt edgeR} in order to maintain existing covariance structure amongst the gene-variety means.
-We estimated sample-specific normalization factors to account for sequencing depth.
-Then, we estimated common a common dispersion for each gene, but shared across the samples.
-Finally, we obtained independent estimates for the gene-variety specific location parameters.
-The location parameters, dispersion parameters, and normalization factors were then treated as the truth for the simulation study, i.e.
-a gene with $\mu_{g3} < \min(\mu_{g1}, \mu_{g2})$ was considered to have LPH while a gene with $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$ was considered to have HPH.
-
 To begin, we applied the {\tt edgeR} package from Bioconductor (citation needed) to the maize data to obtain sample-specific normalization factors, then gene-specific dispersions, and finally gene-variety-specific location parameters.
-We treated all parameters as truth for the simulation study:  i.e.
+We treated these parameters as the truth for the simulation study: i.e.
 a gene with $\mu_{g3} < \min(\mu_{g1}, \mu_{g2})$ was considered to have LPH, while a gene with $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$ was considered to have HPH.
@@
 Existing RNAseq data analysis methods are not specifically designed to identify gene heterosis, but we adapted the approaches in {\tt edgeR}, {\tt baySeq}.
-and {\tt ShrinkBayes} to provide statistics to measure the strength of heterosis of each gene.
-
-% WML: "...
-provide a statistic representing the plausibility of heterosis, i.e.
-higher values indicating higher plausibility." => "...provide statistics to measure the strength of heterosis of each gene."
-
-% WML: I rewrote the paragraph on edgeR.
+and {\tt ShrinkBayes} to provide statistics to measure the strength of extreme heterosis of each gene.
 
@@
 Otherwise, the pseudo-probability is zero.
-
-% NEED MLE
-
-% WML: I rewrote the baySeq paragraph.
 
@@
 Thus, we cannot directly estimate nonlinear quantities such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2}))$.
-To circumvent this limitation, we reparameterized the independence model in Section \ref{s:model} following \cite{ji2014estimation}.
-
-% WML: your use of "following" in this last sentence is not clear to me.
-
-% WML: Here, you use the phi-alpha-delta parameterization.
-You already defined phi and delta in previous sections.
-To avoid a conflict, you could easily change the notation here.
-It doesn't even have to be super intuitive because you only use it in this one paragraph.
-
- Let $\phi_g = (\mu_{g1}+\mu_{g2})/2$ be the parental log-mean, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental difference, and $\delta_g = \mu_{g3} - \phi_g$ be the hybrid effect.
+To circumvent this limitation, we reparameterized the independence model in Section \ref{s:model} using the approach in \cite{ji2014estimation}.
+Let $\gamma_g = (\mu_{g1}+\mu_{g2})/2$ be the parental log-mean, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental difference, and $\zeta_g = \mu_{g3} - \gamma_g$ be the hybrid effect.
 After fitting the model with {\tt ShrinkBayes}, we constructed a pseudo-probability of heterosis for each gene.
-For gene $g$, we used the estimated posterior means $\hat{\delta}_g$ and $\hat{\alpha}_g$ of $\delta_g$ and $\alpha_g$, respectively, along with appropriate posterior probabilities $P(\cdot \mid y)$.
-Specifically, the pseudo-probability of heterosis for gene $g$ is zero if $|\hat{\delta}_f| \le |\hat{\alpha}_f|$, $P(\delta_f + \alpha_f > 0 |y)$ if $\hat{\delta}_f > -\hat{\alpha}_f \ge 0$, $P(\delta_f - \alpha_f > 0 |y)$ if $\hat{\delta}_f > \hat{\alpha}_f \ge 0$, $P(\delta_f - \alpha_f < 0 |y)$ if $\hat{\delta}_f < \hat{\alpha}_f \le 0$, and $P(\delta_f + \alpha_f < 0 |y)$ if $\hat{\delta}_f < -\hat{\alpha}_f \le 0$.
+For gene $g$, we used the estimated posterior means $\hat{\zeta}_g$ and $\hat{\alpha}_g$ of $\zeta_g$ and $\alpha_g$, respectively, along with appropriate posterior probabilities $P(\cdot \mid y)$.
+Specifically, the pseudo-probability of heterosis for gene $g$ is zero if $|\hat{\zeta}_g| \le |\hat{\alpha}_g|$, $P(\zeta_g + \alpha_g > 0 |y)$ if $\hat{\zeta}_g > -\hat{\alpha}_g \ge 0$, $P(\zeta_g - \alpha_g > 0 |y)$ if $\hat{\zeta}_g > \hat{\alpha}_g \ge 0$, $P(\zeta_g - \alpha_g < 0 |y)$ if $\hat{\zeta}_g < \hat{\alpha}_g \le 0$, and $P(\zeta_g + \alpha_g < 0 |y)$ if $\hat{\zeta}_g < -\hat{\alpha}_g \le 0$.
 
@@
 For each of the methods above as well as the method described in Section \ref{s:gene_heterosis}, we sorted genes according to the plausibility of heterosis.
-From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to identify genes with heterosis, as defined in Section \ref{s:sim_data}.
+From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to identify genes with heterosis, as defined in Section \ref{s:sim_data}, \citep{landau2013dispersion}.
 A representative set of ROC curves is shown in Figure \ref{f:roc}.
@@
 
-%\input{stan_models}
-
 \label{lastpage}

### 2014-12-17 ebayes_rnaseq_heterosis.tex
include placeholder real data figs

@@
 To identify genes displaying heterosis using RNAseq data, we build a hierarchical model to borrow information on gene-variety means as well as overdispersion parameters, estimate the parameters using a empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate posterior probabilities for heterosis.
+
+\subsection{RNAseq experiment to identify genes demonstrating heterosis}
+\label{s:experiment}
 
@@
 
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{maize_volcano}}
+\caption{A two-dimensional histogram of the probability of heterosis versus effect size, $\mu_{g3}-\max(\mu_{g1},\mu_{g3})$ if $P(HPH|y)>P(LPH|y)$ or $\mu_{g3}-\min(\mu_{g1},\mu_{g2})$ if $P(HPH|y)<P(LPH|y)$ for the B73 $\times$ Mo17 maize experiment.}
+\label{f:volcano}
+\end{figure}
+
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{normality}}
+\caption{One and two-dimensional histograms for gene-specific parameters from a single MCMC iteration, $\theta_g^{(m)}$,  for the B73 $\times$ Mo17 maize experiment.
+For reference, the one dimensional histograms have the marginal probability density functions from $N(\hat{\eta},\hat{\Sigma})$.}
+\label{f:normality}
+\end{figure}
+
 \section{Discussion}

### 2014-12-17 ebayes_rnaseq_heterosis.tex
start of experiment

@@
 
-% WML: Maybe it would be good to quickly clarify what a formal heterosis RNA-seq experiment is.
-
 To identify genes displaying heterosis using RNAseq data, we build a hierarchical model to borrow information on gene-variety means as well as overdispersion parameters, estimate the parameters using a empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate posterior probabilities for heterosis.
@@
 \label{s:experiment}
+
+An RNAseq experiment to identify genes demonstrating heterosis consists of three genetic varieties: two parental varieties (P1 and P2) and a cross between these two varieties called the hybrid (H).
+For each variety, replicate RNA samples are isolated and assessed for quality.
+cDNA libraries are constructed and the samples are sequenced.
+These libraries are sequenced
+
+loaded on flow cells, and sequenced. The raw sequencing reads
+
+\cite{nettleton2014design} and \cite{paschold2012complementation}
 

### 2014-12-17 ebayes_rnaseq_heterosis.tex
updated plots

@@
 \section{Empirical Bayes identification of gene heterosis from RNAseq counts}
-
-To identify genes displaying heterosis using RNAseq data, we build a hierarchical model to borrow information on gene-variety means as well as overdispersion parameters, estimate the parameters using a empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate posterior probabilities for heterosis.
-
-\subsection{RNAseq experiment to identify genes demonstrating heterosis}
-\label{s:experiment}
+\label{s:method}
 
@@
 For each variety, replicate RNA samples are isolated and assessed for quality.
-cDNA libraries are constructed and the samples are sequenced.
-These libraries are sequenced
+cDNA libraries, i.e.
+short cDNA fragments, are constructed and next generation sequencing technology is used to determine the \emph{reads}, i.e.
+nucleotide sequences in the cDNA libraries.
+These reads are processed using bioinformatic algorithms to match the reads to genes, or specific gene transcripts, exons, microRNAs, etc.
+The result of matching reads to features are summarized by a gene $\times$ sample matrix of counts.
+For more details on RNAseq experiments see \cite{nettleton2014design} and for biological background on the details of these steps see \cite{paschold2012complementation}.
 
-loaded on flow cells, and sequenced. The raw sequencing reads
-
-\cite{nettleton2014design} and \cite{paschold2012complementation}
+To identify genes displaying heterosis using RNAseq counts, we build a hierarchical model to borrow information on gene-variety means as well as overdispersion parameters, estimate the parameters using a empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate empirical Bayes posterior probabilities for heterosis.
 
@@
 
+We calculate an estimated effect size using the posterior means for the gene-specific parameters using the formula 
+\[
+\mbox{estimated effect size} = \left\{ 
+\begin{array}{ll}
+\hat{\mu}_{g3} - \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} < \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
+\hat{\mu}_{g3} - \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} > \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
+0 & \mbox{otherwise}.
+\end{array} 
+\right. 
+\]
+Figure \ref{f:volcano} provides a volcano plot to visualize the probability of heterosis versus estimated effect size.
+
 \begin{figure}
-\centerline{\includegraphics[width=\textwidth]{maize_volcano}}
-\caption{A two-dimensional histogram of the probability of heterosis versus effect size, $\mu_{g3}-\max(\mu_{g1},\mu_{g3})$ if $P(HPH|y)>P(LPH|y)$ or $\mu_{g3}-\min(\mu_{g1},\mu_{g2})$ if $P(HPH|y)<P(LPH|y)$ for the B73 $\times$ Mo17 maize experiment.}
+\centerline{\includegraphics[width=\textwidth]{volcano}}
+\caption{A two-dimensional histogram of the maximum of LPH and HPH probabilities versus estimated effect size, $\mu_{g3}-\max(\mu_{g1},\mu_{g3})$ if $P(HPH|y)>P(LPH|y)$ or $\mu_{g3}-\min(\mu_{g1},\mu_{g2})$ if $P(HPH|y)<P(LPH|y)$ for the B73 $\times$ Mo17 maize experiment.}
 \label{f:volcano}

### 2014-12-17 ebayes_rnaseq_heterosis.tex
changed notation

@@
 
-To identify genes displaying heterosis using RNAseq counts, we build a hierarchical model to borrow information on gene-variety means as well as overdispersion parameters, estimate the parameters using a empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate empirical Bayes posterior probabilities for heterosis.
+To identify genes displaying heterosis using RNAseq counts, we build a hierarchical model to borrow information on gene-variety means and gene-specific overdispersion parameters, estimate the parameters using a empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate empirical Bayes posterior probabilities for heterosis.
 
@@
 As shown in equation \eqref{e:data}, we assume a negative binomial for the counts with a mean that depends on the gene-variety combination through $\mu_{gv}$ and the sequencing depth $c_{vi}$ and a gene-specific over-dispersion $\psi_g$.
-Here $NB(\eta,\psi)$ indicates a negative-binomial with expectation $\eta$ and variance $\eta+\psi\eta^2$ and $ind$ indicates the observations are conditionally independent.
 \begin{equation} 
-Y_{gvi} \stackrel{ind}{\sim} NB(e^{\mu_{gv}+\delta_{vi}},\psi_g) 
+Y_{gvi} \stackrel{ind}{\sim} NB(e^{\mu_{gv}+\gamma_{vi}},\psi_g) 
 \label{e:data}
 \end{equation}
+where $NB(\eta,\psi)$ indicates a negative-binomial with expectation $\eta$ and variance $\eta+\psi\eta^2$ and $ind$ indicates the observations are conditionally independent.
 
@@
 The sample-specific sequencing depth terms are assumed to follow a normal distribution, i.e.
-$\delta_{vi} \stackrel{ind}{\sim} N(0,\tau_\delta^2)$.
-The logarithm of the gene-specific over-dispersion parameters and the gene-specific variety effects are assumed to have a joint normal distribution, i.e.
+$\gamma_{vi} \stackrel{ind}{\sim} N(0,\tau_\gamma^2)$.
+(These terms are the normalization factors that are typically estimated from each sample and then plugged in for further analysis.) The logarithm of the gene-specific over-dispersion parameters and the gene-specific variety effects are assumed to have a joint normal distribution, i.e.
 $\theta_g = [\mu_g, \log(\psi_g)]^\top \stackrel{ind}{\sim} N\left(\eta, \Sigma\right)$
@@
 
-We assume improper uniform priors for $\eta$ and $\tau_\delta\stackrel{ind}{\sim} Ca^+(0,3)$ where 0 and 3 are the location and scale parameters of the untruncated distribution.
+We assume improper uniform priors for $\eta$ and $\tau_\gamma\stackrel{ind}{\sim} Ca^+(0,3)$ where 0 and 3 are the location and scale parameters of the untruncated distribution.
 In the independent model, we also $Ca^+(0,3)$ on the diagonal elements of $\Sigma$ while in the covariance model, we assume a conjugate, proper inverse Wishart with degrees of freedom $V+2$ and an identity scale matrix which results in marginally uniform correlations.
@@
 
-The parameters from the model in Section \ref{s:model} can be group by gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$, and hyperparameter $\phi = (\delta,\eta, \Sigma)$ where $\delta = (\delta_{11},\ldots,\delta_{Vn_V})$.
-We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\phi$, denoted $\hat{\phi}$, and then base gene-specific inference on the posterior conditional on this estimated hyperparameter.
+The parameters from the model in Section \ref{s:model} can be group by gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$, and hyperparameter $\pi = (\gamma,\eta, \Sigma)$ where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$.
+We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\pi$, denoted $\hat{\pi}$, and then base gene-specific inference on the posterior conditional on this estimated hyperparameter.
 Conditional on the hyperparameters, the gene-specific parameters are conditionally independent as in equation \eqref{e:condind}.
@@
 \begin{align}
-p\left(\theta\left|y,\hat{\phi}\right.\right) 
-%&= \prod_{g=1}^G p\left(\theta_g\left|y_g,\hat{\phi}\right.\right) = \prod_{g=1}^G p\left(y_g\left|\theta_g,\hat{\delta}\right.\right) p\left(\theta_g\left|\hat{\phi}\right.\right) \nonumber \\
-&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\delta}_{vi}},\psi_g\right.\right) \right] N\left(\left.\left[\mu_g, \log(\psi_g) \right]^\top\right|\hat{\eta}, \hat{\Sigma} \right)
+p\left(\theta\left|y,\hat{\pi}\right.\right) 
+%&= \prod_{g=1}^G p\left(\theta_g\left|y_g,\hat{\pi}\right.\right) = \prod_{g=1}^G p\left(y_g\left|\theta_g,\hat{\gamma}\right.\right) p\left(\theta_g\left|\hat{\pi}\right.\right) \nonumber \\
+&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\gamma}_{vi}},\psi_g\right.\right) \right] N\left(\left.\left[\mu_g, \log(\psi_g) \right]^\top\right|\hat{\eta}, \hat{\Sigma} \right)
 \label{e:condind}
@@
 
-As we have not integrated out the gene-specific parameters when obtaining the MAP estimate, we actually receive a joint MAP estimate for $\theta$ and $\phi$.
-We explicitly use the MAP estimate for $\phi$ when obtaining empirical Bayes posterior samples for the gene-specific parameters as in equation \eqref{e:condind}.
+As we have not integrated out the gene-specific parameters when obtaining the MAP estimate, we actually receive a joint MAP estimate for $\theta$ and $\pi$.
+We explicitly use the MAP estimate for $\pi$ when obtaining empirical Bayes posterior samples for the gene-specific parameters as in equation \eqref{e:condind}.
 The MAP estimate for $\theta$ can also be used in this step as initial values for the MCMC.
@@
 \begin{align}
-P\left(H_{g1}|y, \hat{\phi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\phi}_{MAP}\right) \nonumber \\
+P\left(H_{g1}|y, \hat{\pi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\pi}_{MAP}\right) \nonumber \\
 &\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right) \label{e:probs}
@@
 where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise.
-The set of probabilities $P\left(H_{g1}|y, \hat{\phi}_{MAP}\right)$ for $g=1,\ldots,G$ can be used to provide a list of genes sorted by the relative strength of evidence for extreme heterosis.
+The set of probabilities $P\left(H_{g1}|y, \hat{\pi}_{MAP}\right)$ for $g=1,\ldots,G$ can be used to provide a list of genes sorted by the relative strength of evidence for extreme heterosis.
 This list can be used by geneticists to prioritize future experiments to understand the genetic mechanism for heterosis.
@@
 To circumvent this limitation, we reparameterized the independence model in Section \ref{s:model} using the approach in \cite{ji2014estimation}.
-Let $\gamma_g = (\mu_{g1}+\mu_{g2})/2$ be the parental log-mean, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental difference, and $\zeta_g = \mu_{g3} - \gamma_g$ be the hybrid effect.
+Let $\phi_g = (\mu_{g1}+\mu_{g2})/2$ be the parental log-mean, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental difference, and $\delta_g = \mu_{g3} - \phi_g$ be the hybrid effect.
 After fitting the model with {\tt ShrinkBayes}, we constructed a pseudo-probability of heterosis for each gene.
-For gene $g$, we used the estimated posterior means $\hat{\zeta}_g$ and $\hat{\alpha}_g$ of $\zeta_g$ and $\alpha_g$, respectively, along with appropriate posterior probabilities $P(\cdot \mid y)$.
-Specifically, the pseudo-probability of heterosis for gene $g$ is zero if $|\hat{\zeta}_g| \le |\hat{\alpha}_g|$, $P(\zeta_g + \alpha_g > 0 |y)$ if $\hat{\zeta}_g > -\hat{\alpha}_g \ge 0$, $P(\zeta_g - \alpha_g > 0 |y)$ if $\hat{\zeta}_g > \hat{\alpha}_g \ge 0$, $P(\zeta_g - \alpha_g < 0 |y)$ if $\hat{\zeta}_g < \hat{\alpha}_g \le 0$, and $P(\zeta_g + \alpha_g < 0 |y)$ if $\hat{\zeta}_g < -\hat{\alpha}_g \le 0$.
+For gene $g$, we used the estimated posterior means $\hat{\delta}_g$ and $\hat{\alpha}_g$ of $\delta_g$ and $\alpha_g$, respectively, along with appropriate posterior probabilities $P(\cdot \mid y)$.
+Specifically, the pseudo-probability of heterosis for gene $g$ is zero if $|\hat{\delta}_g| \le |\hat{\alpha}_g|$, $P(\delta_g + \alpha_g > 0 |y)$ if $\hat{\delta}_g > -\hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g > 0 |y)$ if $\hat{\delta}_g > \hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g < 0 |y)$ if $\hat{\delta}_g < \hat{\alpha}_g \le 0$, and $P(\delta_g + \alpha_g < 0 |y)$ if $\hat{\delta}_g < -\hat{\alpha}_g \le 0$.
 
@@
 The pattern appears to hold with the {\tt RStan} and {\tt ShrinkBayes} approaches outperforming the other two.
-While there does not appear to be a significant difference between the independence model using RStan or ShrinkBayes, there does appear to be a significant decreased performance, at least for smaller sample sizes, when using the model where a covariance is estimated.
+While there does not appear to be a significant difference between the independence model using {\tt RStan} or {\tt ShrinkBayes}, there does appear to be a significant decreased performance, at least for smaller sample sizes, when using the model where a covariance is estimated.
 
-\section{Heterosis in maize ??}
+\section{Heterosis in maize experiment}
 \label{s:corn}

### 2014-12-17 ebayes_rnaseq_heterosis.tex
through volcano plot

@@
 
-We calculate an estimated effect size using the posterior means for the gene-specific parameters using the formula 
-\[
+We used our method to analyze a maize data set on gene expression in parental lines B73 and Mo17 and the hybrid genotype (B73$\times$Mo17).
+The data are available in the Supplemental Materials.
+Each variety had four biological replicates measured with Illumina methodology and equipment.
+Reads were mapped to the whole reference genome using the short reads aligner NOVOALIGN.
+For more specific details, please see \cite{paschold2012complementation}.
+
+A total of 39,656 genes among $V=3$ varieties each with $n_v=4$ replicates were available for analysis.
+As before, we eliminated any genes with three or more zeros for any variety and any genes with mean expression less than one across all samples resulting in a total of $G=27,888$ genes present in the analysis.
+
+We found MAP estimates for both the independence and covariance models as shown in Table \ref{t:map}. 
+% Need MAP estimates table and discussion
+
+Independently for each gene, we ran an MCMC conditional on the hyperparameters to obtain empirical Bayes posterior distributions for all gene-specific parameters.
+Each MCMC was run using 4 chains with initial values randomly drawn by {\tt Stan} for a total of 2000 iterations, each chain with 1000 iterations of burn-in, and the remaining 1000 draws for each chain thinned by a factor of 4 (for storage reasons).
+The potential scale reduction factor, $\hat{R}$, was computed for each parameter in the in the chain \citep{Gelm:Rubi:infe:1992, Broo:Gelm:gene:1997}.
+If the maximum value of $\hat{R}$ was larger than 1.1, the total length was increased by 500 with the burn-in increased proportionally and the chain was rerun.
+This process was repeated until the maximum $\hat{R}$ was less than 1.1 for all parameters.
+
+With posteriors for all parameters, we can calculate empirical Bayes posterior probabilities for LPH and HPH.
+For each gene, the quantity of interest is likely to be the maximum of these two probabilities.
+For those genes with high probabilities for one of these two, it is of interest to see the magnitude of the effect.
+Thus, we calculate an estimated effect size using the posterior means for the gene-specific parameters using the formula
+\begin{equation}
 \mbox{estimated effect size} = \left\{ 
@@
 \right. 
-\]
-Figure \ref{f:volcano} provides a volcano plot to visualize the probability of heterosis versus estimated effect size.
+\label{e:effect_size}
+\end{equation}
+
+Figure \ref{f:volcano} provides a volcano plot to visualize the maximum of the probabilities for LPH and HPH versus estimated effect size.
+The figure shows a line at zero, for probabilities less than 0.5, due to the estimated effect size being set to 0 for the vast majority of genes.
+Above a probability of 0.5, we see a prototypical volcano pattern with increased extreme heterosis probability corresponding to larger estimated effect sizes and no estimated effect sizes near zero for genes with high extreme heterosis probability.
+Assymetry can also be observed with larger negative effect sizes compared with positive effect sizes which is due to genes whose parental counts are relatively high and hybrid counts near zero.
+Genes with high probability of extreme heterosis and large estimated effect sizes are candidates for further investigation.
 
@@
 \centerline{\includegraphics[width=\textwidth]{volcano}}
-\caption{A two-dimensional histogram of the maximum of LPH and HPH probabilities versus estimated effect size, $\mu_{g3}-\max(\mu_{g1},\mu_{g3})$ if $P(HPH|y)>P(LPH|y)$ or $\mu_{g3}-\min(\mu_{g1},\mu_{g2})$ if $P(HPH|y)<P(LPH|y)$ for the B73 $\times$ Mo17 maize experiment.}
+\caption{A two-dimensional histogram of the maximum of LPH and HPH probabilities versus estimated effect size defined in equation \ref{e:effect_size} for the B73 $\times$ Mo17 maize experiment using the independence model.}
 \label{f:volcano}
@@
 
+The supplementary materials include the maize experiment data ({\tt data.csv}), a script to run our method ({\tt script.R}), and {\tt Stan} models: all genes independent (\verb@\tt model_ind.txt\verb@) and covariance (\verb@model_cov.txt\verb@) models used in obtaining MAP estimates and single gene independent (\verb@\tt sg_model_ind.txt\verb@) and covariance (\verb@sg_model_cov.txt\verb@) models used in performing the MCMC on single genes.
+
 \bibliography{jarad}

### 2014-12-17 ebayes_rnaseq_heterosis.tex
through results

@@
 
+% Compare independence and volcano plots
+
 \begin{figure}
@@
 
+With almost 30,000 genes in our data set, we can evaluate the normality assumption for the gene-specific parameters.
+Figure \ref{f:normality} provides histograms of components of $\theta$ from a single MCMC iteration.
+If the prior assumption of normality is correct, we would expect these histograms to show approximate normality.
+The marginal histograms show a bimodal distribution for $\mu$ with a secondary peak around relatively small values.
+The bivariate histograms amongst the components of $\mu$ show a high correlation but with a heavy-tail toward small values of both components of $\mu$.
+With stricter gene-inclusion criterion, it is possible both the secondary peak and the heavy-tailedness could be eliminated.
+The histogram for $\log(\psi)$ displays a heavy-tail to the right and the bivariate histograms show a pattern of near independence between $\log(\psi)$ and $\mu_g$ until $\mu_g$ passes a threshold at which point $\log(\psi)$ appears to decrease.
+Both of these phenomenon are consistent with observations in the literature that the negative-binomial overdispersion parameter decreases as the mean increases, but the bivariate histograms appear to indicate a nonlinear relationship between the means and the log of the overdispersion parameter.
+
 \begin{figure}
 \centerline{\includegraphics[width=\textwidth]{normality}}
-\caption{One and two-dimensional histograms for gene-specific parameters from a single MCMC iteration, $\theta_g^{(m)}$,  for the B73 $\times$ Mo17 maize experiment.
-For reference, the one dimensional histograms have the marginal probability density functions from $N(\hat{\eta},\hat{\Sigma})$.}
+\caption{Marginal and bivariate histograms for gene-specific parameters from a single MCMC iteration, $\theta_g^{(m)}$,  for the B73 $\times$ Mo17 maize experiment.
+For reference, the marginal histograms have a dashed line indicating the appropriate marginal probability density functions from $N(\hat{\eta},\hat{\Sigma})$.}
 \label{f:normality}
@@
 
+Marginal and bivariate histograms for the model with an estimated covariance look similar although the estimated distribution $N(\hat{\eta},\hat{\Sigma})$ has increased variance for $\mu$ and decreased variance for $\log(\psi)$ as was previously seen in Table \ref{t:map} (results not shown).
+
 \section{Discussion}
 \label{s:discuss}
-
-Put your final comments here.
 

### 2014-12-17 ebayes_rnaseq_heterosis.tex
fixed verbatim removed appendix

@@
 
-The supplementary materials include the maize experiment data ({\tt data.csv}), a script to run our method ({\tt script.R}), and {\tt Stan} models: all genes independent (\verb@\tt model_ind.txt\verb@) and covariance (\verb@model_cov.txt\verb@) models used in obtaining MAP estimates and single gene independent (\verb@\tt sg_model_ind.txt\verb@) and covariance (\verb@sg_model_cov.txt\verb@) models used in performing the MCMC on single genes.
+The supplementary materials include the maize experiment data ({\tt data.csv}), a script to run our method ({\tt script.R}), and {\tt Stan} models: all genes independent (\verb@model_ind.txt@) and covariance (\verb@model_cov.txt@) models used in obtaining MAP estimates and single gene independent (\verb@sg_model_ind.txt@) and covariance (\verb@sg_model_cov.txt@) models used in performing the MCMC on single genes.
 
@@
 
-\appendix
-
-%  To get the journal style of heading for an appendix, mimic the following.
-
-\section{}
+% \appendix
+% %  To get the journal style of heading for an appendix, mimic the following.
+% \section{}
+% \subsection{Material here}
 

### 2014-12-17 ebayes_rnaseq_heterosis.tex
updated alternative methods

@@
 Existing RNAseq data analysis methods are not specifically designed to identify gene heterosis, but we adapted the approaches in {\tt edgeR}, {\tt baySeq}.
-and {\tt ShrinkBayes} to provide statistics to measure the strength of extreme heterosis of each gene.
+and {\tt ShrinkBayes} to provide a measure of the strength of extreme heterosis for each gene such that large values of this measure indicate support for extreme heterosis.
 
-{\tt edgeR} is designed to apply a negative binomial loglinear model to test for differential expression between any two varieties \citep{robinson2007moderated, robinson2010edgeR}, and we used this capability to construct a pseudo-probability of heterosis for each gene.
-We first computed the maximum likelihood estimates of the mean expression effects of all genes using {\tt edgeR}'s built-in Fisher scoring algorithm, and then we used the likelihood ratio test functionality to calculate two legitimate p-values for each gene: one for the null hypothesis that the hybrid's mean expression is the same as that of parent 1 and the second for the analogous test involving parent 2.
-For each gene, if the estimated mean expression level of the hybrid is not between the estimated mean expression levels of the parents, we set the pseudo-probability of heterosis to be one minus the p-value corresponding to the parent with estimated expression level closer to that of the hybrid.
-Otherwise, the pseudo-probability is zero.
+{\tt edgeR} is designed to apply a negative-binomial log-linear model to test for differential expression between any two varieties \citep{robinson2007moderated, robinson2010edgeR}.
+To construct a measure of heterosis, we computed the maximum likelihood estimates of the mean expression effects of all genes using {\tt edgeR}'s built-in Fisher scoring algorithm, and then used the likelihood ratio test functionality to calculate two legitimate p-values for each gene: one for the null hypothesis that the hybrid's mean expression is the same as that of parent 1 and the second for the analogous test involving parent 2.
+For each gene, if the estimated mean expression level of the hybrid is between the estimated mean expression levels of the parents, we set the measure of heterosis to zero.
+Otherwise, we set the measure of heterosis to be one minus the p-value corresponding to the parent with estimated expression level closer to that of the hybrid.
 
-The {\tt baySeq} approach allows for a wider range of hypotheses for each gene, including $H_1: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H_2: \mu_{g1}=\mu_{g2}$, $H_3: \mu_{g1}=\mu_{g3}$, $H_4: \mu_{g2}=\mu_{g3}$, and $H_5:$ all $\mu_g$'s are distinct \citep{hardcastle2010bayseq, hardcastle2012baySeq}.
-In a technique similar to our application of {\tt edgeR}, we used used the posterior probabilities of these hypotheses to construct a pseudo-probability of heterosis for each gene.
-We set this pseudo-probability to zero if the hybrid's maximum likelihood estimate of the mean expression level calculated with {\tt edgeR} is between those of the parents.
-Otherwise, the pseudo-probability is the sum of the posterior probabilities of $H_2$ and $H_5$.
-Note that $H_2$ and $H_5$ are the only hypotheses that allow for heterosis.
+The {\tt baySeq} approach \citep{hardcastle2010bayseq, hardcastle2012baySeq} allows for a wider range of hypotheses for each gene including $H_1: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H_2: \mu_{g1}=\mu_{g2}$, $H_3: \mu_{g1}=\mu_{g3}$, $H_4: \mu_{g2}=\mu_{g3}$, and $H_5:$ all $\mu_g$'s are distinct.
+In a technique similar to our application of {\tt edgeR}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene.
+We set this measure to zero if the hybrid's maximum likelihood estimate of the mean expression level calculated using {\tt edgeR} is between those of the parents.
+Otherwise, the measure is the sum of the posterior probabilities of $H_2$ and $H_5$, the only hypotheses that allow for extreme heterosis.
 
@@
 A limitation of INLA is that it only provides approximations to marginal posteriors for parameters or linear combinations of parameters rather than an approximation to the full posterior.
-Thus, we cannot directly estimate nonlinear quantities such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2}))$.
+Thus, we cannot directly estimate nonlinear quantities such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2})|y)$.
 To circumvent this limitation, we reparameterized the independence model in Section \ref{s:model} using the approach in \cite{ji2014estimation}.
 Let $\phi_g = (\mu_{g1}+\mu_{g2})/2$ be the parental log-mean, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental difference, and $\delta_g = \mu_{g3} - \phi_g$ be the hybrid effect.
-After fitting the model with {\tt ShrinkBayes}, we constructed a pseudo-probability of heterosis for each gene.
-For gene $g$, we used the estimated posterior means $\hat{\delta}_g$ and $\hat{\alpha}_g$ of $\delta_g$ and $\alpha_g$, respectively, along with appropriate posterior probabilities $P(\cdot \mid y)$.
-Specifically, the pseudo-probability of heterosis for gene $g$ is zero if $|\hat{\delta}_g| \le |\hat{\alpha}_g|$, $P(\delta_g + \alpha_g > 0 |y)$ if $\hat{\delta}_g > -\hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g > 0 |y)$ if $\hat{\delta}_g > \hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g < 0 |y)$ if $\hat{\delta}_g < \hat{\alpha}_g \le 0$, and $P(\delta_g + \alpha_g < 0 |y)$ if $\hat{\delta}_g < -\hat{\alpha}_g \le 0$.
+After fitting the model with {\tt ShrinkBayes}, we constructed a measure of heterosis for each gene.
+For gene $g$, we used the estimated posterior means $\hat{\delta}_g$ and $\hat{\alpha}_g$ of $\delta_g$ and $\alpha_g$, respectively, along with appropriate posterior probabilities.
+Specifically, the measure of heterosis for gene $g$ is zero if $|\hat{\delta}_g| \le |\hat{\alpha}_g|$, $P(\delta_g + \alpha_g > 0 |y)$ if $\hat{\delta}_g > -\hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g > 0 |y)$ if $\hat{\delta}_g > \hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g < 0 |y)$ if $\hat{\delta}_g < \hat{\alpha}_g \le 0$, and $P(\delta_g + \alpha_g < 0 |y)$ if $\hat{\delta}_g < -\hat{\alpha}_g \le 0$.
 

### 2014-12-17 ebayes_rnaseq_heterosis.tex
first paragraph of discussion

@@
 \begin{align}
-\label{e:hypotheses}
 H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
 H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}.
+\label{e:hypotheses}
 \end{align}
@@
 For each of the methods above as well as the method described in Section \ref{s:gene_heterosis}, we sorted genes according to the plausibility of heterosis.
-From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to identify genes with heterosis, as defined in Section \ref{s:sim_data}, \citep{landau2013dispersion}.
+From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to identify genes with heterosis, as defined in Section \ref{s:sim_data}  \citep{landau2013dispersion}.
 A representative set of ROC curves is shown in Figure \ref{f:roc}.
@@
 
+Gene expression heterosis is speculated to be one possible explanation for phenotypic
+heterosis of traits like plant height or grain yield.
+Existing methods for identifying gene expression heterosis based on RNAseq data are not directly applicable for detecting heterosis due to the composite null hypothesis in equation \eqref{e:hypotheses}.
+\cite{ji2014estimation} introduced an empirical Bayes approach based on a hierarchical model for microarray data.
+We followed their approach modified slightly to allow for any number of varieties and therefore even more complex hypotheses.
+We presented an empirical Bayes method to obtain MAP estimates for hyperparameters followed by MCMC to estimate gene-specific parameters which can be done efficienctly via parallelization.
+This methodology can be conveniently implemented in the statistical software {\tt Stan} which also allows for much more complex models.
+The empirical Bayes posteriors can be used to estimate posterior probabilities of high and low parent heterosis.
+
 \backmatter %  Please keep this command in your document in this position

### 2014-12-17 ebayes_rnaseq_heterosis.tex
added roadmap

@@
 % \citep{ji2014estimation}
+
+The remainder of the paper proceeds as follows.
+Section \ref{s:method} presents the proposed hierarchical model, an empirical Bayes method of estimating the parameters, and calculation of posterior probabilities of gene heterosis.
+Section \ref{s:simulation} presents a simulation study based on a maize experiment and compares our approach to alternative methods.
+Section \ref{s:maize} analyzes a maize experiment where phenotypic heterosis is well established and identifies genes demonstrating heterosis.
+The data and code necessary to perform this analysis is available in the Supplemental Material.
+Section \ref{s:discussion} summarizes the work and suggests directions for future research.
 
@@
 \section{Heterosis in maize experiment}
-\label{s:corn}
+\label{s:maize}
 
@@
 \section{Discussion}
-\label{s:discuss}
+\label{s:discussion}
 
@@
 The empirical Bayes posteriors can be used to estimate posterior probabilities of high and low parent heterosis.
+
+Through a simulation study, we demonstarted that this method outperformed alternative methods (that are not designed to detect heterosis), and performed comparably with an implementation of our model in ShrinkBayes which estimates the full posterior via INLA.
 

### 2014-12-17 ebayes_rnaseq_heterosis.tex
added bioconductor ref

@@
 
-To begin, we applied the {\tt edgeR} package from Bioconductor (citation needed) to the maize data to obtain sample-specific normalization factors, then gene-specific dispersions, and finally gene-variety-specific location parameters.
+To begin, we applied the {\tt edgeR} package \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor} to the maize data to obtain sample-specific normalization factors, then gene-specific dispersions, and finally gene-variety-specific location parameters.
 We treated these parameters as the truth for the simulation study: i.e.

### 2014-12-17 ebayes_rnaseq_heterosis.tex
added preref to section 4
ji paragraph added

@@
 
-% WML: Since you mainly talk about phenotypic heterosis, I added a little clarification about gene-expression heterosis.
-However, the reader will still be mostly focused on phenotypic heterosis, and more may need to be done to refocus on the transcriptome.
-Do we have good real-world examples of heterosis in the expression important or interesting genes?
-This may be a good question for Dan.
-
-Recently \citep{ji2014estimation} provided an approach to assess heterosis using micro-array data using a hierarchical model for the ...
+Recently, \citep{ji2014estimation} introduced an approach to assess gene heterosis using micro-array data where the response is assumed to be continuous.
+They built a hierarchical model for gene expression that allows borrowing of information across genes for means and variances.
+They introduced an empirical Bayes framework that first estimates model hyperparameters, then estimates the posterior distribution for gene-specific parameters, and finally computes heterosis probabilities based on integrals of regions under this posterior.
 
@@
 To assess the efficacy of our method to identify genes demonstrating heterosis, we used a maize data set with parental varieties B73 and Mo17 and the hybrid variety (B73 $\times$ Mo17) \citep{paschold2012complementation}.
+Details of these data set are available in Section \ref{s:maize}.
 We compared our method to approaches using the R packages {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes}.

### 2014-12-17 ebayes_rnaseq_heterosis.tex
through intro

@@
 
-Recently, \citep{ji2014estimation} introduced an approach to assess gene heterosis using micro-array data where the response is assumed to be continuous.
-They built a hierarchical model for gene expression that allows borrowing of information across genes for means and variances.
-They introduced an empirical Bayes framework that first estimates model hyperparameters, then estimates the posterior distribution for gene-specific parameters, and finally computes heterosis probabilities based on integrals of regions under this posterior.
-
-%\section{Heterosis}
-%\label{s:heterosis}
-
-% Well before heterosis was scientifically described by , humans had been using heterosis for various practical purposes.
-Within the last century, heterosis has been used to improve many species for food, feed, and fuel industries.
-A few relatively recent examples include rice , alfalfa , tomatoes .
-Despite the intensive study and successful utilization of heterosis, the basic genomic mechanisms responsible for heterosis remain unclear .
-% 
-% \cite{shull1952beginnings} coined the term heterosis and defined it as 
-% \begin{quote}
-% the interpretation of increased vigor, size, fruitfulness, speed of development, resistance to disease and to insect pests, or to climatic rigors of any kind, manifested by crossbred organisms as compared with corresponding inbreds, as the specific results of unlikeness in the constitutions of the uniting parental gametes.
-% \end{quote}
-% 
-% For example, the maize hybrid produced from a cross of parental inbred lines B73 and Mo17 is larger in size, has lower maturation time, and has higher grain yield than both of its parents \cite{hallauer2010quantitative}.
-Data from \citep{paschold2012complementation}
-% 
-% \citep{ji2014estimation}
+Recently, \cite{ji2014estimation} introduced an approach to assess gene heterosis using microarray data where the data are assumed to be continuous.
+They built a normal hierarchical model for gene expression that allows borrowing of information across genes for means and variances.
+They introduced an empirical Bayes framework that first estimates model hyperparameters, then estimates the posterior distribution for gene-specific parameters conditional on those hyperparameters, and finally computes heterosis probabilities based on integrals of regions under this posterior.
+This development was necessary due to the composite null hypothesis when testing for heterosis making many available methods ill-suited to heterosis identification.
+When transitioning from continuous to count data, the difficulty in a composite null hypothesis remains.
+We following \citeauthor{ji2014estimation} in constructing a hierarchical model now built off a negative-binomial rather than normal data model.
+We also utilize an empirical Bayes approach based on obtaining \emph{maximum a posterior} estimates of hyperparameters and posterior distribution for gene-specific parameters conditional on those hyperparameters.
 
@@
 
-An RNAseq experiment to identify genes demonstrating heterosis consists of three genetic varieties: two parental varieties (P1 and P2) and a cross between these two varieties called the hybrid (H).
+An RNA sequencing (RNAseq) experiment to identify genes demonstrating heterosis consists of three genetic varieties: two parental varieties (P1 and P2) and a cross between these two varieties called the hybrid (H).
 For each variety, replicate RNA samples are isolated and assessed for quality.
@@
 Once a model is constructed, the software can be used to obtain MAP estimates \cite[Section 50.3,][]{stan-manual:2014} using the {\tt optimizing} function in RStan.
-We found it necessary to use the L-BFGS algorithm \cite[Section 55,][]{stan-manual:2014} presumably due to the large dimension of the optimization.
+We found it necessary to use the L-BFGS algorithm \cite[Section 55,][]{stan-manual:2014} and reasonable starting values obtained via {\tt edgeR} presumably due to the large dimension of the optimization.
 The {\tt Stan} software can also be used to run a Markov chain Monte Carlo (MCMC) algorithm to obtain samples from the posterior using the {\tt sampling} function in RStan.

### 2014-12-17 ebayes_rnaseq_heterosis.tex
first stab at abstract

@@
 \pubyear{YYYY}
-\artmonth{Month}
+\artmonth{MM}
 \doi{x}
@@
 
-%  put the summary for your paper here
-
 \begin{abstract}
-This is the summary for this paper.
+Heterosis, or hybrid vigor, refers to the enhances phenotype of hybrid progeny relative to their inbred parents.
+The heterosis phenomenon is extensively utilized in agriculture though the molecular basis is still unknown.
+In an effort to understand phenotypic heterosis at the molecular level, researchers have begun to compare expression levels of thousands of genes between parental inbred lines and their hybrid offspring to search for evidence of gene expression heterosis in RNA sequencing (RNAseq) experiments.
+Standard statistical approaches for RNAseq experiments have difficulty with the composite null hypothesis for gene heterosis testing.
+To address these shortcomings, we develop a hierarchical model based on negative-binomial data model and a computationally tractible empirical Bayes approach to evaluating gene heterosis hypotheses.
+We demonstrate improvements over alternative methods via a simulation study based on a maize experiment, and then analyze that maize experiment.
+This article has supplementary material online.
 \end{abstract}

### 2014-12-17 ebayes_rnaseq_heterosis.tex
discussion

@@
 
-Through a simulation study, we demonstarted that this method outperformed alternative methods (that are not designed to detect heterosis), and performed comparably with an implementation of our model in ShrinkBayes which estimates the full posterior via INLA.
+Through a simulation study, we demonstrated that this method outperformed alternative methods (that are not designed to detect heterosis), and performed comparably with an implementation of our model in ShrinkBayes which estimates the full posterior via INLA.
+We then demonstrated the use of the methodology on a maize experiment where phenotypic heterosis is well known.
+% Estimates of heterosis to compare with Ji et. al.
+In addition to providing estimates of heterosis probabilities and effect sizes, this real data analysis demonstrated violation of normality assumptions that suggest model improvements to improve power for detection of gene heterosis.
+Developing the details of such an extension is the subject of some of our ongoing and future research.
 

### 2014-12-17 ebayes_rnaseq_heterosis.tex
small changes

@@
 
-{\tt edgeR} is designed to apply a negative-binomial log-linear model to test for differential expression between any two varieties \citep{robinson2007moderated, robinson2010edgeR}.
+The {\tt edgeR} approach is designed to apply a negative-binomial log-linear model to test for differential expression between any two varieties \citep{robinson2007moderated, robinson2010edgeR}.
 To construct a measure of heterosis, we computed the maximum likelihood estimates of the mean expression effects of all genes using {\tt edgeR}'s built-in Fisher scoring algorithm, and then used the likelihood ratio test functionality to calculate two legitimate p-values for each gene: one for the null hypothesis that the hybrid's mean expression is the same as that of parent 1 and the second for the analogous test involving parent 2.
@@
 We followed their approach modified slightly to allow for any number of varieties and therefore even more complex hypotheses.
-We presented an empirical Bayes method to obtain MAP estimates for hyperparameters followed by MCMC to estimate gene-specific parameters which can be done efficienctly via parallelization.
+We presented an empirical Bayes method to obtain MAP estimates for hyperparameters followed by MCMC to estimate gene-specific parameters which can be accomplished efficiently via parallelization.
 This methodology can be conveniently implemented in the statistical software {\tt Stan} which also allows for much more complex models.

### 2014-12-18 ebayes_rnaseq_heterosis.tex
changes

@@
 \begin{abstract}
-Heterosis, or hybrid vigor, refers to the enhances phenotype of hybrid progeny relative to their inbred parents.
-The heterosis phenomenon is extensively utilized in agriculture though the molecular basis is still unknown.
+Heterosis, or hybrid vigor, refers to the enhancements in the phenotype of hybrid progeny relative to their inbred parents.
+Although the heterosis phenomenon is extensively utilized in agriculture, the molecular basis is still unknown.
 In an effort to understand phenotypic heterosis at the molecular level, researchers have begun to compare expression levels of thousands of genes between parental inbred lines and their hybrid offspring to search for evidence of gene expression heterosis in RNA sequencing (RNAseq) experiments.
-Standard statistical approaches for RNAseq experiments have difficulty with the composite null hypothesis for gene heterosis testing.
-To address these shortcomings, we develop a hierarchical model based on negative-binomial data model and a computationally tractible empirical Bayes approach to evaluating gene heterosis hypotheses.
-We demonstrate improvements over alternative methods via a simulation study based on a maize experiment, and then analyze that maize experiment.
+Standard statistical approaches for RNAseq experiments struggle with the composite null hypothesis in gene heterosis testing.
+To address these shortcomings, we develop a hierarchical model based on a negative-binomial data model and a computationally tractable empirical Bayes approach.
+We demonstrate improvements over alternative methods via a simulation study based on a maize experiment and then analyze that maize experiment.
 This article has supplementary material online.
@@
 % Need to udpate as this is directly from Tieming's paper
-Heterosis, or hybrid vigor, refers to the enhances phenotype of hybrid progeny relative to their inbred parents.
-Taking maize as an example, the offspring from crossing the inbred lines B73 and Mo17 are taller, mature faster, and produce greater yields than their parent lines \citep{hallauer1981quantitative, hallauer2010quantitative}.
-Since heterosis was scientifically documented by \cite{darwin1876effects}, it has been successfully manipulated to improve many species for food, feed, and fuel industries, such as rice \citep{yu1997importance}, alfalfa \citep{riday2002heterosis}, tomatoes \citep{krieger2010flowering}, and fish \citep{wohlfarth1993heterosis}.
+Heterosis, or hybrid vigor, refers to the enhancements in the phenotype of hybrid progeny relative to their inbred parents.
+Taking maize as an example, the offspring from crossing the inbred lines B73 and Mo17 grow taller, mature faster, and produce greater yields than their parent lines \citep{hallauer1981quantitative, hallauer2010quantitative}.
+Ever since heterosis was scientifically documented by \cite{darwin1876effects}, it has been successfully manipulated to improve many species for the food, feed, and fuel industries, such as rice \citep{yu1997importance}, alfalfa \citep{riday2002heterosis}, tomatoes \citep{krieger2010flowering}, and fish \citep{wohlfarth1993heterosis}.
 Despite the intensive study and successful utilization of heterosis, the basic genomic mechanisms remain unclear \citep{coors1999genetics, lippman2007heterosis}.
-Researchers speculate that gene-expression heterosis $-$ i.e, enhanced expression of one or more genes in the hybrid relative to the inbred parents $-$ could be among the mechanisms responsible for the phenotypic heterosis \citep{swanson2006all, springer2007allelic}.
+Researchers speculate that gene-expression heterosis $-$ i.e, enhanced expression of one or more genes in the hybrid relative to the inbred parents $-$ is a possible  mechanisms for the phenotypic heterosis \citep{swanson2006all, springer2007allelic}.
 
-Recently, \cite{ji2014estimation} introduced an approach to assess gene heterosis using microarray data where the data are assumed to be continuous.
+Recently, \cite{ji2014estimation} introduced an approach to assess gene heterosis using microarray data under the assumption that these data are continuous.
 They built a normal hierarchical model for gene expression that allows borrowing of information across genes for means and variances.
 They introduced an empirical Bayes framework that first estimates model hyperparameters, then estimates the posterior distribution for gene-specific parameters conditional on those hyperparameters, and finally computes heterosis probabilities based on integrals of regions under this posterior.
-This development was necessary due to the composite null hypothesis when testing for heterosis making many available methods ill-suited to heterosis identification.
-When transitioning from continuous to count data, the difficulty in a composite null hypothesis remains.
-We following \citeauthor{ji2014estimation} in constructing a hierarchical model now built off a negative-binomial rather than normal data model.
-We also utilize an empirical Bayes approach based on obtaining \emph{maximum a posterior} estimates of hyperparameters and posterior distribution for gene-specific parameters conditional on those hyperparameters.
+This development was necessary due to the composite null hypothesis in tests for heterosis.
+These hypotheses, which many available methods do not fully accommodate, remain a challenge in the transition from continuous data to count data.
+Building on the work of \citeauthor{ji2014estimation} with the normal data model, we construct a hierarchical model based on a negative-binomial data model.
+We also utilize an empirical Bayes approach to obtain \emph{maximum a posterior} (MAP) estimates of the hyperparameters and the posterior distributions for the gene-specific parameters conditional on those hyperparameters.
 
 The remainder of the paper proceeds as follows.
-Section \ref{s:method} presents the proposed hierarchical model, an empirical Bayes method of estimating the parameters, and calculation of posterior probabilities of gene heterosis.
+Section \ref{s:method} presents the proposed hierarchical model, an empirical Bayes method of estimating the parameters, and the calculation of posterior probabilities of gene heterosis.
 Section \ref{s:simulation} presents a simulation study based on a maize experiment and compares our approach to alternative methods.
@@
 For each variety, replicate RNA samples are isolated and assessed for quality.
-cDNA libraries, i.e.
-short cDNA fragments, are constructed and next generation sequencing technology is used to determine the \emph{reads}, i.e.
-nucleotide sequences in the cDNA libraries.
+cDNA libraries, or collections of short cDNA fragments, are constructed.
+Then, next generation sequencing technology is used to determine the \emph{reads}, or nucleotide sequences, in the cDNA libraries.
 These reads are processed using bioinformatic algorithms to match the reads to genes, or specific gene transcripts, exons, microRNAs, etc.
 The result of matching reads to features are summarized by a gene $\times$ sample matrix of counts.
-For more details on RNAseq experiments see \cite{nettleton2014design} and for biological background on the details of these steps see \cite{paschold2012complementation}.
+See \cite{nettleton2014design} for more details on RNAseq experiments, and see \cite{paschold2012complementation} for the biological background behind the above steps.
 
-To identify genes displaying heterosis using RNAseq counts, we build a hierarchical model to borrow information on gene-variety means and gene-specific overdispersion parameters, estimate the parameters using a empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate empirical Bayes posterior probabilities for heterosis.
+To use RNAseq counts to identify genes displaying heterosis, we build a hierarchical model to borrow information on gene-variety means and gene-specific overdispersion parameters, estimate the parameters using a empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate empirical Bayes posterior probabilities for heterosis.
 
@@
 \end{equation}
-where $NB(\eta,\psi)$ indicates a negative-binomial with expectation $\eta$ and variance $\eta+\psi\eta^2$ and $ind$ indicates the observations are conditionally independent.
+where $NB(\eta,\psi)$ indicates a negative-binomial with expectation $\eta$ and variance $\eta+\psi\eta^2$, and $ind$ indicates the observations are conditionally independent.
 
-A hierarchical model is constructed for the sequencing depth, overdispersion, and gene-variety parameters.
+We present a hierarchical model for the sequencing depth, overdispersion, and gene-variety parameters.
 The sample-specific sequencing depth terms are assumed to follow a normal distribution, i.e.
 $\gamma_{vi} \stackrel{ind}{\sim} N(0,\tau_\gamma^2)$.
-(These terms are the normalization factors that are typically estimated from each sample and then plugged in for further analysis.) The logarithm of the gene-specific over-dispersion parameters and the gene-specific variety effects are assumed to have a joint normal distribution, i.e.
+(These terms are the normalization factors that practitioners typically estimate for each sample and then treat as constant in further analysis.) The logarithm of the gene-specific over-dispersion parameters and the gene-specific variety effects are assumed to have a joint normal distribution, i.e.
 $\theta_g = [\mu_g, \log(\psi_g)]^\top \stackrel{ind}{\sim} N\left(\eta, \Sigma\right)$
 where $\mu_g = (\mu_{g1},\ldots,\mu_{gV})'$.
-We consider an \emph{independent} model where $\Sigma$ is diagonal with elements $\sigma_1^2,\ldots,\sigma_{V+1}^2$ and a \emph{covariance} model where we estimate all parameters in $\Sigma$.
+We consider an \emph{independent} model in which $\Sigma$ is diagonal with elements $\sigma_1^2,\ldots,\sigma_{V+1}^2$ and a \emph{covariance} model in which we estimate all parameters in $\Sigma$.
 
-We assume improper uniform priors for $\eta$ and $\tau_\gamma\stackrel{ind}{\sim} Ca^+(0,3)$ where 0 and 3 are the location and scale parameters of the untruncated distribution.
-In the independent model, we also $Ca^+(0,3)$ on the diagonal elements of $\Sigma$ while in the covariance model, we assume a conjugate, proper inverse Wishart with degrees of freedom $V+2$ and an identity scale matrix which results in marginally uniform correlations.
+We assume improper uniform priors for $\eta$ and take $\tau_\gamma\stackrel{ind}{\sim} Ca^+(0,3)$, where 0 and 3 are the location and scale parameters, respectively, of the untruncated half-Cauchy distribution.
+In the independent model, we also assign the $Ca^+(0,3)$ prior to the diagonal elements of $\Sigma$.
+In the covariance model, we assume a conjugate, proper inverse Wishart distribution with degrees of freedom $V+2$ and an identity scale matrix, resulting in marginally uniform correlations.
 
@@
 
-The parameters from the model in Section \ref{s:model} can be group by gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$, and hyperparameter $\pi = (\gamma,\eta, \Sigma)$ where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$.
-We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\pi$, denoted $\hat{\pi}$, and then base gene-specific inference on the posterior conditional on this estimated hyperparameter.
-Conditional on the hyperparameters, the gene-specific parameters are conditionally independent as in equation \eqref{e:condind}.
+We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ and hyperparameters $\pi = (\gamma,\eta, \Sigma)$, where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$.
+We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\pi$, denoted $\hat{\pi}$, and then base gene-specific inference on the posterior conditional of this estimated hyperparameter.
+Conditional on the hyperparameters, the gene-specific parameters are independent as in equation \eqref{e:condind}.
 Thus, conditional posterior inference for the gene-specific parameters can be obtained independently and in parallel.
@@
 
-To perform the optimization, we use the statistical software {\tt Stan} \citep{stan-software:2014} run through the RStan interface \citep{rstan-software:2014} in R \citep{R2014}.
-The Stan software allows construction of general Bayesian models with the primary restriction that all unknown parameters must be continuous.
-Once a model is constructed, the software can be used to obtain MAP estimates \cite[Section 50.3,][]{stan-manual:2014} using the {\tt optimizing} function in RStan.
-We found it necessary to use the L-BFGS algorithm \cite[Section 55,][]{stan-manual:2014} and reasonable starting values obtained via {\tt edgeR} presumably due to the large dimension of the optimization.
-The {\tt Stan} software can also be used to run a Markov chain Monte Carlo (MCMC) algorithm to obtain samples from the posterior using the {\tt sampling} function in RStan.
-By default, Stan uses a variant of Hamiltonion Monte Carlo \citep{neal2011mcmc} called the no-U-turn sampler \citep{hoffman2013no}.
+To perform the optimization, we use the statistical software, {\tt Stan} \citep{stan-software:2014}, executed through the RStan interface \citep{rstan-software:2014} in R \citep{R2014}.
+The Stan software allows the construction of general Bayesian models with the primary restriction that all unknown parameters are continuous.
+With a fully constructed model, the software can obtain MAP estimates \cite[Section 50.3,][]{stan-manual:2014} using the {\tt optimizing} function in RStan.
+It was necessary to use the L-BFGS algorithm \cite[Section 55,][]{stan-manual:2014} and reasonable starting values obtained via {\tt edgeR}, presumably due to the large dimension of the optimization.
+The {\tt Stan} software also includes the RStan function, {\tt sampling}, an implementation of a Markov chain Monte Carlo (MCMC) algorithm for obtaining samples from the posterior.
+By default, Stan uses a no-U-turn sampler \citep{hoffman2013no}, a variant of Hamiltonion Monte Carlo \citep{neal2011mcmc}.
 
-As we have not integrated out the gene-specific parameters when obtaining the MAP estimate, we actually receive a joint MAP estimate for $\theta$ and $\pi$.
-We explicitly use the MAP estimate for $\pi$ when obtaining empirical Bayes posterior samples for the gene-specific parameters as in equation \eqref{e:condind}.
-The MAP estimate for $\theta$ can also be used in this step as initial values for the MCMC.
+As we have not integrated out the gene-specific parameters when obtaining the MAP estimate, we actually receive a joint MAP estimate of $\theta$ and $\pi$.
+We explicitly use the MAP estimate of $\pi$ when obtaining empirical Bayes posterior samples for the gene-specific parameters as in equation \eqref{e:condind}.
+The MAP estimate of $\theta$ can also be used in this step as initial values for the MCMC.
 
@@
 As defined in Section \ref{s:intro}, heterosis is increased (or decreased) phenotypic expression in a hybrid relative to its inbred parents.
-Thus, when attempting to identify heterosis, there are three varieties of interest, i.e.
-$V=3$.
+Thus, when attempting to identify heterosis, there are $V=3$ varieties of interest.
 We let $v=1,2$ be the two parental varieties and $v=3$ be the hybrid.
-We define gene heterosis to be the differential expression of the hybrid relative to its parents.
+We define gene heterosis to be the differential expression of a gene in the hybrid relative to its parents.
 For a specific gene $g$, low-parent gene heteorisis (LPH) occurs when $\mu_{g3}< \min(\mu_{g1},\mu_{g2})$, mid-parent gene heterosis occurs when $\mu_{g1}\le \mu_{g3}\le \mu_{g2}$ if $\mu_{g1}\le \mu_{g2}$ (or $\mu_{g1}\ge \mu_{g3}\ge \mu_{g2}$ if $\mu_{g1}\ge \mu_{g2}$), and high-parent gene heterosis (HPH) occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$.
-For simplicity, we define \emph{extreme gene heterosis} (EH) which is the event of LPH or HPH.
-Of interest is to evaluate the hypotheses shown in equation \eqref{e:hypotheses} where $H_0$ indicates no EH and $H_1$ indicates EH.
+For simplicity, we define \emph{extreme gene heterosis} (EH) to be the event of LPH or HPH.
+Equation \eqref{e:hypotheses} shows the most important hypotheses, where $H_0$ indicates no EH and $H_1$ indicates EH.
 \begin{align}
@@
 \end{align}
-We evaluate these hypotheses based on empirical Bayes estimates of their posterior probability, e.g. 
+We evaluate these hypotheses based on empirical Bayes estimates of their posterior probabilities, e.g. 
 \begin{align}
@@
 \end{align}
-where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise.
-The set of probabilities $P\left(H_{g1}|y, \hat{\pi}_{MAP}\right)$ for $g=1,\ldots,G$ can be used to provide a list of genes sorted by the relative strength of evidence for extreme heterosis.
-This list can be used by geneticists to prioritize future experiments to understand the genetic mechanism for heterosis.
+where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise.
+The set of probabilities $P\left(H_{g1}|y, \hat{\pi}_{MAP}\right)$ for $g=1,\ldots,G$ can be used to provide a list of genes sorted by the relative strength of evidence of extreme heterosis.
+Geneticists can use this list to prioritize future experiments to understand the genetic mechanism for heterosis.
 
@@
 To assess the efficacy of our method to identify genes demonstrating heterosis, we used a maize data set with parental varieties B73 and Mo17 and the hybrid variety (B73 $\times$ Mo17) \citep{paschold2012complementation}.
-Details of these data set are available in Section \ref{s:maize}.
+Section \ref{s:maize} describes this dataset in detail.
 We compared our method to approaches using the R packages {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes}.
@@
 Initially, the maize experiment had RNAseq counts for 39,656 genes and 4 replicates per variety.
-We eliminated low count genes by removing any gene with three or more zeros for any variety and genes whose mean count across the varieties was less than one, which resulted in 27,888 genes remaining.
+We eliminated low count genes by removing any gene with three or more zeros for any variety and genes whose mean count across the varieties was less than one, resulting in 27,888 remaining genes.
 
-To begin, we applied the {\tt edgeR} package \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor} to the maize data to obtain sample-specific normalization factors, then gene-specific dispersions, and finally gene-variety-specific location parameters.
+Next, we applied the {\tt edgeR} package \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor} to the maize data to obtain sample-specific normalization factors, then gene-specific dispersions, and finally gene-variety-specific location parameters.
 We treated these parameters as the truth for the simulation study: i.e.
@@
 Using these parameters and normalization factors, we simulated data according to the negative-binomial model in equation \eqref{e:data} independently for each gene.
-For each simulation, low count genes, i.e.
-those with three or more zeros in any variety or whose mean count across the varieties was less than one, were removed.
-A random set of 25,000 genes was retained from those with high enough counts.
-This simulation process was repeated 10 times for each of 4, 8, and 16 replicates per variety (reusing normalization factors when necessary).
+For each simulation, low count genes $-$ those with three or more zeros in any variety or whose mean count across the varieties was less than one $-$ were removed.
+We retained a random subset of 25,000 genes from those with high enough counts.
+We repeated this simulation process 10 times for each of 4, 8, and 16 replicates per variety, reusing normalization factors when necessary.
 
@@
 
-Existing RNAseq data analysis methods are not specifically designed to identify gene heterosis, but we adapted the approaches in {\tt edgeR}, {\tt baySeq}.
-and {\tt ShrinkBayes} to provide a measure of the strength of extreme heterosis for each gene such that large values of this measure indicate support for extreme heterosis.
+Existing RNAseq data analysis methods are not specifically designed to identify gene heterosis, but we adapted the approaches in {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes} to provide a measure of the strength of extreme heterosis for each gene such that large values of this measure indicate support for extreme heterosis.
 
@@
 
-The {\tt baySeq} approach \citep{hardcastle2010bayseq, hardcastle2012baySeq} allows for a wider range of hypotheses for each gene including $H_1: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H_2: \mu_{g1}=\mu_{g2}$, $H_3: \mu_{g1}=\mu_{g3}$, $H_4: \mu_{g2}=\mu_{g3}$, and $H_5:$ all $\mu_g$'s are distinct.
+The {\tt baySeq} approach \citep{hardcastle2010bayseq, hardcastle2012baySeq} allows for a wider range of hypotheses for each gene, including $H_1: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H_2: \mu_{g1}=\mu_{g2}$, $H_3: \mu_{g1}=\mu_{g3}$, $H_4: \mu_{g2}=\mu_{g3}$, and $H_5:$ all $\mu_g$'s are distinct.
 In a technique similar to our application of {\tt edgeR}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene.
@@
 
-For each of the methods above as well as the method described in Section \ref{s:gene_heterosis}, we sorted genes according to the plausibility of heterosis.
+For each of the methods above, as well as the method described in Section \ref{s:gene_heterosis}, we sorted genes according to the plausibility of heterosis.
 From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to identify genes with heterosis, as defined in Section \ref{s:sim_data}  \citep{landau2013dispersion}.
@@
 \end{figure}
-For this simulation, we can see the approaches based on the model in Section \ref{s:model}, i.e.
-both RStan approaches and the ShrinkBayes approach, provide the best true positive rate for a given false positive rate.
-Also, as expected, as the sample size increases our ability to distinguish genes with heterosis from genes without improves.
+For this simulation, we can see that the approaches based on the model in Section \ref{s:model} $-$ i.e.
+both RStan approaches and the ShrinkBayes approach $-$ provide the best true positive rate for a given false positive rate.
+Also, as expected, as the sample size increases, our ability to distinguish genes with heterosis from genes without improves.
 
-Figure \ref{f:auc} provides the area under the ROC below a false positive rate of 0.1 across the 10 simulations for each of the 3 different sample sizes. 
+Figure \ref{f:auc} provides the area under the ROC curve below a false positive rate of 0.1 across the 10 simulations for each of the 3 different sample sizes. 
 \begin{figure}
@@
 \end{figure}
-The pattern appears to hold with the {\tt RStan} and {\tt ShrinkBayes} approaches outperforming the other two.
-While there does not appear to be a significant difference between the independence model using {\tt RStan} or {\tt ShrinkBayes}, there does appear to be a significant decreased performance, at least for smaller sample sizes, when using the model where a covariance is estimated.
+The pattern appears to hold, with the {\tt RStan} and {\tt ShrinkBayes} techniques outperforming the other two.
+While there does not appear to be a significant difference between {\tt ShrinkBayes} and the independence model using {\tt RStan}, performance appears to decrease for the model where a covariance is estimated, at least for smaller sample sizes.
 
@@
 Each variety had four biological replicates measured with Illumina methodology and equipment.
-Reads were mapped to the whole reference genome using the short reads aligner NOVOALIGN.
-For more specific details, please see \cite{paschold2012complementation}.
+Reads were mapped to the whole reference genome using the short reads aligner, NOVOALIGN.
+For more specifics, please see \cite{paschold2012complementation}.
 
-A total of 39,656 genes among $V=3$ varieties each with $n_v=4$ replicates were available for analysis.
-As before, we eliminated any genes with three or more zeros for any variety and any genes with mean expression less than one across all samples resulting in a total of $G=27,888$ genes present in the analysis.
+A total of 39,656 genes among $V=3$ varieties, each with $n_v=4$ replicates, were available for analysis.
+As before, we eliminated genes with three or more zeros for any variety and genes with mean expression less than one across all samples, resulting in a total of $G=27,888$ genes present in the analysis.
 
@@
 Independently for each gene, we ran an MCMC conditional on the hyperparameters to obtain empirical Bayes posterior distributions for all gene-specific parameters.
-Each MCMC was run using 4 chains with initial values randomly drawn by {\tt Stan} for a total of 2000 iterations, each chain with 1000 iterations of burn-in, and the remaining 1000 draws for each chain thinned by a factor of 4 (for storage reasons).
-The potential scale reduction factor, $\hat{R}$, was computed for each parameter in the in the chain \citep{Gelm:Rubi:infe:1992, Broo:Gelm:gene:1997}.
-If the maximum value of $\hat{R}$ was larger than 1.1, the total length was increased by 500 with the burn-in increased proportionally and the chain was rerun.
+Each MCMC was run using 4 chains with initial values randomly drawn by {\tt Stan}.
+Each chain ran for a total of 2000 iterations, with 1000 iterations of burn-in and the remaining 1000 draws thinned by a factor of 4 for storage reasons.
+The potential scale reduction factor, $\hat{R}$, was computed for each parameter \citep{Gelm:Rubi:infe:1992, Broo:Gelm:gene:1997}.
+If the maximum value of $\hat{R}$ was larger than 1.1, the total length was increased by 500, with the burn-in increased proportionally, and the chain was rerun.
 This process was repeated until the maximum $\hat{R}$ was less than 1.1 for all parameters.
@@
 For each gene, the quantity of interest is likely to be the maximum of these two probabilities.
-For those genes with high probabilities for one of these two, it is of interest to see the magnitude of the effect.
-Thus, we calculate an estimated effect size using the posterior means for the gene-specific parameters using the formula
+For each gene with a high probability of either HPH or LPH, the magnitude of the effect is of interest.
+Thus, we calculate estimated effect sizes using the estimated posterior means of the gene-specific parameters using the formula
 \begin{equation}
@@
 
-Figure \ref{f:volcano} provides a volcano plot to visualize the maximum of the probabilities for LPH and HPH versus estimated effect size.
-The figure shows a line at zero, for probabilities less than 0.5, due to the estimated effect size being set to 0 for the vast majority of genes.
-Above a probability of 0.5, we see a prototypical volcano pattern with increased extreme heterosis probability corresponding to larger estimated effect sizes and no estimated effect sizes near zero for genes with high extreme heterosis probability.
-Assymetry can also be observed with larger negative effect sizes compared with positive effect sizes which is due to genes whose parental counts are relatively high and hybrid counts near zero.
-Genes with high probability of extreme heterosis and large estimated effect sizes are candidates for further investigation.
+Figure \ref{f:volcano} provides a volcano plot to visualize the maximum of the probabilities of LPH and HPH versus estimated effect size.
+The figure shows a line at zero for probabilities less than 0.5 because the effect size is set to 0 for the vast majority of genes.
+Above a probability of 0.5, we see a prototypical volcano pattern, with increased extreme heterosis probability corresponding to larger estimated effect sizes and no estimated effect sizes near zero for genes with high extreme heterosis probability.
+We also see asymmetry, with larger negative effect sizes than positive effect sizes, because many genes have hybrid counts near zero and relatively high parental counts.
+Genes with high probabilities of extreme heterosis and large estimated effect sizes are candidates for further investigation.
 

### 2014-12-18 ebayes_rnaseq_heterosis.tex
WML: second round of edits complete. Comments at the top of ebayes_rnaseq_heterosis.tex.

@@
+% WML: Throughout, I made several scattered grammatical and clarity fixes without documentation since I couldn't contribute much else.
+I have far fewer comments this time around, so I just grouped them all together here at the top.
+%
+% In section 2.1, your first mention of the sequencing depth terms uses c_{vi}, but those sequencing depths are \gamma_{vi} everywhere else. I didn't change this.
+%
+% I still can't compile jarad.bib. The console tells me that it couldn't find jarad.aux.
+%
+% Please decide between "over-dispersion" and "overdispersion".
+%
+% The definition of MPH is unchanged, and I wonder if we even need to include it.
+%
+% From sections 3 onward, I think you should stick to the present tense.
+%
+% First paragraph of the discussion: "We followed their approach modified slightly to allow for any number of varieties and therefore even more complex hypotheses." This sentence suggests to me that the paper actually took this additional step, which it didn't.
+I think that we should be clear that we only made this future goal more reachable.
+%
+% The short paragraph in the Supplementary Materials section was unclear. I tried to clear it up, but I may have changed the original intended meaning.
+%
+% WML: End of comments.
+
 \documentclass[useAMS,usenatbib,referee]{biom}
@@
 \centerline{\includegraphics[width=\textwidth]{exampleROC0_1}}
-\caption{Repersentative ROC curves for false positive rates below 0.1 for the approaches using {\tt edgeR}, {\tt baySeq},  and {\tt ShrinkBayes} described in Section \ref{s:alternative} and the method using {\tt RStan} assuming independence and with a covariance described in Section \ref{s:gene_heterosis}.}
+\caption{Repersentative ROC curves for false positive rates below 0.1 for the approaches using {\tt edgeR}, {\tt baySeq},  and {\tt ShrinkBayes} described in Section \ref{s:alternative}, along with the independent and covariance methods using {\tt RStan} described in Section \ref{s:gene_heterosis}.}
 \label{f:roc}
@@
 \centerline{\includegraphics[width=\textwidth]{auc-facet-TRUE}}
-\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different sample sizes for the approaches using {\tt edgeR}, {\tt baySeq},  and {\tt ShrinkBayes} described in Section \ref{s:alternative} and the method using {\tt RStan} assuming independence and with a covariance described in Section \ref{s:gene_heterosis}.
+\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different sample sizes for the approaches using {\tt edgeR}, {\tt baySeq},  and {\tt ShrinkBayes} described in Section \ref{s:alternative} , along with the independent and covariance methods using {\tt RStan} described in Section \ref{s:gene_heterosis}.
 Each line is a different simulation while the blue box indicates mean AUC and its standard error across the simulations.}
@@
 \centerline{\includegraphics[width=\textwidth]{volcano}}
-\caption{A two-dimensional histogram of the maximum of LPH and HPH probabilities versus estimated effect size defined in equation \ref{e:effect_size} for the B73 $\times$ Mo17 maize experiment using the independence model.}
+\caption{A two-dimensional histogram of the maximum of the LPH and HPH probabilities versus estimated effect size defined in equation \ref{e:effect_size} for the B73 $\times$ Mo17 maize experiment using the independence model.}
 \label{f:volcano}
@@
 \centerline{\includegraphics[width=\textwidth]{normality}}
-\caption{Marginal and bivariate histograms for gene-specific parameters from a single MCMC iteration, $\theta_g^{(m)}$,  for the B73 $\times$ Mo17 maize experiment.
+\caption{Marginal and bivariate histograms for gene-specific parameters from a single MCMC sample, $\theta_g^{(m)}$,  for the B73 $\times$ Mo17 maize experiment.
 For reference, the marginal histograms have a dashed line indicating the appropriate marginal probability density functions from $N(\hat{\eta},\hat{\Sigma})$.}
@@
 
-Gene expression heterosis is speculated to be one possible explanation for phenotypic
-heterosis of traits like plant height or grain yield.
-Existing methods for identifying gene expression heterosis based on RNAseq data are not directly applicable for detecting heterosis due to the composite null hypothesis in equation \eqref{e:hypotheses}.
+Geneticists speculate that gene expression heterosis is one possible explanation of phenotypic
+heterosis of traits, such as plant height or grain yield.
+Because of the unwieldy composite null hypothesis in equation \eqref{e:hypotheses}, existing methods for identifying gene expression heterosis based on RNAseq data are not directly applicable for detecting heterosis genes.
 \cite{ji2014estimation} introduced an empirical Bayes approach based on a hierarchical model for microarray data.
 We followed their approach modified slightly to allow for any number of varieties and therefore even more complex hypotheses.
-We presented an empirical Bayes method to obtain MAP estimates for hyperparameters followed by MCMC to estimate gene-specific parameters which can be accomplished efficiently via parallelization.
-This methodology can be conveniently implemented in the statistical software {\tt Stan} which also allows for much more complex models.
+We presented an empirical Bayes method to obtain MAP estimates for hyperparameters, followed by an MCMC to estimate gene-specific parameters that can be accomplished efficiently via parallelization.
+This methodology can be conveniently implemented in the statistical software, {\tt Stan}, which also allows for far more complex models.
 The empirical Bayes posteriors can be used to estimate posterior probabilities of high and low parent heterosis.
 
-Through a simulation study, we demonstrated that this method outperformed alternative methods (that are not designed to detect heterosis), and performed comparably with an implementation of our model in ShrinkBayes which estimates the full posterior via INLA.
-We then demonstrated the use of the methodology on a maize experiment where phenotypic heterosis is well known.
+Through a simulation study, we demonstrated that this method outperformed alternative methods (that are not designed to detect heterosis), and performed comparably well with an implementation of our model in {\tt ShrinkBayes}, which estimates the full posterior via INLA.
+We then demonstrated the use of the methodology on a maize experiment in which phenotypic heterosis is well known.
 % Estimates of heterosis to compare with Ji et. al.
-In addition to providing estimates of heterosis probabilities and effect sizes, this real data analysis demonstrated violation of normality assumptions that suggest model improvements to improve power for detection of gene heterosis.
-Developing the details of such an extension is the subject of some of our ongoing and future research.
+In addition to providing estimates of heterosis probabilities and effect sizes, this real data analysis exposed violations of our normality assumptions, indicating room to improve the model and increase the power to detect gene heterosis.
+Some of our ongoing and future research will develop the details of such an extension.
 
@@
 
-The authors thank Andrew Lithio for help implementing our model in {\tt ShrinkBayes}.
+The authors thank Andrew Lithio for help in implementing our model in {\tt ShrinkBayes}.
 Research reported in this publication was supported by National Institute of General Medical Sciences of the National Institutes of Health under award number R01GM109458.
@@
 
-The supplementary materials include the maize experiment data ({\tt data.csv}), a script to run our method ({\tt script.R}), and {\tt Stan} models: all genes independent (\verb@model_ind.txt@) and covariance (\verb@model_cov.txt@) models used in obtaining MAP estimates and single gene independent (\verb@sg_model_ind.txt@) and covariance (\verb@sg_model_cov.txt@) models used in performing the MCMC on single genes.
+The supplementary materials include the maize experiment data ({\tt data.csv}), a script to run our method ({\tt script.R}), and the {\tt Stan} models: the all-genes-independent (\verb@model_ind.txt@) and covariance (\verb@model_cov.txt@) models used in obtaining MAP estimates, along with the single-gene-independent (\verb@sg_model_ind.txt@) and covariance (\verb@sg_model_cov.txt@) models used in performing the MCMC on single genes.
 

### 2014-12-18 ebayes_rnaseq_heterosis.tex
after will's comments

@@
-% WML: Throughout, I made several scattered grammatical and clarity fixes without documentation since I couldn't contribute much else.
-I have far fewer comments this time around, so I just grouped them all together here at the top.
-%
-% In section 2.1, your first mention of the sequencing depth terms uses c_{vi}, but those sequencing depths are \gamma_{vi} everywhere else. I didn't change this.
-%
-% I still can't compile jarad.bib. The console tells me that it couldn't find jarad.aux.
-%
-% Please decide between "over-dispersion" and "overdispersion".
-%
-% The definition of MPH is unchanged, and I wonder if we even need to include it.
-%
-% From sections 3 onward, I think you should stick to the present tense.
-%
-% First paragraph of the discussion: "We followed their approach modified slightly to allow for any number of varieties and therefore even more complex hypotheses." This sentence suggests to me that the paper actually took this additional step, which it didn't.
-I think that we should be clear that we only made this future goal more reachable.
-%
-% The short paragraph in the Supplementary Materials section was unclear. I tried to clear it up, but I may have changed the original intended meaning.
-%
-% WML: End of comments.
-
 \documentclass[useAMS,usenatbib,referee]{biom}
@@
 \graphicspath{{include/}}
-
-%%%%% PLACE YOUR OWN MACROS HERE %%%%%
 
@@
 Let $Y_{gvi}$ be the count for gene $g=1,\ldots,G$, variety $v=1,\ldots,V$, and replicate $i=1,\ldots,n_v$.
-As shown in equation \eqref{e:data}, we assume a negative binomial for the counts with a mean that depends on the gene-variety combination through $\mu_{gv}$ and the sequencing depth $c_{vi}$ and a gene-specific over-dispersion $\psi_g$.
+As shown in equation \eqref{e:data}, we assume a negative binomial for the counts with a mean that depends on the gene-variety combination through $\mu_{gv}$ and the sequencing depth $\gamma_{vi}$ and a gene-specific overdispersion $\psi_g$.
 \begin{equation} 
@@
 $\gamma_{vi} \stackrel{ind}{\sim} N(0,\tau_\gamma^2)$.
-(These terms are the normalization factors that practitioners typically estimate for each sample and then treat as constant in further analysis.) The logarithm of the gene-specific over-dispersion parameters and the gene-specific variety effects are assumed to have a joint normal distribution, i.e.
+(These terms are the normalization factors that practitioners typically estimate for each sample and then treat as constant in further analysis.) The logarithm of the gene-specific overdispersion parameters and the gene-specific variety effects are assumed to have a joint normal distribution, i.e.
 $\theta_g = [\mu_g, \log(\psi_g)]^\top \stackrel{ind}{\sim} N\left(\eta, \Sigma\right)$
@@
 We define gene heterosis to be the differential expression of a gene in the hybrid relative to its parents.
-For a specific gene $g$, low-parent gene heteorisis (LPH) occurs when $\mu_{g3}< \min(\mu_{g1},\mu_{g2})$, mid-parent gene heterosis occurs when $\mu_{g1}\le \mu_{g3}\le \mu_{g2}$ if $\mu_{g1}\le \mu_{g2}$ (or $\mu_{g1}\ge \mu_{g3}\ge \mu_{g2}$ if $\mu_{g1}\ge \mu_{g2}$), and high-parent gene heterosis (HPH) occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$.
+For a specific gene $g$, low-parent gene heteorisis (LPH) occurs when $\mu_{g3}< \min(\mu_{g1},\mu_{g2})$
+%, mid-parent gene heterosis occurs when $\mu_{g1}\le \mu_{g3}\le \mu_{g2}$ if $\mu_{g1}\le \mu_{g2}$ (or $\mu_{g1}\ge \mu_{g3}\ge \mu_{g2}$ if $\mu_{g1}\ge \mu_{g2}$), 
+and high-parent gene heterosis (HPH) occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$.
 For simplicity, we define \emph{extreme gene heterosis} (EH) to be the event of LPH or HPH.
@@
 \cite{ji2014estimation} introduced an empirical Bayes approach based on a hierarchical model for microarray data.
-We followed their approach modified slightly to allow for any number of varieties and therefore even more complex hypotheses.
+We followed their approach modified slightly to allow for any number of varieties and therefore, possibly, even more complex hypotheses.
 We presented an empirical Bayes method to obtain MAP estimates for hyperparameters, followed by an MCMC to estimate gene-specific parameters that can be accomplished efficiently via parallelization.

### 2014-12-18 ebayes_rnaseq_heterosis.tex
updating of shrinkbayes

@@
 
-The {\tt ShrinkBayes} package implements fully Bayesian inference through integrated nested Laplace approximation (INLA) \citep{rue2009approximate,van2014shrinkbayes}.
+The {\tt ShrinkBayes} package implements empirical Bayesian inference via integrated nested Laplace approximation (INLA) \citep{rue2009approximate,van2014shrinkbayes}.
 A limitation of INLA is that it only provides approximations to marginal posteriors for parameters or linear combinations of parameters rather than an approximation to the full posterior.
@@
 
-Geneticists speculate that gene expression heterosis is one possible explanation of phenotypic
-heterosis of traits, such as plant height or grain yield.
+Geneticists speculate that gene expression heterosis is one possible explanation of phenotypic heterosis of traits, such as plant height or grain yield.
 Because of the unwieldy composite null hypothesis in equation \eqref{e:hypotheses}, existing methods for identifying gene expression heterosis based on RNAseq data are not directly applicable for detecting heterosis genes.
 \cite{ji2014estimation} introduced an empirical Bayes approach based on a hierarchical model for microarray data.
-We followed their approach modified slightly to allow for any number of varieties and therefore, possibly, even more complex hypotheses.
-We presented an empirical Bayes method to obtain MAP estimates for hyperparameters, followed by an MCMC to estimate gene-specific parameters that can be accomplished efficiently via parallelization.
-This methodology can be conveniently implemented in the statistical software, {\tt Stan}, which also allows for far more complex models.
+We followed their approach, modified slightly to allow for any number of varieties and therefore even more complex hypotheses.
+We developed an empirical Bayes approach based on obtaining MAP estimates for hyperparameters followed by an MCMC to estimate gene-specific parameters.
+This methodology can be conveniently implemented in the statistical software, {\tt Stan}.
 The empirical Bayes posteriors can be used to estimate posterior probabilities of high and low parent heterosis.

### 2014-12-18 ebayes_rnaseq_heterosis.tex
added blinding command

@@
 
+\newcommand{\blind}{1} % 0 = not blind, 1 = blind
+
 \title{Empirical Bayes analysis for detection of gene heterosis in RNAseq data}
 
+\if0\blind{
 \author{Jarad Niemi$^*$\email{niemi@iastate.edu}, 
@@
 Department of Statistics, Iowa State University, Ames, Iowa, U.S.A.}
+} \fi
 
@@
 
+\if0\blind{
 \section*{Acknowledgements}
@@
 The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health.
+} \fi
 

### 2014-12-18 ebayes_rnaseq_heterosis.tex
added map estimates

@@
 The sample-specific sequencing depth terms are assumed to follow a normal distribution, i.e.
-$\gamma_{vi} \stackrel{ind}{\sim} N(0,\tau_\gamma^2)$.
+$\gamma_{vi} \stackrel{ind}{\sim} N(0,\tau^2)$.
 (These terms are the normalization factors that practitioners typically estimate for each sample and then treat as constant in further analysis.) The logarithm of the gene-specific overdispersion parameters and the gene-specific variety effects are assumed to have a joint normal distribution, i.e.
@@
 
-We assume improper uniform priors for $\eta$ and take $\tau_\gamma\stackrel{ind}{\sim} Ca^+(0,3)$, where 0 and 3 are the location and scale parameters, respectively, of the untruncated half-Cauchy distribution.
+We assume improper uniform priors for $\eta$ and take $\tau\stackrel{ind}{\sim} Ca^+(0,3)$, where 0 and 3 are the location and scale parameters, respectively, of the untruncated half-Cauchy distribution.
 In the independent model, we also assign the $Ca^+(0,3)$ prior to the diagonal elements of $\Sigma$.
@@
 
-We found MAP estimates for both the independence and covariance models as shown in Table \ref{t:map}. 
-% Need MAP estimates table and discussion
+We found MAP estimates for both the independent and covariance models as shown in Table \ref{t:map}. 
+\input{include/map}
+The MAP estimates for the hierarchical expression means, $\eta_i = E[\mu_{gi}]$ for $i=1,2,3$, are consistent between the the independent and covariance model while the mean for the overdispersion parameter, $\eta_4=E[\log(\psi)]$, is a bit large in the covariance model.
+The standard deviation around the expression means are approximately twice as large in the covariance model as in the independent model while the standard deviation for the overdispersion is half as large.
+The sequencing depth terms are consistent between the two models and indicate that there is about a 25\% increase in counts between the least and most thoroughly sequenced samples.
+In the covariance model, there appears to be moderate correlation amongst the expression means and the expression means are negatively correlated with the (log) overdispersion parameter.
 

### 2014-12-19 ebayes_rnaseq_heterosis.tex
updated discussion of map estimates

--- 2014-12-18 ebayes_rnaseq_heterosis.tex
@@
 
-We found MAP estimates for both the independence and covariance models as shown in Table \ref{t:map}. 
-% Need MAP estimates table and discussion
+We found MAP estimates for both the independent and covariance models as shown in Table \ref{t:map}. 
+\input{include/map}
+The MAP estimates for the hierarchical means, $\eta$, and standard deviations, $\sigma$, are consistent between the the independent and covariance model with, perhaps, a slightly larger mean and smaller standard deviation for the logarithm of the overdispersion parameter, $\log(\psi_g)$ in the covariance model relative to the independent model.
+In the covariance model, there appears to be high correlation amongst the expression means, $\eta_i$ for $i=1,2,3$, and the expression means are negatively correlated with the (log) overdispersion parameter.
+The sequencing depth terms, and their variance, are consistent between the two models, with the estimates for the covariance model being consistently 0.02 larger than the independent model, and both indicating that there is about a 25\% increase in counts between the least and most thoroughly sequenced samples.
 

### 2014-12-19 ebayes_rnaseq_heterosis.tex
fixed up some of the real data analysis

@@
 The potential scale reduction factor, $\hat{R}$, was computed for each parameter \citep{Gelm:Rubi:infe:1992, Broo:Gelm:gene:1997}.
-If the maximum value of $\hat{R}$ was larger than 1.1, the total length was increased by 500, with the burn-in increased proportionally, and the chain was rerun.
+If the maximum $\hat{R}$ was larger than 1.1 for any parameter, the total number of iterations was increased by 500, with the burn-in increased proportionally, and the chain was rerun.
 This process was repeated until the maximum $\hat{R}$ was less than 1.1 for all parameters.
@@
 For each gene with a high probability of either HPH or LPH, the magnitude of the effect is of interest.
-Thus, we calculate estimated effect sizes using the estimated posterior means of the gene-specific parameters using the formula
+Thus, we calculate estimated effect sizes using the estimated posterior means of the gene-specific parameters using the formula in equation \ref{e:effect_size} where a 0 indicates an estimated mean expression for the hybrid that is between those of the parents, i.e.
+not extreme heterosis.
 \begin{equation}

### 2014-12-22 ebayes_rnaseq_heterosis.tex
Added comments to model section.

--- 2014-12-18 ebayes_rnaseq_heterosis.tex
@@
 
-We assume improper uniform priors for $\eta$ and take $\tau\stackrel{ind}{\sim} Ca^+(0,3)$, where 0 and 3 are the location and scale parameters, respectively, of the untruncated half-Cauchy distribution.
-In the independent model, we also assign the $Ca^+(0,3)$ prior to the diagonal elements of $\Sigma$.
+We assume improper uniform priors for $\eta$ and take 
+%In fact, the prior on \tau is just Unif(0,5)
+$\tau\stackrel{ind}{\sim} Ca^+(0,3)$, where 0 and 3 are the location and scale parameters, respectively, of the untruncated half-Cauchy distribution.
+In the independent model, we also assign the $Ca^+(0,3)$ prior to the
+diagonal elements
+%square root of diagonal elements?
+of $\Sigma$.
 In the covariance model, we assume a conjugate, proper inverse Wishart distribution with degrees of freedom $V+2$ and an identity scale matrix, resulting in marginally uniform correlations.

### 2015-03-02 ebayes_rnaseq_heterosis.tex
added publisher
post Dan edits, but pre comment fixes

@@
 
-\newcommand{\blind}{1} % 0 = not blind, 1 = blind
+\newcommand{\blind}{0} % 0 = not blind, 1 = blind
 
-\title{Empirical Bayes analysis for detection of gene heterosis in RNAseq data}
+\title{Empirical Bayes analysis of RNA-seq data for detection of gene expression heterosis}
 
@@
 Heterosis, or hybrid vigor, refers to the enhancements in the phenotype of hybrid progeny relative to their inbred parents.
-Although the heterosis phenomenon is extensively utilized in agriculture, the molecular basis is still unknown.
-In an effort to understand phenotypic heterosis at the molecular level, researchers have begun to compare expression levels of thousands of genes between parental inbred lines and their hybrid offspring to search for evidence of gene expression heterosis in RNA sequencing (RNAseq) experiments.
-Standard statistical approaches for RNAseq experiments struggle with the composite null hypothesis in gene heterosis testing.
-To address these shortcomings, we develop a hierarchical model based on a negative-binomial data model and a computationally tractable empirical Bayes approach.
-We demonstrate improvements over alternative methods via a simulation study based on a maize experiment and then analyze that maize experiment.
+Although the heterosis phenomenon is extensively utilized in agriculture, the molecular basis of heterosis is still largely unknown.
+In an effort to understand phenotypic heterosis at the molecular level, researchers have begun to measure transcript abundance levels of thousands of genes in parental inbred lines and their hybrid offspring using RNA sequencing (RNA-seq) technology.
+The resulting data allow researchers to search for evidence of gene expression heterosis as one potential molecular mechanism underlying heterosis of agriculturally important traits.
+The null hypotheses of greatest interest in testing for gene expression heterosis are composite null hypotheses that are difficult to test with standard statistical approaches for RNA-seq analysis.
+To address these shortcomings, we develop a hierarchical model based on a negativeb binomial data model and draw inferences using a computationally tractable empirical Bayes approach to inference.
+We demonstrate improvements over alternative methods via a simulation study based on a maize experiment and then analyze that maize experiment with our newly proposed methodology.
 This article has supplementary material online.
@@
 \begin{keywords}
-Empirical Bayes; Gene expression; Heterosis; Hierarchical model; Negative binomial; RNAseq.
+Empirical Bayes; Gene expression; Heterosis; Hierarchical model; Negative binomial; RNA-seq.
 \end{keywords}
@@
 
-% Need to udpate as this is directly from Tieming's paper
-Heterosis, or hybrid vigor, refers to the enhancements in the phenotype of hybrid progeny relative to their inbred parents.
-Taking maize as an example, the offspring from crossing the inbred lines B73 and Mo17 grow taller, mature faster, and produce greater yields than their parent lines \citep{hallauer1981quantitative, hallauer2010quantitative}.
-Ever since heterosis was scientifically documented by \cite{darwin1876effects}, it has been successfully manipulated to improve many species for the food, feed, and fuel industries, such as rice \citep{yu1997importance}, alfalfa \citep{riday2002heterosis}, tomatoes \citep{krieger2010flowering}, and fish \citep{wohlfarth1993heterosis}.
-Despite the intensive study and successful utilization of heterosis, the basic genomic mechanisms remain unclear \citep{coors1999genetics, lippman2007heterosis}.
-Researchers speculate that gene-expression heterosis $-$ i.e, enhanced expression of one or more genes in the hybrid relative to the inbred parents $-$ is a possible  mechanisms for the phenotypic heterosis \citep{swanson2006all, springer2007allelic}.
+Heterosis, or hybrid vigor, occurs when hybrid progeny display phenotypes that are superior to the phenotypes of their parents.
+The heterosis phenomenon was scientifically documented in plants by \cite{darwin1876effects}and has long been used to improve agricultural production.
+One classic example involves hybrid maize offspring that are taller, faster to mature, and yield considerably more grain than their inbred parents \citep{hallauer1981quantitative, hallauer2010quantitative}.
+Despite intensive study and successful use of heterosis in agriculture, the basic molecular genetic mechanisms remain largely unknown \citep{coors1999genetics, lippman2007heterosis}.
+One potential explanation, known as gene expression heterosis, is enhanced expression of one or more genes in hybrids compared to their inbred parents \citep{swanson2006all, springer2007allelic}.
 
-Recently, \cite{ji2014estimation} introduced an approach to assess gene heterosis using microarray data under the assumption that these data are continuous.
-They built a normal hierarchical model for gene expression that allows borrowing of information across genes for means and variances.
+Recently, \cite{ji2014estimation} introduced an approach to assess gene expression heterosis using microarray data under the assumption that these data are continuous.
+They built a normal hierarchical model for microarray measurements of transcript abundance that allows borrowing of information across genes to estimate means and variances.
 They introduced an empirical Bayes framework that first estimates model hyperparameters, then estimates the posterior distribution for gene-specific parameters conditional on those hyperparameters, and finally computes heterosis probabilities based on integrals of regions under this posterior.
-This development was necessary due to the composite null hypothesis in tests for heterosis.
-These hypotheses, which many available methods do not fully accommodate, remain a challenge in the transition from continuous data to count data.
+This development was necessary due to the composite null hypotheses in tests for heterosis.
+These hypotheses, which many available methods do not fully accommodate, remain a challenge in the transition from continuous measurements of transcript abundance to count-based measurements that arise from RNA sequencing (RNA-seq) technology.
 Building on the work of \citeauthor{ji2014estimation} with the normal data model, we construct a hierarchical model based on a negative-binomial data model.
@@
 The remainder of the paper proceeds as follows.
-Section \ref{s:method} presents the proposed hierarchical model, an empirical Bayes method of estimating the parameters, and the calculation of posterior probabilities of gene heterosis.
+Section \ref{s:method} presents the proposed hierarchical model, an empirical Bayes method of estimating the parameters, and the calculation of posterior probabilities of gene expression heterosis.
 Section \ref{s:simulation} presents a simulation study based on a maize experiment and compares our approach to alternative methods.
-Section \ref{s:maize} analyzes a maize experiment where phenotypic heterosis is well established and identifies genes demonstrating heterosis.
-The data and code necessary to perform this analysis is available in the Supplemental Material.
+Section \ref{s:maize} analyzes a maize experiment where phenotypic heterosis is well established and identifies genes demonstrating expression heterosis.
+The data and code necessary to perform this analysis are available in the Supplemental Material.
 Section \ref{s:discussion} summarizes the work and suggests directions for future research.
 
-\section{Empirical Bayes identification of gene heterosis from RNAseq counts}
+\section{Empirical Bayes identification of gene expression heterosis from RNA-seq read counts}
 \label{s:method}
 
-An RNA sequencing (RNAseq) experiment to identify genes demonstrating heterosis consists of three genetic varieties: two parental varieties (P1 and P2) and a cross between these two varieties called the hybrid (H).
+An RNA sequencing (RNA-seq) experiment to identify genes demonstrating expression heterosis consists of three genetic varieties: two parental varieties (P1 and P2) and a cross between these two varieties called the hybrid (H).
 For each variety, replicate RNA samples are isolated and assessed for quality.
-cDNA libraries, or collections of short cDNA fragments, are constructed.
+Complementary DNA (cDNA) libraries, consisting of short cDNA fragments derived from RNA, are constructed.
 Then, next generation sequencing technology is used to determine the \emph{reads}, or nucleotide sequences, in the cDNA libraries.
 These reads are processed using bioinformatic algorithms to match the reads to genes, or specific gene transcripts, exons, microRNAs, etc.
-The result of matching reads to features are summarized by a gene $\times$ sample matrix of counts.
-See \cite{nettleton2014design} for more details on RNAseq experiments, and see \cite{paschold2012complementation} for the biological background behind the above steps.
+The results of read processing are summarized by a gene $\times$ sample matrix of counts.
+See \cite{nettleton2014design}
+% JBN: cite the book 
+for more details on RNA-seq experiments and data from a statistical perspective, and see \cite{paschold2012complementation} for the biological background behind the use of RNA-seq to study gene expression heterosis.
 
-To use RNAseq counts to identify genes displaying heterosis, we build a hierarchical model to borrow information on gene-variety means and gene-specific overdispersion parameters, estimate the parameters using a empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate empirical Bayes posterior probabilities for heterosis.
+To use RNA-seq counts to identify genes displaying expression heterosis, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the parameters using an empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate empirical Bayes posterior probabilities for heterosis.
 
-\subsection{Hierarchical model for RNAseq counts}
+\subsection{Hierarchical model for RNA-seq counts}
 \label{s:model}
 
-Let $Y_{gvi}$ be the count for gene $g=1,\ldots,G$, variety $v=1,\ldots,V$, and replicate $i=1,\ldots,n_v$.
-As shown in equation \eqref{e:data}, we assume a negative binomial for the counts with a mean that depends on the gene-variety combination through $\mu_{gv}$ and the sequencing depth $\gamma_{vi}$ and a gene-specific overdispersion $\psi_g$.
+Let $Y_{gvi}$ be the count for gene $g=1,\ldots,G$, variety $v=1,\ldots,V$, and replicate $i=1,\ldots,n_v$.  We assume
 \begin{equation} 
-Y_{gvi} \stackrel{ind}{\sim} NB(e^{\mu_{gv}+\gamma_{vi}},\psi_g) 
+Y_{gvi} \stackrel{ind}{\sim} NB(e^{\mu_{gv}+\gamma_{vi}},\psi_g), 
 \label{e:data}
 \end{equation}
-where $NB(\eta,\psi)$ indicates a negative-binomial with expectation $\eta$ and variance $\eta+\psi\eta^2$, and $ind$ indicates the observations are conditionally independent.
+where $NB(\eta,\psi)$ indicates a negative binomial distribution with expectation $\eta$ and variance $\eta+\psi\eta^2$, and $ind$ indicates the observations are conditionally independent.
+As shown in equation \eqref{e:data}, our data model involves gene-specific overdispersion $\psi_g$ and a mean that depends on the gene-variety combination through $\mu_{gv}$ and on the sample though  $\gamma_{vi}$.
+The $\mu_{gv}$ terms are of primary scientific interest; the $\gamma_{vi}$ terms are normalization factors that account for differences in the thoroughness of sequencing from sample to sample.
+Practitioners typically compute normalization factors for each sample and then treat them as known constants in further analysis.
+In contrast, we include normalization factors in our model as unknowns that we estimate along with other parameters.
 
-We present a hierarchical model for the sequencing depth, overdispersion, and gene-variety parameters.
+We present a hierarchical model for the normalization factors, overdispersions, and gene-variety parameters.
 The sample-specific sequencing depth terms are assumed to follow a normal distribution, i.e.
@@
 where $\mu_g = (\mu_{g1},\ldots,\mu_{gV})'$.
-We consider an \emph{independent} model in which $\Sigma$ is diagonal with elements $\sigma_1^2,\ldots,\sigma_{V+1}^2$ and a \emph{covariance} model in which we estimate all parameters in $\Sigma$.
+We consider an \emph{independence} model in which $\Sigma$ is diagonal with elements $\sigma_1^2,\ldots,\sigma_{V+1}^2$ and a \emph{dependence} model in which $\Sigma$ is a general positive definite matrix.
+% JBN: Switch mu and eta throughout
 
@@
 $\tau\stackrel{ind}{\sim} Ca^+(0,3)$, where 0 and 3 are the location and scale parameters, respectively, of the untruncated half-Cauchy distribution.
-In the independent model, we also assign the $Ca^+(0,3)$ prior to the
+In the independence model, we also assign the $Ca^+(0,3)$ prior to the
 diagonal elements
@@
 of $\Sigma$.
-In the covariance model, we assume a conjugate, proper inverse Wishart distribution with degrees of freedom $V+2$ and an identity scale matrix, resulting in marginally uniform correlations.
+In the dependence model, we assume a conjugate, proper inverse Wishart distribution with degrees of freedom $V+2$ and an identity scale matrix, resulting in marginally uniform correlations.
 
@@
 We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ and hyperparameters $\pi = (\gamma,\eta, \Sigma)$, where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$.
-We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\pi$, denoted $\hat{\pi}$, and then base gene-specific inference on the posterior conditional of this estimated hyperparameter.
-Conditional on the hyperparameters, the gene-specific parameters are independent as in equation \eqref{e:condind}.
-Thus, conditional posterior inference for the gene-specific parameters can be obtained independently and in parallel.
+We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\pi$, denoted $\hat{\pi}$, and then base gene-specific inference on the posterior conditional on $\hat{\pi}$.
+Conditional on the hyperparameters, the gene-specific parameters are independent, i.e.,
 \begin{align}
@@
 %&= \prod_{g=1}^G p\left(\theta_g\left|y_g,\hat{\pi}\right.\right) = \prod_{g=1}^G p\left(y_g\left|\theta_g,\hat{\gamma}\right.\right) p\left(\theta_g\left|\hat{\pi}\right.\right) \nonumber \\
-&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\gamma}_{vi}},\psi_g\right.\right) \right] N\left(\left.\left[\mu_g, \log(\psi_g) \right]^\top\right|\hat{\eta}, \hat{\Sigma} \right)
+&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\gamma}_{vi}},\psi_g\right.\right) \right] N\left(\left.\left[\mu_g, \log(\psi_g) \right]^\top\right|\hat{\eta}, \hat{\Sigma} \right).
 \label{e:condind}
 \end{align}
+Thus, conditional posterior inference for the gene-specific parameters can be obtained independently and in parallel.
 
-To perform the optimization, we use the statistical software, {\tt Stan} \citep{stan-software:2014}, executed through the RStan interface \citep{rstan-software:2014} in R \citep{R2014}.
+To perform the optimization required for the MAP estimate $\hat{\pi}$, we use the statistical software, {\tt Stan} \citep{stan-software:2014}, executed through the RStan interface \citep{rstan-software:2014} in R \citep{R2014}.
 The Stan software allows the construction of general Bayesian models with the primary restriction that all unknown parameters are continuous.
 With a fully constructed model, the software can obtain MAP estimates \cite[Section 50.3,][]{stan-manual:2014} using the {\tt optimizing} function in RStan.
-It was necessary to use the L-BFGS algorithm \cite[Section 55,][]{stan-manual:2014} and reasonable starting values obtained via {\tt edgeR}, presumably due to the large dimension of the optimization.
+Presumably due to the large dimension of the optimization, it was necessary to use the L-BFGS algorithm \cite[Section 55,][]{stan-manual:2014} and reasonable starting values obtained via {\tt edgeR} \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor}.
 The {\tt Stan} software also includes the RStan function, {\tt sampling}, an implementation of a Markov chain Monte Carlo (MCMC) algorithm for obtaining samples from the posterior.
@@
 
-\subsection{Gene heterosis}
+\subsection{Gene expression heterosis}
 \label{s:gene_heterosis}
 
-As defined in Section \ref{s:intro}, heterosis is increased (or decreased) phenotypic expression in a hybrid relative to its inbred parents.
-Thus, when attempting to identify heterosis, there are $V=3$ varieties of interest.
-We let $v=1,2$ be the two parental varieties and $v=3$ be the hybrid.
-We define gene heterosis to be the differential expression of a gene in the hybrid relative to its parents.
-For a specific gene $g$, low-parent gene heteorisis (LPH) occurs when $\mu_{g3}< \min(\mu_{g1},\mu_{g2})$
-%, mid-parent gene heterosis occurs when $\mu_{g1}\le \mu_{g3}\le \mu_{g2}$ if $\mu_{g1}\le \mu_{g2}$ (or $\mu_{g1}\ge \mu_{g3}\ge \mu_{g2}$ if $\mu_{g1}\ge \mu_{g2}$), 
-and high-parent gene heterosis (HPH) occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$.
-For simplicity, we define \emph{extreme gene heterosis} (EH) to be the event of LPH or HPH.
-Equation \eqref{e:hypotheses} shows the most important hypotheses, where $H_0$ indicates no EH and $H_1$ indicates EH.
+To identify gene expression heterosis, we consider $V=3$ varieties, where  $v=1,2$ signify the two parental varieties and $v=3$ represents the hybrid.
+For a specific gene $g$, low-parent  heteorisis (LPH) of gene expression occurs when $\mu_{g3}< \min(\mu_{g1},\mu_{g2})$
+%, mid-parent gene expression heterosis occurs when $\mu_{g1}\le \mu_{g3}\le \mu_{g2}$ if $\mu_{g1}\le \mu_{g2}$ (or $\mu_{g1}\ge \mu_{g3}\ge \mu_{g2}$ if $\mu_{g1}\ge \mu_{g2}$), 
+and high-parent  heterosis (HPH) of gene expression occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$.
+For simplicity, we define \emph{ heterosis} of gene expression to be the event of LPH or HPH.
+The most important hypotheses are
 \begin{align}
 H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
-H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}.
+H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3},
 \label{e:hypotheses}
 \end{align}
-We evaluate these hypotheses based on empirical Bayes estimates of their posterior probabilities, e.g. 
+where $H_{g0}$ indicates no heterosis for gene $g$ and $H_{g1}$ indicates  heterosis for gene $g$. 
+We evaluate these hypotheses based on empirical Bayes estimates of their posterior probabilities, e.g., 
 \begin{align}
 P\left(H_{g1}|y, \hat{\pi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\pi}_{MAP}\right) \nonumber \\
-&\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right) \label{e:probs}
+&\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right), \label{e:probs}
 \end{align}
 where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise.
-The set of probabilities $P\left(H_{g1}|y, \hat{\pi}_{MAP}\right)$ for $g=1,\ldots,G$ can be used to provide a list of genes sorted by the relative strength of evidence of extreme heterosis.
-Geneticists can use this list to prioritize future experiments to understand the genetic mechanism for heterosis.
+The set of probabilities $P\left(H_{g1}|y, \hat{\pi}_{MAP}\right)$ for $g=1,\ldots,G$ can be used to provide a list of genes sorted by the relative strength of evidence of  heterosis.
+Geneticists can use this list to prioritize future experiments to understand the molecular genetic mechanisms for heterosis.
 
-\section{Simulation study based on maize experiment}
+\section{Simulation study based on a maize experiment}
 \label{s:simulation}
@@
 
-To assess the efficacy of our method to identify genes demonstrating heterosis, we used a maize data set with parental varieties B73 and Mo17 and the hybrid variety (B73 $\times$ Mo17) \citep{paschold2012complementation}.
-Section \ref{s:maize} describes this dataset in detail.
-We compared our method to approaches using the R packages {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes}.
+To assess the efficacy of our method to identify genes demonstrating heterosis, we used a maize data set with parental varieties B73 and Mo17 and the hybrid variety (B73 $\times$ Mo17) \citep{paschold2012complementation} to determine realistic parameter values for a simulation study.
+Section \ref{s:maize} describes the maize dataset in detail.
+We compared our method to approaches using the R packages {\tt edgeR}, {\tt baySeq} \citep{hardcastle2010bayseq, hardcastle2012baySeq}, and {\tt ShrinkBayes} \citep{van2014shrinkbayes}.
 
@@
 
-Initially, the maize experiment had RNAseq counts for 39,656 genes and 4 replicates per variety.
-We eliminated low count genes by removing any gene with three or more zeros for any variety and genes whose mean count across the varieties was less than one, resulting in 27,888 remaining genes.
+We applied the {\tt edgeR} package \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor} to the maize data on 27,888 genes with average count at least one and at most two zero read counts for each variety across four replications.
+This analysis produced  sample-specific normalization factors () % JBN: state the normalization factors, "I suppose you used the edgeR default, which is TMM, but I think this should be stated explicitly."
+, gene-specific dispersion estimates, and gene-variety-specific location parameter estimates.
+We treated these estimates as the true parameter values for the simulation study so that our simulated datasets mimicked the existing covariance structure among the gene-variety means of the original maize data.
 
-Next, we applied the {\tt edgeR} package \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor} to the maize data to obtain sample-specific normalization factors, then gene-specific dispersions, and finally gene-variety-specific location parameters.
-We treated these parameters as the truth for the simulation study: i.e.
-a gene with $\mu_{g3} < \min(\mu_{g1}, \mu_{g2})$ was considered to have LPH, while a gene with $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$ was considered to have HPH.
-This technique ensured that our simulated datasets maintained the existing covariance structure among the gene-variety means of the original maize data.
-
-Using these parameters and normalization factors, we simulated data according to the negative-binomial model in equation \eqref{e:data} independently for each gene.
-For each simulation, low count genes $-$ those with three or more zeros in any variety or whose mean count across the varieties was less than one $-$ were removed.
-We retained a random subset of 25,000 genes from those with high enough counts.
+Using these parameters and normalization factors, we simulated data according to the negative-binomial model in \eqref{e:data} independently for each gene.
+For each simulation, we analyzed a subset of 25,000 genes selected randomly from genes with simulated counts at least one on average and with at most two zeros for each variety across replications.
 We repeated this simulation process 10 times for each of 4, 8, and 16 replicates per variety, reusing normalization factors when necessary.
+% From Dan "Did we consider changing the number of zeros allowed when the number of reps is higher?
+I like the idea of simplifying all this by just picking an average count threshold that his high enough to avoid problems.
+I think that may be what others have done in the past."
 
@@
 
-Existing RNAseq data analysis methods are not specifically designed to identify gene heterosis, but we adapted the approaches in {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes} to provide a measure of the strength of extreme heterosis for each gene such that large values of this measure indicate support for extreme heterosis.
+Existing RNA-seq data analysis methods are not specifically designed to identify gene expression heterosis, but we adapted the approaches in {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes} to provide a measure of the strength of heterosis for each gene such that large values of this measure indicate support for heterosis.
 
-The {\tt edgeR} approach is designed to apply a negative-binomial log-linear model to test for differential expression between any two varieties \citep{robinson2007moderated, robinson2010edgeR}.
-To construct a measure of heterosis, we computed the maximum likelihood estimates of the mean expression effects of all genes using {\tt edgeR}'s built-in Fisher scoring algorithm, and then used the likelihood ratio test functionality to calculate two legitimate p-values for each gene: one for the null hypothesis that the hybrid's mean expression is the same as that of parent 1 and the second for the analogous test involving parent 2.
-For each gene, if the estimated mean expression level of the hybrid is between the estimated mean expression levels of the parents, we set the measure of heterosis to zero.
-Otherwise, we set the measure of heterosis to be one minus the p-value corresponding to the parent with estimated expression level closer to that of the hybrid.
+The {\tt edgeR} approach can be used to test for differential expression between any two varieties based on the fit of a negative binomial log-linear model \citep{robinson2007moderated, robinson2010edgeR}.
+To construct a measure of heterosis, we computed the maximum likelihood estimates of the $\mu_{gv}$ parameters for all genes using {\tt edgeR}'s built-in Fisher scoring algorithm, and then used likelihood ratio tests to calculate two $p$-values for each gene: $p_{g1}$  for testing $H_{g01}:\mu_{g1}=\mu_{g3}$ and $p_{g2}$ for testing $H_{g02}:\mu_{g2}=\mu_{g3}$.
+Then, for each gene, we define a new $p$-value denoted as $p_g$ and determined as $p_g=1$ if the estimate of $\mu_{g3}$ falls between the estimates of $\mu_{g1}$ and $\mu_{g2}$ and $p_g=\max\{p_{g1},p_{g2}\}/2$ otherwise.
+(For all relevant signifcance thresholds $\omega$ near 0, it can be shown that rejecting the null hypothesis of no heterosis of expression for gene $g$ whenever $p_g\leq \omega$ results in a test that is asymptotically size $\omega$.)  We then use $1-p_g$ as a final measure positively associated with strength of evidence for gene expression heterosis.
 
-The {\tt baySeq} approach \citep{hardcastle2010bayseq, hardcastle2012baySeq} allows for a wider range of hypotheses for each gene, including $H_1: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H_2: \mu_{g1}=\mu_{g2}$, $H_3: \mu_{g1}=\mu_{g3}$, $H_4: \mu_{g2}=\mu_{g3}$, and $H_5:$ all $\mu_g$'s are distinct.
+The {\tt baySeq} approach allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct.
 In a technique similar to our application of {\tt edgeR}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene.
-We set this measure to zero if the hybrid's maximum likelihood estimate of the mean expression level calculated using {\tt edgeR} is between those of the parents.
-Otherwise, the measure is the sum of the posterior probabilities of $H_2$ and $H_5$, the only hypotheses that allow for extreme heterosis.
+We set this measure to zero if the maximum likelihood estimate, calculated using {\tt edgeR}, of $\mu_{g3}$ is between the maximum likelihood estimates of $\mu_{g1}$ and $\mu_{g2}$.
+Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the only hypotheses that allow for heterosis.
 
@@
 A limitation of INLA is that it only provides approximations to marginal posteriors for parameters or linear combinations of parameters rather than an approximation to the full posterior.
-Thus, we cannot directly estimate nonlinear quantities such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2})|y)$.
+Thus, we cannot directly estimate posterior probabilities involving nonlinear functions of $\mu_{g1}$, $\mu_{g2}$, and $\mu_{g3}$, such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2})|y)$.
 To circumvent this limitation, we reparameterized the independence model in Section \ref{s:model} using the approach in \cite{ji2014estimation}.
-Let $\phi_g = (\mu_{g1}+\mu_{g2})/2$ be the parental log-mean, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental difference, and $\delta_g = \mu_{g3} - \phi_g$ be the hybrid effect.
+Let $\phi_g = (\mu_{g1}+\mu_{g2})/2$ be the average of parental effects, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental effect difference, and $\delta_g = \mu_{g3} - \phi_g$ be the hybrid effect relative to the average parental effect.
 After fitting the model with {\tt ShrinkBayes}, we constructed a measure of heterosis for each gene.
@@
 
-For each of the methods above, as well as the method described in Section \ref{s:gene_heterosis}, we sorted genes according to the plausibility of heterosis.
-From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to identify genes with heterosis, as defined in Section \ref{s:sim_data}  \citep{landau2013dispersion}.
-A representative set of ROC curves is shown in Figure \ref{f:roc}.
+For each of the methods above, as well as the method described in Section \ref{s:gene_heterosis}, we sorted genes according to the computed measure of the strength of evidence for heterosis.
+From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to distinguish genes with heterosis, as defined in Section \ref{s:sim_data}  \citep{landau2013dispersion},
+% From Dan, "why this citation?"
+from those without heterosis. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
 \begin{figure}[htbp]
 \centerline{\includegraphics[width=\textwidth]{exampleROC0_1}}
-\caption{Repersentative ROC curves for false positive rates below 0.1 for the approaches using {\tt edgeR}, {\tt baySeq},  and {\tt ShrinkBayes} described in Section \ref{s:alternative}, along with the independent and covariance methods using {\tt RStan} described in Section \ref{s:gene_heterosis}.}
+\caption{Representative ROC curves for false positive rates below 0.1 for the approaches using {\tt edgeR}, {\tt baySeq},  and {\tt ShrinkBayes} described in Section \ref{s:alternative}, along with the independence and dependence methods using {\tt RStan} described in Section \ref{s:gene_heterosis}.}
 \label{f:roc}
@@
 For this simulation, we can see that the approaches based on the model in Section \ref{s:model} $-$ i.e.
-both RStan approaches and the ShrinkBayes approach $-$ provide the best true positive rate for a given false positive rate.
+both {\tt RStan} approaches and the {\tt ShrinkBayes} approach $-$ provide the best true positive rate for a given false positive rate.
 Also, as expected, as the sample size increases, our ability to distinguish genes with heterosis from genes without improves.
@@
 \centerline{\includegraphics[width=\textwidth]{auc-facet-TRUE}}
-\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different sample sizes for the approaches using {\tt edgeR}, {\tt baySeq},  and {\tt ShrinkBayes} described in Section \ref{s:alternative} , along with the independent and covariance methods using {\tt RStan} described in Section \ref{s:gene_heterosis}.
-Each line is a different simulation while the blue box indicates mean AUC and its standard error across the simulations.}
+\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different sample sizes for the approaches using {\tt edgeR}, {\tt baySeq},  and {\tt ShrinkBayes} described in Section \ref{s:alternative} , along with the independence and dependence methods using {\tt RStan} described in Section \ref{s:gene_heterosis}.
+Each line is a different simulation while the blue box indicates mean AUC and mean AUC plus and minus one standard error across the simulations.}
 \label{f:auc}
@@
 The pattern appears to hold, with the {\tt RStan} and {\tt ShrinkBayes} techniques outperforming the other two.
-While there does not appear to be a significant difference between {\tt ShrinkBayes} and the independence model using {\tt RStan}, performance appears to decrease for the model where a covariance is estimated, at least for smaller sample sizes.
+While there does not appear to be a significant difference between {\tt ShrinkBayes} and the independence model using {\tt RStan}, performance for {\tt RStan} appears to decrease for the dependence model where an unstructured covariance is estimated, at least for smaller sample sizes.
 
-\section{Heterosis in maize experiment}
+\section{Searching for gene expression heterosis in the maize experiment}
 \label{s:maize}
 
-We used our method to analyze a maize data set on gene expression in parental lines B73 and Mo17 and the hybrid genotype (B73$\times$Mo17).
+We used our method to analyze the maize data set, generated by \cite{paschold2012complementation}, on gene expression in parental lines B73 and Mo17 and the hybrid genotype (B73$\times$Mo17).
 The data are available in the Supplemental Materials.
@@
 
-A total of 39,656 genes among $V=3$ varieties, each with $n_v=4$ replicates, were available for analysis.
-As before, we eliminated genes with three or more zeros for any variety and genes with mean expression less than one across all samples, resulting in a total of $G=27,888$ genes present in the analysis.
+As explained in Section~\ref{s:sim_data}, we analyzed a set of $G=27,888$ genes for $V=3$ varieties and $n_v=4$ replicates ($v=1,2,3$).
 
-We found MAP estimates for both the independent and covariance models as shown in Table \ref{t:map}. 
+We found MAP estimates for both the independence and dependence models as shown in Table \ref{t:map}. 
 \input{include/map}
-The MAP estimates for the hierarchical means, $\eta$, and standard deviations, $\sigma$, are consistent between the the independent and covariance model with, perhaps, a slightly larger mean and smaller standard deviation for the logarithm of the overdispersion parameter, $\log(\psi_g)$ in the covariance model relative to the independent model.
-In the covariance model, there appears to be high correlation amongst the expression means, $\eta_i$ for $i=1,2,3$, and the expression means are negatively correlated with the (log) overdispersion parameter.
-The sequencing depth terms, and their variance, are consistent between the two models, with the estimates for the covariance model being consistently 0.02 larger than the independent model, and both indicating that there is about a 25\% increase in counts between the least and most thoroughly sequenced samples.
+The MAP estimates for the hierarchical means, $\eta$, and standard deviations, $\sigma$, are consistent between the independence and dependence models with, perhaps, a slightly larger mean and smaller standard deviation for the logarithm of the overdispersion parameter, $\log(\psi_g)$, in the dependence model relative to the independence model.
+In the dependence model, there appears to be high correlation amongst $\mu_{g1}$, $\mu_{g2}$, and $\mu_{g3}$, and negative correlation between $\log(\psi_g)$ and each of $\mu_{g1}$, $\mu_{g2}$, and $\mu_{g3}$.
+The normalization factors, and their variance, are consistent between the two models, with the estimates for the dependence model being consistently 0.02 larger than the estimates for the independence model, and both indicating that there is about a 25\% increase in counts between the least and most thoroughly sequenced samples.
 
-Independently for each gene, we ran an MCMC conditional on the hyperparameters to obtain empirical Bayes posterior distributions for all gene-specific parameters.
-Each MCMC was run using 4 chains with initial values randomly drawn by {\tt Stan}.
-Each chain ran for a total of 2000 iterations, with 1000 iterations of burn-in and the remaining 1000 draws thinned by a factor of 4 for storage reasons.
+Independently for each gene, we ran four Markov chains conditional on the estimated hyperparameters to obtain approximate posterior distributions for all gene-specific parameters.
+Each Markov chain used initial values randomly drawn by {\tt Stan} and was run for a total of 2000 iterations, with 1000 iterations of burn-in and the remaining 1000 draws thinned by a factor of 4 for storage reasons.
+% From Dan "Do we need to be more specific about what method was used to generate the chains?"
 The potential scale reduction factor, $\hat{R}$, was computed for each parameter \citep{Gelm:Rubi:infe:1992, Broo:Gelm:gene:1997}.
@@
 For each gene with a high probability of either HPH or LPH, the magnitude of the effect is of interest.
-Thus, we calculate estimated effect sizes using the estimated posterior means of the gene-specific parameters using the formula in equation \ref{e:effect_size} where a 0 indicates an estimated mean expression for the hybrid that is between those of the parents, i.e.
-not extreme heterosis.
+Thus, we calculate
 \begin{equation}
-\mbox{estimated effect size} = \left\{ 
+\mbox{estimated effect size}_g = \left\{ 
 \begin{array}{ll}
@@
 \hat{\mu}_{g3} - \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} > \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
-0 & \mbox{otherwise}.
+0 & \mbox{otherwise},
 \end{array} 
@@
 \end{equation}
+where $\hat{\mu}_{gi}$ is the estimated posterior mean of $\mu_{gi}$ for $i=1,2,3$.
 
@@
 The figure shows a line at zero for probabilities less than 0.5 because the effect size is set to 0 for the vast majority of genes.
-Above a probability of 0.5, we see a prototypical volcano pattern, with increased extreme heterosis probability corresponding to larger estimated effect sizes and no estimated effect sizes near zero for genes with high extreme heterosis probability.
+Above a probability of 0.5, we see a prototypical volcano pattern, with increased  heterosis probability corresponding to larger estimated effect sizes and no estimated effect sizes near zero for genes with high heterosis probability.
 We also see asymmetry, with larger negative effect sizes than positive effect sizes, because many genes have hybrid counts near zero and relatively high parental counts.
-Genes with high probabilities of extreme heterosis and large estimated effect sizes are candidates for further investigation.
+Genes with high estimated posterior probabilities of heterosis and large estimated effect sizes are candidates for further investigation.
 
@@
 Figure \ref{f:normality} provides histograms of components of $\theta$ from a single MCMC iteration.
+% From Dan, "Why use draws from a single Markov chain iteration?
+Ideally, we want to see if the true underlying components of theta follow the distributions we have specified.
+Wouldn't posterior means tend to be closer to the truth than single draws?"
 If the prior assumption of normality is correct, we would expect these histograms to show approximate normality.
@@
 Geneticists speculate that gene expression heterosis is one possible explanation of phenotypic heterosis of traits, such as plant height or grain yield.
-Because of the unwieldy composite null hypothesis in equation \eqref{e:hypotheses}, existing methods for identifying gene expression heterosis based on RNAseq data are not directly applicable for detecting heterosis genes.
+Because of the unwieldy composite null hypothesis in \eqref{e:hypotheses}, existing methods for identifying gene expression heterosis based on RNA-seq data are not directly applicable for detecting heterosis genes.
 \cite{ji2014estimation} introduced an empirical Bayes approach based on a hierarchical model for microarray data.
-We followed their approach, modified slightly to allow for any number of varieties and therefore even more complex hypotheses.
-We developed an empirical Bayes approach based on obtaining MAP estimates for hyperparameters followed by an MCMC to estimate gene-specific parameters.
+We followed their approach, modified to allow for RNA-seq read counts as measures of transcript abundance.
+We developed an empirical Bayes approach based on obtaining MAP estimates for hyperparameters followed by MCMC to estimate gene-specific parameters.
 This methodology can be conveniently implemented in the statistical software, {\tt Stan}.
@@
 % Estimates of heterosis to compare with Ji et. al.
-In addition to providing estimates of heterosis probabilities and effect sizes, this real data analysis exposed violations of our normality assumptions, indicating room to improve the model and increase the power to detect gene heterosis.
+In addition to providing estimates of heterosis probabilities and effect sizes, this real data analysis exposed violations of our normality assumptions, indicating room to improve the model and increase the power to detect gene expression heterosis.
 Some of our ongoing and future research will develop the details of such an extension.
@@
 
+% From Dan, "There are some capitalization issues in the references section.
+I think you need to protect "RNA" with curly braces so that it stays capitalized.
+Also, in Hallauer et al.
+(2010), the last initial should presumably be "D." rather than "d."."
+
 % \appendix

### 2015-03-02 ebayes_rnaseq_heterosis.tex
\psi -> e^{\psi}

@@
 \begin{equation} 
-Y_{gvi} \stackrel{ind}{\sim} NB(e^{\mu_{gv}+\gamma_{vi}},\psi_g), 
+Y_{gvi} \stackrel{ind}{\sim} NB\left(e^{\mu_{gv}+\gamma_{vi}},e^{\psi_g}\right), 
 \label{e:data}
 \end{equation}
-where $NB(\eta,\psi)$ indicates a negative binomial distribution with expectation $\eta$ and variance $\eta+\psi\eta^2$, and $ind$ indicates the observations are conditionally independent.
+where $NB(\eta,e^\psi)$ indicates a negative binomial distribution with expectation $\eta$ and variance $\eta+e^\psi\eta^2$, and $ind$ indicates the observations are conditionally independent.
 As shown in equation \eqref{e:data}, our data model involves gene-specific overdispersion $\psi_g$ and a mean that depends on the gene-variety combination through $\mu_{gv}$ and on the sample though  $\gamma_{vi}$.
@@
 (These terms are the normalization factors that practitioners typically estimate for each sample and then treat as constant in further analysis.) The logarithm of the gene-specific overdispersion parameters and the gene-specific variety effects are assumed to have a joint normal distribution, i.e.
-$\theta_g = [\mu_g, \log(\psi_g)]^\top \stackrel{ind}{\sim} N\left(\eta, \Sigma\right)$
+$\theta_g = [\mu_g, \psi_g]^\top \stackrel{ind}{\sim} N\left(\eta, \Sigma\right)$
 where $\mu_g = (\mu_{g1},\ldots,\mu_{gV})'$.
@@
 
-We assume improper uniform priors for $\eta$ and take 
-%In fact, the prior on \tau is just Unif(0,5)
-$\tau\stackrel{ind}{\sim} Ca^+(0,3)$, where 0 and 3 are the location and scale parameters, respectively, of the untruncated half-Cauchy distribution.
+We assume improper uniform priors for $\eta$ and take $\tau\stackrel{ind}{\sim} Ca^+(0,1)$, where 0 and 1 are the location and scale parameters, respectively, of the untruncated half-Cauchy distribution.
 In the independence model, we also assign the $Ca^+(0,3)$ prior to the
@@
 %&= \prod_{g=1}^G p\left(\theta_g\left|y_g,\hat{\pi}\right.\right) = \prod_{g=1}^G p\left(y_g\left|\theta_g,\hat{\gamma}\right.\right) p\left(\theta_g\left|\hat{\pi}\right.\right) \nonumber \\
-&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\gamma}_{vi}},\psi_g\right.\right) \right] N\left(\left.\left[\mu_g, \log(\psi_g) \right]^\top\right|\hat{\eta}, \hat{\Sigma} \right).
+&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right.\right) \right] N\left(\left.\left[\mu_g, \psi_g \right]^\top\right|\hat{\eta}, \hat{\Sigma} \right).
 \label{e:condind}
@@
 \input{include/map}
-The MAP estimates for the hierarchical means, $\eta$, and standard deviations, $\sigma$, are consistent between the independence and dependence models with, perhaps, a slightly larger mean and smaller standard deviation for the logarithm of the overdispersion parameter, $\log(\psi_g)$, in the dependence model relative to the independence model.
-In the dependence model, there appears to be high correlation amongst $\mu_{g1}$, $\mu_{g2}$, and $\mu_{g3}$, and negative correlation between $\log(\psi_g)$ and each of $\mu_{g1}$, $\mu_{g2}$, and $\mu_{g3}$.
+The MAP estimates for the hierarchical means, $\eta$, and standard deviations, $\sigma$, are consistent between the independence and dependence models with, perhaps, a slightly larger mean and smaller standard deviation for the logarithm of the overdispersion parameter, $\psi_g$, in the dependence model relative to the independence model.
+In the dependence model, there appears to be high correlation amongst $\mu_{g1}$, $\mu_{g2}$, and $\mu_{g3}$, and negative correlation between $\psi_g$ and each of $\mu_{g1}$, $\mu_{g2}$, and $\mu_{g3}$.
 The normalization factors, and their variance, are consistent between the two models, with the estimates for the dependence model being consistently 0.02 larger than the estimates for the independence model, and both indicating that there is about a 25\% increase in counts between the least and most thoroughly sequenced samples.
@@
 With stricter gene-inclusion criterion, it is possible both the secondary peak and the heavy-tailedness could be eliminated.
-The histogram for $\log(\psi)$ displays a heavy-tail to the right and the bivariate histograms show a pattern of near independence between $\log(\psi)$ and $\mu_g$ until $\mu_g$ passes a threshold at which point $\log(\psi)$ appears to decrease.
+The histogram for $\psi$ displays a heavy-tail to the right and the bivariate histograms show a pattern of near independence between $\psi$ and $\mu_g$ until $\mu_g$ passes a threshold at which point $\psi$ appears to decrease.
 Both of these phenomenon are consistent with observations in the literature that the negative-binomial overdispersion parameter decreases as the mean increases, but the bivariate histograms appear to indicate a nonlinear relationship between the means and the log of the overdispersion parameter.
@@
 
-Marginal and bivariate histograms for the model with an estimated covariance look similar although the estimated distribution $N(\hat{\eta},\hat{\Sigma})$ has increased variance for $\mu$ and decreased variance for $\log(\psi)$ as was previously seen in Table \ref{t:map} (results not shown).
+Marginal and bivariate histograms for the model with an estimated covariance look similar although the estimated distribution $N(\hat{\eta},\hat{\Sigma})$ has increased variance for $\mu$ and decreased variance for $\psi$ as was previously seen in Table \ref{t:map} (results not shown).
 

### 2015-03-03 ebayes_rnaseq_heterosis.tex
remove comment
model section updated to PAD

@@
 The results of read processing are summarized by a gene $\times$ sample matrix of counts.
-See \cite{nettleton2014design}
-% JBN: cite the book 
-for more details on RNA-seq experiments and data from a statistical perspective, and see \cite{paschold2012complementation} for the biological background behind the use of RNA-seq to study gene expression heterosis.
+See \cite{datta2014statistical} for more details on RNA-seq experiments and data from a statistical perspective, and see \cite{paschold2012complementation} for the biological background behind the use of RNA-seq to study gene expression heterosis.
 
@@
 \end{equation}
-where $NB(\eta,e^\psi)$ indicates a negative binomial distribution with expectation $\eta$ and variance $\eta+e^\psi\eta^2$, and $ind$ indicates the observations are conditionally independent.
+where $NB(\xi,e^\psi)$ indicates a negative binomial distribution with expectation $\xi$ and variance $\xi+e^\psi\xi^2$, and $ind$ indicates the observations are conditionally independent.
 As shown in equation \eqref{e:data}, our data model involves gene-specific overdispersion $\psi_g$ and a mean that depends on the gene-variety combination through $\mu_{gv}$ and on the sample though  $\gamma_{vi}$.
 The $\mu_{gv}$ terms are of primary scientific interest; the $\gamma_{vi}$ terms are normalization factors that account for differences in the thoroughness of sequencing from sample to sample.
-Practitioners typically compute normalization factors for each sample and then treat them as known constants in further analysis.
-In contrast, we include normalization factors in our model as unknowns that we estimate along with other parameters.
 
-We present a hierarchical model for the normalization factors, overdispersions, and gene-variety parameters.
-The sample-specific sequencing depth terms are assumed to follow a normal distribution, i.e.
-$\gamma_{vi} \stackrel{ind}{\sim} N(0,\tau^2)$.
-(These terms are the normalization factors that practitioners typically estimate for each sample and then treat as constant in further analysis.) The logarithm of the gene-specific overdispersion parameters and the gene-specific variety effects are assumed to have a joint normal distribution, i.e.
-$\theta_g = [\mu_g, \psi_g]^\top \stackrel{ind}{\sim} N\left(\eta, \Sigma\right)$
-where $\mu_g = (\mu_{g1},\ldots,\mu_{gV})'$.
-We consider an \emph{independence} model in which $\Sigma$ is diagonal with elements $\sigma_1^2,\ldots,\sigma_{V+1}^2$ and a \emph{dependence} model in which $\Sigma$ is a general positive definite matrix.
-% JBN: Switch mu and eta throughout
+Following \cite{ji2014estimation}, we reparameterize the gene-variety mean structure into the gene-specific parental average ($\phi_g$), half-parental difference ($\alpha_g$), and hybrid difference ($\delta_g$).
+Specifically, if $V=3$ with $v=1,2$ indicate the two parental varieties and $v=3$ indicates the hybrid, then
+\[ \phi_g = \frac{\mu_{g1}+\mu_{g2}}{2} \qquad \alpha_g = \frac{\mu_{g1}-\mu_{g2}}{2} \qquad \delta_g = \phi_g-\mu_{g3}. \]
 
-We assume improper uniform priors for $\eta$ and take $\tau\stackrel{ind}{\sim} Ca^+(0,1)$, where 0 and 1 are the location and scale parameters, respectively, of the untruncated half-Cauchy distribution.
-In the independence model, we also assign the $Ca^+(0,3)$ prior to the
-diagonal elements
-%square root of diagonal elements?
-of $\Sigma$.
-In the dependence model, we assume a conjugate, proper inverse Wishart distribution with degrees of freedom $V+2$ and an identity scale matrix, resulting in marginally uniform correlations.
+We present a hierarchical model for the normalization factors, overdispersions, and gene-specific mean parameters.
+These normalization factors, parental averages, and overdispersion parameters are assumed to follow a normal distribution.
+\[ 
+\gamma_{vi} \stackrel{ind}{\sim} N(0,\sigma_\gamma^2) \qquad 
+\phi_g \stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2) \qquad 
+\psi_g \stackrel{ind}{\sim} N(0,\sigma_\psi^2) \qquad 
+\]
+To allow for more shrinkage with the half-parental difference and hybrid difference, we use a Laplace (or double exponential prior) and thus implement a Bayesian LASSO \citep{park2008bayesian}. Thus
+\[ 
+\alpha_g \stackrel{ind}{\sim} La(\eta_\alpha, \sigma_\alpha) \qquad
+\delta_g \stackrel{ind}{\sim} La(\eta_\delta, \sigma_\delta) 
+\]
+where $\alpha\sim La(\eta,\sigma)$ has a probability density function given by $p(\alpha|\eta,\sigma) = exp(-|\alpha-\eta|/\sigma)/2\sigma$, expectation $\eta$ and variance $2\sigma^2$.
+We assume independence amongst the normalization factors, parental averages, half parental differences, hybrid differences, and overdispersion parameters.
+
+We assume improper uniform priors for $\eta_\phi$, $\eta_alpha$, and $\eta_\delta$ and independent half-cauchy priors for all standard deviations.
 
@@
 
-We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ and hyperparameters $\pi = (\gamma,\eta, \Sigma)$, where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$.
-We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\pi$, denoted $\hat{\pi}$, and then base gene-specific inference on the posterior conditional on $\hat{\pi}$.
+We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$ and hyperparameters $\pi = (\gamma, \eta, \sigma)$, where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, $\eta=(\eta_\phi,\eta_alpha,\eta_delta)$, and $\sigma=(\sigma_gamma,\sigma_\phi,\sigma_\alpha,\sigma_delta,\sigma_psi)$.
+We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\pi$, denoted $\hat{\pi}$, and then base gene-specific inference on the posterior conditional on this MAP estimate.
 Conditional on the hyperparameters, the gene-specific parameters are independent, i.e.,

### 2015-03-03 ebayes_rnaseq_heterosis.tex
fixed typos

@@
 Following \cite{ji2014estimation}, we reparameterize the gene-variety mean structure into the gene-specific parental average ($\phi_g$), half-parental difference ($\alpha_g$), and hybrid difference ($\delta_g$).
-Specifically, if $V=3$ with $v=1,2$ indicate the two parental varieties and $v=3$ indicates the hybrid, then
-\[ \phi_g = \frac{\mu_{g1}+\mu_{g2}}{2} \qquad \alpha_g = \frac{\mu_{g1}-\mu_{g2}}{2} \qquad \delta_g = \phi_g-\mu_{g3}. \]
+Specifically, for heterosis $V=3$ and let $v=1,2$ indicate the two parental varieties and $v=3$ indicate the hybrid, then
+\[ \phi_g = \frac{\mu_{g1}+\mu_{g2}}{2}, \quad 
+\alpha_g = \frac{\mu_{g1}-\mu_{g2}}{2}, \quad \mbox{and} \quad
+\delta_g = \phi_g-\mu_{g3}. \]
 
 We present a hierarchical model for the normalization factors, overdispersions, and gene-specific mean parameters.
-These normalization factors, parental averages, and overdispersion parameters are assumed to follow a normal distribution.
+The normalization factors, parental averages, and overdispersion parameters are assumed to follow a normal distribution, i.e.
 \[ 
-\gamma_{vi} \stackrel{ind}{\sim} N(0,\sigma_\gamma^2) \qquad 
-\phi_g \stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2) \qquad 
-\psi_g \stackrel{ind}{\sim} N(0,\sigma_\psi^2) \qquad 
+\gamma_{vi} \stackrel{ind}{\sim} N(0,\sigma_\gamma^2), \quad 
+\phi_g \stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2), \quad \mbox{and}\quad 
+\psi_g \stackrel{ind}{\sim} N(0,\sigma_\psi^2). 
 \]
-To allow for more shrinkage with the half-parental difference and hybrid difference, we use a Laplace (or double exponential prior) and thus implement a Bayesian LASSO \citep{park2008bayesian}. Thus
+To allow for more shrinkage with the half-parental and hybrid differences, we use a Laplace (or double exponential prior) and thus implement a Bayesian LASSO \citep{park2008bayesian}, i.e. 
 \[ 
-\alpha_g \stackrel{ind}{\sim} La(\eta_\alpha, \sigma_\alpha) \qquad
+\alpha_g \stackrel{ind}{\sim} La(\eta_\alpha, \sigma_\alpha) \qquad \mbox{and} \qquad
 \delta_g \stackrel{ind}{\sim} La(\eta_\delta, \sigma_\delta) 
 \]
-where $\alpha\sim La(\eta,\sigma)$ has a probability density function given by $p(\alpha|\eta,\sigma) = exp(-|\alpha-\eta|/\sigma)/2\sigma$, expectation $\eta$ and variance $2\sigma^2$.
+where $\alpha\sim La(\eta,\sigma)$ has a probability density function given by $p(\alpha|\eta,\sigma) = \exp(-|\alpha-\eta|/\sigma)/2\sigma$, expectation $\eta$, and variance $2\sigma^2$.
 We assume independence amongst the normalization factors, parental averages, half parental differences, hybrid differences, and overdispersion parameters.
 
-We assume improper uniform priors for $\eta_\phi$, $\eta_alpha$, and $\eta_\delta$ and independent half-cauchy priors for all standard deviations.
+We assume improper uniform priors for $\eta_\phi$, $\eta_\alpha$, and $\eta_\delta$ and independent half-cauchy priors for all standard deviations \citep{gelman2006prior}.
 
@@
 
-We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$ and hyperparameters $\pi = (\gamma, \eta, \sigma)$, where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, $\eta=(\eta_\phi,\eta_alpha,\eta_delta)$, and $\sigma=(\sigma_gamma,\sigma_\phi,\sigma_\alpha,\sigma_delta,\sigma_psi)$.
+We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$ and hyperparameters $\pi = (\gamma, \eta, \sigma)$, where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, $\eta=(\eta_\phi,\eta_\alpha,\eta_\delta)$, and $\sigma=(\sigma_\gamma,\sigma_\phi,\sigma_\alpha,\sigma_\delta,\sigma_psi)$.
 We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\pi$, denoted $\hat{\pi}$, and then base gene-specific inference on the posterior conditional on this MAP estimate.

### 2015-03-03 ebayes_rnaseq_heterosis.tex
eBayes updated

@@
 
-We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$ and hyperparameters $\pi = (\gamma, \eta, \sigma)$, where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, $\eta=(\eta_\phi,\eta_\alpha,\eta_\delta)$, and $\sigma=(\sigma_\gamma,\sigma_\phi,\sigma_\alpha,\sigma_\delta,\sigma_psi)$.
+We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$ and hyperparameters $\pi = (\gamma, \eta, \sigma)$, where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, $\eta=(\eta_\phi,\eta_\alpha,\eta_\delta)$, and $\sigma=(\sigma_\gamma,\sigma_\phi,\sigma_\alpha,\sigma_\delta,\sigma_\psi)$.
 We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\pi$, denoted $\hat{\pi}$, and then base gene-specific inference on the posterior conditional on this MAP estimate.
 Conditional on the hyperparameters, the gene-specific parameters are independent, i.e.,
-\begin{align}
+\begin{equation}
+\begin{split}
 p\left(\theta\left|y,\hat{\pi}\right.\right) 
 %&= \prod_{g=1}^G p\left(\theta_g\left|y_g,\hat{\pi}\right.\right) = \prod_{g=1}^G p\left(y_g\left|\theta_g,\hat{\gamma}\right.\right) p\left(\theta_g\left|\hat{\pi}\right.\right) \nonumber \\
-&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right.\right) \right] N\left(\left.\left[\mu_g, \psi_g \right]^\top\right|\hat{\eta}, \hat{\Sigma} \right).
+%&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right.\right) \right] p(\theta_g|\hat{\eta},\hat{\sigma}) \\
+\propto \prod_{g=1}^G & \left[ \,\prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi};e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right)  \right. \\
+&\times 
+\left. \vphantom{\prod_{v=1}^V}
+N(\phi_g;\hat{\eta}_\phi, \hat{\sigma}_\phi^2)
+La(\alpha_g;\hat{\eta}_\alpha, \hat{\sigma}_\alpha)
+La(\delta_g;\hat{\eta}_\delta, \hat{\sigma}_\delta)
+N(\psi_g;\hat{\eta}_\psi, \hat{\sigma}_\psi^2) 
+\right]
 \label{e:condind}
-\end{align}
+\end{split}
+\end{equation}
 Thus, conditional posterior inference for the gene-specific parameters can be obtained independently and in parallel.
@@
 We explicitly use the MAP estimate of $\pi$ when obtaining empirical Bayes posterior samples for the gene-specific parameters as in equation \eqref{e:condind}.
-The MAP estimate of $\theta$ can also be used in this step as initial values for the MCMC.
+The MAP estimate of $\theta$ could be used as initial values in the MCMC, although we did not find this necessary.
 
@@
 
-To identify gene expression heterosis, we consider $V=3$ varieties, where  $v=1,2$ signify the two parental varieties and $v=3$ represents the hybrid.
-For a specific gene $g$, low-parent  heteorisis (LPH) of gene expression occurs when $\mu_{g3}< \min(\mu_{g1},\mu_{g2})$
-%, mid-parent gene expression heterosis occurs when $\mu_{g1}\le \mu_{g3}\le \mu_{g2}$ if $\mu_{g1}\le \mu_{g2}$ (or $\mu_{g1}\ge \mu_{g3}\ge \mu_{g2}$ if $\mu_{g1}\ge \mu_{g2}$), 
-and high-parent  heterosis (HPH) of gene expression occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$.
-For simplicity, we define \emph{ heterosis} of gene expression to be the event of LPH or HPH.
+For a specific gene $g$, low-parent heteorisis (LPH) of gene expression occurs when average expression in the hybrid is less than the average expression of either parent, i.e.
+$\delta_g < -|\alpha_g|$, and high-parent heterosis (HPH) occurs when average expression in the hybrid is greater than the average expression of either parent, i.e.
+$\delta_g > |\alpha_g|$.
+% and high-parent  heterosis (HPH) of gene expression occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$.
+For simplicity, we define \emph{heterosis} of gene expression to be the event of LPH or HPH.
 The most important hypotheses are
-\begin{align}
-H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
-H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3},
-\label{e:hypotheses}
-\end{align}
-where $H_{g0}$ indicates no heterosis for gene $g$ and $H_{g1}$ indicates  heterosis for gene $g$. 
-We evaluate these hypotheses based on empirical Bayes estimates of their posterior probabilities, e.g., 
-\begin{align}
-P\left(H_{g1}|y, \hat{\pi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\pi}_{MAP}\right) \nonumber \\
-&\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right), \label{e:probs}
-\end{align}
-where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise.
-The set of probabilities $P\left(H_{g1}|y, \hat{\pi}_{MAP}\right)$ for $g=1,\ldots,G$ can be used to provide a list of genes sorted by the relative strength of evidence of  heterosis.
+% \begin{align}
+% H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
+% H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3},
+% \label{e:hypotheses}
+% \end{align}
+% where $H_{g0}$ indicates no heterosis for gene $g$ and $H_{g1}$ indicates  heterosis for gene $g$. 
+% We evaluate these hypotheses based on empirical Bayes estimates of their posterior probabilities, e.g., 
+% \begin{align}
+% P\left(H_{g1}|y, \hat{\pi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\pi}_{MAP}\right) \nonumber \\
+% &\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right), \label{e:probs}
+% \end{align}
+% where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise.
+We evaluate these probabilities based on empirical Bayes estimates of their posterior probabilities, e.g., 
+\begin{equation}
+%P(LPH_g|y) \approx 
+P(LPH_g|y,\hat{\pi}_{MAP}) = P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi}_{MAP}\right) = \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right), 
+\label{e:probs}
+\end{equation}
+where $\left(\delta_g^{(m)},\alpha_g^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise.
+We construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities.
 Geneticists can use this list to prioritize future experiments to understand the molecular genetic mechanisms for heterosis.

### 2015-03-03 ebayes_rnaseq_heterosis.tex
macros for tt

@@
 \newcommand{\blind}{0} % 0 = not blind, 1 = blind
+
+%\newcommand{\Stan}{{\tt Stan}}
+%\newcommand{\Stan}{Stan}
+\newcommand{\RStan}{{\tt RStan}}
+\newcommand{\edgeR}{{\tt edgeR}}
+\newcommand{\baySeq}{{\tt baySeq}}
+\newcommand{\ShrinkBayes}{{\tt ShrinkBayes}}
 
@@
 
-To perform the optimization required for the MAP estimate $\hat{\pi}$, we use the statistical software, {\tt Stan} \citep{stan-software:2014}, executed through the RStan interface \citep{rstan-software:2014} in R \citep{R2014}.
-The Stan software allows the construction of general Bayesian models with the primary restriction that all unknown parameters are continuous.
-With a fully constructed model, the software can obtain MAP estimates \cite[Section 50.3,][]{stan-manual:2014} using the {\tt optimizing} function in RStan.
-Presumably due to the large dimension of the optimization, it was necessary to use the L-BFGS algorithm \cite[Section 55,][]{stan-manual:2014} and reasonable starting values obtained via {\tt edgeR} \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor}.
-The {\tt Stan} software also includes the RStan function, {\tt sampling}, an implementation of a Markov chain Monte Carlo (MCMC) algorithm for obtaining samples from the posterior.
-By default, Stan uses a no-U-turn sampler \citep{hoffman2013no}, a variant of Hamiltonion Monte Carlo \citep{neal2011mcmc}.
-
 As we have not integrated out the gene-specific parameters when obtaining the MAP estimate, we actually receive a joint MAP estimate of $\theta$ and $\pi$.
@@
 
+\subsection{Implementation in {\tt Stan}}
+
+To perform the optimization required for the MAP estimate $\hat{\pi}$, we use the statistical software, {\tt Stan} \citep{stan-software:2014}, executed through the \RStan{} interface \citep{rstan-software:2014} in R \citep{R2014}.
+The Stan software allows the construction of general Bayesian models with the primary restriction that all unknown parameters are continuous.
+With a fully constructed model, the software can obtain MAP estimates \cite[Section 50.3,][]{stan-manual:2014} using the {\tt optimizing} function in \RStan{}.
+Presumably due to the large dimension of the optimization, it was necessary to use the L-BFGS algorithm \cite[Section 55,][]{stan-manual:2014} and reasonable starting values obtained via \edgeR{} \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor}.
+The {\tt Stan} software also includes the \RStan{} function, {\tt sampling}, an implementation of a Markov chain Monte Carlo (MCMC) algorithm for obtaining samples from the posterior.
+By default, Stan uses a no-U-turn sampler \citep{hoffman2013no}, a variant of Hamiltonion Monte Carlo \citep{neal2011mcmc}.
+
 \section{Simulation study based on a maize experiment}
@@
 Section \ref{s:maize} describes the maize dataset in detail.
-We compared our method to approaches using the R packages {\tt edgeR}, {\tt baySeq} \citep{hardcastle2010bayseq, hardcastle2012baySeq}, and {\tt ShrinkBayes} \citep{van2014shrinkbayes}.
+We compared our method to approaches using the R packages \edgeR{}, \baySeq{} \citep{hardcastle2010bayseq, hardcastle2012baySeq}, and \ShrinkBayes{} \citep{van2014shrinkbayes}.
 
@@
 
-We applied the {\tt edgeR} package \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor} to the maize data on 27,888 genes with average count at least one and at most two zero read counts for each variety across four replications.
+We applied the \edgeR{} package \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor} to the maize data on 27,888 genes with average count at least one and at most two zero read counts for each variety across four replications.
 This analysis produced  sample-specific normalization factors () % JBN: state the normalization factors, "I suppose you used the edgeR default, which is TMM, but I think this should be stated explicitly."
@@
 
-Existing RNA-seq data analysis methods are not specifically designed to identify gene expression heterosis, but we adapted the approaches in {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes} to provide a measure of the strength of heterosis for each gene such that large values of this measure indicate support for heterosis.
+Existing RNA-seq data analysis methods are not specifically designed to identify gene expression heterosis, but we adapted the approaches in \edgeR{}, {\tt baySeq}, and \ShrinkBayes{} to provide a measure of the strength of heterosis for each gene such that large values of this measure indicate support for heterosis.
 
-The {\tt edgeR} approach can be used to test for differential expression between any two varieties based on the fit of a negative binomial log-linear model \citep{robinson2007moderated, robinson2010edgeR}.
-To construct a measure of heterosis, we computed the maximum likelihood estimates of the $\mu_{gv}$ parameters for all genes using {\tt edgeR}'s built-in Fisher scoring algorithm, and then used likelihood ratio tests to calculate two $p$-values for each gene: $p_{g1}$  for testing $H_{g01}:\mu_{g1}=\mu_{g3}$ and $p_{g2}$ for testing $H_{g02}:\mu_{g2}=\mu_{g3}$.
+The \edgeR{} approach can be used to test for differential expression between any two varieties based on the fit of a negative binomial log-linear model \citep{robinson2007moderated, robinson2010edgeR}.
+To construct a measure of heterosis, we computed the maximum likelihood estimates of the $\mu_{gv}$ parameters for all genes using \edgeR{}'s built-in Fisher scoring algorithm, and then used likelihood ratio tests to calculate two $p$-values for each gene: $p_{g1}$  for testing $H_{g01}:\mu_{g1}=\mu_{g3}$ and $p_{g2}$ for testing $H_{g02}:\mu_{g2}=\mu_{g3}$.
 Then, for each gene, we define a new $p$-value denoted as $p_g$ and determined as $p_g=1$ if the estimate of $\mu_{g3}$ falls between the estimates of $\mu_{g1}$ and $\mu_{g2}$ and $p_g=\max\{p_{g1},p_{g2}\}/2$ otherwise.
@@
 The {\tt baySeq} approach allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct.
-In a technique similar to our application of {\tt edgeR}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene.
-We set this measure to zero if the maximum likelihood estimate, calculated using {\tt edgeR}, of $\mu_{g3}$ is between the maximum likelihood estimates of $\mu_{g1}$ and $\mu_{g2}$.
+In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene.
+We set this measure to zero if the maximum likelihood estimate, calculated using \edgeR{}, of $\mu_{g3}$ is between the maximum likelihood estimates of $\mu_{g1}$ and $\mu_{g2}$.
 Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the only hypotheses that allow for heterosis.
 
-The {\tt ShrinkBayes} package implements empirical Bayesian inference via integrated nested Laplace approximation (INLA) \citep{rue2009approximate,van2014shrinkbayes}.
+The \ShrinkBayes{} package implements empirical Bayesian inference via integrated nested Laplace approximation (INLA) \citep{rue2009approximate,van2014shrinkbayes}.
 A limitation of INLA is that it only provides approximations to marginal posteriors for parameters or linear combinations of parameters rather than an approximation to the full posterior.
@@
 Let $\phi_g = (\mu_{g1}+\mu_{g2})/2$ be the average of parental effects, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental effect difference, and $\delta_g = \mu_{g3} - \phi_g$ be the hybrid effect relative to the average parental effect.
-After fitting the model with {\tt ShrinkBayes}, we constructed a measure of heterosis for each gene.
+After fitting the model with \ShrinkBayes{}, we constructed a measure of heterosis for each gene.
 For gene $g$, we used the estimated posterior means $\hat{\delta}_g$ and $\hat{\alpha}_g$ of $\delta_g$ and $\alpha_g$, respectively, along with appropriate posterior probabilities.
@@
 \centerline{\includegraphics[width=\textwidth]{exampleROC0_1}}
-\caption{Representative ROC curves for false positive rates below 0.1 for the approaches using {\tt edgeR}, {\tt baySeq},  and {\tt ShrinkBayes} described in Section \ref{s:alternative}, along with the independence and dependence methods using {\tt RStan} described in Section \ref{s:gene_heterosis}.}
+\caption{Representative ROC curves for false positive rates below 0.1 for the approaches using \edgeR{}, {\tt baySeq},  and \ShrinkBayes{} described in Section \ref{s:alternative}, along with the independence and dependence methods using \RStan{} described in Section \ref{s:gene_heterosis}.}
 \label{f:roc}
@@
 For this simulation, we can see that the approaches based on the model in Section \ref{s:model} $-$ i.e.
-both {\tt RStan} approaches and the {\tt ShrinkBayes} approach $-$ provide the best true positive rate for a given false positive rate.
+both \RStan{} approaches and the \ShrinkBayes{} approach $-$ provide the best true positive rate for a given false positive rate.
 Also, as expected, as the sample size increases, our ability to distinguish genes with heterosis from genes without improves.
@@
 \centerline{\includegraphics[width=\textwidth]{auc-facet-TRUE}}
-\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different sample sizes for the approaches using {\tt edgeR}, {\tt baySeq},  and {\tt ShrinkBayes} described in Section \ref{s:alternative} , along with the independence and dependence methods using {\tt RStan} described in Section \ref{s:gene_heterosis}.
+\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different sample sizes for the approaches using \edgeR{}, {\tt baySeq},  and \ShrinkBayes{} described in Section \ref{s:alternative} , along with the independence and dependence methods using \RStan{} described in Section \ref{s:gene_heterosis}.
 Each line is a different simulation while the blue box indicates mean AUC and mean AUC plus and minus one standard error across the simulations.}
@@
 \end{figure}
-The pattern appears to hold, with the {\tt RStan} and {\tt ShrinkBayes} techniques outperforming the other two.
-While there does not appear to be a significant difference between {\tt ShrinkBayes} and the independence model using {\tt RStan}, performance for {\tt RStan} appears to decrease for the dependence model where an unstructured covariance is estimated, at least for smaller sample sizes.
+The pattern appears to hold, with the \RStan{} and \ShrinkBayes{} techniques outperforming the other two.
+While there does not appear to be a significant difference between \ShrinkBayes{} and the independence model using \RStan{}, performance for \RStan{} appears to decrease for the dependence model where an unstructured covariance is estimated, at least for smaller sample sizes.
 
@@
 
-Through a simulation study, we demonstrated that this method outperformed alternative methods (that are not designed to detect heterosis), and performed comparably well with an implementation of our model in {\tt ShrinkBayes}, which estimates the full posterior via INLA.
+Through a simulation study, we demonstrated that this method outperformed alternative methods (that are not designed to detect heterosis), and performed comparably well with an implementation of our model in \ShrinkBayes{}, which estimates the full posterior via INLA.
 We then demonstrated the use of the methodology on a maize experiment in which phenotypic heterosis is well known.
@@
 
-The authors thank Andrew Lithio for help in implementing our model in {\tt ShrinkBayes}.
+The authors thank Andrew Lithio for help in implementing our model in \ShrinkBayes{}.
 Research reported in this publication was supported by National Institute of General Medical Sciences of the National Institutes of Health under award number R01GM109458.

### 2015-03-03 ebayes_rnaseq_heterosis.tex
through simulation study

@@
 %\newcommand{\Stan}{{\tt Stan}}
-%\newcommand{\Stan}{Stan}
+\newcommand{\Stan}{Stan}
 \newcommand{\RStan}{{\tt RStan}}
@@
 \newcommand{\ShrinkBayes}{{\tt ShrinkBayes}}
+\newcommand{\RNAseq}{RNAseq}
 
-\title{Empirical Bayes analysis of RNA-seq data for detection of gene expression heterosis}
+\title{Empirical Bayes analysis of \RNAseq{} data for detection of gene expression heterosis}
 
@@
 Although the heterosis phenomenon is extensively utilized in agriculture, the molecular basis of heterosis is still largely unknown.
-In an effort to understand phenotypic heterosis at the molecular level, researchers have begun to measure transcript abundance levels of thousands of genes in parental inbred lines and their hybrid offspring using RNA sequencing (RNA-seq) technology.
+In an effort to understand phenotypic heterosis at the molecular level, researchers have begun to measure transcript abundance levels of thousands of genes in parental inbred lines and their hybrid offspring using RNA sequencing (\RNAseq{}) technology.
 The resulting data allow researchers to search for evidence of gene expression heterosis as one potential molecular mechanism underlying heterosis of agriculturally important traits.
-The null hypotheses of greatest interest in testing for gene expression heterosis are composite null hypotheses that are difficult to test with standard statistical approaches for RNA-seq analysis.
+The null hypotheses of greatest interest in testing for gene expression heterosis are composite null hypotheses that are difficult to test with standard statistical approaches for \RNAseq{} analysis.
 To address these shortcomings, we develop a hierarchical model based on a negativeb binomial data model and draw inferences using a computationally tractable empirical Bayes approach to inference.
@@
 \begin{keywords}
-Empirical Bayes; Gene expression; Heterosis; Hierarchical model; Negative binomial; RNA-seq.
+Empirical Bayes; Gene expression; Heterosis; Hierarchical model; Negative binomial; \RNAseq{}.
 \end{keywords}
@@
 This development was necessary due to the composite null hypotheses in tests for heterosis.
-These hypotheses, which many available methods do not fully accommodate, remain a challenge in the transition from continuous measurements of transcript abundance to count-based measurements that arise from RNA sequencing (RNA-seq) technology.
+These hypotheses, which many available methods do not fully accommodate, remain a challenge in the transition from continuous measurements of transcript abundance to count-based measurements that arise from RNA sequencing (\RNAseq{}) technology.
 Building on the work of \citeauthor{ji2014estimation} with the normal data model, we construct a hierarchical model based on a negative-binomial data model.
@@
 
-\section{Empirical Bayes identification of gene expression heterosis from RNA-seq read counts}
+\section{Empirical Bayes identification of gene expression heterosis from \RNAseq{} read counts}
 \label{s:method}
 
-An RNA sequencing (RNA-seq) experiment to identify genes demonstrating expression heterosis consists of three genetic varieties: two parental varieties (P1 and P2) and a cross between these two varieties called the hybrid (H).
+An RNA sequencing (\RNAseq{}) experiment to identify genes demonstrating expression heterosis consists of three genetic varieties: two parental varieties (P1 and P2) and a cross between these two varieties called the hybrid (H).
 For each variety, replicate RNA samples are isolated and assessed for quality.
@@
 The results of read processing are summarized by a gene $\times$ sample matrix of counts.
-See \cite{datta2014statistical} for more details on RNA-seq experiments and data from a statistical perspective, and see \cite{paschold2012complementation} for the biological background behind the use of RNA-seq to study gene expression heterosis.
+See \cite{datta2014statistical} for more details on \RNAseq{} experiments and data from a statistical perspective, and see \cite{paschold2012complementation} for the biological background behind the use of \RNAseq{} to study gene expression heterosis.
 
-To use RNA-seq counts to identify genes displaying expression heterosis, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the parameters using an empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate empirical Bayes posterior probabilities for heterosis.
+To use \RNAseq{} counts to identify genes displaying expression heterosis, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the parameters using an empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate empirical Bayes posterior probabilities for heterosis.
 
-\subsection{Hierarchical model for RNA-seq counts}
+\subsection{Hierarchical model for \RNAseq{} counts}
 \label{s:model}
@@
 where $\left(\delta_g^{(m)},\alpha_g^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise.
+HPH probability is defined similarly with the inequality reversed and without the negative sign.
 We construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities.
@@
 
-\subsection{Implementation in {\tt Stan}}
+\subsection{Implementation in \Stan{}}
 
@@
 By default, Stan uses a no-U-turn sampler \citep{hoffman2013no}, a variant of Hamiltonion Monte Carlo \citep{neal2011mcmc}.
+
+\subsection{Implementation in \ShrinkBayes{}}
+
+\ShrinkBayes{} utilizes integrated nested Laplace approximation (INLA) \citep{rue2009approximate} in combination with empirical Bayes ideas \citep{van2014shrinkbayes}.
+One limitation with inferential methods based on INLA is that all distributions, except for the data distribution, must have tails as light or lighter than the normal density.
+Thus, we cannot implement the Laplace priors for the half-parental difference ($\alpha_g$) and the hybrid difference ($\delta_g$) and instead use normal priors in this situation.
+An additional limitation is that INLA provides approximations to marginal posteriors for parameters or linear combinations of parameters, but not an approximation to the full posterior.
+Since we are interested in non-linear quantities such as $P(\delta_g > |\alpha_g||y)$, we cannot compute these directly using \ShrinkBayes{}.
+Instead, for \ShrinkBayes{}, we calculate heterosis probabilities conditional on posterior means for $\hat{\delta}_g$ and $\hat{\alpha}_g$, e.g.
+\[ 
+% P_{\ShrinkBayes{}}(LPH_g|y) \equiv \left\{ \begin{array}{ll}
+% P(\delta_g>-\alpha_g|y) & \mbox{if } \hat{\delta}_g > -\hat{\alpha}_g \ge 0 \\
+% P(\delta_g>\phantom{-}\alpha_g|y) & \mbox{if } \hat{\delta}_g > \phantom{-}\hat{\alpha}_g \ge 0 \\
+% 0 & \mbox{otherwise}
+% \end{array} \right.
+P_{\ShrinkBayes{}}(LPH_g|y) \equiv P(\delta_g>-\alpha_g|y)\mathrm{I}\left(\hat{\delta}_g > -\hat{\alpha}_g \ge 0\right) + P(\delta_g>\alpha_g|y) \mathrm{I}\left(\hat{\delta}_g > \hat{\alpha}_g \ge 0\right).
+%P_{\ShrinkBayes{}}(HPH|y) \equiv \left\{ \begin{array}{ll}
+%P(\delta_g<-\alpha_g|y) & \mbox{if } \hat{\delta}_g > -\hat{\alpha}_g \ge 0 \\
+%(\delta_g<\phantom{-}\alpha_g|y) & \mbox{if } \hat{\delta}_g > \phantom{-}\hat{\alpha}_g \ge 0 \\
+%0 & \mbox{otherwise}
+%\end{array} \right.
+\]
+HPH probability is defined similarly with all inequalities reversed. As before, we construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities.
+%the heterosis probability for gene $g$ is zero if $|\hat{\delta}_g| \le |\hat{\alpha}_g|$, $P(\delta_g>\alpha_g|y)$ if $\hat{\delta}_g > -\hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g > 0 |y)$ if $\hat{\delta}_g > \hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g < 0 |y)$ if $\hat{\delta}_g < \hat{\alpha}_g \le 0$, and $P(\delta_g + \alpha_g < 0 |y)$ if $\hat{\delta}_g < -\hat{\alpha}_g \le 0$.
 
@@
 
-Existing RNA-seq data analysis methods are not specifically designed to identify gene expression heterosis, but we adapted the approaches in \edgeR{}, {\tt baySeq}, and \ShrinkBayes{} to provide a measure of the strength of heterosis for each gene such that large values of this measure indicate support for heterosis.
+Existing \RNAseq{} data analysis methods are not specifically designed to identify gene expression heterosis, but we adapted the approaches in \edgeR{} and {\tt baySeq} to provide a measure of the strength of heterosis for each gene such that large values of this measure indicate support for heterosis.
 
-The \edgeR{} approach can be used to test for differential expression between any two varieties based on the fit of a negative binomial log-linear model \citep{robinson2007moderated, robinson2010edgeR}.
+\edgeR{} can be used to test for differential expression between any two varieties based on the fit of a negative binomial log-linear model \citep{robinson2007moderated, robinson2010edgeR}.
 To construct a measure of heterosis, we computed the maximum likelihood estimates of the $\mu_{gv}$ parameters for all genes using \edgeR{}'s built-in Fisher scoring algorithm, and then used likelihood ratio tests to calculate two $p$-values for each gene: $p_{g1}$  for testing $H_{g01}:\mu_{g1}=\mu_{g3}$ and $p_{g2}$ for testing $H_{g02}:\mu_{g2}=\mu_{g3}$.
@@
 
-The {\tt baySeq} approach allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct.
+{\tt baySeq} allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct.
 In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene.
@@
 
-The \ShrinkBayes{} package implements empirical Bayesian inference via integrated nested Laplace approximation (INLA) \citep{rue2009approximate,van2014shrinkbayes}.
-A limitation of INLA is that it only provides approximations to marginal posteriors for parameters or linear combinations of parameters rather than an approximation to the full posterior.
-Thus, we cannot directly estimate posterior probabilities involving nonlinear functions of $\mu_{g1}$, $\mu_{g2}$, and $\mu_{g3}$, such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2})|y)$.
-To circumvent this limitation, we reparameterized the independence model in Section \ref{s:model} using the approach in \cite{ji2014estimation}.
-Let $\phi_g = (\mu_{g1}+\mu_{g2})/2$ be the average of parental effects, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental effect difference, and $\delta_g = \mu_{g3} - \phi_g$ be the hybrid effect relative to the average parental effect.
-After fitting the model with \ShrinkBayes{}, we constructed a measure of heterosis for each gene.
-For gene $g$, we used the estimated posterior means $\hat{\delta}_g$ and $\hat{\alpha}_g$ of $\delta_g$ and $\alpha_g$, respectively, along with appropriate posterior probabilities.
-Specifically, the measure of heterosis for gene $g$ is zero if $|\hat{\delta}_g| \le |\hat{\alpha}_g|$, $P(\delta_g + \alpha_g > 0 |y)$ if $\hat{\delta}_g > -\hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g > 0 |y)$ if $\hat{\delta}_g > \hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g < 0 |y)$ if $\hat{\delta}_g < \hat{\alpha}_g \le 0$, and $P(\delta_g + \alpha_g < 0 |y)$ if $\hat{\delta}_g < -\hat{\alpha}_g \le 0$.
-
 \subsection{Results}
@@
 For each of the methods above, as well as the method described in Section \ref{s:gene_heterosis}, we sorted genes according to the computed measure of the strength of evidence for heterosis.
-From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to distinguish genes with heterosis, as defined in Section \ref{s:sim_data}  \citep{landau2013dispersion},
+From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to distinguish genes with heterosis, as defined in Section \ref{s:sim_data},
 % From Dan, "why this citation?"
-from those without heterosis. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
+from those without heterosis \citep{landau2013dispersion}. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
 \begin{figure}[htbp]
@@
 \end{figure}
-For this simulation, we can see that the approaches based on the model in Section \ref{s:model} $-$ i.e.
-both \RStan{} approaches and the \ShrinkBayes{} approach $-$ provide the best true positive rate for a given false positive rate.
+For this simulation, we can see that the approaches based on the model in Section \ref{s:model}, i.e.
+\RStan{} and \ShrinkBayes{}, provide the best true positive rate for a given false positive rate.
 Also, as expected, as the sample size increases, our ability to distinguish genes with heterosis from genes without improves.
@@
 The pattern appears to hold, with the \RStan{} and \ShrinkBayes{} techniques outperforming the other two.
-While there does not appear to be a significant difference between \ShrinkBayes{} and the independence model using \RStan{}, performance for \RStan{} appears to decrease for the dependence model where an unstructured covariance is estimated, at least for smaller sample sizes.
+%While there does not appear to be a significant difference between \ShrinkBayes{} and the independence model using \RStan{}, performance for \RStan{} appears to decrease for the dependence model where an unstructured covariance is estimated, at least for smaller sample sizes.
 
@@
 Geneticists speculate that gene expression heterosis is one possible explanation of phenotypic heterosis of traits, such as plant height or grain yield.
-Because of the unwieldy composite null hypothesis in \eqref{e:hypotheses}, existing methods for identifying gene expression heterosis based on RNA-seq data are not directly applicable for detecting heterosis genes.
+Because of the unwieldy composite null hypothesis in \eqref{e:hypotheses}, existing methods for identifying gene expression heterosis based on \RNAseq{} data are not directly applicable for detecting heterosis genes.
 \cite{ji2014estimation} introduced an empirical Bayes approach based on a hierarchical model for microarray data.
-We followed their approach, modified to allow for RNA-seq read counts as measures of transcript abundance.
+We followed their approach, modified to allow for \RNAseq{} read counts as measures of transcript abundance.
 We developed an empirical Bayes approach based on obtaining MAP estimates for hyperparameters followed by MCMC to estimate gene-specific parameters.

### 2015-03-03 ebayes_rnaseq_heterosis.tex
simulation details and new plots

@@
 We applied the \edgeR{} package \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor} to the maize data on 27,888 genes with average count at least one and at most two zero read counts for each variety across four replications.
-This analysis produced  sample-specific normalization factors () % JBN: state the normalization factors, "I suppose you used the edgeR default, which is TMM, but I think this should be stated explicitly."
-, gene-specific dispersion estimates, and gene-variety-specific location parameter estimates.
+This analysis produced sample-specific normalization factors using the TMM method \citep{robinson2010scaling},
+%(0.080, -0.041, -0.054, 0.000, -0.036, -0.147, 0.093, 0.059, -0.004, -0.057, 0.067,  0.041) 
+gene-specific dispersion estimates, and gene-variety-specific location parameter estimates \citep{mccarthy2012differential}.
 We treated these estimates as the true parameter values for the simulation study so that our simulated datasets mimicked the existing covariance structure among the gene-variety means of the original maize data.
@@
 For each gene with a high probability of either HPH or LPH, the magnitude of the effect is of interest.
-Thus, we calculate
+We calculate
 \begin{equation}
-\mbox{estimated effect size}_g = \left\{ 
+% \mbox{estimated effect size}_g = \left\{ 
+% \begin{array}{ll}
+% \hat{\mu}_{g3} - \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} < \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
+% \hat{\mu}_{g3} - \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} > \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
+% 0 & \mbox{otherwise},
+% \end{array} 
+% \right. 
+\mbox{estimated effect size}_g \equiv \left\{ 
 \begin{array}{ll}
-\hat{\mu}_{g3} - \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} < \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
-\hat{\mu}_{g3} - \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} > \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
+\hat{\delta}_g - |\hat{\alpha}_g| & \mbox{if } \hat{\delta}_g > \phantom{-}|\alpha_g| \\
+\hat{\delta}_g + |\hat{\alpha}_g| & \mbox{if } \hat{\delta}_g < -|\alpha_g| \\
 0 & \mbox{otherwise},
@@
 \end{equation}
-where $\hat{\mu}_{gi}$ is the estimated posterior mean of $\mu_{gi}$ for $i=1,2,3$.
+where $\hat{\mu}_{gi}$ is the estimated posterior mean of $\mu_{gv}$ for $v=1,2,3$.
+Thus, the estimated effect size is the difference between hybrid mean and the nearest parent with negative values indicating LPH and positive values indicating HPH.
+If the hybrid mean is estimated to be between the parents, this effect is defined to be zero.
 
-Figure \ref{f:volcano} provides a volcano plot to visualize the maximum of the probabilities of LPH and HPH versus estimated effect size.
-The figure shows a line at zero for probabilities less than 0.5 because the effect size is set to 0 for the vast majority of genes.
+Figure \ref{f:volcano} provides a volcano plot, in this case a bivariate histogram, to visualize the maximum of the probabilities of LPH and HPH versus estimated effect size.
+The figure shows a peak at zero for probabilities below 0.5 with estimated effect sizes of zero.
 Above a probability of 0.5, we see a prototypical volcano pattern, with increased heterosis probability corresponding to larger estimated effect sizes and no estimated effect sizes near zero for genes with high heterosis probability.
-We also see asymmetry, with larger negative effect sizes than positive effect sizes, because many genes have hybrid counts near zero and relatively high parental counts.
+We also see asymmetry, with larger negative effect sizes than positive effect sizes due to genes with hybrid counts near zero and relatively high parental counts.
 Genes with high estimated posterior probabilities of heterosis and large estimated effect sizes are candidates for further investigation.
 
-% Compare independence and volcano plots
-
 \begin{figure}
-\centerline{\includegraphics[width=\textwidth]{volcano}}
-\caption{A two-dimensional histogram of the maximum of the LPH and HPH probabilities versus estimated effect size defined in equation \ref{e:effect_size} for the B73 $\times$ Mo17 maize experiment using the independence model.}
+\centerline{\includegraphics[width=\textwidth]{maize_volcano_dexp}}
+\caption{A bivariaet histogram of the maximum of the LPH and HPH probabilities versus estimated effect size defined in equation \ref{e:effect_size} for the B73 $\times$ Mo17 maize experiment.}
 \label{f:volcano}
@@
 
-With almost 30,000 genes in our data set, we can evaluate the normality assumption for the gene-specific parameters.
-Figure \ref{f:normality} provides histograms of components of $\theta$ from a single MCMC iteration.
-% From Dan, "Why use draws from a single Markov chain iteration?
-Ideally, we want to see if the true underlying components of theta follow the distributions we have specified.
-Wouldn't posterior means tend to be closer to the truth than single draws?"
-If the prior assumption of normality is correct, we would expect these histograms to show approximate normality.
-The marginal histograms show a bimodal distribution for $\mu$ with a secondary peak around relatively small values.
-The bivariate histograms amongst the components of $\mu$ show a high correlation but with a heavy-tail toward small values of both components of $\mu$.
-With stricter gene-inclusion criterion, it is possible both the secondary peak and the heavy-tailedness could be eliminated.
-The histogram for $\psi$ displays a heavy-tail to the right and the bivariate histograms show a pattern of near independence between $\psi$ and $\mu_g$ until $\mu_g$ passes a threshold at which point $\psi$ appears to decrease.
-Both of these phenomenon are consistent with observations in the literature that the negative-binomial overdispersion parameter decreases as the mean increases, but the bivariate histograms appear to indicate a nonlinear relationship between the means and the log of the overdispersion parameter.
+With almost 30,000 genes in our data set, we can evaluate the distribution assumptions for the gene-specific parameters.
+Figure \ref{f:scatterplot} provides marginal and bivariate histograms for the gene-specific parameters.
 
 \begin{figure}
-\centerline{\includegraphics[width=\textwidth]{normality}}
-\caption{Marginal and bivariate histograms for gene-specific parameters from a single MCMC sample, $\theta_g^{(m)}$,  for the B73 $\times$ Mo17 maize experiment.
-For reference, the marginal histograms have a dashed line indicating the appropriate marginal probability density functions from $N(\hat{\eta},\hat{\Sigma})$.}
-\label{f:normality}
+\centerline{\includegraphics[width=\textwidth]{scatterplot_dexp}}
+\caption{Marginal and bivariate histograms for gene-specific parameter posterior meansfor the B73 $\times$ Mo17 maize experiment.
+For reference, the marginal histograms have a dashed line indicating the appropriate marginal probability density functions in the hierarchical model with the MAP estimates.}
+\label{f:scatterplot}
 \end{figure}
@@
 Geneticists speculate that gene expression heterosis is one possible explanation of phenotypic heterosis of traits, such as plant height or grain yield.
-Because of the unwieldy composite null hypothesis in \eqref{e:hypotheses}, existing methods for identifying gene expression heterosis based on \RNAseq{} data are not directly applicable for detecting heterosis genes.
+Existing methods for identifying differential gene expression based on \RNAseq{} data are not directly applicable for detecting heterosis genes.
 \cite{ji2014estimation} introduced an empirical Bayes approach based on a hierarchical model for microarray data.

### 2015-03-06 monte_carlo_integration.R
monte carlo integration

@@ -1,0 +1,36 @@
+library(plyr)
+
+G = 10
+v = 3
+nv = rep(4,3)
+
+rlaplace = function(n, mean, scale) {
+  u = runif(n, -.5, .5)
+  mean - scale * sign(u) * log(1-2*abs(u))
+}
+
+# Priors
+prior = list(eta_phi     = 1,
+             eta_alpha   = 0,
+             eta_delta   = 0,
+             eta_psi     = -2,
+             sigma_phi   = 1,
+             sigma_alpha =0.1,
+             sigma_delta = 0.01,
+             sigma_psi   = 0.1)
+
+# Parameters
+parameters = with(prior, list(phi   = rnorm   (G, eta_phi,   sigma_phi  ),
+                  alpha = rlaplace(G, eta_alpha, sigma_alpha),
+                  delta = rlaplace(G, eta_delta, sigma_delta),
+                  psi   = rnorm   (G, eta_psi,   sigma_psi  )))
+
+# Data
+d = ddply(data.frame(gene=1:G), .(gene), function(x) {
+  mutate(data.frame(parent = rep(1:3, times = nv)),
+         sample = 1:sum(nv),
+         mu = with(parameters, c(phi[g]+alpha[g], phi[g]-alpha[g], phi[g]+delta[g])),
+         y = # need negative binomial with mu parameterization
+  )
+  
+})

### 2015-03-07 monte_carlo_integration.R
moved to sandbox

@@ -1,36 +1,0 @@
-library(plyr)
-
-G = 10
-v = 3
-nv = rep(4,3)
-
-rlaplace = function(n, mean, scale) {
-  u = runif(n, -.5, .5)
-  mean - scale * sign(u) * log(1-2*abs(u))
-}
-
-# Priors
-prior = list(eta_phi     = 1,
-             eta_alpha   = 0,
-             eta_delta   = 0,
-             eta_psi     = -2,
-             sigma_phi   = 1,
-             sigma_alpha =0.1,
-             sigma_delta = 0.01,
-             sigma_psi   = 0.1)
-
-# Parameters
-parameters = with(prior, list(phi   = rnorm   (G, eta_phi,   sigma_phi  ),
-                  alpha = rlaplace(G, eta_alpha, sigma_alpha),
-                  delta = rlaplace(G, eta_delta, sigma_delta),
-                  psi   = rnorm   (G, eta_psi,   sigma_psi  )))
-
-# Data
-d = ddply(data.frame(gene=1:G), .(gene), function(x) {
-  mutate(data.frame(parent = rep(1:3, times = nv)),
-         sample = 1:sum(nv),
-         mu = with(parameters, c(phi[g]+alpha[g], phi[g]-alpha[g], phi[g]+delta[g])),
-         y = # need negative binomial with mu parameterization
-  )
-  
-})

### 2015-03-07 sandbox/normal_approximation.tex
moved to sandbox

--- 2015-03-03 ebayes_rnaseq_heterosis.tex
@@
-\documentclass[useAMS,usenatbib,referee]{biom}
-
-\usepackage{amsmath}
-\usepackage[figuresright]{rotating}
-
-\graphicspath{{include/}}
-
-\def\bSig\mathbf{\Sigma}
-\newcommand{\VS}{V\&S}
-\newcommand{\tr}{\mbox{tr}}
-
-\newcommand{\blind}{0} % 0 = not blind, 1 = blind
-
-%\newcommand{\Stan}{{\tt Stan}}
-\newcommand{\Stan}{Stan}
-\newcommand{\RStan}{{\tt RStan}}
-\newcommand{\edgeR}{{\tt edgeR}}
-\newcommand{\baySeq}{{\tt baySeq}}
-\newcommand{\ShrinkBayes}{{\tt ShrinkBayes}}
-\newcommand{\RNAseq}{RNAseq}
-
-\title{Empirical Bayes analysis of \RNAseq{} data for detection of gene expression heterosis}
-
-\if0\blind{
-\author{Jarad Niemi$^*$\email{niemi@iastate.edu}, 
-Eric Mittman, 
-Will Landau, and 
-Dan Nettleton \\
-Department of Statistics, Iowa State University, Ames, Iowa, U.S.A.}
-} \fi
+\documentclass{article}
 
@@
 
-\date{{\it Received December} 2014}
-
-\pagerange{\pageref{firstpage}--\pageref{lastpage}} 
-\volume{VV}
-\pubyear{YYYY}
-\artmonth{MM}
-\doi{x}
-
-\label{firstpage}
-
-\begin{abstract}
-Heterosis, or hybrid vigor, refers to the enhancements in the phenotype of hybrid progeny relative to their inbred parents.
-Although the heterosis phenomenon is extensively utilized in agriculture, the molecular basis of heterosis is still largely unknown.
-In an effort to understand phenotypic heterosis at the molecular level, researchers have begun to measure transcript abundance levels of thousands of genes in parental inbred lines and their hybrid offspring using RNA sequencing (\RNAseq{}) technology.
-The resulting data allow researchers to search for evidence of gene expression heterosis as one potential molecular mechanism underlying heterosis of agriculturally important traits.
-The null hypotheses of greatest interest in testing for gene expression heterosis are composite null hypotheses that are difficult to test with standard statistical approaches for \RNAseq{} analysis.
-To address these shortcomings, we develop a hierarchical model based on a negativeb binomial data model and draw inferences using a computationally tractable empirical Bayes approach to inference.
-We demonstrate improvements over alternative methods via a simulation study based on a maize experiment and then analyze that maize experiment with our newly proposed methodology.
-This article has supplementary material online.
-\end{abstract}
-
-\begin{keywords}
-Empirical Bayes; Gene expression; Heterosis; Hierarchical model; Negative binomial; \RNAseq{}.
-\end{keywords}
-
-\maketitle
-
-%  A maximum of six (6) tables or figures combined is often required.''
-
-\section{Introduction}
-\label{s:intro}
-
-Heterosis, or hybrid vigor, occurs when hybrid progeny display phenotypes that are superior to the phenotypes of their parents.
-The heterosis phenomenon was scientifically documented in plants by \cite{darwin1876effects}and has long been used to improve agricultural production.
-One classic example involves hybrid maize offspring that are taller, faster to mature, and yield considerably more grain than their inbred parents \citep{hallauer1981quantitative, hallauer2010quantitative}.
-Despite intensive study and successful use of heterosis in agriculture, the basic molecular genetic mechanisms remain largely unknown \citep{coors1999genetics, lippman2007heterosis}.
-One potential explanation, known as gene expression heterosis, is enhanced expression of one or more genes in hybrids compared to their inbred parents \citep{swanson2006all, springer2007allelic}.
-
-Recently, \cite{ji2014estimation} introduced an approach to assess gene expression heterosis using microarray data under the assumption that these data are continuous.
-They built a normal hierarchical model for microarray measurements of transcript abundance that allows borrowing of information across genes to estimate means and variances.
-They introduced an empirical Bayes framework that first estimates model hyperparameters, then estimates the posterior distribution for gene-specific parameters conditional on those hyperparameters, and finally computes heterosis probabilities based on integrals of regions under this posterior.
-This development was necessary due to the composite null hypotheses in tests for heterosis.
-These hypotheses, which many available methods do not fully accommodate, remain a challenge in the transition from continuous measurements of transcript abundance to count-based measurements that arise from RNA sequencing (\RNAseq{}) technology.
-Building on the work of \citeauthor{ji2014estimation} with the normal data model, we construct a hierarchical model based on a negative-binomial data model.
-We also utilize an empirical Bayes approach to obtain \emph{maximum a posterior} (MAP) estimates of the hyperparameters and the posterior distributions for the gene-specific parameters conditional on those hyperparameters.
-
-The remainder of the paper proceeds as follows.
-Section \ref{s:method} presents the proposed hierarchical model, an empirical Bayes method of estimating the parameters, and the calculation of posterior probabilities of gene expression heterosis.
-Section \ref{s:simulation} presents a simulation study based on a maize experiment and compares our approach to alternative methods.
-Section \ref{s:maize} analyzes a maize experiment where phenotypic heterosis is well established and identifies genes demonstrating expression heterosis.
-The data and code necessary to perform this analysis are available in the Supplemental Material.
-Section \ref{s:discussion} summarizes the work and suggests directions for future research.
-
-\section{Empirical Bayes identification of gene expression heterosis from \RNAseq{} read counts}
-\label{s:method}
-
-An RNA sequencing (\RNAseq{}) experiment to identify genes demonstrating expression heterosis consists of three genetic varieties: two parental varieties (P1 and P2) and a cross between these two varieties called the hybrid (H).
-For each variety, replicate RNA samples are isolated and assessed for quality.
-Complementary DNA (cDNA) libraries, consisting of short cDNA fragments derived from RNA, are constructed.
-Then, next generation sequencing technology is used to determine the \emph{reads}, or nucleotide sequences, in the cDNA libraries.
-These reads are processed using bioinformatic algorithms to match the reads to genes, or specific gene transcripts, exons, microRNAs, etc.
-The results of read processing are summarized by a gene $\times$ sample matrix of counts.
-See \cite{datta2014statistical} for more details on \RNAseq{} experiments and data from a statistical perspective, and see \cite{paschold2012complementation} for the biological background behind the use of \RNAseq{} to study gene expression heterosis.
-
-To use \RNAseq{} counts to identify genes displaying expression heterosis, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the parameters using an empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate empirical Bayes posterior probabilities for heterosis.
-
-\subsection{Hierarchical model for \RNAseq{} counts}
-\label{s:model}
-
-Let $Y_{gvi}$ be the count for gene $g=1,\ldots,G$, variety $v=1,\ldots,V$, and replicate $i=1,\ldots,n_v$.  We assume
+Model 
 \begin{equation} 
@@
 \label{e:data}
+The probability mass function is 
+\[ p(y_{gvi}|\mu_{gv},\gamma_{vi},\psi_g) = \frac{}{} \]
+
 \end{equation}
-where $NB(\xi,e^\psi)$ indicates a negative binomial distribution with expectation $\xi$ and variance $\xi+e^\psi\xi^2$, and $ind$ indicates the observations are conditionally independent.
-As shown in equation \eqref{e:data}, our data model involves gene-specific overdispersion $\psi_g$ and a mean that depends on the gene-variety combination through $\mu_{gv}$ and on the sample though  $\gamma_{vi}$.
-The $\mu_{gv}$ terms are of primary scientific interest; the $\gamma_{vi}$ terms are normalization factors that account for differences in the thoroughness of sequencing from sample to sample.
-
-Following \cite{ji2014estimation}, we reparameterize the gene-variety mean structure into the gene-specific parental average ($\phi_g$), half-parental difference ($\alpha_g$), and hybrid difference ($\delta_g$).
-Specifically, for heterosis $V=3$ and let $v=1,2$ indicate the two parental varieties and $v=3$ indicate the hybrid, then
+with $E[Y_{gvi}] = \eta_{gvi} =e^{\mu_{gv}+\gamma_{vi}}$ and $V[Y_{gvi}] = \eta_{gvi} + e^{\psi_g}\eta_{gvi}^2$.
 \[ \phi_g = \frac{\mu_{g1}+\mu_{g2}}{2}, \quad 
@@
 \delta_g = \phi_g-\mu_{g3}. \]
-
-We present a hierarchical model for the normalization factors, overdispersions, and gene-specific mean parameters.
-The normalization factors, parental averages, and overdispersion parameters are assumed to follow a normal distribution, i.e.
-\[ 
-\gamma_{vi} \stackrel{ind}{\sim} N(0,\sigma_\gamma^2), \quad 
-\phi_g \stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2), \quad \mbox{and}\quad 
-\psi_g \stackrel{ind}{\sim} N(0,\sigma_\psi^2). 
-\]
-To allow for more shrinkage with the half-parental and hybrid differences, we use a Laplace (or double exponential prior) and thus implement a Bayesian LASSO \citep{park2008bayesian}, i.e. 
-\[ 
-\alpha_g \stackrel{ind}{\sim} La(\eta_\alpha, \sigma_\alpha) \qquad \mbox{and} \qquad
-\delta_g \stackrel{ind}{\sim} La(\eta_\delta, \sigma_\delta) 
-\]
-where $\alpha\sim La(\eta,\sigma)$ has a probability density function given by $p(\alpha|\eta,\sigma) = \exp(-|\alpha-\eta|/\sigma)/2\sigma$, expectation $\eta$, and variance $2\sigma^2$.
-We assume independence amongst the normalization factors, parental averages, half parental differences, hybrid differences, and overdispersion parameters.
-
-We assume improper uniform priors for $\eta_\phi$, $\eta_\alpha$, and $\eta_\delta$ and independent half-cauchy priors for all standard deviations \citep{gelman2006prior}.
-
-\subsection{Empirical Bayes}
-\label{s:ebayes}
-
-We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$ and hyperparameters $\pi = (\gamma, \eta, \sigma)$, where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, $\eta=(\eta_\phi,\eta_\alpha,\eta_\delta)$, and $\sigma=(\sigma_\gamma,\sigma_\phi,\sigma_\alpha,\sigma_\delta,\sigma_\psi)$.
-We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\pi$, denoted $\hat{\pi}$, and then base gene-specific inference on the posterior conditional on this MAP estimate.
-Conditional on the hyperparameters, the gene-specific parameters are independent, i.e.,
-\begin{equation}
-\begin{split}
-p\left(\theta\left|y,\hat{\pi}\right.\right) 
-%&= \prod_{g=1}^G p\left(\theta_g\left|y_g,\hat{\pi}\right.\right) = \prod_{g=1}^G p\left(y_g\left|\theta_g,\hat{\gamma}\right.\right) p\left(\theta_g\left|\hat{\pi}\right.\right) \nonumber \\
-%&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right.\right) \right] p(\theta_g|\hat{\eta},\hat{\sigma}) \\
-\propto \prod_{g=1}^G & \left[ \,\prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi};e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right)  \right. \\
-&\times 
-\left. \vphantom{\prod_{v=1}^V}
-N(\phi_g;\hat{\eta}_\phi, \hat{\sigma}_\phi^2)
-La(\alpha_g;\hat{\eta}_\alpha, \hat{\sigma}_\alpha)
-La(\delta_g;\hat{\eta}_\delta, \hat{\sigma}_\delta)
-N(\psi_g;\hat{\eta}_\psi, \hat{\sigma}_\psi^2) 
-\right]
-\label{e:condind}
-\end{split}
-\end{equation}
-Thus, conditional posterior inference for the gene-specific parameters can be obtained independently and in parallel.
-
-As we have not integrated out the gene-specific parameters when obtaining the MAP estimate, we actually receive a joint MAP estimate of $\theta$ and $\pi$.
-We explicitly use the MAP estimate of $\pi$ when obtaining empirical Bayes posterior samples for the gene-specific parameters as in equation \eqref{e:condind}.
-The MAP estimate of $\theta$ could be used as initial values in the MCMC, although we did not find this necessary.
-
-\subsection{Gene expression heterosis}
-\label{s:gene_heterosis}
-
-For a specific gene $g$, low-parent heteorisis (LPH) of gene expression occurs when average expression in the hybrid is less than the average expression of either parent, i.e.
-$\delta_g < -|\alpha_g|$, and high-parent heterosis (HPH) occurs when average expression in the hybrid is greater than the average expression of either parent, i.e.
-$\delta_g > |\alpha_g|$.
-% and high-parent  heterosis (HPH) of gene expression occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$.
-For simplicity, we define \emph{heterosis} of gene expression to be the event of LPH or HPH.
-The most important hypotheses are
-% \begin{align}
-% H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
-% H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3},
-% \label{e:hypotheses}
-% \end{align}
-% where $H_{g0}$ indicates no heterosis for gene $g$ and $H_{g1}$ indicates  heterosis for gene $g$. 
-% We evaluate these hypotheses based on empirical Bayes estimates of their posterior probabilities, e.g., 
-% \begin{align}
-% P\left(H_{g1}|y, \hat{\pi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\pi}_{MAP}\right) \nonumber \\
-% &\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right), \label{e:probs}
-% \end{align}
-% where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise.
-We evaluate these probabilities based on empirical Bayes estimates of their posterior probabilities, e.g., 
-\begin{equation}
-%P(LPH_g|y) \approx 
-P(LPH_g|y,\hat{\pi}_{MAP}) = P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi}_{MAP}\right) = \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right), 
-\label{e:probs}
-\end{equation}
-where $\left(\delta_g^{(m)},\alpha_g^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise.
-HPH probability is defined similarly with the inequality reversed and without the negative sign.
-We construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities.
-Geneticists can use this list to prioritize future experiments to understand the molecular genetic mechanisms for heterosis.
-
-\subsection{Implementation in \Stan{}}
-
-To perform the optimization required for the MAP estimate $\hat{\pi}$, we use the statistical software, {\tt Stan} \citep{stan-software:2014}, executed through the \RStan{} interface \citep{rstan-software:2014} in R \citep{R2014}.
-The Stan software allows the construction of general Bayesian models with the primary restriction that all unknown parameters are continuous.
-With a fully constructed model, the software can obtain MAP estimates \cite[Section 50.3,][]{stan-manual:2014} using the {\tt optimizing} function in \RStan{}.
-Presumably due to the large dimension of the optimization, it was necessary to use the L-BFGS algorithm \cite[Section 55,][]{stan-manual:2014} and reasonable starting values obtained via \edgeR{} \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor}.
-The {\tt Stan} software also includes the \RStan{} function, {\tt sampling}, an implementation of a Markov chain Monte Carlo (MCMC) algorithm for obtaining samples from the posterior.
-By default, Stan uses a no-U-turn sampler \citep{hoffman2013no}, a variant of Hamiltonion Monte Carlo \citep{neal2011mcmc}.
-
-\subsection{Implementation in \ShrinkBayes{}}
-
-\ShrinkBayes{} utilizes integrated nested Laplace approximation (INLA) \citep{rue2009approximate} in combination with empirical Bayes ideas \citep{van2014shrinkbayes}.
-One limitation with inferential methods based on INLA is that all distributions, except for the data distribution, must have tails as light or lighter than the normal density.
-Thus, we cannot implement the Laplace priors for the half-parental difference ($\alpha_g$) and the hybrid difference ($\delta_g$) and instead use normal priors in this situation.
-An additional limitation is that INLA provides approximations to marginal posteriors for parameters or linear combinations of parameters, but not an approximation to the full posterior.
-Since we are interested in non-linear quantities such as $P(\delta_g > |\alpha_g||y)$, we cannot compute these directly using \ShrinkBayes{}.
-Instead, for \ShrinkBayes{}, we calculate heterosis probabilities conditional on posterior means for $\hat{\delta}_g$ and $\hat{\alpha}_g$, e.g.
-\[ 
-% P_{\ShrinkBayes{}}(LPH_g|y) \equiv \left\{ \begin{array}{ll}
-% P(\delta_g>-\alpha_g|y) & \mbox{if } \hat{\delta}_g > -\hat{\alpha}_g \ge 0 \\
-% P(\delta_g>\phantom{-}\alpha_g|y) & \mbox{if } \hat{\delta}_g > \phantom{-}\hat{\alpha}_g \ge 0 \\
-% 0 & \mbox{otherwise}
-% \end{array} \right.
-P_{\ShrinkBayes{}}(LPH_g|y) \equiv P(\delta_g>-\alpha_g|y)\mathrm{I}\left(\hat{\delta}_g > -\hat{\alpha}_g \ge 0\right) + P(\delta_g>\alpha_g|y) \mathrm{I}\left(\hat{\delta}_g > \hat{\alpha}_g \ge 0\right).
-%P_{\ShrinkBayes{}}(HPH|y) \equiv \left\{ \begin{array}{ll}
-%P(\delta_g<-\alpha_g|y) & \mbox{if } \hat{\delta}_g > -\hat{\alpha}_g \ge 0 \\
-%(\delta_g<\phantom{-}\alpha_g|y) & \mbox{if } \hat{\delta}_g > \phantom{-}\hat{\alpha}_g \ge 0 \\
-%0 & \mbox{otherwise}
-%\end{array} \right.
-\]
-HPH probability is defined similarly with all inequalities reversed. As before, we construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities.
-%the heterosis probability for gene $g$ is zero if $|\hat{\delta}_g| \le |\hat{\alpha}_g|$, $P(\delta_g>\alpha_g|y)$ if $\hat{\delta}_g > -\hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g > 0 |y)$ if $\hat{\delta}_g > \hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g < 0 |y)$ if $\hat{\delta}_g < \hat{\alpha}_g \le 0$, and $P(\delta_g + \alpha_g < 0 |y)$ if $\hat{\delta}_g < -\hat{\alpha}_g \le 0$.
-
-\section{Simulation study based on a maize experiment}
-\label{s:simulation}
-
-% \subsection{Coverage for our model}
-% \input{../temp/analyze-heterosis/FIGURES/coverage}
-
-To assess the efficacy of our method to identify genes demonstrating heterosis, we used a maize data set with parental varieties B73 and Mo17 and the hybrid variety (B73 $\times$ Mo17) \citep{paschold2012complementation} to determine realistic parameter values for a simulation study.
-Section \ref{s:maize} describes the maize dataset in detail.
-We compared our method to approaches using the R packages \edgeR{}, \baySeq{} \citep{hardcastle2010bayseq, hardcastle2012baySeq}, and \ShrinkBayes{} \citep{van2014shrinkbayes}.
-
-\subsection{Constructing simulated data}
-\label{s:sim_data}
-
-We applied the \edgeR{} package \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor} to the maize data on 27,888 genes with average count at least one and at most two zero read counts for each variety across four replications.
-This analysis produced sample-specific normalization factors using the TMM method \citep{robinson2010scaling},
-%(0.080, -0.041, -0.054, 0.000, -0.036, -0.147, 0.093, 0.059, -0.004, -0.057, 0.067,  0.041) 
-gene-specific dispersion estimates, and gene-variety-specific location parameter estimates \citep{mccarthy2012differential}.
-We treated these estimates as the true parameter values for the simulation study so that our simulated datasets mimicked the existing covariance structure among the gene-variety means of the original maize data.
-
-Using these parameters and normalization factors, we simulated data according to the negative-binomial model in \eqref{e:data} independently for each gene.
-For each simulation, we analyzed a subset of 25,000 genes selected randomly from genes with simulated counts at least one on average and with at most two zeros for each variety across replications.
-We repeated this simulation process 10 times for each of 4, 8, and 16 replicates per variety, reusing normalization factors when necessary.
-% From Dan "Did we consider changing the number of zeros allowed when the number of reps is higher?
-I like the idea of simplifying all this by just picking an average count threshold that his high enough to avoid problems.
-I think that may be what others have done in the past."
-
-\subsection{Alternative methods}
-\label{s:alternative}
-
-Existing \RNAseq{} data analysis methods are not specifically designed to identify gene expression heterosis, but we adapted the approaches in \edgeR{} and {\tt baySeq} to provide a measure of the strength of heterosis for each gene such that large values of this measure indicate support for heterosis.
-
-\edgeR{} can be used to test for differential expression between any two varieties based on the fit of a negative binomial log-linear model \citep{robinson2007moderated, robinson2010edgeR}.
-To construct a measure of heterosis, we computed the maximum likelihood estimates of the $\mu_{gv}$ parameters for all genes using \edgeR{}'s built-in Fisher scoring algorithm, and then used likelihood ratio tests to calculate two $p$-values for each gene: $p_{g1}$  for testing $H_{g01}:\mu_{g1}=\mu_{g3}$ and $p_{g2}$ for testing $H_{g02}:\mu_{g2}=\mu_{g3}$.
-Then, for each gene, we define a new $p$-value denoted as $p_g$ and determined as $p_g=1$ if the estimate of $\mu_{g3}$ falls between the estimates of $\mu_{g1}$ and $\mu_{g2}$ and $p_g=\max\{p_{g1},p_{g2}\}/2$ otherwise.
-(For all relevant signifcance thresholds $\omega$ near 0, it can be shown that rejecting the null hypothesis of no heterosis of expression for gene $g$ whenever $p_g\leq \omega$ results in a test that is asymptotically size $\omega$.)  We then use $1-p_g$ as a final measure positively associated with strength of evidence for gene expression heterosis.
-
-{\tt baySeq} allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct.
-In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene.
-We set this measure to zero if the maximum likelihood estimate, calculated using \edgeR{}, of $\mu_{g3}$ is between the maximum likelihood estimates of $\mu_{g1}$ and $\mu_{g2}$.
-Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the only hypotheses that allow for heterosis.
-
-\subsection{Results}
-
-For each of the methods above, as well as the method described in Section \ref{s:gene_heterosis}, we sorted genes according to the computed measure of the strength of evidence for heterosis.
-From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to distinguish genes with heterosis, as defined in Section \ref{s:sim_data},
-% From Dan, "why this citation?"
-from those without heterosis \citep{landau2013dispersion}. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
-\begin{figure}[htbp]
-\centerline{\includegraphics[width=\textwidth]{exampleROC0_1}}
-\caption{Representative ROC curves for false positive rates below 0.1 for the approaches using \edgeR{}, {\tt baySeq},  and \ShrinkBayes{} described in Section \ref{s:alternative}, along with the independence and dependence methods using \RStan{} described in Section \ref{s:gene_heterosis}.}
-\label{f:roc}
-\end{figure}
-For this simulation, we can see that the approaches based on the model in Section \ref{s:model}, i.e.
-\RStan{} and \ShrinkBayes{}, provide the best true positive rate for a given false positive rate.
-Also, as expected, as the sample size increases, our ability to distinguish genes with heterosis from genes without improves.
-
-Figure \ref{f:auc} provides the area under the ROC curve below a false positive rate of 0.1 across the 10 simulations for each of the 3 different sample sizes. 
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{auc-facet-TRUE}}
-\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different sample sizes for the approaches using \edgeR{}, {\tt baySeq},  and \ShrinkBayes{} described in Section \ref{s:alternative} , along with the independence and dependence methods using \RStan{} described in Section \ref{s:gene_heterosis}.
-Each line is a different simulation while the blue box indicates mean AUC and mean AUC plus and minus one standard error across the simulations.}
-\label{f:auc}
-\end{figure}
-The pattern appears to hold, with the \RStan{} and \ShrinkBayes{} techniques outperforming the other two.
-%While there does not appear to be a significant difference between \ShrinkBayes{} and the independence model using \RStan{}, performance for \RStan{} appears to decrease for the dependence model where an unstructured covariance is estimated, at least for smaller sample sizes.
-
-\section{Searching for gene expression heterosis in the maize experiment}
-\label{s:maize}
-
-We used our method to analyze the maize data set, generated by \cite{paschold2012complementation}, on gene expression in parental lines B73 and Mo17 and the hybrid genotype (B73$\times$Mo17).
-The data are available in the Supplemental Materials.
-Each variety had four biological replicates measured with Illumina methodology and equipment.
-Reads were mapped to the whole reference genome using the short reads aligner, NOVOALIGN.
-For more specifics, please see \cite{paschold2012complementation}.
-
-As explained in Section~\ref{s:sim_data}, we analyzed a set of $G=27,888$ genes for $V=3$ varieties and $n_v=4$ replicates ($v=1,2,3$).
-
-We found MAP estimates for both the independence and dependence models as shown in Table \ref{t:map}. 
-\input{include/map}
-The MAP estimates for the hierarchical means, $\eta$, and standard deviations, $\sigma$, are consistent between the independence and dependence models with, perhaps, a slightly larger mean and smaller standard deviation for the logarithm of the overdispersion parameter, $\psi_g$, in the dependence model relative to the independence model.
-In the dependence model, there appears to be high correlation amongst $\mu_{g1}$, $\mu_{g2}$, and $\mu_{g3}$, and negative correlation between $\psi_g$ and each of $\mu_{g1}$, $\mu_{g2}$, and $\mu_{g3}$.
-The normalization factors, and their variance, are consistent between the two models, with the estimates for the dependence model being consistently 0.02 larger than the estimates for the independence model, and both indicating that there is about a 25\% increase in counts between the least and most thoroughly sequenced samples.
-
-Independently for each gene, we ran four Markov chains conditional on the estimated hyperparameters to obtain approximate posterior distributions for all gene-specific parameters.
-Each Markov chain used initial values randomly drawn by {\tt Stan} and was run for a total of 2000 iterations, with 1000 iterations of burn-in and the remaining 1000 draws thinned by a factor of 4 for storage reasons.
-% From Dan "Do we need to be more specific about what method was used to generate the chains?"
-The potential scale reduction factor, $\hat{R}$, was computed for each parameter \citep{Gelm:Rubi:infe:1992, Broo:Gelm:gene:1997}.
-If the maximum $\hat{R}$ was larger than 1.1 for any parameter, the total number of iterations was increased by 500, with the burn-in increased proportionally, and the chain was rerun.
-This process was repeated until the maximum $\hat{R}$ was less than 1.1 for all parameters.
-
-With posteriors for all parameters, we can calculate empirical Bayes posterior probabilities for LPH and HPH.
-For each gene, the quantity of interest is likely to be the maximum of these two probabilities.
-For each gene with a high probability of either HPH or LPH, the magnitude of the effect is of interest.
-We calculate
-\begin{equation}
-% \mbox{estimated effect size}_g = \left\{ 
-% \begin{array}{ll}
-% \hat{\mu}_{g3} - \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} < \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
-% \hat{\mu}_{g3} - \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} > \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
-% 0 & \mbox{otherwise},
-% \end{array} 
-% \right. 
-\mbox{estimated effect size}_g \equiv \left\{ 
-\begin{array}{ll}
-\hat{\delta}_g - |\hat{\alpha}_g| & \mbox{if } \hat{\delta}_g > \phantom{-}|\alpha_g| \\
-\hat{\delta}_g + |\hat{\alpha}_g| & \mbox{if } \hat{\delta}_g < -|\alpha_g| \\
-0 & \mbox{otherwise},
-\end{array} 
-\right. 
-\label{e:effect_size}
-\end{equation}
-where $\hat{\mu}_{gi}$ is the estimated posterior mean of $\mu_{gv}$ for $v=1,2,3$.
-Thus, the estimated effect size is the difference between hybrid mean and the nearest parent with negative values indicating LPH and positive values indicating HPH.
-If the hybrid mean is estimated to be between the parents, this effect is defined to be zero.
-
-Figure \ref{f:volcano} provides a volcano plot, in this case a bivariate histogram, to visualize the maximum of the probabilities of LPH and HPH versus estimated effect size.
-The figure shows a peak at zero for probabilities below 0.5 with estimated effect sizes of zero.
-Above a probability of 0.5, we see a prototypical volcano pattern, with increased heterosis probability corresponding to larger estimated effect sizes and no estimated effect sizes near zero for genes with high heterosis probability.
-We also see asymmetry, with larger negative effect sizes than positive effect sizes due to genes with hybrid counts near zero and relatively high parental counts.
-Genes with high estimated posterior probabilities of heterosis and large estimated effect sizes are candidates for further investigation.
-
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{maize_volcano_dexp}}
-\caption{A bivariaet histogram of the maximum of the LPH and HPH probabilities versus estimated effect size defined in equation \ref{e:effect_size} for the B73 $\times$ Mo17 maize experiment.}
-\label{f:volcano}
-\end{figure}
-
-With almost 30,000 genes in our data set, we can evaluate the distribution assumptions for the gene-specific parameters.
-Figure \ref{f:scatterplot} provides marginal and bivariate histograms for the gene-specific parameters.
-
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{scatterplot_dexp}}
-\caption{Marginal and bivariate histograms for gene-specific parameter posterior meansfor the B73 $\times$ Mo17 maize experiment.
-For reference, the marginal histograms have a dashed line indicating the appropriate marginal probability density functions in the hierarchical model with the MAP estimates.}
-\label{f:scatterplot}
-\end{figure}
-
-Marginal and bivariate histograms for the model with an estimated covariance look similar although the estimated distribution $N(\hat{\eta},\hat{\Sigma})$ has increased variance for $\mu$ and decreased variance for $\psi$ as was previously seen in Table \ref{t:map} (results not shown).
-
-\section{Discussion}
-\label{s:discussion}
-
-Geneticists speculate that gene expression heterosis is one possible explanation of phenotypic heterosis of traits, such as plant height or grain yield.
-Existing methods for identifying differential gene expression based on \RNAseq{} data are not directly applicable for detecting heterosis genes.
-\cite{ji2014estimation} introduced an empirical Bayes approach based on a hierarchical model for microarray data.
-We followed their approach, modified to allow for \RNAseq{} read counts as measures of transcript abundance.
-We developed an empirical Bayes approach based on obtaining MAP estimates for hyperparameters followed by MCMC to estimate gene-specific parameters.
-This methodology can be conveniently implemented in the statistical software, {\tt Stan}.
-The empirical Bayes posteriors can be used to estimate posterior probabilities of high and low parent heterosis.
-
-Through a simulation study, we demonstrated that this method outperformed alternative methods (that are not designed to detect heterosis), and performed comparably well with an implementation of our model in \ShrinkBayes{}, which estimates the full posterior via INLA.
-We then demonstrated the use of the methodology on a maize experiment in which phenotypic heterosis is well known.
-% Estimates of heterosis to compare with Ji et. al.
-In addition to providing estimates of heterosis probabilities and effect sizes, this real data analysis exposed violations of our normality assumptions, indicating room to improve the model and increase the power to detect gene expression heterosis.
-Some of our ongoing and future research will develop the details of such an extension.
-
-\backmatter %  Please keep this command in your document in this position
-
-\if0\blind{
-\section*{Acknowledgements}
-
-The authors thank Andrew Lithio for help in implementing our model in \ShrinkBayes{}.
-Research reported in this publication was supported by National Institute of General Medical Sciences of the National Institutes of Health under award number R01GM109458.
-The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health.
-} \fi
-
-%  If your paper refers to supplementary web material, then you MUST
-%  include this section!!  See Instructions for Authors at the journal
-%  website http://www.biometrics.tibs.org
-
-\section*{Supplementary Materials}
-
-The supplementary materials include the maize experiment data ({\tt data.csv}), a script to run our method ({\tt script.R}), and the {\tt Stan} models: the all-genes-independent (\verb@model_ind.txt@) and covariance (\verb@model_cov.txt@) models used in obtaining MAP estimates, along with the single-gene-independent (\verb@sg_model_ind.txt@) and covariance (\verb@sg_model_cov.txt@) models used in performing the MCMC on single genes.
-
-\bibliography{jarad}
-\bibliographystyle{biom}
-
-% From Dan, "There are some capitalization issues in the references section.
-I think you need to protect "RNA" with curly braces so that it stays capitalized.
-Also, in Hallauer et al.
-(2010), the last initial should presumably be "D." rather than "d."."
-
-% \appendix
-% %  To get the journal style of heading for an appendix, mimic the following.
-% \section{}
-% \subsection{Material here}
-
-\label{lastpage}
+Then the likelihood is 
+\[ L(\phi_g,\alpha_g,\delta_g,\psi_g) \propto \]
 

### 2015-03-07 sandbox/sim_data.R
sim_data function

@@ -1,0 +1,36 @@
+library(plyr)
+
+G = 10
+v = 3
+nv = rep(4,3)
+
+rlaplace = function(n, mean, scale) {
+  u = runif(n, -.5, .5)
+  mean - scale * sign(u) * log(1-2*abs(u))
+}
+
+# Priors
+prior = list(eta_phi     = 1,
+             eta_alpha   = 0,
+             eta_delta   = 0,
+             eta_psi     = -2,
+             sigma_phi   = 1,
+             sigma_alpha =0.1,
+             sigma_delta = 0.01,
+             sigma_psi   = 0.1)
+
+# Parameters
+parameters = with(prior, list(phi   = rnorm   (G, eta_phi,   sigma_phi  ),
+                              alpha = rlaplace(G, eta_alpha, sigma_alpha),
+                              delta = rlaplace(G, eta_delta, sigma_delta),
+                              psi   = rnorm   (G, eta_psi,   sigma_psi  )))
+
+# Data
+d = ddply(data.frame(gene=1:G), .(gene), function(x) {
+  mutate(data.frame(parent = rep(1:3, times = nv)),
+         sample = 1:sum(nv),
+         mu = with(parameters, c(phi[g]+alpha[g], phi[g]-alpha[g], phi[g]+delta[g])),
+         y = # need negative binomial with mu parameterization
+  )
+  
+})

### 2015-03-07 sandbox/Laplace.R
laplace distribution

@@ -1,0 +1,36 @@
+#' Laplace (double exponential distribution)
+#' 
+#' Density, distribution function, quantile function, and random generation for the Laplace 
+#' (double exponential) distribution with mean equal to mean and scale equal to sd.
+#' 
+#' @usage dlaplace(x, mean = 0, scale = 1, log=FALSE)
+#' @usage rlaplace(n, mean = 0, scale = 1)
+#' 
+#' @param x vector of quantiles.
+#' @param n number of observations. If length(x) > 1, the length is taken to be the number required.
+#' @param mean vector of means.
+#' @param scale vector of scales.
+#' 
+#' @details The Laplace distribution has density ...
+#' 
+#' @return dlaplace gives the density, plaplace gives the distribution function, 
+#'   qlaplace gives the quantile function, and rlaplace generates random deviates. 
+#'   
+#'   The length of the result is determined by n for rlaplace, and is the maximum 
+#'   of the lengths of the numerical arguments for the other functions.
+#'   
+#'   The numerical arguments other than n are recycled to the length of the result. 
+#'   Only the first elements of the logical arguments are used.
+#'   
+#' @examples dlaplace(0)
+#' 
+dlaplace = function(x, mean=0, scale=1, log=FALSE) {
+  logd = -abs(x-mean)/scale - log(2*scale)
+  if (log) return(logd)
+  return(exp(logd))
+}
+
+rlaplace = function(n, mean=0, scale=1) {
+  u = runif(n)-.5
+  mean - scale * sign(u)*log(1-2*abs(u))
+}

### 2015-03-07 sandbox/sim_data.R
simulate heterosis data

@@ -1,36 +1,0 @@
-library(plyr)
-
-G = 10
-v = 3
-nv = rep(4,3)
-
-rlaplace = function(n, mean, scale) {
-  u = runif(n, -.5, .5)
-  mean - scale * sign(u) * log(1-2*abs(u))
-}
-
-# Priors
-prior = list(eta_phi     = 1,
-             eta_alpha   = 0,
-             eta_delta   = 0,
-             eta_psi     = -2,
-             sigma_phi   = 1,
-             sigma_alpha =0.1,
-             sigma_delta = 0.01,
-             sigma_psi   = 0.1)
-
-# Parameters
-parameters = with(prior, list(phi   = rnorm   (G, eta_phi,   sigma_phi  ),
-                              alpha = rlaplace(G, eta_alpha, sigma_alpha),
-                              delta = rlaplace(G, eta_delta, sigma_delta),
-                              psi   = rnorm   (G, eta_psi,   sigma_psi  )))
-
-# Data
-d = ddply(data.frame(gene=1:G), .(gene), function(x) {
-  mutate(data.frame(parent = rep(1:3, times = nv)),
-         sample = 1:sum(nv),
-         mu = with(parameters, c(phi[g]+alpha[g], phi[g]-alpha[g], phi[g]+delta[g])),
-         y = # need negative binomial with mu parameterization
-  )
-  
-})

### 2015-03-07 sandbox/sim_heterosis_data.R

@@ -1,0 +1,45 @@
+#' Simulate heterosis data
+#' 
+#' Simulates heterosis data from either hyperparameters (and thus simulating gene-specific parameters)
+#' or gene-specific parameters.
+#' 
+#' @param G integer. The number of genes to simulate.
+#' @param nv integer (or integer vector). indicating the number of samples per genotype
+#' @param parameters. a list containing either hyperparameters or (gene-specific) parameters
+#' @param distributions. a list containing the distribution to use for the gene-specific parameters
+#' 
+#' @return list containing the hyperparameters, gene-specific parameters, and data
+#' 
+
+sim_heterosis_data = function(G=10, nv=4, parameters=NULL, distributions=NULL) {
+  require(plyr)
+  
+  if (length(nv==1)) nv = rep(nv,3)
+  
+  if (is.null(parameters)) 
+    hyperparameters = data.frame(parameter      = c("phi","alpha","delta","psi"),
+                                 location = c(4.6,0,0,-2),
+                                 scale    = c(1.8,.1,.01,.1))
+  
+  # Simulate gene-specific parameters
+  if (is.null(distributions)) 
+    parameters = rdply(G, {
+      with(hyperparameters, 
+           data.frame(phi   = rnorm   (1, location[parameter == "phi"  ], scale[parameter == "phi"  ]),
+                      alpha = rlaplace(1, location[parameter == "alpha"], scale[parameter == "alpha"]),
+                      delta = rlaplace(1, location[parameter == "delta"], scale[parameter == "delta"]),
+                      psi   = rnorm   (1, location[parameter == "psi"  ], scale[parameter == "psi"  ])))
+    }, .id="gene")
+  
+  data = ddply(data.frame(gene=1:G), .(gene), function(x) {
+    g = x$gene
+    mutate(data.frame(parent = rep(1:3, times = nv)),
+           sample = 1:sum(nv),
+           eta = with(parameters, c(phi[g]+alpha[g], phi[g]-alpha[g], phi[g]+delta[g])),
+           y = rnbinom(length(eta), size = 1/exp(parameters$psi[g]), mu = exp(eta)))
+  })
+  
+  return(list(data            = data, 
+              parameters      = parameters,
+              hyperparameters = hyperparameters))
+}

### 2015-03-09 sandbox/normal_approximation.tex
weekend

@@
 \documentclass{article}
+
+\usepackage{units, amsmath}
+
+\newcommand{\mGamma}{\mathrm{\Gamma}}
 
@@
 
-Model 
+\section{Model}
 \begin{equation} 
@@
 \label{e:data}
-The probability mass function is 
-\[ p(y_{gvi}|\mu_{gv},\gamma_{vi},\psi_g) = \frac{}{} \]
+\end{equation}
+with probability mass function 
+\[ p(y_{gvi}|\mu_{g},\gamma_{vi},\psi_g) =\frac{\mGamma(e^{\psi_g})}{y_{gvi}!\mathrm{\Gamma(e^{\psi_g})}}  p_{gv}^{y_{gvi}}(1-p_{gv})^{e^{\psi_g}} \] 
+where $\mu_g = (\mu_{g1},\mu_{g2},\mu_{g3})$ and 
+\[ p_{gv} = \frac{e^{\mu_{gv}+\gamma_{vi}}}{e^{\psi_g}+e^{\mu_{gv}+\gamma_{vi}}}. \]
+This parameterization of the negative binomial has 
+\[ E[Y_{gvi}] = \eta_{gvi} =e^{\mu_{gv}+\gamma_{vi}} \quad \mbox{and} \quad V[Y_{gvi}] = \eta_{gvi} + \eta_{gvi}^2/e^{\psi_g}. \]
+Let $y_{g} = (y_{g11},\ldots,y_{g1n_1},y_{g21},\ldots,y_{g2n_2},y_{g31},\ldots,y_{g3n_3})$, then 
+\[ 
+p(y_g|\mu_{gv} ,\gamma_{vi}, \psi_g) = \prod_{v=1}^3 \prod_{i=1}^{n_v} p(y_{gvi}|\mu_{g},\gamma_{vi},\psi_g) 
+\]
 
-\end{equation}
-with $E[Y_{gvi}] = \eta_{gvi} =e^{\mu_{gv}+\gamma_{vi}}$ and $V[Y_{gvi}] = \eta_{gvi} + e^{\psi_g}\eta_{gvi}^2$.
+\subsection{Reparameterization}
+
+Rather than using a parameterization for the variety specific means, we reparameterize for the purpose of constructing a hierarhical model that performs appropriate shrinkage. Let
+
 \[ \phi_g = \frac{\mu_{g1}+\mu_{g2}}{2}, \quad 
@@
 \delta_g = \phi_g-\mu_{g3}. \]
-Then the likelihood is 
-\[ L(\phi_g,\alpha_g,\delta_g,\psi_g) \propto \]
+So, 
+\[ \theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)^\top = T \mu_g = 
+\left[ \begin{array}{cccc}
+\nicefrac{1}{2} & \nicefrac{1}{2} & 0 & 0 \\
+\nicefrac{1}{2} & -\nicefrac{1}{2} & 0 & 0\\
+\nicefrac{1}{2} & \nicefrac{1}{2} & -1 & 0\\
+0 & 0 & 0 & 1\\
+\end{array}\right] \left[ \begin{array}{c}
+\mu_{g1} \\ \mu_{g2} \\ \mu_{g3} \\ \psi_g 
+\end{array} \right]
+\]
+We assume the distribution $\theta_g \sim p(\theta_g|\pi)$.
+
+\subsection{Observed Fisher information}
+
+Let $\mathcal{I}(\mu_g)$ be the Fisher information for $\mu_g$, then 
+\begin{align*}
+\log p(y|\mu_g) 
+&= \sum_{v=1}^3 \sum_{i=1}^{n_v} \log \mGamma(e^{\psi_g}+y_{gvi}) - \log(y_{gvi}!) -\mGamma(e^{\psi_g}) + y_{gvi}\log(p_{gv}) + e^{\psi_g} \log(1-p_{gv}) \\
+\frac{\partial}{\partial \mu_{gv}} \log p(y|\mu_g)  
+&= \sum_{v=1}^3 \sum_{i=1}^{n_v} \frac{y_{gvi}}{p_{gv}} \frac{\partial p_{gv}}{\partial \mu_{gv}} - \frac{e^{\psi_g}}{1-p_{gv}}  \frac{\partial p_{gv}}{\partial \mu_{gv}}  \\
+&= \sum_{v=1}^3 \sum_{i=1}^{n_v} \frac{y_{gvi}}{p_{gv}} \frac{\partial p_{gv}}{\partial \mu_{gv}} - \frac{e^{\psi_g}}{1-p_{gv}}  \frac{\partial p_{gv}}{\partial \mu_{gv}}  \\
+%
+%
+%
+%
+%
+%\frac{\partial}{\partial \mu_{g1}}  \log p(y|\mu_g)  
+%&= \sum_{v=1}^3 \sum_{i=1}^{n_v} \nabla \left(\log \mGamma(e^{\psi_g}+y_{gvi}) - \log(y_{gvi}!) -\mGamma(e^{\psi_g})  + y_{gvi}\log(p_{gv}) + e^{\psi_g} \log(1-p_{gv}) \right) \\
+\end{align*}
+
+\subsection{Normal approximation}
+
+We would like to approximate the conditional 
+\[ p(\theta_g| y,\pi) \propto p(y|\theta_g,\pi) = \prod_{v=1}^V \prod_{i=1}^{n_v} NB(y_{gvi}; \]
 

### 2015-03-09 sandbox/find_map_estimates.R

@@ -1,0 +1,18 @@
+library(Rcpp)
+library(reshape2)
+
+source("sim_heterosis_data.R")
+
+# Get data
+all = sim_heterosis_data(G=2)
+
+# Reshape data (put this in a function?)
+count = dcast(all$data[,c("gene","sample","y")], gene ~ sample, value.var='y')
+variety = all$data$parent[1:ncol(count)-1] # hack
+
+
+hyperparameters = melt(all$hyperparameters)$value
+
+sourceCpp("monte_carlo_integral.cpp")
+
+monte_carlo_integral(as.matrix(count[,-1]), variety, hyperparameters, 1)

### 2015-03-09 sandbox/monte_carlo_integral.cpp

@@ -1,0 +1,71 @@
+#include <Rcpp.h>
+using namespace Rcpp;
+
+// Below is a simple example of exporting a C++ function to R. You can
+// source this function into an R session using the Rcpp::sourceCpp 
+// function (or via the Source button on the editor toolbar)
+
+// For more on using Rcpp click the Help button on the editor toolbar
+
+// [[Rcpp::export]]
+double monte_carlo_integral(IntegerMatrix count, 
+                            IntegerVector variety, 
+                            NumericVector hyperparameters,
+                            int n_sims) {
+  // count is the observed count (gene x sample)
+  // variety is the variety identifier for the columns
+  // hyperparameters are the values for the hyperparameters
+  // sims number of Monte Carlo simulations to use
+  int G = count.nrow();
+  int n = count.ncol();
+  
+  // argument error handling
+  if (variety.size() != n) throw std::range_error("Length of variety and number of columns of count do not agree.");
+  if (hyperparameters.size() != 8) throw std::range_error("Hyperparameters must have length 8.");
+  
+  // set up hyperparameters
+  double phi_location   = hyperparameters[0],
+         alpha_location = hyperparameters[1],
+         delta_location = hyperparameters[2],
+         psi_location   = hyperparameters[3],
+         phi_scale      = hyperparameters[4],
+         alpha_scale    = hyperparameters[5], 
+         delta_scale    = hyperparameters[6],
+         psi_scale      = hyperparameters[7];
+         
+  double integral = 0, partial_integral, log_mass, phi, alpha, delta, psi, mu[3], gene_avg;
+  
+  for (int g=0; g<G; g++) {
+    partial_integral = 0;
+    
+    gene_avg = 0;
+    for (int ii=0; ii<n; ii++) gene_avg += count(g,ii);
+    gene_avg /= n;
+    
+    for (int s=0; s<n_sims; s++) {
+      //phi   = rnorm(1, phi_location,   phi_scale)[0]; 
+      phi   = rnorm(1, log(gene_avg),   .1)[0]; 
+      alpha = rnorm(1, alpha_location, alpha_scale)[0]; // change to laplace
+      delta = rnorm(1, delta_location, delta_scale)[0]; // change to laplace
+      psi   = rnorm(1, psi_location+2,   psi_scale)[0];
+      
[…]

### 2015-03-09 sandbox/monte_carlo_integration.R

@@ -1,0 +1,6 @@
+source("sim_heterosis_data.R")
+
+# Get data
+d = sim_heterosis_data()
+
+# Find MAP estimates

### 2015-03-09 sandbox/integration.tex
methods of integration

\documentclass{article}

\usepackage{amsmath}

\newcommand{\argmax}{\mbox{argmax}}

\begin{document}

\section{MAP estimate}

Let $\theta_g$ be the gene-specific parameters and $\pi$ be the hyperparameters. The MAP estimate for $\pi$ is 
\begin{align*}
\argmax_{\pi} \log p(\pi|y) 
&= \argmax_{\pi} \log\left( \int \prod_{g=1}^G \left[ \, \prod_{v=1}^V \prod_{i=1}^{n_v} NB(y_{gvi}|f(\theta_g)) p(\theta_g|\pi) \right] p(\pi) d\theta \right) \\
&= \argmax_{\pi}\sum_{g=1}^G \log \left[ \, \int \prod_{v=1}^V \prod_{i=1}^{n_v} NB(y_{gvi}| f(\theta_g)) p(\theta_g|\pi) d\theta_g  \right] + \log(p(\pi)) \\
&= \argmax_{\pi}\sum_{g=1}^G \log(I_g) + \log(p(\pi)) 
\end{align*}
where 
 \[
 I_g = \int \prod_{v=1}^V \prod_{i=1}^{n_v} NB(y_{gvi}| f(\theta_g)) p(\theta_g|\pi) d\theta_g
 \]

\subsection{Quadrature}

Using quadrature, we would have a set of points $\theta_g^{(k)}$ that would depend on $\pi$ and associated weight $w_k$. We would approximate the integral via 
\[
I_g \approx \sum_{k=1}^K w_k NB(y_{gvi}| f(\theta_g^{(k)})).
\]

\subsection{Monte Carlo integration}

Alternatively, we can use Monte Carlo integration which would sample $\theta_g^{(k)} \sim p(\theta_g|\pi)$ and approximate the integral via 
\[
I_g \approx \frac{1}{K} \sum_{k=1}^K NB(y_{gvi}| f(\theta_g^{(k)})).
\]

\subsection{Importance sampling}

Importance sampling uses an alternative sampling distribution $q(\theta_g|\pi)$. Then we sample $\theta_g^{(k)} \sim q(\theta_g|\pi)$ and approximate the integral via 
\[
I_g \approx \sum_{k=1}^K w_k NB(y_{gvi}| f(\theta_g^{(k)})).
\]
where $w_k = p(\theta^{(k)}|\pi)/p(\theta^{(k)}|\pi)$.

If we make the proposal gene specific, i.e. $q_g(\theta_g|\pi)$, then we can sample using a proposal that is tuned to the data. For example, one component of $\theta_g$ is $\phi_g$, the mean parental response, i.e. $E[y_{g1i}+y_{g2i}]/2$. This parameter is extremely variable across $g$. It certainly makes sense to choose $\phi_g$ in a range near the empirical parental average, i.e.

\[ \hat{\phi}_g = \frac{1}{2}\left[ \frac{1}{n_1} \sum_{i=1}^{n_1} y_{g1i} + \frac{1}{n_2} \sum_{i=1}^{n_2} y_{g2i} \right] \]

\end{document}

### 2015-03-09 sandbox/monte_carlo_integral.cpp
implement logsumexp

@@ -7,6 +7,20 @@ using namespace Rcpp;
 
 // For more on using Rcpp click the Help button on the editor toolbar
 
+double logsumexp(double x[], int n) {
+  double max_val = x[0], sum = 0.0;
+  int i;
+  
+  for (i = 1; i< n; i++)
+    if (x[i] > max_val) 
+      max_val = x[i];
+      
+  for (i=0; i<n; i++)
+    sum += exp(x[i] - max_val);
+    
+  return log(sum) + max_val;
+}
+
 // [[Rcpp::export]]
 double monte_carlo_integral(IntegerMatrix count, 
                             IntegerVector variety, 
@@ -33,7 +47,7 @@ double monte_carlo_integral(IntegerMatrix count,
          delta_scale    = hyperparameters[6],
          psi_scale      = hyperparameters[7];
          
-  double integral = 0, partial_integral, log_mass, phi, alpha, delta, psi, mu[3], gene_avg;
+  double integral = 0, partial_integral, log_mass[n_sims], phi, alpha, delta, psi, mu[3], gene_avg;
   
   for (int g=0; g<G; g++) {
     partial_integral = 0;
@@ -43,28 +57,25 @@ double monte_carlo_integral(IntegerMatrix count,
     gene_avg /= n;
     
     for (int s=0; s<n_sims; s++) {
-      //phi   = rnorm(1, phi_location,   phi_scale)[0]; 
-      phi   = rnorm(1, log(gene_avg),   .1)[0]; 
+      phi   = rnorm(1, phi_location,   phi_scale)[0]; 
+      //phi   = rnorm(1, log(gene_avg),   .1)[0]; 
       alpha = rnorm(1, alpha_location, alpha_scale)[0]; // change to laplace
       delta = rnorm(1, delta_location, delta_scale)[0]; // change to laplace
-      psi   = rnorm(1, psi_location+2,   psi_scale)[0];
+      psi   = rnorm(1, psi_location,   psi_scale)[0];
       
       mu[0] = phi+alpha; // make sure these agree with definition
       mu[1] = phi-alpha;
       mu[2] = phi+delta;
       
-      log_mass = 0;
+      log_mass[s] = 0;
       for (int i=0; i<n; i++) {
-        log_mass += dnbinom_mu(count(g,i), 1/exp(psi), exp(mu[variety[i]]), 1);
[…]

### 2015-03-09 sandbox/monte_carlo_integral.cpp
allows for importance sampling

@@ -7,6 +7,26 @@ using namespace Rcpp;
 
 // For more on using Rcpp click the Help button on the editor toolbar
 
+void rpadp(double *padp, NumericVector mean, NumericVector scale) {
+  for (int i=0; i<4; i++) 
+    padp[i] = rnorm(1, mean[i], scale[i])[0];
+}
+
+double dpadp(double *padp, NumericVector mean, NumericVector scale) {
+  double sum=0;
+  
+  for (int i=0; i<4; i++)
+    sum += R::dnorm(padp[i], mean[i], scale[i], 1);
+
+  return sum;
+}
+
+void calculate_mu(double mu[], double padp[], IntegerVector variety) {
+  mu[0] = padp[0]+padp[1]; // make sure these agree with definition
+  mu[1] = padp[0]-padp[1];
+  mu[2] = padp[0]+padp[2];
+}
+
 double logsumexp(double x[], int n) {
   double max_val = x[0], sum = 0.0;
   int i;
@@ -24,58 +44,47 @@ double logsumexp(double x[], int n) {
 // [[Rcpp::export]]
 double monte_carlo_integral(IntegerMatrix count, 
                             IntegerVector variety, 
-                            NumericVector hyperparameters,
+                            NumericVector mean,
+                            NumericVector scale,
                             int n_sims) {
   // count is the observed count (gene x sample)
   // variety is the variety identifier for the columns
-  // hyperparameters are the values for the hyperparameters
-  // sims number of Monte Carlo simulations to use
-  int G = count.nrow();
-  int n = count.ncol();
+  // mean and scale contain the hyperparameters for the phi-alpha-delta-psi distributions
+  // n_sims number of Monte Carlo simulations to use
+  int G = count.nrow(); // number of genes
+  int n = count.ncol(); // number of samples
+  double log_n_sims = log( (double) n_sims);
   
   // argument error handling
   if (variety.size() != n) throw std::range_error("Length of variety and number of columns of count do not agree.");
-  if (hyperparameters.size() != 8) throw std::range_error("Hyperparameters must have length 8.");
-  
-  // set up hyperparameters
-  double phi_location   = hyperparameters[0],
-         alpha_location = hyperparameters[1],
-         delta_location = hyperparameters[2],
[…]

### 2015-03-09 sandbox/find_map_estimates.R
beginning of function to find MAP estimates

@@ -2,17 +2,24 @@ library(Rcpp)
 library(reshape2)
 
 source("sim_heterosis_data.R")
+source("Laplace.R")
 
 # Get data
-all = sim_heterosis_data(G=2)
+all = sim_heterosis_data(G=3)
 
 # Reshape data (put this in a function?)
 count = dcast(all$data[,c("gene","sample","y")], gene ~ sample, value.var='y')
 variety = all$data$parent[1:ncol(count)-1] # hack
 
 
-hyperparameters = melt(all$hyperparameters)$value
+hyperparameters = melt(all$hyperparameters)
 
 sourceCpp("monte_carlo_integral.cpp")
 
-monte_carlo_integral(as.matrix(count[,-1]), variety, hyperparameters, 1)
+integral = rep(NA,20)
+for (i in 1:length(integral))
+  integral[i] = monte_carlo_integral(as.matrix(count[,-1]), variety-1, 
+                     with(hyperparameters, value[variable=="location"]), 
+                     with(hyperparameters, value[variable=="scale"]), 
+                     4)
+

### 2015-03-09 sandbox/find_map_estimates.R
more genes

@@ -5,7 +5,7 @@ source("sim_heterosis_data.R")
 source("Laplace.R")
 
 # Get data
-all = sim_heterosis_data(G=3)
+all = sim_heterosis_data(G=100)
 
 # Reshape data (put this in a function?)
 count = dcast(all$data[,c("gene","sample","y")], gene ~ sample, value.var='y')
@@ -16,10 +16,8 @@ hyperparameters = melt(all$hyperparameters)
 
 sourceCpp("monte_carlo_integral.cpp")
 
-integral = rep(NA,20)
-for (i in 1:length(integral))
-  integral[i] = monte_carlo_integral(as.matrix(count[,-1]), variety-1, 
-                     with(hyperparameters, value[variable=="location"]), 
-                     with(hyperparameters, value[variable=="scale"]), 
-                     4)
+log_like = function(pi) {
+  -monte_carlo_integral(as.matrix(count[,-1]), variety-1, pi[1:4], exp(pi[5:8]), 1000)
+}
 
+o = optim(rep(0,8), log_like)

### 2015-03-09 sandbox/monte_carlo_integral.cpp

@@ -67,7 +67,7 @@ double monte_carlo_integral(IntegerMatrix count,
     mu[3];             // array to contain values for mu derived from padp
   
   for (int g=0; g<G; g++) {
-    Rprintf("%i: ", g); 
+    //Rprintf("%i: ", g); 
     
     for (int s=0; s<n_sims; s++) {
       rpadp(padp, mean, scale);        // random draws for phi, alpha, delta, and psi
@@ -83,8 +83,8 @@ double monte_carlo_integral(IntegerMatrix count,
     }
     
     integral += logsumexp(log_mass, n_sims) - log_n_sims;
-    Rprintf(", total=%f", logsumexp(log_mass, n_sims) - log_n_sims);
-    Rprintf("\n");
+    //Rprintf(", total=%f", logsumexp(log_mass, n_sims) - log_n_sims);
+    //Rprintf("\n");
   }     
          
   return integral;

### 2015-03-10 sandbox/find_map_estimates.R
dnbinom function not working

@@ -17,7 +17,7 @@ hyperparameters = melt(all$hyperparameters)
 sourceCpp("monte_carlo_integral.cpp")
 
 log_like = function(pi) {
-  -monte_carlo_integral(as.matrix(count[,-1]), variety-1, pi[1:4], exp(pi[5:8]), 1000)
+  -monte_carlo_integral(t(as.matrix(count[,-1])), variety-1, pi[1:4], exp(pi[5:8]), 1000)
 }
 
 o = optim(rep(0,8), log_like)

### 2015-03-10 sandbox/monte_carlo_integral.cpp

@@ -21,10 +21,14 @@ double dpadp(double *padp, NumericVector mean, NumericVector scale) {
   return sum;
 }
 
-void calculate_mu(double mu[], double padp[], IntegerVector variety) {
-  mu[0] = padp[0]+padp[1]; // make sure these agree with definition
-  mu[1] = padp[0]-padp[1];
-  mu[2] = padp[0]+padp[2];
+void calculate_mu(NumericVector mu, double padp[], IntegerVector variety) {
+  double tmp_mu[3];
+  tmp_mu[0] = padp[0]+padp[1]; // make sure these agree with definition
+  tmp_mu[1] = padp[0]-padp[1];
+  tmp_mu[2] = padp[0]+padp[2];
+  
+  for (int i=0; i<mu.length(); i++)
+    mu[i] = tmp_mu[variety[i]];
 }
 
 double logsumexp(double x[], int n) {
@@ -41,18 +45,30 @@ double logsumexp(double x[], int n) {
   return log(sum) + max_val;
 }
 
+
+// [[Rcpp::export]]
+NumericVector dnbinom_mu(NumericMatrix::Column x, double size, NumericVector mu, int lg) {
+  // Vectorized for mu (mean)
+  NumericVector result(mu.length());
+  for (int i=0; i<mu.length(); i++) 
+    result[i] = dnbinom_mu(x[i], size, mu[i], lg);
+  return result;
+}
+
+
+
 // [[Rcpp::export]]
 double monte_carlo_integral(IntegerMatrix count, 
                             IntegerVector variety, 
                             NumericVector mean,
                             NumericVector scale,
                             int n_sims) {
-  // count is the observed count (gene x sample)
+  // count is the observed count (sample x gene, column major for quick access)
   // variety is the variety identifier for the columns
   // mean and scale contain the hyperparameters for the phi-alpha-delta-psi distributions
   // n_sims number of Monte Carlo simulations to use
-  int G = count.nrow(); // number of genes
-  int n = count.ncol(); // number of samples
+  int G = count.ncol(); // number of genes
+  int n = count.nrow(); // number of samples
   double log_n_sims = log( (double) n_sims);
   
   // argument error handling
@@ -63,8 +79,9 @@ double monte_carlo_integral(IntegerMatrix count,
          
   double integral = 0, // value of the integral (updated for each gene)
[…]

### 2015-03-12 sandbox/find_map_estimates.R
1/K

@@ -10,8 +10,8 @@ source("Laplace.R")
 sourceCpp("monte_carlo_integral.cpp", verbose=TRUE)
 
 # Get data
-set.seed(1)
-all = sim_heterosis_data(G=2)
+set.seed(3)
+all = sim_heterosis_data(G=25)
 R = rbind(c(.5, .5, 0),
           c(.5,-.5, 0),
           c(.5, .5,-1))
@@ -23,11 +23,11 @@ X = with(all$data[all$data$gene==1,],
 data_summary = ddply(mutate(all$data, gene=as.factor(gene), parent=as.factor(parent)),
           .(gene), 
           function(x) {
-            m = glm.nb(formula = y~0+X, data=x)
+            m = glm.nb(formula = y~0+X, data=x, control=glm.control(maxit=1000))
             data.frame(
               parameter = c('phi', 'alpha', 'delta', 'psi'),
               estimate = c(coef(m), -log(summary(m)$theta)),
-              sd = sqrt(c(diag(vcov(m, summary(m)$theta)),NA)))
+              sd = sqrt(c(diag(summary(m)$cov.scaled),NA)))
           })
 
 # Reshape data (put this in a function?)
@@ -39,25 +39,31 @@ hyperparameters = melt(all$hyperparameters)
 
 # sourceCpp("monte_carlo_integral.cpp", verbose=TRUE)
 
-log_like = function(pi) {
-  monte_carlo_integral(t(as.matrix(count[,-1])), variety-1, pi[1:4], exp(pi[5:8]), 
-                        with(data_summary, estimate[parameter=='phi']),
-                        with(data_summary,       sd[parameter=='phi']),
-                        10)
-}
 
-r = ddply(data.frame(n_sims = 10^seq(1,4)), .(n_sims), function(x) {
-  rdply(.n=10,data.frame(integral = monte_carlo_integral(t(as.matrix(count[,-1])), variety-1, pi[1:4], exp(pi[5:8]), 
+r = ddply(expand.grid(n_sims = 10^seq(1,4), n_genes=c(25)), 
+          .(n_sims, n_genes), 
+          function(x) {
+  rdply(.n=100,data.frame(integral = monte_carlo_integral(t(as.matrix(count[1:x$n_genes,-1])), variety-1, pi[1:4], exp(pi[5:8]), 
                                              with(data_summary, estimate[parameter=='phi']),
                                              with(data_summary,       sd[parameter=='phi']),
[…]

### 2015-03-31 sandbox/integration.tex
integrations notes
1/K
added em

@@
 \[
-I_g \approx \sum_{k=1}^K w_k NB(y_{gvi}| f(\theta_g^{(k)})).
+I_g \approx \sum_{k=1}^K w_k \prod_{v=1}^V \prod_{i=1}^{n_v}  NB(y_{gvi}| f(\theta_g^{(k)})).
 \]
@@
 \[
-I_g \approx \frac{1}{K} \sum_{k=1}^K NB(y_{gvi}| f(\theta_g^{(k)})).
+I_g \approx \frac{1}{K} \sum_{k=1}^K \prod_{v=1}^V \prod_{i=1}^{n_v}  NB(y_{gvi}| f(\theta_g^{(k)})).
 \]
@@
 \[
-I_g \approx \sum_{k=1}^K w_k NB(y_{gvi}| f(\theta_g^{(k)})).
+I_g \approx \frac{1}{K} \sum_{k=1}^K w_k \prod_{v=1}^V \prod_{i=1}^{n_v} NB(y_{gvi}| f(\theta_g^{(k)})).
 \]
-where $w_k = p(\theta^{(k)}|\pi)/p(\theta^{(k)}|\pi)$.
+where $w_k = p(\theta^{(k)}|\pi)/q(\theta^{(k)}|\pi)$.
 
@@
 
+\section{Expectation-Maximization}
+
+An EM algorithm to obtain MAP estimates for $\pi$ is 
+\begin{enumerate}
+\item Expectation
+\[ Q(\pi|\pi^{(t)}) = E_{\theta|y,\pi^{(t)}}[\log p(\theta|y)] \]
+\item Maximization
+\[ \pi^{(t+1)} = \argmax_{\pi} Q(\pi|\pi^{(t)})  \]
+\end{enumerate}
+
+Below, the expectation is always with respect to $p(\theta|y,\pi^{(t)})$.
+
+The expectation can be broken up into its constitutive components, i.e.
+
+\[
+E[\log p(\theta|y)]= E[C] + E[\log p(y|\theta)] + E[\log p(\theta|\pi)] + E[\log p(\pi)] 
+\]
+
+In the maximization step, we are maximizing with respect to $\pi$ and just $C + \log p(y|\theta)$ is irrelevant
+if we are obtaining MLEs (instead of MAPs) then just ignore $\log p(\pi)$ as well.
+
+\subsection{MLEs}
+
+First consider just $E[\log p(\theta|\pi)]$ where 
+\begin{equation}
+p(\theta|\pi) = \prod_{g=1}^G p(\theta_g|\pi) = \prod_{g=1}^G p(\phi_g|\mu_\phi, \sigma_\phi) p(\alpha_g|\mu_\alpha, \sigma_\alpha) p(\delta_g|\mu_\delta, \sigma_\delta) p(\psi_g|\mu_\psi, \sigma_\psi) \label{e:theta}
+\end{equation}
+
+Thus, the expectation is
+
+\begin{align*}
+E[\log p(\theta|\pi)]
+=& E_{\theta|y,\pi^{(t)}} \left[ \sum_{g=1}^G \log p(\phi_g|\mu_\phi, \sigma_\phi) +\log p(\alpha_g|\mu_\alpha, \sigma_\alpha) +\log p(\delta_g|\mu_\delta, \sigma_\delta) +\log p(\psi_g|\mu_\psi, \sigma_\psi) \right] \\
+=& E_{\phi|y,\pi^{(t)}} \left[ \sum_{g=1}^G \log p(\phi_g|\mu_\phi, \sigma_\phi) \right] +  E_{\alpha|y,\pi^{(t)}} \left[ \sum_{g=1}^G \log p(\alpha_g|\mu_\alpha, \sigma_\alpha) \right] \\
+&+  E_{\delta|y,\pi^{(t)}} \left[ \sum_{g=1}^G \log p(\delta_g|\mu_\delta, \sigma_\delta) \right] +  E_{\psi|y,\pi^{(t)}} \left[ \sum_{g=1}^G \log p(\psi_g|\mu_\psi, \sigma_\psi) \right]
+\end{align*}
+
+\subsubsection{Normal model}
+
+For simplicity, define the following 
+\[
+\hat{\mu}_{\phi,g}^{(t)} = E[\phi_g|y,\pi^{(t)}] \qquad \mbox{and} \qquad
+\hat{\sigma}_{\phi,g}^{2(t)} = V[\phi_g|y,\pi^{(t)}] 
+\]
+
+For a normal model, e.g. $\phi_g|\pi \stackrel{ind}{\sim} N(\mu_\phi,\sigma_\phi^2)$, we have 
+\[ 
+\begin{array}{rl}
+\multicolumn{2}{l}{
+E_{\phi|y,\pi^{(t)}} \left[ \sum_{g=1}^G \log p(\phi_g|\mu_\phi, \sigma_\phi) \right] = 
+E_{\phi|y,\pi^{(t)}} \left[ -\frac{G}{2}\log(2\pi \sigma_\phi^2) - \frac{1}{2\sigma_\phi^2} \sum_{g=1}^G (\phi_g-\mu_\phi)^2 \right]} \\
+&= -\frac{G}{2}\log(2\pi \sigma_\phi^2) - \frac{1}{2\sigma_\phi^2} \sum_{g=1}^G E\left[\left.\left(\phi_g-\mu_\phi\right)^2\right|y,\pi^{(t)}\right]\\
+&= -\frac{G}{2}\log(2\pi \sigma_\phi^2) - \frac{1}{2\sigma_\phi^2} \sum_{g=1}^G E\left[\left.\left(\phi_g-\hat{\mu}_{\phi,g}^{(t)}+\hat{\mu}_{\phi,g}^{(t)}-\mu_\phi\right)^2\right|y,\pi^{(t)}\right]\\
+&= -\frac{G}{2}\log(2\pi \sigma_\phi^2) - \frac{1}{2\sigma_\phi^2} \sum_{g=1}^G 
+E\left[\left.\left(\phi_g-\hat{\mu}_{\phi,g}^{(t)}\right)^2+2\left(\phi_g-\hat{\mu}_{\phi,g}^{(t)}\right)\left(\hat{\mu}_{\phi,g}^{(t)}-\mu_\phi\right)+\left(\hat{\mu}_{\phi,g}^{(t)}-\mu_\phi\right)^2\right|y,\pi^{(t)}\right]\\
+&= -\frac{G}{2}\log(2\pi \sigma_\phi^2) - \frac{1}{2\sigma_\phi^2} \sum_{g=1}^G \left[ \hat{\sigma}_{\phi,g}^{2(t)} + \left(\hat{\mu}_{\phi,g}^{(t)}-\mu_\phi\right)^2 \right]\\
+&= -\frac{G}{2}\log(2\pi \sigma_\phi^2) - \frac{1}{2\sigma_\phi^2} \sum_{g=1}^G \hat{\sigma}_{\phi,g}^{2(t)} - \frac{1}{2\sigma_\phi^2} \sum_{g=1}^G \left(\hat{\mu}_{\phi,g}^{(t)}-\mu_\phi\right)^2 
+\end{array}
+\]
+
+This is maximized when 
+\[ \mu_\phi^{(t+1)} = \frac{1}{G} \sum_{g=1}^G \hat{\mu}_{\phi,g}^{(t)} \qquad \mbox{and} \qquad 
+\sigma_\phi^{2(t+1)} = \frac{1}{G} \sum_{g=1}^G \left[ \hat{\sigma}_{\phi,g}^{2(t)} + \left(\hat{\mu}_{\phi,g}^{(t)}-\mu^{(t+1)}_\phi\right)^2 \right].
+\]
+
+\subsubsection{Laplace model}
+
+For simplicity, define the following 
+\[
+\hat{\mu}_{\alpha,g}^{(t)} = E[\alpha_g|y,\pi^{(t)}] \qquad \mbox{and} \qquad
+\hat{\sigma}_{\alpha,g}^{2(t)} = V[\alpha_g|y,\pi^{(t)}] 
+\]
+
+For a Laplace model, e.g. $\alpha_g|\pi \stackrel{ind}{\sim} La(\mu_\alpha,\sigma_\alpha)$, we have 
+\[ 
+\begin{array}{rl}
+\multicolumn{2}{l}{
+E_{\alpha|y,\pi^{(t)}} \left[ \sum_{g=1}^G \log p(\alpha_g|\mu_\alpha, \sigma_\alpha) \right] = 
+E_{\alpha|y,\pi^{(t)}} \left[ -G\log(2\sigma_\alpha) - \frac{1}{\sigma_\alpha} \sum_{g=1}^G |\alpha_g-\mu_\alpha| \right]} \\
+&= -G\log(2\sigma_\alpha) - \frac{1}{\sigma_\alpha} \sum_{g=1}^G E_{\alpha|y,\pi^{(t)}} \left[|\alpha_g-\mu_\alpha| \right] \\
+\end{array}
+\]
+Since the median minimizes absolute deviation and thus it is likely that 
+\[ 
+\mu_\alpha^{(t+1)} = \mbox{median}_g\, \hat{\mu}_{\alpha,g}^{(t)}
+\]
+This doesn't make complete sense since there i no reason that the posterior mean need to be used here, perhaps the posterior median? 
+or perhaps, by analogy,
+\[
+\mu_\alpha^{(t+1)} = \frac{1}{G} \sum_{g=1}^G \hat{\mu}_{\alpha,g}^{(t)} \qquad \mbox{and} \qquad
+\sigma_\alpha^{2(t+1)} = \frac{1}{G} \sum_{g=1}^G E\left[\left|\alpha_g-\hat{\mu}_{\alpha,g}^{(t)}\right|\right] + \left|\hat{\mu}_{\alpha,g}^{(t)}-\mu^{(t+1)}_\alpha\right| .
+\] 
+why the average?
+
 \end{document}

### 2015-04-01 sandbox/integration.tex
updated with intuition

--- 2015-03-09 sandbox/integration.tex
@@
 
+\section{Expectation-Maximization}
+
+An EM algorithm to obtain MAP estimates for $\pi$ is 
+\begin{enumerate}
+\item Expectation
+\[ Q(\pi|\pi^{(t)}) = E_{\theta|y,\pi^{(t)}}[\log p(\theta|y)] \]
+\item Maximization
+\[ \pi^{(t+1)} = \argmax_{\pi} Q(\pi|\pi^{(t)})  \]
+\end{enumerate}
+
+Below, the expectation is always with respect to $p(\theta|y,\pi^{(t)})$.
+
+The expectation can be broken up into its constitutive components, i.e.
+
+\[
+E[\log p(\theta|y)]= E[C] + E[\log p(y|\theta)] + E[\log p(\theta|\pi)] + E[\log p(\pi)] 
+\]
+
+In the maximization step, we are maximizing with respect to $\pi$ and just $C + \log p(y|\theta)$ is irrelevant
+if we are obtaining MLEs (instead of MAPs) then just ignore $\log p(\pi)$ as well.
+
+\subsection{MLEs}
+
+First consider just $E[\log p(\theta|\pi)]$ where 
+\begin{equation}
+p(\theta|\pi) = \prod_{g=1}^G p(\theta_g|\pi) = \prod_{g=1}^G p(\phi_g|\mu_\phi, \sigma_\phi) p(\alpha_g|\mu_\alpha, \sigma_\alpha) p(\delta_g|\mu_\delta, \sigma_\delta) p(\psi_g|\mu_\psi, \sigma_\psi) \label{e:theta}
+\end{equation}
+
+Thus, the expectation is
+
+\begin{align*}
+E[\log p(\theta|\pi)]
+=& E_{\theta|y,\pi^{(t)}} \left[ \sum_{g=1}^G \log p(\phi_g|\mu_\phi, \sigma_\phi) +\log p(\alpha_g|\mu_\alpha, \sigma_\alpha) +\log p(\delta_g|\mu_\delta, \sigma_\delta) +\log p(\psi_g|\mu_\psi, \sigma_\psi) \right] \\
+=& E_{\phi|y,\pi^{(t)}} \left[ \sum_{g=1}^G \log p(\phi_g|\mu_\phi, \sigma_\phi) \right] +  E_{\alpha|y,\pi^{(t)}} \left[ \sum_{g=1}^G \log p(\alpha_g|\mu_\alpha, \sigma_\alpha) \right] \\
+&+  E_{\delta|y,\pi^{(t)}} \left[ \sum_{g=1}^G \log p(\delta_g|\mu_\delta, \sigma_\delta) \right] +  E_{\psi|y,\pi^{(t)}} \left[ \sum_{g=1}^G \log p(\psi_g|\mu_\psi, \sigma_\psi) \right]
+\end{align*}
+
+\subsubsection{Normal model}
+
+For simplicity, define the following 
+\[
+\hat{\phi}_{g}^{(t)} = E[\phi_g|y,\pi^{(t)}] \qquad \mbox{and} \qquad
+\hat{\sigma}_{\phi_g}^{2(t)} = V[\phi_g|y,\pi^{(t)}] 
+\]
+
+For a normal model, e.g. $\phi_g|\pi \stackrel{ind}{\sim} N(\mu_\phi,\sigma_\phi^2)$, we have 
+\[ 
+\begin{array}{rl}
+\multicolumn{2}{l}{
+E_{\phi|y,\pi^{(t)}} \left[ \sum_{g=1}^G \log p(\phi_g|\mu_\phi, \sigma_\phi) \right] = 
+E_{\phi|y,\pi^{(t)}} \left[ -\frac{G}{2}\log(2\pi \sigma_\phi^2) - \frac{1}{2\sigma_\phi^2} \sum_{g=1}^G (\phi_g-\mu_\phi)^2 \right]} \\
+&= -\frac{G}{2}\log(2\pi \sigma_\phi^2) - \frac{1}{2\sigma_\phi^2} \sum_{g=1}^G E\left[\left(\phi_g-\mu_\phi\right)^2\right]\\
+&= -\frac{G}{2}\log(2\pi \sigma_\phi^2) - \frac{1}{2\sigma_\phi^2} \sum_{g=1}^G E\left[\left(\phi_g-\hat{\phi}_{g}^{(t)}+\hat{\phi}_{g}^{(t)}-\mu_\phi\right)^2\right]\\
+&= -\frac{G}{2}\log(2\pi \sigma_\phi^2) - \frac{1}{2\sigma_\phi^2} \sum_{g=1}^G 
+E\left[\left(\phi_g-\hat{\phi}_{g}^{(t)}\right)^2+2\left(\phi_g-\hat{\phi}_{g}^{(t)}\right)\left(\hat{\phi}_{g}^{(t)}-\mu_\phi\right)+\left(\hat{\phi}_{g}^{(t)}-\mu_\phi\right)^2\right]\\
+&= -\frac{G}{2}\log(2\pi \sigma_\phi^2) - \frac{1}{2\sigma_\phi^2} \sum_{g=1}^G \left[ \hat{\sigma}_{\phi_g}^{2(t)} + \left(\hat{\phi}_{g}^{(t)}-\mu_\phi\right)^2 \right]\\
+&= -\frac{G}{2}\log(2\pi \sigma_\phi^2) - \frac{1}{2\sigma_\phi^2} \sum_{g=1}^G \hat{\sigma}_{\phi_g}^{2(t)} - \frac{1}{2\sigma_\phi^2} \sum_{g=1}^G \left(\hat{\phi}_{g}^{(t)}-\mu_\phi\right)^2 
+\end{array}
+\]
+
+This is maximized when 
+\[ \mu_\phi^{(t+1)} = \frac{1}{G} \sum_{g=1}^G \hat{\phi}_{g}^{(t)} \]
+and
+\[ 
+\sigma_\phi^{2(t+1)} 
+= \frac{1}{G} \sum_{g=1}^G E_{\phi|y,\pi^{(t)}}\left[\left.\left(\phi_g-\mu_\phi^{(t+1)}\right)^2\right|y,\pi^{(t)}\right] 
+= \frac{1}{G} \sum_{g=1}^G \left[ \hat{\sigma}_{\phi_g}^{2(t)} + \left(\hat{\phi}_{g}^{(t)}-\mu^{(t+1)}_\phi\right)^2 \right].
+\]
+
+\subsubsection{Laplace model}
+
+For simplicity, define $\hat{\alpha}_{g}^{(t)}$ as a point estimate (median?) for $\alpha_g$ based on $\pi^{(t)}$.
+% \[ \hat{\alpha}_{g}^{(t)} = E[\alpha_g|y,\pi^{(t)}] \qquad \mbox{and} \qquad
+% \hat{\sigma}_{\alpha,g}^{2(t)} = V[\alpha_g|y,\pi^{(t)}]  \]
+
+For a Laplace model, e.g. $\alpha_g|\pi \stackrel{ind}{\sim} La(\mu_\alpha,\sigma_\alpha)$, we have 
+\[ 
+\begin{array}{rl}
+\multicolumn{2}{l}{
+E_{\alpha|y,\pi^{(t)}} \left[ \sum_{g=1}^G \log p(\alpha_g|\mu_\alpha, \sigma_\alpha) \right] = 
+E_{\alpha|y,\pi^{(t)}} \left[ -G\log(2\sigma_\alpha) - \frac{1}{\sigma_\alpha} \sum_{g=1}^G |\alpha_g-\mu_\alpha| \right]} \\
+&= -G\log(2\sigma_\alpha) - \frac{1}{\sigma_\alpha} \sum_{g=1}^G E \left[|\alpha_g-\mu_\alpha| \right] \\
+&\ge -G\log(2\sigma_\alpha) - \frac{1}{\sigma_\alpha} \sum_{g=1}^G E\left[\left|\alpha_g-\hat{\alpha}_{g}^{(t)}\right|\right] + \left|\hat{\alpha}_{g}^{(t)}-\mu_\alpha\right] \\
+\end{array}
+\]
+Since the median minimizes absolute deviation and thus it is likely that 
+\[ 
+\mu_\alpha^{(t+1)} = \mbox{median}_g\, \hat{\alpha}_{g}^{(t)}
+\]
+Now the choice for what to use for $\hat{\alpha}_{g}^{(t)}$ appears pretty arbitrary.
+Perhaps we should use the posterior median since it will minimize $E\left[\left|\alpha_g-\hat{\alpha}_{g}^{(t)}\right|\right]$.
+
+\[
+\mu_\alpha^{(t+1)} = \frac{1}{G} \sum_{g=1}^G \hat{\alpha}_{g}^{(t)} \]
+and 
+\[
+\sigma_\alpha^{2(t+1)} 
+= \frac{1}{G} \sum_{g=1}^G E_{\alpha|y,\pi^{(t)}}\left[\left.\left|\alpha_g-\mu_\alpha^{(t+1)}\right|\right|y,\pi^{(t)}\right] 
+\le  \frac{1}{G} \sum_{g=1}^G E_{\alpha|y,\pi^{(t)}}\left[\left|\alpha_g-\hat{\alpha}_{g}^{(t)}\right|\right] + \left|\hat{\alpha}_{g}^{(t)}-\mu^{(t+1)}_\alpha\right| .
+\] 
+We will probably just need to use a Monte Carlo approximation to this expectation.
+
 \end{document}

### 2015-04-03 sandbox/integration.tex
em

@@
 &= -G\log(2\sigma_\alpha) - \frac{1}{\sigma_\alpha} \sum_{g=1}^G E \left[|\alpha_g-\mu_\alpha| \right] \\
-&\ge -G\log(2\sigma_\alpha) - \frac{1}{\sigma_\alpha} \sum_{g=1}^G E\left[\left|\alpha_g-\hat{\alpha}_{g}^{(t)}\right|\right] + \left|\hat{\alpha}_{g}^{(t)}-\mu_\alpha\right] \\
+&\ge -G\log(2\sigma_\alpha) - \frac{1}{\sigma_\alpha} \sum_{g=1}^G E\left[\left|\alpha_g-\hat{\alpha}_{g}^{(t)}\right|\right] + \left|\hat{\alpha}_{g}^{(t)}-\mu_\alpha\right| \\
 \end{array}
 \]
-Since the median minimizes absolute deviation and thus it is likely that 
+To maximize this final quantity, we take $\hat{\alpha}_{g}^{(t)}$ to be the posterior median and 
 \[ 
-\mu_\alpha^{(t+1)} = \mbox{median}_g\, \hat{\alpha}_{g}^{(t)}
+\mu_\alpha^{(t+1)} = \mbox{median}_g\, \hat{\alpha}_{g}^{(t)}.
 \]
-Now the choice for what to use for $\hat{\alpha}_{g}^{(t)}$ appears pretty arbitrary.
-Perhaps we should use the posterior median since it will minimize $E\left[\left|\alpha_g-\hat{\alpha}_{g}^{(t)}\right|\right]$.
-
-\[
-\mu_\alpha^{(t+1)} = \frac{1}{G} \sum_{g=1}^G \hat{\alpha}_{g}^{(t)} \]
-and 
+This maximization provides a lower bound on the quantity that we actually want to maximize.
+For whatever value of $\mu_\alpha^{(t+1)}$ is chosen (ideally we choose the one that provides the maximum), then
 \[
@@
 = \frac{1}{G} \sum_{g=1}^G E_{\alpha|y,\pi^{(t)}}\left[\left.\left|\alpha_g-\mu_\alpha^{(t+1)}\right|\right|y,\pi^{(t)}\right] 
-\le  \frac{1}{G} \sum_{g=1}^G E_{\alpha|y,\pi^{(t)}}\left[\left|\alpha_g-\hat{\alpha}_{g}^{(t)}\right|\right] + \left|\hat{\alpha}_{g}^{(t)}-\mu^{(t+1)}_\alpha\right| .
+%\le  \frac{1}{G} \sum_{g=1}^G E_{\alpha|y,\pi^{(t)}}\left[\left|\alpha_g-\hat{\alpha}_{g}^{(t)}\right|\right] + \left|\hat{\alpha}_{g}^{(t)}-\mu^{(t+1)}_\alpha\right| .
 \] 
-We will probably just need to use a Monte Carlo approximation to this expectation.
+We will probably just need to use a Monte Carlo (MCMC) approximation to this expectation.
 

### 2015-04-03 sandbox/em.R

@@ -1,0 +1,233 @@
+library(reshape2)
+library(ggplot2)
+library(plyr)
+library(lme4)
+library(distr)
+library(rstan)
+
+# Compile Stan models
+normal_em = stan_model("normal_em.stan")
+laplace_em = stan_model("laplace_em.stan")
+nb_normal = stan_model("nb_normal.stan")
+
+
+####################################################
+# Normal random effects example
+####################################################
+set.seed(1)
+
+G = 1000
+n = 10
+
+truth = data.frame(mu=0, tau=1, sigma=1)
+
+d = ddply(data.frame(gene=1:G), .(gene), function(x) {
+  theta = rnorm(1, truth$mu, truth$tau)
+  data.frame(theta = theta, y = rnorm(n, theta, truth$sigma))
+})
+
+sm = ddply(d, .(gene), summarize, mean=mean(y), sd=sd(y))
+
+# Truth
+o = lmer(y~(1|gene), d, REML=FALSE)
+(truth = data.frame(parameter=c("mu","tau","sigma"), value=c(fixef(o), as.data.frame(VarCorr(o))$sdcor)))
+# alternatively
+mean(d$y)           # == truth[1]
+
+sqrt(mean(sm$sd^2)) # == truth[3]
+
+# EM
+mu = 0
+tau = 1
+sigma = truth$value[truth$parameter=='sigma'] # only writing the EM for mu and tau
+
+n_iter = 10
+mu_em = rep(mu,10)
+tau_em= rep(tau,10)
+
+for (i in 2:n_iter) {
+  theta_var  = 1/(n/sigma^2+1/tau^2)
+  theta_mean = theta_var*(n*sm$mean/sigma^2+mu/tau^2)
+    
+  # Maximization
+  mu = mean(theta_mean)
+  tau = sqrt(mean(theta_var + (theta_mean-mu)^2)) # add the variance and deviation from the mean
+  
+  # Save values
+  mu_em[i] = mu
+  tau_em[i] = tau
+}
+
+em = data.frame(iteration=1:length(mu_em),
+                mu = mu_em,
+                tau = tau_em,
+                method = "em")
+
+#######################################
+# MCEM
+#######################################
+# Initial values
+mu =0 
+tau = 1
+
+n_iter = 10
+mu_em_stan  = rep(mu,10)
+tau_em_stan = rep(tau,10)
+
[…]

### 2015-04-03 sandbox/em_full.R

@@ -1,0 +1,86 @@
+library(rstan)
+library(reshape2)
+library(doMC)
+library(dplyr)
+library(ggplot2)
+
+source("Laplace.R")
+source("sim_heterosis_data.R")
+
+registerDoMC()
+
+m = stan_model("sg_model_pad_dexp.txt")
+
+d = sim_heterosis_data(G=100)
+
+
+# Set initial values for hyperparameters
+gene_names = c("phi","alpha","delta","psi")
+hyper = with(d$hyperparameters, as.list(c(location, scale)))
+names(hyper) = c(paste("eta_",   gene_names, sep=''),
+                 paste("sigma_", gene_names, sep=''))
+n_hyper = 8
+
+truth = data.frame(variable=names(hyper), value=with(d$hyperparameter, c(location,scale)))
+
+# 
+n_iter = 50
+hyper_keep = matrix(NA, nrow=n_iter, ncol=n_hyper)
+
+# Run EM
+for (i in 1:n_iter) {
+  cat(i, unlist(hyper), "\n")
+  
+  # MCMC for gene specific parameters
+  mcmc = dlply(d$data, .(gene), function(x) {
+    sampling(m, c(list(S=nrow(x), variety=x$parent, count=x$y, c=rep(1, nrow(x))), hyper), gene_names, iter=100+(i+2)^2, chains=1)
+  }, .parallel=TRUE)
+  for(j in 1:length(mcmc)) attr(mcmc[[j]],"name") <- names(mcmc)[j]
+
+  # Get initial values (for next step)
+  
+  
+  # Extract samples to calculate M step
+  samps = ldply(mcmc, function(x) {
+    tmp = extract(x, gene_names)
+    data.frame(gene = attr(x, 'name'), phi = tmp$phi, alpha = tmp$alpha, delta = tmp$delta, psi = tmp$psi)
+  })
+  
+  # Calculate summary statistics
+  sm = ddply(samps, .(gene), summarize,
+             phi_mean = mean(phi),
+             phi_var  = var(phi),
+             psi_mean = mean(psi),
+             psi_var  = var(psi),
+             alpha_median = median(alpha),
+             delta_median = median(delta), 
+             .parallel=TRUE)
+
+  # Update hyperparameters
+  hyper$eta_phi = mean(sm$phi_mean)
[…]

### 2015-04-03 sandbox/sim_heterosis_data.R

@@ -19,7 +19,7 @@ sim_heterosis_data = function(G=10, nv=4, parameters=NULL, distributions=NULL) {
   if (is.null(parameters)) 
     hyperparameters = data.frame(parameter      = c("phi","alpha","delta","psi"),
                                  location = c(4.6,0,0,-2),
-                                 scale    = c(1.8,.1,.01,.1))
+                                 scale    = c(1.8,.1,.1,.1))
   
   # Simulate gene-specific parameters
   if (is.null(distributions)) 

### 2015-04-03 sandbox/find_map_estimates.R
not much

@@ -1,4 +1,4 @@
-library(Rcpp)
+ library(Rcpp)
 library(reshape2)
 library(plyr)
 library(ggplot2)

### 2015-04-03 sandbox/get_last.R

@@ -1,0 +1,6 @@
+get_last = function(stanfit) {
+  # Takes a stanfit object and returns the last samples in a format
+  # that can be used by the 'init' argument in the 'sampling' function
+  
+  
+}

### 2015-04-03 sandbox/em_full.R
save only hyper

@@ -11,7 +11,7 @@ registerDoMC()
 
 m = stan_model("sg_model_pad_dexp.txt")
 
-d = sim_heterosis_data(G=100)
+d = sim_heterosis_data(G=1000)
 
 
 # Set initial values for hyperparameters
@@ -24,7 +24,7 @@ n_hyper = 8
 truth = data.frame(variable=names(hyper), value=with(d$hyperparameter, c(location,scale)))
 
 # 
-n_iter = 50
+n_iter = 100
 hyper_keep = matrix(NA, nrow=n_iter, ncol=n_hyper)
 
 # Run EM
@@ -72,6 +72,8 @@ for (i in 1:n_iter) {
   
   # Save hyperparameters
   for (j in 1:n_hyper) hyper_keep[i,j] = hyper[[j]]
+  
+  save(hyper, "em_full.RData")
 }
 
 # Plot hyperparameters

### 2015-04-03 sandbox/em_full.R
set cores to 8

@@ -1,13 +1,14 @@
 library(rstan)
 library(reshape2)
 library(doMC)
-library(dplyr)
+#library(dplyr)
 library(ggplot2)
+library(plyr)
 
 source("Laplace.R")
 source("sim_heterosis_data.R")
 
-registerDoMC()
+registerDoMC(cores=8)
 
 m = stan_model("sg_model_pad_dexp.txt")
 

### 2015-04-03 sandbox/em_full.R
direct output to file

@@ -12,7 +12,8 @@ registerDoMC(cores=8)
 
 m = stan_model("sg_model_pad_dexp.txt")
 
-d = sim_heterosis_data(G=1000)
+G = 1000
+d = sim_heterosis_data(G=G)
 
 
 # Set initial values for hyperparameters
@@ -28,9 +29,10 @@ truth = data.frame(variable=names(hyper), value=with(d$hyperparameter, c(locatio
 n_iter = 100
 hyper_keep = matrix(NA, nrow=n_iter, ncol=n_hyper)
 
+cat("MCEM with", G, "genes.\n", file="em_full.Ro")
 # Run EM
 for (i in 1:n_iter) {
-  cat(i, unlist(hyper), "\n")
+  start_time = proc.time()
   
   # MCMC for gene specific parameters
   mcmc = dlply(d$data, .(gene), function(x) {
@@ -74,7 +76,10 @@ for (i in 1:n_iter) {
   # Save hyperparameters
   for (j in 1:n_hyper) hyper_keep[i,j] = hyper[[j]]
   
-  save(hyper, "em_full.RData")
+  save(hyper, file="em_full.RData")
+  
+  cat(i, unlist(hyper), "time=", proc.time()-start_time, "\n",
+      file="em_full.Ro", append=TRUE)
 }
 
 # Plot hyperparameters

### 2015-05-05 sandbox/em.R
testing em_de_poisson: not really converging?

@@ -11,20 +11,24 @@ models$het_negbin  = stan_model('stan_models/het_negbin.stan')
 
 # calculates hyper parameters for normal distribution
 normal_hyper = function(samples, learn_mean) {
-  eta = ifelse(learn_mean, mean(samples, 0))
+  eta = ifelse(learn_mean, mean(samples), 0)
   sigma = sqrt(mean((samples-eta)^2))
   return(list(eta=eta,sigma=sigma))
 }
 
 # calculates hyper parameters for laplace distribution
 laplace_hyper = function(samples, learn_mean) {
-  eta = ifelse(learn_mean, median(samples,0))
+  eta = ifelse(learn_mean, median(samples), 0)
   sigma = mean(abs(samples-eta))
   return(list(eta=eta,sigma=sigma))
 }
 
-# MCEM algorithm for 
-em_de_poisson = function(dat, niter, learn_alpha_mean = FALSE, alpha_dist = 'normal') {
+
+
+
+# MCEM algorithm for differential expression using Poisson model
+em_de_poisson = function(dat, niter, learn_alpha_mean = FALSE, dist = list(alpha = 'normal'),
+                         .parallel=FALSE, verbose=1) {
   parameter_names = c("phi","alpha")
   
   # Build X matrix
@@ -40,14 +44,14 @@ em_de_poisson = function(dat, niter, learn_alpha_mean = FALSE, alpha_dist = 'nor
   hyper$iteration = 0
   hyper$eta_phi = mean(initial$phi)
   hyper$sigma_phi = sd(initial$phi)
-  if (alpha_dist == 'normal') {
-    hyper$eta_alpha = ifelse(learn_eta_alpha, mean(initial$alpha), 0)
+  if (dist$alpha == 'normal') {
+    hyper$eta_alpha = ifelse(learn_alpha_mean, mean(initial$alpha), 0)
     hyper$sigma_alpha = sqrt(mean((initial$alpha-hyper$eta_alpha)^2))
-  } else if (alpha_dist == 'laplace') {
-    hyper$eta_alpha = ifelse(learn_eta_alpha, median(initial$alpha), 0)
+  } else if (dist$alpha == 'laplace') {
+    hyper$eta_alpha = ifelse(learn_alpha_mean, median(initial$alpha), 0)
     hyper$sigma_alpha = mean(abs(initial$alpha-hyper$eta_alpha))
   } else {
-    stop(paste('Alpha distribution',alpha_dist,'not implemented.'))
[…]

### 2015-05-05 sandbox/run_em.R

@@ -6,11 +6,12 @@ registerDoMC(4)
 
 source("Laplace.R")
 source("sim_heterosis_data.R")
+source("em.R")
 
 truth = sim_heterosis_data(10)
-d = subset(d$data, variety != 3)
+d = subset(truth$data, variety != 3)
 
-tmp = em_de_poisson(d, 100)
+tmp = em_de_poisson(d, 10, .parallel=TRUE)
 
 ggplot(melt(tmp, id.var='iteration', variable.name='parameter'), aes(x=iteration, y=value)) + 
   geom_point() +

### 2015-05-05 slides/KState/JaradNiemi_KState.Rnw
KState talk

\documentclass{beamer}

\usecolortheme[RGB={0,0,144}]{structure}
\usetheme{AnnArbor}\usecolortheme{beaver}

\usepackage{verbatim,xmpmulti,color,multicol,multirow}
\setlength{\unitlength}{\textwidth}  % measure in textwidths
\usepackage[normalem]{ulem}

\graphicspath{{../figs/}}

%\usepackage{beamerthemesplit}
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{enumerate items}[default]
\setbeamertemplate{enumerate subitem}{\alph{enumii}.}
\setbeamertemplate{enumerate subsubitem}{\roman{enumiii}.}
\setkeys{Gin}{width=0.6\textwidth}

\title[EBayes for heterosis of RNAseq]{Empirical Bayes Analysis for Detection of Gene Heterosis in RNAseq Data}
\author[Jarad Niemi]{Dr. Jarad Niemi \\ joint with Eric Mittman, Will Landau, and Dr. Dan Nettleton}
\institute[Iowa State]{Iowa State University}
\date{\today}

\begin{document}

%\section{Temp??} \begin{comment}

<<chunk_options, echo=FALSE, message=FALSE>>=
opts_chunk$set(fig.width=6, 
               fig.height=5, 
               out.width='.8\\linewidth', 
               fig.align='center', 
               size='tiny',
               echo=FALSE)
library(ggplot2)
@

\frame{\titlepage}

\begin{frame}
\frametitle{Outline}

\begin{itemize}
\item Data
  \begin{itemize}
  \item Phenotypic heterosis
  \item RNAseq
  \end{itemize}
\item Methodology
  \begin{itemize}
  \item Model
  \item Empircal Bayes
  \item Stan
  \end{itemize}
\item Results
  \begin{itemize}
  \item Simulation study vs {\tt edgeR}, {\tt baySeq}, {\tt ShrinkBayes}
  \item Real data analysis
  \end{itemize}
\end{itemize}

\end{frame}

\section{Data}
\subsection{Phenotypic heterosis}

\begin{frame}
\frametitle{Phenotypic heterosis in maize}
\setkeys{Gin}{width=0.8\textwidth}
\begin{center}
\includegraphics{heterosis_corn}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Phenotypic heterosis in cattle}
\setkeys{Gin}{width=0.8\textwidth}

\vspace{-0.1in}

\begin{center}
\includegraphics{heterosis_cattle1}

\includegraphics{heterosis_cattle2}
\end{center}
\end{frame}

\subsection{RNAseq experiment}
\begin{frame}
\frametitle{RNAseq experiment}
\setkeys{Gin}{width=0.55\textwidth}

\begin{center}
\includegraphics{rnaseq_experiment}
\end{center}

\end{frame}

\begin{frame}
\frametitle{Phenotypic heterosis in RNAseq}
\setkeys{Gin}{width=0.8\textwidth}
\begin{center}
\includegraphics{heterosis_rnaseq}
\end{center}
\end{frame}

\section{Method}
\subsection{Model}

\begin{frame}
\frametitle{RNAseq model}

Let $Y_{gvi}$ be the count for 
\begin{itemize}
\item gene $g=1,\ldots,G$, 
\item variety $v=1,\ldots,V$, and 
\item replicate $i=1,\ldots,n_v$. 
\end{itemize}

\vspace{0.2in} \pause

We assume
\[
Y_{gvi} \stackrel{ind}{\sim} NB\left(e^{\mu_{gv}+\gamma_{vi}},e^{\psi_g}\right) 
\]
\pause
where 
\begin{itemize}
\item $E[Y_{gvi}] = \zeta_{gvi} = e^{\mu_{gv}+\gamma_{vi}}$, \pause 
\item $V[Y_{gvi}] = \zeta_{gvi}+e^{\psi_g}\zeta_{gvi}^2$, \pause and 
\item $\gamma_{vi}$ controls sequencing depth (normalization factors).
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Reparameterization}
\setkeys{Gin}{width=\textwidth}

Convert the variety specific parameters for parent 1 ($\mu_{g1}$), parent 2 ($\mu_{g2}$), and the hybrid ($\mu_{g3}$) \pause to

\begin{itemize}[<+->]
\item Parental average: $\phi_g = (\mu_{g1}+\mu_{g2})/2$
\item Half-parental difference: $\alpha_g = (\mu_{g2}-\mu_{g1})/2$
\item Hybrid effect: $\delta_g = \mu_{g3}-\phi_g$
\end{itemize}

\vspace{0.2in} \pause

and thus 
\[ \begin{array}{rl}
\mu_{g1} &= \phi_g + \alpha_g \\
\mu_{g2} &= \phi_g - \alpha_g \\
\mu_{g3} &= \phi_g + \delta_g
\end{array}\]

\pause
\begin{center}
\includegraphics{pad}
\end{center}

\end{frame}

\begin{frame}
\frametitle{Gene heterosis}
\setkeys{Gin}{width=\textwidth}

\begin{center}
\includegraphics{pad}
\end{center}
\pause
We define \alert{gene heterosis}:

\[ \begin{array}{rll}
H_{g,HPH}: & \delta_g > \phantom{-}|\alpha_g|  & \mbox{high parent heterosis} \pause \\
H_{g,LPH}: & \delta_g < -|\alpha_g|  & \mbox{low parent heterosis} \pause \\
H_{g0}: & |\delta_g| \le  |\alpha_g| 
\end{array} \]

\pause

We are interested in identifying genes with high posterior probabilities for LPH and HPH, i.e. 
\begin{itemize}
\item $P(H_{g,HPH}|y) = P(\delta_g > \phantom{-}|\alpha_g|\, |y)$ and 
\item $P(H_{g,LPH}|y) = P(\delta_g < -|\alpha_g|\, |y)$. 
\end{itemize}

\end{frame}

\begin{frame}[fragile]
\frametitle{Hierarchical model}
\setkeys{Gin}{width=\textwidth}
\begin{itemize}
\item Data model:
\[ Y_{gvi} \stackrel{ind}{\sim} NB\left(e^{\mu_{gv}+\gamma_{vi}},e^{\psi_g}\right) \]
\[ \mu_{g1} = \phi_g + \alpha_g, \mu_{g2} = \phi_g-\alpha_g, \mu_{g3} = \phi_g + \delta_g \]

\pause

\item Hierarchical structure:
\begin{columns}
\begin{column}{0.5\linewidth}
\[ \begin{array}{rl}
\phi_g &\stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2) \\
\alpha_g &\stackrel{ind}{\sim} La(\eta_\alpha, \sigma_\alpha) \\
\delta_g &\stackrel{ind}{\sim} La(\eta_\delta, \sigma_\delta) \\
\psi_g &\stackrel{ind}{\sim} N(\eta_\psi, \sigma_\psi^2) 
\end{array} \]
\end{column} \pause
\begin{column}{0.5\linewidth}
<<>>=
source("../../sandbox/Laplace.R")
x = seq(-4,4,by=0.1)
d = data.frame(rbind(data.frame(density="normal", x=x, y=dnorm(x)),
                     data.frame(density="Laplace", x=x, y=dlaplace(x,0,1/sqrt(2)))))
ggplot(d, aes(x=x,y=y,color=density)) + 
  geom_line(size=2) + 
  labs(y="f(x)")
@
\end{column}
\end{columns}
\end{itemize}
\end{frame}

% \begin{frame}[fragile]
% \frametitle{Normal vs Laplace}
% <<>>=
% source("../../sandbox/Laplace.R")
% x = seq(-4,4,by=0.1)
% d = data.frame(rbind(data.frame(density="normal", x=x, y=dnorm(x)),
%                      data.frame(density="Laplace", x=x, y=dlaplace(x,0,sqrt(2)))))
% ggplot(d, aes(x=x,y=y,color=density)) + 
%   geom_line() + 
%   labs(y="f(x)")
% @
% \end{frame}

\subsection{Empirical Bayes}
\begin{frame}
\frametitle{Empirical Bayes}

\begin{enumerate}
\item Estimate hyperparameter, $\pi=(\eta,\sigma,\gamma)$:

  \begin{enumerate}
  \item Use {\tt edgeR} to obtain $\hat{\gamma}$ and $\hat{\theta}_g = (\hat{\phi}_g,\hat{\alpha}_g,\hat{\delta}_g,\hat{\psi}_g)$.
  \item Use moment matching, to obtain $\hat{\eta}$ and $\hat{\sigma}$.
  \end{enumerate}

\pause
\item Empirical Bayes posterior

\[ p(\theta|y, \hat{\pi}) = \prod_{g=1}^G p(\theta_g|y_g,\hat{\pi}) \]
where $y_g$ is all observations from gene $g$ \pause and 
{\footnotesize
\[ p(\theta_g|y_g,\hat{\pi}) \propto \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi};e^{\mu_{gv}+\hat{\gamma}_{vi}}, e^{\psi_g}\right) \right] p\left(\theta_g;\hat{\eta},\hat{\sigma}\right) \]
}
\end{enumerate}
\end{frame}

\subsection{Stan}
\begin{frame}
\frametitle{Stan}
\setkeys{Gin}{width=1\textwidth}

\begin{center}
\includegraphics{stan}
\end{center}

For more information, go to \url{http://mc-stan.org/}.

\end{frame}

\begin{frame}[containsverbatim]
\frametitle{}

{\tiny
\begin{verbatim}
data {
  int<lower=1> S;
  int<lower=0> count[S]; 
  int<lower=1,upper=3> variety[S];
  real   eta_phi; real   eta_alpha; real   eta_delta; real   eta_psi; 
  real sigma_phi; real sigma_alpha; real sigma_delta; real sigma_psi;
  vector[S] gamma;                     // lane sequencing depth
}
transformed data {
  matrix[S,3] X;
  for (s in 1:S) {
    if (variety[s] == 1) { X[s,1] <-  1; X[s,2] <- -1; X[s,3] <-  0; }
    if (variety[s] == 2) { X[s,1] <-  1; X[s,2] <-  1; X[s,3] <-  0; }
    if (variety[s] == 3) { X[s,1] <-  1; X[s,2] <-  0; X[s,3] <-  1; }
  }
}
parameters {
  real phi; real alpha; real delta; real psi;          
}
transformed parameters {
  vector[3] pad;

  pad[1] <- phi;
  pad[2] <- alpha;
  pad[3] <- delta;
}
model {
  phi   ~ normal(            eta_phi,   sigma_phi);
  alpha ~ double_exponential(eta_alpha, sigma_alpha); // Laplace
  delta ~ double_exponential(eta_delta, sigma_delta); // Laplace
  psi   ~ normal(            eta_psi,   sigma_psi);

  count ~ neg_binomial_2_log(X*pad+gamma, 1/exp(psi));
}
\end{verbatim}
}

\end{frame}

\begin{frame}[fragile]
\frametitle{RStan}

<<echo=TRUE, eval=FALSE>>=
# set up data
# ...
# ...

# get hyperparameter estimates
hyperparameter = get_indep_est(counts.t,group,geneid.t)

# change data to long format
data_stan = w.to.l(counts.t,group)

# compile Stan model (this can take some time, but only needs to be done once)
sg_model = stan_model("sg_model.txt")

# use initial values to obtain hyperparameter estimates
# ...
# ...

# loop over all genes
posterior = dlply(stan_data, .(gene), function(x) {
  dat = list(x, hyperparameter)
  sampling(sg_model, dat, chains=4, iter=2000)
}, .parallel=TRUE)
@

\end{frame}

\section{Results}
\subsection{Simulation study}
\begin{frame}
\frametitle{Simulation study}

Compared our method to {\tt edgeR}, {\tt baySeq}, {\tt ShrinkBayes} (not fair)

\vspace{0.2in} \pause

Simulation study:
\begin{enumerate}[<+->]
\item Used {\tt edgeR} to estimate $\theta_g$ and $\gamma$ (about 30\% have heterosis)
\item Simulated data from a negative binomial model using the estimated $\theta_g$ and $\gamma$
\item Used each method to rank genes according to plausibility of heterosis
\item Compared methods based on receiving operating characteristic (ROC) curves
\end{enumerate}
\end{frame}

\subsection{ROC curves}
\begin{frame}
\frametitle{Example ROC curve}
\setkeys{Gin}{width=\textwidth}
\begin{center}
\includegraphics{exampleROC0_1}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Area under ROC curve}
\setkeys{Gin}{width=\textwidth}
\begin{center}
\includegraphics{auc-facet-TRUE}
\end{center}
\end{frame}

\subsection{Real data analysis}
\begin{frame}
\frametitle{}
{\tiny
\input{../../include/map-laplace}
}
\end{frame}

\begin{frame}
\frametitle{Volcano plot for real data}
\setkeys{Gin}{width=0.6\textwidth}
\begin{center}
\includegraphics{maize_volcano}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Current work}

\begin{itemize}[<+->]
%\item Monte Carlo Expectation Maximization algorithm to obtain improved hyperparameter ($\pi$) estimates.
\item Session P03 - 3:50pm Monday \\
Eric Mittman \\
Hierarchical Bayesian Modeling of RNA-seq Data with Stan 
\item Session P04 - 10am Tuesday \\
William Landau \\
A Hierarchical Modeling Strategy for Identifying Gene Expression Heterosis using Parallel Computing with Graphics Processing Units (GPUs) 
\end{itemize}

\vspace{0.2in} \pause

{\small
This research was supported by National Institute of General Medical Sciences (NIGMS) of the National Institutes of Health and the joint National Science Foundation / NIGMS Mathematical Biology Program under award number R01GM109458. The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health or the National Science Foundation.
}
\end{frame}

\end{document}

### 2015-05-13 ebayes_rnaseq_heterosis.tex
starting to update for eBayes

@@
 \newcommand{\RNAseq}{RNAseq}
+
+\renewcommand{\gamma}{c}
 
@@
 
-We present a hierarchical model for the normalization factors, overdispersions, and gene-specific mean parameters.
-The normalization factors, parental averages, and overdispersion parameters are assumed to follow a normal distribution, i.e.
+We present a hierarchical model for the gene-specific mean parameters and overdispersion parameters.
+The parental averages and overdispersion parameters are assumed to follow a normal distribution, i.e.
 \[ 
-\gamma_{vi} \stackrel{ind}{\sim} N(0,\sigma_\gamma^2), \quad 
+%\gamma_{vi} \stackrel{ind}{\sim} N(0,\sigma_\gamma^2), \quad 
 \phi_g \stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2), \quad \mbox{and}\quad 
@@
 \]
-where $\alpha\sim La(\eta,\sigma)$ has a probability density function given by $p(\alpha|\eta,\sigma) = \exp(-|\alpha-\eta|/\sigma)/2\sigma$, expectation $\eta$, and variance $2\sigma^2$.
+where $\alpha\sim La(\eta,\sigma)$ has a probability density function given by $p(\alpha|\eta,\sigma) = \exp(-|\alpha-\eta|/\sigma)/2\sigma$ with expectation $\eta$ and variance $2\sigma^2$.
 We assume independence amongst the normalization factors, parental averages, half parental differences, hybrid differences, and overdispersion parameters.

### 2015-05-13 ebayes_rnaseq_heterosis.tex
to eBayes

@@
 where $\alpha\sim La(\eta,\sigma)$ has a probability density function given by $p(\alpha|\eta,\sigma) = \exp(-|\alpha-\eta|/\sigma)/2\sigma$ with expectation $\eta$ and variance $2\sigma^2$.
-We assume independence amongst the normalization factors, parental averages, half parental differences, hybrid differences, and overdispersion parameters.
+We assume independence amongst the parental averages, half parental differences, hybrid differences, and overdispersion parameters.
 
-We assume improper uniform priors for $\eta_\phi$, $\eta_\alpha$, and $\eta_\delta$ and independent half-cauchy priors for all standard deviations \citep{gelman2006prior}.
+%We assume improper uniform priors for $\eta_\phi$, $\eta_\alpha$, and $\eta_\delta$ and independent half-cauchy priors for all standard deviations \citep{gelman2006prior}.
 
@@
 
-We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$ and hyperparameters $\pi = (\gamma, \eta, \sigma)$, where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, $\eta=(\eta_\phi,\eta_\alpha,\eta_\delta)$, and $\sigma=(\sigma_\gamma,\sigma_\phi,\sigma_\alpha,\sigma_\delta,\sigma_\psi)$.
-We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\pi$, denoted $\hat{\pi}$, and then base gene-specific inference on the posterior conditional on this MAP estimate.
+We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$ and hyperparameters $\pi = (\gamma, \eta, \sigma)$, where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, $\eta=(\eta_\phi,\eta_\alpha,\eta_\delta)$, and $\sigma=(\sigma_\phi,\sigma_\alpha,\sigma_\delta,\sigma_\psi)$.
+We obtain estimates for the hyperparameters and then base gene-specific inference on the posterior conditional on these estimates.
+
+To obtain estimates of the hyperparameters, we use \edgeR to provide estimates of gene-specific parameters, $\hat{\theta}$ and normalization factors, $\hat{\gamma}$.
+
 Conditional on the hyperparameters, the gene-specific parameters are independent, i.e., 

### 2015-05-14 ebayes_rnaseq_heterosis.tex
thru 2.2

@@
 
-\renewcommand{\gamma}{c}
+%\renewcommand{\gamma}{c}
 
@@
 
-We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$ and hyperparameters $\pi = (\gamma, \eta, \sigma)$, where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, $\eta=(\eta_\phi,\eta_\alpha,\eta_\delta)$, and $\sigma=(\sigma_\phi,\sigma_\alpha,\sigma_\delta,\sigma_\psi)$.
+We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$, normalization factors $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, and hyperparameters $\pi = (\eta, \sigma)$ where $\eta=(\eta_\phi,\eta_\alpha,\eta_\delta)$ and $\sigma=(\sigma_\phi,\sigma_\alpha,\sigma_\delta,\sigma_\psi)$.
 We obtain estimates for the hyperparameters and then base gene-specific inference on the posterior conditional on these estimates.
 
-To obtain estimates of the hyperparameters, we use \edgeR to provide estimates of gene-specific parameters, $\hat{\theta}$ and normalization factors, $\hat{\gamma}$.
+To obtain normalization factors $\hat{\gamma}$, we use the weighted trimmed mean of $M$ values method of \cite{robinson2010scaling}.
+We then use the methods of \cite{mccarthy2012differential}, to obtain genewise dispersion estimates, $\hat{\phi}_g$, and their generalized linear model methods to obtain estimates for the remaining gene-specific parameters, i.e.
+$\hat{\phi}_g$, $\hat{\alpha}_g$, and $\hat{\delta}_g$.
+Using the gene-specific parameter estimates, we estimate hyperparameters for the location and scale parameters in the hierarchical model using a central method of moments approach, e.g.
+$\hat{\eta}_\delta = \sum_{g=1}^G \hat{\delta}_g/G$ and $2\hat{\sigma}_\delta^2 = \sum_{g=1}^G (\hat{\delta}_g - \hat{\eta}_\delta)^2/(G-1)$.
 
-Conditional on the hyperparameters, the gene-specific parameters are independent, i.e., 
+Conditional on the estimated normalization factors $\hat{\gamma}$ and hyperparameters $\hat{\pi}$, we perform a Bayesian analysis to re-estimate the gene-specific parameters and describe their uncertainty.
+Conditional on $\hat{\gamma}$ and $\hat{\pi}$ the gene-specific parameters are independent, i.e.
 \begin{equation}
 \begin{split}
-p\left(\theta\left|y,\hat{\pi}\right.\right) 
+p\left(\theta\left|y,\hat{\pi},\hat{\gamma}\right.\right) 
 %&= \prod_{g=1}^G p\left(\theta_g\left|y_g,\hat{\pi}\right.\right) = \prod_{g=1}^G p\left(y_g\left|\theta_g,\hat{\gamma}\right.\right) p\left(\theta_g\left|\hat{\pi}\right.\right) \nonumber \\
@@
 \end{equation}
-Thus, conditional posterior inference for the gene-specific parameters can be obtained independently and in parallel.
+Thus, conditional posterior inference across the genes can be obtained independently and in parallel.
 
-As we have not integrated out the gene-specific parameters when obtaining the MAP estimate, we actually receive a joint MAP estimate of $\theta$ and $\pi$.
-We explicitly use the MAP estimate of $\pi$ when obtaining empirical Bayes posterior samples for the gene-specific parameters as in equation \eqref{e:condind}.
-The MAP estimate of $\theta$ could be used as initial values in the MCMC, although we did not find this necessary.
+To perform this Bayesian analysis, we used the statistical software \Stan which implements a variant of Markov chain Monte Carlo (MCMC) called Hamiltonion Monte Carlo (HMC) \citep{neal2011mcmc} to obtaining samples from the posterior in equation \eqref{e:condind}.
+We ran 4 simultaneous chains with random initial starting values for 1000 burnin (and tuning) iterations followed by another 1000 iterations thinned by 4 (to reduce storage space) for inference.
+We monitored convergence using the potential scale reduction factor (PSRF) \citep{Gelm:Rubi:infe:1992} for $\phi_g$, $\alpha_g$, $\delta_g$, and $\psi_g$.
+If the maximum PSRF was greater than 1.1, we reran the chain with an additional 500 iterations (250 burnin and 250 for inference) and repeated this process until all PSRF's were less than 1.1.
 
@@
 
-For a specific gene $g$, low-parent heteorisis (LPH) of gene expression occurs when average expression in the hybrid is less than the average expression of either parent, i.e.
+For a specific gene $g$, low-parent heterosis (LPH) of gene expression occurs when average expression in the hybrid is less than the average expression of either parent, i.e.
 $\delta_g < -|\alpha_g|$, and high-parent heterosis (HPH) occurs when average expression in the hybrid is greater than the average expression of either parent, i.e.
@@
 %P(LPH_g|y) \approx 
-P(LPH_g|y,\hat{\pi}_{MAP}) = P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi}_{MAP}\right) = \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right), 
+P(LPH_g|y,\hat{\pi},\hat{\gamma}) = P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi},\hat{\gamma}\right) = \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right),
 \label{e:probs}
@@
 
-To perform the optimization required for the MAP estimate $\hat{\pi}$, we use the statistical software, {\tt Stan} \citep{stan-software:2014}, executed through the \RStan{} interface \citep{rstan-software:2014} in R \citep{R2014}.
+%To perform the optimization required for the MAP estimate $\hat{\pi}$, we use the statistical software, {\tt Stan} \citep{stan-software:2014}, executed through the \RStan{} interface \citep{rstan-software:2014} in R \citep{R2014}.
 The Stan software allows the construction of general Bayesian models with the primary restriction that all unknown parameters are continuous.
@@
 Presumably due to the large dimension of the optimization, it was necessary to use the L-BFGS algorithm \cite[Section 55,][]{stan-manual:2014} and reasonable starting values obtained via \edgeR{} \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor}.
-The {\tt Stan} software also includes the \RStan{} function, {\tt sampling}, an implementation of a Markov chain Monte Carlo (MCMC) algorithm for obtaining samples from the posterior.
-By default, Stan uses a no-U-turn sampler \citep{hoffman2013no}, a variant of Hamiltonion Monte Carlo \citep{neal2011mcmc}.
 

### 2015-05-14 ebayes_rnaseq_heterosis.tex
remove sec

@@
 
-\subsection{Implementation in \Stan{}}
+% \subsection{Implementation in \Stan{}}
 
@@
 
-The supplementary materials include the maize experiment data ({\tt data.csv}), a script to run our method ({\tt script.R}), and the {\tt Stan} models: the all-genes-independent (\verb@model_ind.txt@) and covariance (\verb@model_cov.txt@) models used in obtaining MAP estimates, along with the single-gene-independent (\verb@sg_model_ind.txt@) and covariance (\verb@sg_model_cov.txt@) models used in performing the MCMC on single genes.
+The supplementary materials include the maize experiment data ({\tt data.csv}), a script to run our method ({\tt script.R}), and the {\tt Stan} model (\verb@sg_model_ind.txt@).  
+%the all-genes-independent (\verb@model_ind.txt@) and covariance (\verb@model_cov.txt@) models used in obtaining MAP estimates, along with the single-gene-independent (\verb@sg_model_ind.txt@) and covariance (\verb@sg_model_cov.txt@) models used in performing the MCMC on single genes.
 
@@
 \bibliographystyle{biom}
-
-% From Dan, "There are some capitalization issues in the references section.
-I think you need to protect "RNA" with curly braces so that it stays capitalized.
-Also, in Hallauer et al.
-(2010), the last initial should presumably be "D." rather than "d."."
 

### 2015-05-15 supp/script.R
start of edgeR analysis

@@ -1,2 +1,71 @@
-# We need an R script to perform the analysis
+library(rstan)
+library(plyr)
+library(dplyr)
+library(reshape2)
+library(MASS)
 
+# Compile stan model
+model = stan_model("model.stan")
+
+d = read.csv("data_matrix.csv", row.names=1)
+
+# trim genes for average count 
+d = d[which(rowMeans(d)>0),] 
+
+# use edgeR 
+library(edgeR)
+
+# phi, alpha, delta parameterization
+design = cbind(1,
+               rep(c(1,-1,0), each=4),
+               rep(c(0, 0,1), each=4))
+
+# GLM fit
+dge = d %>% 
+  DGEList() %>%
+  calcNormFactors %>%
+  estimateCommonDisp(design) %>%
+  estimateGLMTagwiseDisp(design) %>%
+  glmFit(design)
+
+
+
+
+# long format
+m = melt(d, 
+         id.var = "GeneID",
+         variable.name="sample", 
+         value.name="count")
+m$variety = factor(gsub("_[1-4]", 
+                        "", 
+                        as.character(m$sample)),
+                   levels=c("B73","Mo17","B73xMo17"))
+
+ind = ddply(m, .(GeneID), function(x) {
+  mo = glm.nb(count+1~variety, data=x)
+  as.data.frame(coef(mo))
+}, .inform=TRUE)
+
+r = glm(count+1~variety,data=test, family="poisson")
+
+tmp = 
+m %>% 
+  group_by(GeneID, variety) %>%
+  summarize(mean=mean(count)) %>%
+  dcast(GeneID~variety, value.var="mean")
+
+zeros = 
+tmp %>% 
+  rowwise() %>%
+  summarize(any = B73==0 | Mo17 == 0 | B73xMo17 == 0)
+mean(zeros$any)
+sum(zeros$any)
+
+
+tmp[zeros$any,] 
+  
+
+
+rm = rowMeans(d[,-1])
+
+mean(rowM)

### 2015-05-28 ebayes_rnaseq_heterosis.tex
laplace

@@
 We found MAP estimates for both the independence and dependence models as shown in Table \ref{t:map}. 
-\input{include/map}
+\input{include/map-laplace}
 The MAP estimates for the hierarchical means, $\eta$, and standard deviations, $\sigma$, are consistent between the independence and dependence models with, perhaps, a slightly larger mean and smaller standard deviation for the logarithm of the overdispersion parameter, $\psi_g$, in the dependence model relative to the independence model.

### 2015-05-28 ebayes_rnaseq_heterosis.tex
need to update results

@@
 
-We present a hierarchical model for the gene-specific mean parameters and overdispersion parameters.
+We assume a hierarchical model for the gene-specific mean parameters and overdispersion parameters.
 The parental averages and overdispersion parameters are assumed to follow a normal distribution, i.e.
@@
 
-To obtain normalization factors $\hat{\gamma}$, we use the weighted trimmed mean of $M$ values method of \cite{robinson2010scaling}.
+To obtain normalization factors $\hat{\gamma}$, we use the weighted trimmed mean of $M$ values (TMM) method of \cite{robinson2010scaling}.
 We then use the methods of \cite{mccarthy2012differential}, to obtain genewise dispersion estimates, $\hat{\phi}_g$, and their generalized linear model methods to obtain estimates for the remaining gene-specific parameters, i.e.
@@
 
-We used our method to analyze the maize data set, generated by \cite{paschold2012complementation}, on gene expression in parental lines B73 and Mo17 and the hybrid genotype (B73$\times$Mo17).
+We used our method to analyze a maize data set \citep{paschold2012complementation} on gene expression in parental lines B73 and Mo17 and the hybrid genotype (B73$\times$Mo17).
 The data are available in the Supplemental Materials.
@@
 
-The supplementary materials include the maize experiment data ({\tt data.csv}), a script to run our method ({\tt script.R}), and the {\tt Stan} model (\verb@sg_model_ind.txt@).  
+The supplementary materials include three plain text files: the maize experiment data ({\tt data.csv}), a script to run our method ({\tt script.R}) on this data, and the {\tt Stan} model ({\tt model.stan}).
 %the all-genes-independent (\verb@model_ind.txt@) and covariance (\verb@model_cov.txt@) models used in obtaining MAP estimates, along with the single-gene-independent (\verb@sg_model_ind.txt@) and covariance (\verb@sg_model_cov.txt@) models used in performing the MCMC on single genes.

### 2015-05-28 supp/script.R
updating script

@@ -2,25 +2,27 @@ library(rstan)
 library(plyr)
 library(dplyr)
 library(reshape2)
-library(MASS)
+library(MASS) #?
+library(edgeR)
 
 # Compile stan model
 model = stan_model("model.stan")
 
-d = read.csv("data_matrix.csv", row.names=1)
+d = read.csv("data.csv", row.names=1)
+variety = factor(gsub("_[1-4]", "", names(d)), levels=c("B73","Mo17","B73xMo17"))
 
-# trim genes for average count 
-d = d[which(rowMeans(d)>0),] 
+# trim genes for average count greater than 1
+d = d[which(rowMeans(d)>1),] 
 
-# use edgeR 
-library(edgeR)
-
+#######################################
+# Find hyperparameter estimates
+#######################################
 # phi, alpha, delta parameterization
 design = cbind(1,
-               rep(c(1,-1,0), each=4),
-               rep(c(0, 0,1), each=4))
+               c(1,-1,0)[as.numeric(variety)],
+               variety == 'B73xMo17')
 
-# GLM fit
+# GLM fit using edgeR
 fit = d %>% 
   DGEList() %>%
   calcNormFactors %>%
@@ -28,3 +30,32 @@ fit = d %>%
   estimateGLMTagwiseDisp(design) %>%
   glmFit(design)
 
+# Calculate gene-specific estimates for phi, alpha, and delta
+hat = data.frame(gene = 1:length(fit$dispersion),
+                 phi   = fit$coefficients[,1] + mean(fit$offset[1,]),
+                 alpha = fit$coefficients[,2],
+                 delta = fit$coefficients[,3],
+                 psi   = log(fit$dispersion))
+
+ss = hat %>%
+  melt(id.vars='gene', variable.name = 'parameter') %>%
+  group_by(parameter) %>%
+  summarize(mean=mean(value), sd=sd(value))
+
+# Fix standard deviations for Laplace priors
+ss$sd[2:3] = ss$sd[2:3]/sqrt(2)
+
+
+######################################
+# Run individual gene analyses
+######################################
+# This will take a long time, so parallelize if possible
+if (parallel <- require(doMC)) {
+  registerDoMC()
+} 
+
+analysis = d %>% 
+  mutate(gene = 1:nrow(d)) %>%
+  melt(id.vars = 'gene', variable.name = 'sampleID') %>%
+  group_by(gene) 
+  

### 2015-06-15 slides/EM/em.Rnw
EM slides

\documentclass[handout]{beamer}

\usecolortheme[RGB={0,0,144}]{structure}
\usetheme{AnnArbor}\usecolortheme{beaver}

\usepackage{verbatim,xmpmulti,color,multicol,multirow}
\setlength{\unitlength}{\textwidth}  % measure in textwidths
\usepackage[normalem]{ulem}

\graphicspath{{../figs/}}

%\usepackage{beamerthemesplit}
\setbeamertemplate{navigation symbols}{}
%\setbeamercolor{alerted text}{fg=red}
%\setbeamertemplate{block body theorem}{bg=orange}
\setkeys{Gin}{width=0.6\textwidth}

\title{Monte Carlo Expectation Maximization for RNAseq}
\author[Jarad Niemi]{Dr. Jarad Niemi}
\institute[Iowa State]{Iowa State University}
\date{\today}

\begin{document}

%\section{Temp??} \begin{comment}

<<chunk_options, echo=FALSE, message=FALSE>>=
opts_chunk$set(fig.width=6, fig.height=5, out.width='.8\\linewidth', fig.align='center', size='tiny')
library(plyr)
library(ggplot2)
library(reshape2)
@

\frame{\titlepage}

\begin{frame}
\frametitle{Outline}
\begin{itemize}
\item EM
\end{itemize}
\end{frame}

\section{Expectation maximization}
\subsection{Hierarchical models}
\begin{frame}
\frametitle{Hierarchical models}
Generically 
\[ y|\theta \sim p(y|\theta) \qquad \theta|\pi \sim p(\theta|\pi) \]

\vspace{0.2in} \pause

Conditional independencies:
\begin{itemize}[<+->]
\item Data model 
\[ p(y|\theta) = \prod_{g=1}^G p(y_g|\theta_g) \]
\item Hierarchical distribution
\[ p(\theta|\pi) = \prod_{g=1}^G p(\theta_g|\pi) \]
\end{itemize}
where $G$ is large, e.g. $\sim 10^4$. 
\end{frame}

\begin{frame}
\frametitle{Expectation-Maximization algorithm}

Suppose we wish to find 
\[ \hat{\pi}_{MLE} = \mbox{argmax}_\pi p(y|\pi) \quad \mbox{where} \quad p(y|\pi) = \int p(y|\theta) p(\theta|\pi) d\theta \]

\pause

An expectation-maximization algorithm is 
\begin{enumerate}
\item Expectation: 
\[ \begin{array}{rll}
Q(\pi|\pi^{(t)}) &= E_{\theta|y,\pi^{(t)}}\left[\log L(\pi|y,\theta) \right] \pause \\
& = E_{\theta|y,\pi^{(t)}}\left[\log p(y|\theta)\right] &+ E_{\theta|y,\pi^{(t)}}\left[\log p(\theta|\pi)\right] \pause \\
& = C &+ E_{\theta|y,\pi^{(t)}}\left[\log p(\theta|\pi)\right] 
\end{array} \]
\item \pause Maximization:
\[ \pi^{(t+1)} = \mbox{argmax}_{\pi} Q(\pi|\pi^{(t)}) \pause = \mbox{argmax}_{\pi} E_{\theta|y,\pi^{(t)}}\left[\log p(\theta|\pi)\right]  \]
\end{enumerate}

\pause

Theory states that 
\[ \pi^{(t)} \to \hat{\pi}_{MLE} \quad \mbox{as } t\to\infty. \]

\end{frame}

\begin{frame}
\frametitle{Normal with hierarchical mean}

Let $y_{gi} \stackrel{ind}{\sim} N(\theta_g, s_g^2)$ and $\theta_g \sim N(\mu,\tau)$ for $i=1,\ldots,n_g=n$ and $g=1,\ldots,G$. \pause Let $\pi=(\mu,\tau)$,  
\[ \begin{array}{rl}
m_g^{(t)} &= E\left[\theta_g\left|y,\pi^{(t)}\right.\right] = v_g^{(t)}\left[\frac{\mu^{(t)}}{\tau^{(t)}} + \frac{\overline{y}_g}{s_g^2/n_g} \right] \\
v_g^{(t)} &= V\left[\theta_g\left|y,\pi^{(t)}\right.\right] = \left[\frac{1}{\tau^{(t)}} + \frac{1}{s_g^2/n_g} \right]^{-1}
\end{array} \]

Then, we have 
{\small
\[ \begin{array}{rl}
E_{\theta\left|y,\pi^{(t)}\right.}\left[\log p(\theta|\pi)\right]
%&= \log \prod_{g=1}^G (2\pi \tau)^{-1/2} \exp\left(-\frac{1}{2\tau} (\theta_g-\mu)^2 \right) \\
%&= \sum_{g=1}^G \left[ -\log(2\pi \tau)/2  -\frac{1}{2\tau} (\theta_g-\mu)^2\right] \\
&= -\frac{G\log(2\pi \tau)}{2}-\frac{1}{2\tau}\sum_{g=1}^G  E_{\theta\left|y,\pi^{(t)}\right.}\left[(\theta_g-\mu)^2\right] \\
&= -\frac{G\log(2\pi \tau)}{2}-\frac{1}{2\tau}\sum_{g=1}^G v_g^{(t)} - \frac{1}{2\tau}\sum_{g=1}^G (m_g^{(t)}-\mu)^2 \\
\end{array} \]
}
\pause
which is maximized when 
{\small
\[ \begin{array}{rl}
\mu^{(t+1)} &= \frac{1}{G} \sum_{g=1}^G m_g^{(t)} \pause \\ 
\tau^{(t+1)} &= \frac{1}{G} \sum_{g=1}^G  E_{\theta\left|y,\pi^{(t)}\right.}\left[(\theta_g-\mu)^2\right] =  \frac{1}{G} \sum_{g=1}^G \left[ v_g^{(t)} + \left(m_g^{(t)}-\mu^{(t+1)}\right)^2 \right]. 
\end{array}\]
}
\end{frame}

\begin{frame}[fragile]

<<normal_hierarchical_mean, cache=TRUE, echo=FALSE>>=
G = 1000
truth = data.frame(mu = 0, tau = 1)
d = ddply(data.frame(gene=1:G, theta = with(truth, rnorm(G, mu, sqrt(tau)))),
          .(gene), function(x) data.frame(gene=x$gene, y=rnorm(4, x$theta)))
sm = ddply(d, .(gene), summarize, n = length(y), ybar = mean(y))
mu = tau = 1 # Initial values
n_iter = 10
keep = data.frame(iteration = 1:n_iter, mu = rep(NA, n_iter), tau = rep(NA, n_iter))
for (i in 1:n_iter) {
  sm$v = 1/(1/tau+sm$n)
  sm$m = sm$v * (mu/tau + sm$ybar*sm$n)
  mu = mean(sm$m)
  tau = mean(sm$v + (sm$m-mu)^2)
  keep$mu[i] = mu; keep$tau[i] = tau
}
ggplot(melt(keep, id.var="iteration", variable.name="parameter"), 
       aes(iteration, value)) +
  geom_point() + 
  facet_wrap(~parameter, scales="free_y") +
  geom_hline(data=melt(truth, variable.name='parameter'), aes(yintercept=value), color='red')
@

\end{frame}

\subsection{Monte Carlo EM}
\begin{frame}
\frametitle{Monte Carlo EM}
If the expectation step is not analytically tractible, we can replace the expectation with a Monte Carlo approximation, \pause e.g. 
\[ 
E_{\theta|y,\pi^{(t)}}\left[\log p(\theta|\pi)\right] \approx \frac{1}{M} \sum_{m=1}^M \log p\left(\left.\theta^{(m)}\right|\pi\right)
\]
where $\theta^{(m)} \sim p(\theta|y,\pi^{(t)})$. \pause since
\[
\frac{1}{M} \sum_{m=1}^M \log p\left(\left.\theta^{(m)}\right|\pi\right) \to E_{\theta|y,\pi^{(t)}}\left[\log p(\theta|\pi)\right]
\]
as $m\to \infty$.
\end{frame}

\begin{frame}
\frametitle{Parallelism in hierarchical models}

In hierarchical models, we have 
\[ p\left(\theta\left|y,\pi^{(t)}\right.\right) = \prod_{g=1}^G p\left(\theta_g\left|y_g,\pi^{(t)}\right.\right)\]
and thus we can simulate $\theta_g$ independently across $g$. \pause Also, since we have 
\[ E_{\theta|y,\pi^{(t)}}[\log p(\theta|\pi)] = \sum_{g=1}^G E_{\theta|y,\pi^{(t)}}[\log p(\theta_g|\pi)] = \sum_{g=1}^G E_{\theta_{\alert{g}}|y,\pi^{(t)}}[\log p(\theta_g|\pi)] \]
we can approximate the expectation via 
\[
E_{\theta|y,\pi^{(t)}}[\log p(\theta|\pi)] \approx \frac{1}{M} \sum_{m=1}^M \sum_{g=1}^G \log p\left(\left.\theta_g^{(m)}\right|\pi\right).
\]

\end{frame}

\begin{frame}
\frametitle{Normal with hierarchical mean}

This Monte Carlo approximation to the desired expectation

{\small
\[ \begin{array}{rl}
E_{\theta|y,\pi^{(t)}}[\log p(\theta|\pi)] &= -\frac{G\log(2\pi \tau)}{2}-\frac{1}{2\tau}\sum_{g=1}^G  E_{\theta_g\left|y,\pi^{(t)}\right.}\left[(\theta_g-\mu)^2\right] \\
&\approx -\frac{G\log(2\pi \tau)}{2}-\frac{1}{2\tau}\sum_{g=1}^G \frac{1}{M} \sum_{m=1}^M (\theta_g^{(m)}-\mu)^2
\end{array} \]
}

is maximized with

\[ \begin{array}{rl}
\mu^{(t+1)} &= \frac{1}{MG} \sum_{m=1}^M \sum_{g=1}^G \theta_g^{(m)} \\
\tau^{(t+1)} &= \frac{1}{MG} \sum_{m=1}^M \sum_{g=1}^G \left(\theta_g^{(m)} - \mu^{(t+1)} \right)^2
\end{array} \]
\end{frame}

\end{document}

### 2015-06-16 slides/EM/em.Rnw
through mcmcem

@@
 
-\title{Monte Carlo Expectation Maximization for RNAseq}
+\title[MCMCEM for RNAseq]{Monte Carlo Expectation Maximization for RNAseq}
 \author[Jarad Niemi]{Dr. Jarad Niemi}
@@
 library(reshape2)
+library(rstan)
 @
@@
 
-<<normal_hierarchical_mean, cache=TRUE, echo=FALSE>>=
-G = 1000
+<<normal_hierarchical_mean_em, cache=TRUE, echo=FALSE>>=
+G = 100
 truth = data.frame(mu = 0, tau = 1)
@@
 n_iter = 10
-keep = data.frame(iteration = 1:n_iter, mu = rep(NA, n_iter), tau = rep(NA, n_iter))
+keep = data.frame(iteration = 1:n_iter, mu = rep(NA, n_iter), tau = rep(NA, n_iter), method="em")
 for (i in 1:n_iter) {
@@
 }
-ggplot(melt(keep, id.var="iteration", variable.name="parameter"), 
+@
+
+<<, dependson='normal_hierarchical_mean_em', echo=FALSE>>=
+ggplot(melt(keep, id.var=c("iteration","method"), variable.name="parameter"), 
        aes(iteration, value)) +
@@
 
-This Monte Carlo approximation to the desired expectation
+The Monte Carlo approximation to the desired expectation is
 
@@
 \[ \begin{array}{rl}
-\mu^{(t+1)} &= \frac{1}{MG} \sum_{m=1}^M \sum_{g=1}^G \theta_g^{(m)} \\
-\tau^{(t+1)} &= \frac{1}{MG} \sum_{m=1}^M \sum_{g=1}^G \left(\theta_g^{(m)} - \mu^{(t+1)} \right)^2
+\mu^{(t+1)} &= \frac{1}{G} \sum_{g=1}^G \frac{1}{M}\sum_{m=1}^M \theta_g^{(m)} \\
+\tau^{(t+1)} &= \frac{1}{G} \sum_{g=1}^G  \frac{1}{M}\sum_{m=1}^M\left(\theta_g^{(m)} - \mu^{(t+1)} \right)^2
 \end{array} \]
+\pause 
+where $\theta_g^{(m)} \stackrel{ind}{\sim} p\left(\theta_g\left| y_g,\pi^{(t)}\right.\right)$.
+\end{frame}
+
+\begin{frame}[fragile]
+<<normal_hierarchical_mean_mcem, dependson="normal_hierarchical_mean_em", cache=TRUE, echo=FALSE>>=
+M = 1000
+mu = tau = 1 # Initial values
+keep2 = data.frame(iteration = 1:n_iter, mu = rep(NA, n_iter), tau = rep(NA, n_iter), method='mcem')
+for (i in 1:n_iter) {
+  sm$v = 1/(1/tau+sm$n)
+  sm$m = sm$v * (mu/tau + sm$ybar*sm$n)
+  draws = ddply(sm, .(gene), function(x) {
+    data.frame(theta = rnorm(M, x$m, sqrt(x$v)))
+  })
+  mu = mean(draws$theta)
+  tau = mean((draws$theta-mu)^2)
+  keep2$mu[i] = mu; keep2$tau[i] = tau
+}
+@
+
+<<dependson=c('normal_hierarchical_mean_em','normal_hierarchical_mean_mcem'), echo=FALSE>>=
+ggplot(melt(rbind(keep,keep2), id.var=c("iteration","method"), variable.name="parameter"), 
+       aes(iteration, value, color=method)) +
+  geom_point() + 
+  facet_wrap(~parameter, scales="free_y") +
+  geom_hline(data=melt(truth, variable.name='parameter'), aes(yintercept=value))
+@
+\end{frame}
+
+\subsection{MCMC EM}
+\begin{frame}
+\frametitle{Markov chain Monte Carlo EM}
+
+It will not always be easy to sample from $p\left(\theta\left|y,\pi^{(t)}\right.\right)$, so we may want to use an MCMC approach to sample from this distribution.
+\pause The resulting MCMCEM algorithm is identical to the MCEM algorithm, i.e.
+
+\[ \begin{array}{rl}
+\mu^{(t+1)} &= \frac{1}{G} \sum_{g=1}^G \frac{1}{M}\sum_{m=1}^M \theta_g^{(m)} \\
+\tau^{(t+1)} &= \frac{1}{G} \sum_{g=1}^G  \frac{1}{M}\sum_{m=1}^M\left(\theta_g^{(m)} - \mu^{(t+1)} \right)^2
+\end{array} \]
+\pause where 
+\[ \theta_g^{(m)} \sim p\left(\theta_g\left|y_g,\pi^{(t)}\right.\right) \]
+but now $\theta_g^{(m)}$ and $\theta_g^{(m+1)}$ are not independent \pause (they are still independent across $g$).
+
+\end{frame}
+
+\begin{frame}[fragile]
+<<normal_hierarchical_mean_mcmcem, dependson="normal_hierarchical_mean_em", cache=TRUE, echo=FALSE, results='hide'>>=
+normal_hierarchical_mean_model = "
+data {
+  int<lower=1> n;
+  real y[n];
+  real mu;
+  real<lower=0> s;
+  real<lower=0> tau;
+}
+parameters {
+  real theta;
+}
+model {
+  theta ~ normal(mu,sqrt(tau));
+  y ~ normal(theta, s);
+}
+"
+
+m = stan_model(model_code = normal_hierarchical_mean_model)
+
+mu = tau = 1 # Initial values
+keep3 = data.frame(iteration = 1:n_iter, mu = rep(NA, n_iter), tau = rep(NA, n_iter), method='mcmcem')
+for (i in 1:n_iter) {
+  # Run MCMC using rstan
+  draws = ddply(d, .(gene), function(x) {
+    s = sampling(m, data = list(n = length(x$y), y=x$y, mu=mu, s=1, tau=tau), iter=M/2)
+    data.frame(theta = as.numeric(extract(s, 'theta')$theta))
+  })
+
+  mu = mean(draws$theta)
+  tau = mean((draws$theta-mu)^2)
+  keep3$mu[i] = mu; keep3$tau[i] = tau
+}
+@
+
+<<dependson=c('normal_hierarchical_mean_em','normal_hierarchical_mean_mcem'), echo=FALSE>>=
+ggplot(melt(rbind(keep,keep2,keep3), id.var=c("iteration","method"), variable.name="parameter"), 
+       aes(iteration, value, color=method)) +
+  geom_point() + 
+  facet_wrap(~parameter, scales="free_y") +
+  geom_hline(data=melt(truth, variable.name='parameter'), aes(yintercept=value))
+@
+
+\end{frame}
+
+\section{Applications}
+\begin{frame}
+\frametitle{Negative binomial example}
+
+Let's suppose $y_{gi} \stackrel{ind}{\sim} NB(\theta_g,\psi_g)$, i.e. 
+\[ E[y_{gi}] = \eta_g = e^{\theta_g} \quad\mbox{and}\quad V[y_{gi}] = \eta_g + e^{\psi_g}\eta_g \]
+\pause 
+and we assume a hierarchical distribution for $\theta_g$ and $\psi_g$:
+\[ 
+\theta_g \stackrel{ind}{\sim} N(\mu_\theta,\tau_\theta) \quad\mbox{and, independently,}\quad \psi_g \stackrel{ind}{\sim} N(\mu_\psi, \tau_\psi).
+\]
+\end{frame}
+
+\begin{frame}
+\frametitle{MCMCEM for negative binomial hierarchical model}
+
+Let $\pi=(\mu_\theta, \tau_\theta, \mu_\psi, \tau_\psi)$, then 
+\[ 
+E_{\theta,\psi|y,\pi^{(t)}}[\log p(\theta,\psi|\pi)] = E_{\theta|y,\pi^{(t)}}[\log p(\theta|\mu_\theta,\tau_\theta)] + E_{\psi|y,\pi^{(t)}}[\log p(\psi|\mu_\psi,\tau_\psi)]
+ \]
+ which is maximized with
+\[ \begin{array}{rl}
+\mu_\theta^{(t+1)} &= \frac{1}{G} \sum_{g=1}^G \frac{1}{M}\sum_{m=1}^M \theta_g^{(m)} \\
+\tau_\theta^{(t+1)} &= \frac{1}{G} \sum_{g=1}^G  \frac{1}{M}\sum_{m=1}^M\left(\theta_g^{(m)} - \mu_\theta^{(t+1)} \right)^2 \\
+\mu_\psi^{(t+1)} &= \frac{1}{G} \sum_{g=1}^G \frac{1}{M}\sum_{m=1}^M \psi_g^{(m)} \\
+\tau_\psi^{(t+1)} &= \frac{1}{G} \sum_{g=1}^G  \frac{1}{M}\sum_{m=1}^M\left(\psi_g^{(m)} - \mu_\psi^{(t+1)} \right)^2
+\end{array} \]
+\pause where 
+\[ \theta_g^{(m)},\psi_g^{(m)} \sim p\left(\theta_g,\psi_g\left|y_g,\pi^{(t)}\right.\right) \]
+via MCMC. 
+\end{frame}
+
+\begin{frame}[fragile]
+
+<<negative_binomial_data, cache=TRUE>>=
+G = 100
+n = 4
+truth = data.frame(mu_theta = 3, tau_theta = 1, mu_psi = 0, tau_psi = 1)
+par = data.frame(gene=1:G, 
+                 theta = with(truth, rnorm(G, mu_theta, sqrt(tau_theta))),
+                 psi   = with(truth, rnorm(G, mu_psi, sqrt(tau_psi))))
+d = ddply(par, .(gene), function(x) {
+  data.frame(y = rnbinom(n, mu = exp(x$theta), size=1/exp(x$psi)))
+})
+@
+
+<<negative_binomial_stan, cache=TRUE>>=
+negative_binomial_model = "
+data {
+  int<lower=1> n;
+  int<lower=0> count;
+  real mu_theta;
+  real<lower=0> tau_theta;
+  real mu_psi;
+  real<lower=0> tau_psi;
+}
+parameters {
+  real theta;
+  real psi;
+}
+model {
+  theta ~ normal(mu_theta, sqrt(tau_theta));
+  psi   ~ normal(mu_psi  , sqrt(tau_psi  ));
+  count ~ neg_binomial_2_log(theta, 1/psi);
+}
+"
+nb_m = stan_model(model_code = negative_binomial_model)
+@
+
+<<negative_binomial_mcmcem, dependson=c('negative_binomial_data','negative_binomial_stan'), cache=TRUE, echo=FALSE, results='hide'>>=
+M = 1000
+mu_theta = 0; tau_theta=1, mu_psi = 0; mu_psi = 1 # Initial values
+n_iter = 10
+keep = data.frame(iteration = 1:n_iter, 
+                  mu_theta = rep(NA, n_iter), 
+                  tau_theta = rep(NA, n_iter), 
+                  mu_psi = rep(NA,n_iter),
+                  tau_psi = rep(NA,n_iter))
+for (i in 1:n_iter) {
+  # Run MCMC using rstan
+  draws = ddply(d, .(gene), function(x) {
+    s = sampling(nb_m, data = list(n = length(x$y), y=x$y, 
+                                   mu_theta=mu_theta, tau_psi=tau_theta, 
+                                   mu_psi=mu_psi, tau_psi=tau_psi))
+    e = extract(s, c('theta','psi'))
+    data.frame(theta = as.numeric(e$theta),
+               psi = as.numeric(e$psi))
+  })
+
+  mu_theta = mean(draws$theta)
+  tau_theta = mean((draws$theta-mu_theta)^2)
+  mu_psi = mean(draws$psi)
+  tau_psi = mean((draws$psi-mu_psi)^2)
+  keep$mu_theta[i] = mu_theta; keep$tau_theta[i] = tau_theta
+  keep$mu_psi[i] = mu_psi; keep$tau_psi[i] = tau_psi
+}
+@
+
+<<dependson=c('normal_hierarchical_mean_em','normal_hierarchical_mean_mcem'), echo=FALSE>>=
+ggplot(keep, id.var=c("iteration"), variable.name="parameter"), 
+       aes(iteration, value, color=method)) +
+  geom_point() + 
+  facet_wrap(~parameter, scales="free_y") +
+  geom_hline(data=melt(truth, variable.name='parameter'), aes(yintercept=value))
+@
+
 \end{frame}

### 2015-06-16 slides/EM/em.Rnw
mcmcem for nb

@@
   int<lower=1> n;
-  int<lower=0> count;
+  int<lower=0> count[n];
   real mu_theta;
@@
   psi   ~ normal(mu_psi  , sqrt(tau_psi  ));
-  count ~ neg_binomial_2_log(theta, 1/psi);
+  count ~ neg_binomial_2_log(theta, 1/exp(psi));
 }
@@
 <<negative_binomial_mcmcem, dependson=c('negative_binomial_data','negative_binomial_stan'), cache=TRUE, echo=FALSE, results='hide'>>=
-M = 1000
-mu_theta = 0; tau_theta=1, mu_psi = 0; mu_psi = 1 # Initial values
-n_iter = 10
+mu_theta = 0; tau_theta=1; mu_psi = 0; tau_psi = 1 # Initial values
+n_iter = 100
 keep = data.frame(iteration = 1:n_iter, 
@@
   draws = ddply(d, .(gene), function(x) {
-    s = sampling(nb_m, data = list(n = length(x$y), y=x$y, 
-                                   mu_theta=mu_theta, tau_psi=tau_theta, 
+    s = sampling(nb_m, data = list(n = length(x$y), count=x$y, 
+                                   mu_theta=mu_theta, tau_theta=tau_theta, 
                                    mu_psi=mu_psi, tau_psi=tau_psi))
@@
 <<dependson=c('normal_hierarchical_mean_em','normal_hierarchical_mean_mcem'), echo=FALSE>>=
-ggplot(keep, id.var=c("iteration"), variable.name="parameter"), 
-       aes(iteration, value, color=method)) +
+ggplot(melt(keep, id.var=c("iteration"), variable.name="parameter"),
+       aes(iteration, value)) +
   geom_point() + 

### 2015-06-17 est_testing/prior_het_prob.tex
moved file to proper location

\documentclass{article}
\usepackage{amsmath,amsfonts}
\usepackage{setspace}
\begin{document}

\doublespace

Calculating prior probability of heterosis when
\[\alpha \sim Laplace(0,\sigma_\alpha),\; \delta \sim Laplace(0,\sigma_\delta)\]

First, note that $X = |\alpha| \sim Exp(\sigma_\alpha)$ and $Y = |\delta| \sim Exp(\sigma_\delta)$. Note also the pivotal quantities $X/\sigma_{\alpha}, Y/\sigma_{\delta} \sim Exp(1)$. Since we interested in the probability of the event $\{Y > X\} = \{Y/X>1\}$, consider the pivotal quantity, $Z = \frac{Y/\sigma_\delta}{X/\sigma_\alpha}$. By identifying $\frac{\sigma_\alpha}{\sigma_\delta} = c$, it is clear that $P(Z>c) = P(X/Y > 1)$.We can calculate the density of $Z$ by the transformation method as follows:

\[f_Z(z) = \int t\cdot f_{X,Y}(t,zt) dt = \int  t\cdot e^{-t(z+1)} dt = \frac{1}{(z+1)^2},\]

where the last equality comes from recognizing the integrand as the kernel of a gamma distribution.\\

Integrating this density, we find that $P(Z>c) = 1/(c+1)$. Given a probability $p$, $c$ can be calculated by $c = 1/p - 1$.\\

\end{document}

### 2015-06-17 slides/EM/em.Rnw
split plot

@@
 @
+\end{frame}
 
+\begin{frame}[fragile]
 <<dependson=c('normal_hierarchical_mean_em','normal_hierarchical_mean_mcem'), echo=FALSE>>=

### 2015-06-17 slides/EM/em.Rnw
heterosis with normal priors

@@
 <<negative_binomial_mcmcem, dependson=c('negative_binomial_data','negative_binomial_stan'), cache=TRUE, echo=FALSE, results='hide'>>=
-mu_theta = 0; tau_theta=1; mu_psi = 0; tau_psi = 1 # Initial values
+mu_theta = 0; tau_theta=1; mu_psi = -2; tau_psi = 1 # Initial values
 n_iter = 100
@@
 
+\section{Heterosis}
+\begin{frame}[fragile]
+\frametitle{Heterosis}
+
+<<>>=
+heterosis_model = "
+data {
+  int<lower=1> n;
+  int<lower=0> count[n];
+  int<lower=1, upper=3> variety[n];
+  real mu_phi;
+  real mu_alpha;
+  real mu_delta;
+  real mu_psi;
+  real<lower=0> tau_phi;
+  real<lower=0> tau_alpha;
+  real<lower=0> tau_delta;
+  real<lower=0> tau_psi;
+}
+transformed data {
+  matrix[n,3] X;
+  for (i in 1:n) {
+    if (variety[i] == 1) { X[i,1] <-  1; X[i,2] <- -1; X[i,3] <-  0; }
+    if (variety[i] == 2) { X[i,1] <-  1; X[i,2] <-  1; X[i,3] <-  0; }
+    if (variety[i] == 3) { X[i,1] <-  1; X[i,2] <-  0; X[i,3] <-  1; }
+  }
+}
+parameters {
+  real phi;
+  real alpha;
+  real delta;
+  real psi;
+}
+transformed parameters {
+  vector[3] pad;
+
+  pad[1] <- phi;
+  pad[2] <- alpha;
+  pad[3] <- delta;
+}
+model {
+  phi   ~ normal(mu_phi  , sqrt(tau_phi  ));
+  alpha ~ normal(mu_alpha, sqrt(tau_alpha));
+  delta ~ normal(mu_delta, sqrt(tau_delta));
+  psi   ~ normal(mu_psi  , sqrt(tau_psi  ));
+  count ~ neg_binomial_2_log(X*pad, 1/exp(psi));
+}
+"
+h_m = stan_model(model_code = heterosis_model)
+@
+\end{frame}
+
+<<heterosis_data, cache=TRUE>>=
+G = 10
+n = 4
+truth = data.frame(mu_phi   = 3, tau_phi   = 1, mu_psi = 0, tau_psi = 1,
+                   mu_alpha = 0, tau_alpha = 1, mu_delta = 0, tau_delta = .1^2)
+par = data.frame(gene=1:G, 
+                 phi   = with(truth, rnorm(G, mu_phi  , sqrt(tau_phi  ))),
+                 alpha = with(truth, rnorm(G, mu_alpha, sqrt(tau_alpha))),
+                 delta = with(truth, rnorm(G, mu_delta, sqrt(tau_delta))),
+                 psi   = with(truth, rnorm(G, mu_psi  , sqrt(tau_psi  ))))
+X = cbind(1, rep(c(-1,1,0), each=n), rep(c(0,0,1), each=n))
+d = ddply(par, .(gene), function(x) {
+  pad = with(x, c(phi,alpha,delta))
+  data.frame(y = rnbinom(3*n, mu = exp(X%*%pad), size=1/exp(x$psi)),
+             variety = rep(1:3, each=n))
+})
+@
+
+<<heterosis_mcmcem, dependson=c('negative_binomial_data','negative_binomial_stan'), cache=TRUE, echo=FALSE, results='hide'>>=
+current = as.list(truth)
+n_iter = 10
+keep = data.frame(iteration = 1:n_iter, 
+                  mu_phi    = rep(NA, n_iter), 
+                  mu_alpha  = rep(NA, n_iter), 
+                  mu_delta  = rep(NA, n_iter), 
+                  mu_psi    = rep(NA,n_iter),
+                  tau_phi   = rep(NA, n_iter),
+                  tau_alpha = rep(NA, n_iter), 
+                  tau_delta = rep(NA, n_iter),
+                  tau_psi   = rep(NA,n_iter))
+
+for (i in 1:n_iter) {
+  # Run MCMC using rstan
+  draws = ddply(d, .(gene), function(x) {
+    s = sampling(h_m, data = c(list(n = length(x$y), count=x$y, variety = x$variety), current))
+    e = extract(s, c('phi','alpha','delta','psi'))
+    data.frame(phi   = as.numeric(e$phi),
+               alpha = as.numeric(e$alpha),
+               delta = as.numeric(e$delta),
+               psi   = as.numeric(e$psi))
+  })
+
+  current = list(mu_phi   = mean(draws$phi)  , tau_phi   = mean((draws$phi-mu_phi)^2),
+                 mu_alpha = mean(draws$alpha), tau_alpha = mean((draws$alpha-mu_alpha)^2),
+                 mu_delta = mean(draws$delta), tau_delta = mean((draws$delta-mu_delta)^2),
+                 mu_psi   = mean(draws$psi)  , tau_psi   = mean((draws$psi-mu_psi)^2))
+  keep$mu_phi[i]   = current$mu_phi;   keep$tau_phi[i]   = current$tau_phi
+  keep$mu_alpha[i] = current$mu_alpha; keep$tau_alpha[i] = current$tau_alpha
+  keep$mu_delta[i] = current$mu_delta; keep$tau_delta[i] = current$tau_delta
+  keep$mu_psi[i]   = current$mu_psi;   keep$tau_psi[i]   = current$tau_psi
+}
+@
+\end{frame}
+
+\begin{frame}[fragile]
+<<dependson=c('normal_hierarchical_mean_em','normal_hierarchical_mean_mcem'), echo=FALSE>>=
+ggplot(melt(keep, id.var=c("iteration"), variable.name="parameter"),
+       aes(iteration, value)) +
+  geom_point() + 
+  facet_wrap(~parameter, scales="free_y", nrow=2) +
+  geom_hline(data=melt(truth, variable.name='parameter'), aes(yintercept=value))
+@
+
+\end{frame}
+
 \end{document}

### 2015-06-17 slides/EM/em.Rnw
increase G and n_iter

@@
 <<heterosis_data, cache=TRUE>>=
-G = 10
+G = 100
 n = 4
@@
 current = as.list(truth)
-n_iter = 10
+n_iter = 100
 keep = data.frame(iteration = 1:n_iter, 

### 2015-06-22 slides/EM/em.Rnw
+dplyr

--- 2015-06-17 slides/EM/em.Rnw
@@
 library(plyr)
+library(dplyr)
 library(ggplot2)
@@
 
+\section{Heterosis}
+\begin{frame}[fragile]
+\frametitle{Heterosis}
+
+<<heterosis_model, cache=TRUE>>=
+heterosis_model = "
+data {
+  int<lower=1> n;
+  int<lower=0> count[n];
+  int<lower=1, upper=3> variety[n];
+  real mu_phi;
+  real mu_alpha;
+  real mu_delta;
+  real mu_psi;
+  real<lower=0> tau_phi;
+  real<lower=0> tau_alpha;
+  real<lower=0> tau_delta;
+  real<lower=0> tau_psi;
+}
+transformed data {
+  matrix[n,3] X;
+  for (i in 1:n) {
+    if (variety[i] == 1) { X[i,1] <-  1; X[i,2] <- -1; X[i,3] <-  0; }
+    if (variety[i] == 2) { X[i,1] <-  1; X[i,2] <-  1; X[i,3] <-  0; }
+    if (variety[i] == 3) { X[i,1] <-  1; X[i,2] <-  0; X[i,3] <-  1; }
+  }
+}
+parameters {
+  real phi;
+  real alpha;
+  real delta;
+  real psi;
+}
+transformed parameters {
+  vector[3] pad;
+
+  pad[1] <- phi;
+  pad[2] <- alpha;
+  pad[3] <- delta;
+}
+model {
+  phi   ~ normal(mu_phi  , sqrt(tau_phi  ));
+  alpha ~ normal(mu_alpha, sqrt(tau_alpha));
+  delta ~ normal(mu_delta, sqrt(tau_delta));
+  psi   ~ normal(mu_psi  , sqrt(tau_psi  ));
+  count ~ neg_binomial_2_log(X*pad, 1/exp(psi));
+}
+"
+h_m = stan_model(model_code = heterosis_model)
+@
+\end{frame}
+
+\begin{frame}[fragile]
+<<heterosis_data, cache=TRUE>>=
+G = 100
+n = 4
+truth = data.frame(mu_phi   = 3, tau_phi   = 1, mu_psi = 0, tau_psi = 1,
+                   mu_alpha = 0, tau_alpha = 1, mu_delta = 0, tau_delta = .1^2)
+par = data.frame(gene=1:G, 
+                 phi   = with(truth, rnorm(G, mu_phi  , sqrt(tau_phi  ))),
+                 alpha = with(truth, rnorm(G, mu_alpha, sqrt(tau_alpha))),
+                 delta = with(truth, rnorm(G, mu_delta, sqrt(tau_delta))),
+                 psi   = with(truth, rnorm(G, mu_psi  , sqrt(tau_psi  ))))
+X = cbind(1, rep(c(-1,1,0), each=n), rep(c(0,0,1), each=n))
+d = ddply(par, .(gene), function(x) {
+  pad = with(x, c(phi,alpha,delta))
+  data.frame(y = rnbinom(3*n, mu = exp(X%*%pad), size=1/exp(x$psi)),
+             variety = rep(1:3, each=n))
+})
+@
+\end{frame}
+
+\begin{frame}[fragile]
+<<heterosis_mcmcem, dependson=c('heterosis_data','heterosis_model'), cache=TRUE, echo=FALSE, results='hide'>>=
+current = truth
+n_iter = 100
+keep = data.frame(iteration = 1:n_iter, 
+                  mu_phi    = rep(NA, n_iter), 
+                  mu_alpha  = rep(NA, n_iter), 
+                  mu_delta  = rep(NA, n_iter), 
+                  mu_psi    = rep(NA,n_iter),
+                  tau_phi   = rep(NA, n_iter),
+                  tau_alpha = rep(NA, n_iter), 
+                  tau_delta = rep(NA, n_iter),
+                  tau_psi   = rep(NA,n_iter))
+
+for (i in 1:n_iter) {
+  # Run MCMC using rstan
+  draws = ddply(d, .(gene), function(x) {
+    s = sampling(h_m, data = c(list(n = length(x$y), count=x$y, variety = x$variety), as.list(current)))
+    e = extract(s, c('phi','alpha','delta','psi'))
+    data.frame(phi   = as.numeric(e$phi),
+               alpha = as.numeric(e$alpha),
+               delta = as.numeric(e$delta),
+               psi   = as.numeric(e$psi))
+  })
+
+  current = data.frame(mu_phi = mean(draws$phi)) %>%
+    mutate(mu_alpha = mean(draws$alpha),
+           mu_delta = mean(draws$delta),
+           mu_psi   = mean(draws$psi),
+           tau_phi    = mean((draws$phi   - current$mu_phi)^2),
+           tau_alpha  = mean((draws$alpha - current$mu_alpha)^2),
+           tau_delta  = mean((draws$delta - current$mu_delta)^2),
+           tau_psi    = mean((draws$psi   - current$mu_psi)^2))
+
+  keep$mu_phi[i]   = current$mu_phi;   keep$tau_phi[i]   = current$tau_phi
+  keep$mu_alpha[i] = current$mu_alpha; keep$tau_alpha[i] = current$tau_alpha
+  keep$mu_delta[i] = current$mu_delta; keep$tau_delta[i] = current$tau_delta
+  keep$mu_psi[i]   = current$mu_psi;   keep$tau_psi[i]   = current$tau_psi
+}
+@
+
+<<dependson=c('normal_hierarchical_mean_em','normal_hierarchical_mean_mcem'), echo=FALSE>>=
+ggplot(melt(keep, id.var=c("iteration"), variable.name="parameter"),
+       aes(iteration, value)) +
+  geom_point() + 
+  facet_wrap(~parameter, scales="free_y", nrow=2) +
+  geom_hline(data=melt(truth, variable.name='parameter'), aes(yintercept=value))
+@
+\end{frame}
+
 \end{document}

### 2015-07-13 supp/results/combine.R
script to combine results of all simulations

@@ -1,0 +1,14 @@
+library(plyr)
+
+combined = ddply(expand.grid(r = c(4,8), i = 1:10), .(r,i), function(x) {  
+  d = readRDS(paste("results/results-",x$r,"-",x$i,".rds", sep=''))
+  data.frame(geneid = rownames(d),
+             sim = x$i, 
+             sample.size = x$r, 
+             phph = d$prob_HPH - d$prob_LPH, # temporary fix until model.stan gets fixed
+             plph = d$prob_LPH)
+})
+
+combined$r = combined$i = NULL
+
+

### 2015-07-13 supp/results/combine.R
file to combine results

@@ -1,7 +1,9 @@
+
+
 library(plyr)
 
 combined = ddply(expand.grid(r = c(4,8), i = 1:10), .(r,i), function(x) {  
-  d = readRDS(paste("results/results-",x$r,"-",x$i,".rds", sep=''))
+  d = readRDS(paste("results-",x$r,"-",x$i,".rds", sep=''))
   data.frame(geneid = rownames(d),
              sim = x$i, 
              sample.size = x$r, 
@@ -11,4 +13,4 @@ combined = ddply(expand.grid(r = c(4,8), i = 1:10), .(r,i), function(x) {
 
 combined$r = combined$i = NULL
 
-
+saveRDS(combined, file="eBayes_edgeR_stan.rds")

### 2015-07-14 ebayes_rnaseq_heterosis.tex
update volcano plot and name
up to discussion

@@
 The null hypotheses of greatest interest in testing for gene expression heterosis are composite null hypotheses that are difficult to test with standard statistical approaches for \RNAseq{} analysis.
-To address these shortcomings, we develop a hierarchical model based on a negativeb binomial data model and draw inferences using a computationally tractable empirical Bayes approach to inference.
+To address these shortcomings, we develop a hierarchical negative binomial model and draw inferences using a computationally tractable empirical Bayes approach to inference.
 We demonstrate improvements over alternative methods via a simulation study based on a maize experiment and then analyze that maize experiment with our newly proposed methodology.
@@
 These hypotheses, which many available methods do not fully accommodate, remain a challenge in the transition from continuous measurements of transcript abundance to count-based measurements that arise from RNA sequencing (\RNAseq{}) technology.
-Building on the work of \citeauthor{ji2014estimation} with the normal data model, we construct a hierarchical model based on a negative-binomial data model.
+Building on the work of \citeauthor{ji2014estimation} with the normal data model, we construct a hierarchical model based on a negative binomial data model.
 We also utilize an empirical Bayes approach to obtain \emph{maximum a posterior} (MAP) estimates of the hyperparameters and the posterior distributions for the gene-specific parameters conditional on those hyperparameters.
@@
 
-We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$, normalization factors $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, and hyperparameters $\pi = (\eta, \sigma)$ where $\eta=(\eta_\phi,\eta_\alpha,\eta_\delta)$ and $\sigma=(\sigma_\phi,\sigma_\alpha,\sigma_\delta,\sigma_\psi)$.
+We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$, normalization factors $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, and hyperparameters $\pi = (\eta, \sigma)$ where $\eta=(\eta_\phi,\eta_\alpha,\eta_\delta, \eta_\psi)$ and $\sigma=(\sigma_\phi,\sigma_\alpha,\sigma_\delta,\sigma_\psi)$.
 We obtain estimates for the hyperparameters and then base gene-specific inference on the posterior conditional on these estimates.
@@
 To obtain normalization factors $\hat{\gamma}$, we use the weighted trimmed mean of $M$ values (TMM) method of \cite{robinson2010scaling}.
-We then use the methods of \cite{mccarthy2012differential}, to obtain genewise dispersion estimates, $\hat{\phi}_g$, and their generalized linear model methods to obtain estimates for the remaining gene-specific parameters, i.e.
+We then use the methods of \cite{mccarthy2012differential}, to obtain genewise dispersion estimates, $\hat{\psi}_g$, and their generalized linear model methods to obtain estimates for the remaining gene-specific parameters, i.e.
 $\hat{\phi}_g$, $\hat{\alpha}_g$, and $\hat{\delta}_g$.
-Using the gene-specific parameter estimates, we estimate hyperparameters for the location and scale parameters in the hierarchical model using a central method of moments approach, e.g.
-$\hat{\eta}_\delta = \sum_{g=1}^G \hat{\delta}_g/G$ and $2\hat{\sigma}_\delta^2 = \sum_{g=1}^G (\hat{\delta}_g - \hat{\eta}_\delta)^2/(G-1)$.
+Using $\hat{\theta}_g = (\hat{\phi}_g,\hat{\alpha}_g,\hat{\delta}_g,\hat{\psi}_g)$, we estimate hyperparameters for the location and scale parameters in the hierarchical model using a central method of moments approach, e.g.
+$\hat{\eta}_\phi = \sum_{g=1}^G \hat{\phi}_g/G$ and $\hat{\sigma}_\phi^2 = \sum_{g=1}^G (\hat{\phi}_g - \hat{\eta}_\phi)^2/(G-1)$ (and similarly for $\hat{\eta}_\psi$ and $\hat{\sigma}_\psi$) and $\hat{\eta}_\alpha = \sum_{g=1}^G \hat{\alpha}_g/G$ and $2\hat{\sigma}_\alpha^2 = \sum_{g=1}^G (\hat{\alpha}_g - \hat{\eta}_\alpha)^2/(G-1)$ (and similarly for $\hat{\eta}_\delta$ and $\hat{\sigma}_\delta$).
 
@@
 
-To perform this Bayesian analysis, we used the statistical software \Stan which implements a variant of Markov chain Monte Carlo (MCMC) called Hamiltonion Monte Carlo (HMC) \citep{neal2011mcmc} to obtaining samples from the posterior in equation \eqref{e:condind}.
+To perform this Bayesian analysis, we used the statistical software \Stan{} which implements a variant of Markov chain Monte Carlo (MCMC) called Hamiltonion Monte Carlo (HMC) \citep{neal2011mcmc} to obtaining samples from the posterior in equation \eqref{e:condind}.
 We ran 4 simultaneous chains with random initial starting values for 1000 burnin (and tuning) iterations followed by another 1000 iterations thinned by 4 (to reduce storage space) for inference.
-We monitored convergence using the potential scale reduction factor (PSRF) \citep{Gelm:Rubi:infe:1992} for $\phi_g$, $\alpha_g$, $\delta_g$, and $\psi_g$.
-If the maximum PSRF was greater than 1.1, we reran the chain with an additional 500 iterations (250 burnin and 250 for inference) and repeated this process until all PSRF's were less than 1.1.
+We monitored convergence using the potential scale reduction factor and effective sample size (ESS) for $\phi_g$, $\alpha_g$, $\delta_g$, and $\psi_g$ \citep{Gelm:Rubi:infe:1992}.
+If the minimum ESS was less than 1000, we reran the chains with double the iterations for both burnin and inference.
+We continued this restarting and doubling until we obtained minimum ESS greater than 1000 for all parameters.
 
@@
 
-\subsection{Implementation in \ShrinkBayes{}}
+\subsection{Implementation in \ShrinkBayes{}} \label{s:shrinkbayes}
 
@@
 
-We applied the \edgeR{} package \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor} to the maize data on 27,888 genes with average count at least one and at most two zero read counts for each variety across four replications.
-This analysis produced sample-specific normalization factors using the TMM method \citep{robinson2010scaling},
-%(0.080, -0.041, -0.054, 0.000, -0.036, -0.147, 0.093, 0.059, -0.004, -0.057, 0.067,  0.041) 
-gene-specific dispersion estimates, and gene-variety-specific location parameter estimates \citep{mccarthy2012differential}.
-We treated these estimates as the true parameter values for the simulation study so that our simulated datasets mimicked the existing covariance structure among the gene-variety means of the original maize data.
+We used the methods described at the beginning of Section \ref{s:ebayes} to obtain normalization factors $\hat{\gamma}$ and gene-specific parameter estimates $\hat{\theta}_g$ for all genes using the \edgeR{} package \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor} applied to the maize data on 27,888 genes with average count at least one and at most two zero read counts for each variety across the four replicates.
+This analysis produced sample-specific normalization factors of $\hat{\gamma}=
+(0.074, -0.059, -0.074, -0.014, -0.014, -0.124,  0.093,  0.063,  0.021, -0.037,  0.049,  0.021)$.
+The  gene-specific parameter estimates were treated as the true parameter values for the simulation study so that our simulated datasets mimicked the existing structure among the gene-variety means of the original maize data.
+If $\hat{\delta}_g<-|\hat{\alpha}_g$ ($\hat{\delta}_g >|\hat{\alpha}_g|$)  the gene was considered to have LPH (HPH).
 
-Using these parameters and normalization factors, we simulated data according to the negative-binomial model in \eqref{e:data} independently for each gene.
+Using these parameters and normalization factors, we simulated data according to the negative binomial model in \eqref{e:data} independently for each gene.
 For each simulation, we analyzed a subset of 25,000 genes selected randomly from genes with simulated counts at least one on average and with at most two zeros for each variety across replications.
@@
 To construct a measure of heterosis, we computed the maximum likelihood estimates of the $\mu_{gv}$ parameters for all genes using \edgeR{}'s built-in Fisher scoring algorithm, and then used likelihood ratio tests to calculate two $p$-values for each gene: $p_{g1}$  for testing $H_{g01}:\mu_{g1}=\mu_{g3}$ and $p_{g2}$ for testing $H_{g02}:\mu_{g2}=\mu_{g3}$.
-Then, for each gene, we define a new $p$-value denoted as $p_g$ and determined as $p_g=1$ if the estimate of $\mu_{g3}$ falls between the estimates of $\mu_{g1}$ and $\mu_{g2}$ and $p_g=\max\{p_{g1},p_{g2}\}/2$ otherwise.
-(For all relevant signifcance thresholds $\omega$ near 0, it can be shown that rejecting the null hypothesis of no heterosis of expression for gene $g$ whenever $p_g\leq \omega$ results in a test that is asymptotically size $\omega$.)  We then use $1-p_g$ as a final measure positively associated with strength of evidence for gene expression heterosis.
+Then, for each gene, we define a new $p$-value denoted as $p_g$ and set to $p_g=1$ if the estimate of $\mu_{g3}$ falls between the estimates of $\mu_{g1}$ and $\mu_{g2}$ and $p_g=\max\{p_{g1},p_{g2}\}/2$ otherwise.
+(For all relevant significance thresholds $\omega$ near 0, it can be shown that rejecting the null hypothesis of no heterosis of expression for gene $g$ whenever $p_g\leq \omega$ results in a test that is asymptotically size $\omega$.)  We then use $1-p_g$ as a final measure positively associated with strength of evidence for gene expression heterosis.
 
@@
 We set this measure to zero if the maximum likelihood estimate, calculated using \edgeR{}, of $\mu_{g3}$ is between the maximum likelihood estimates of $\mu_{g1}$ and $\mu_{g2}$.
-Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the only hypotheses that allow for heterosis.
+Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the two hypotheses that allow for extreme heterosis.
 
@@
 \centerline{\includegraphics[width=\textwidth]{exampleROC0_1}}
-\caption{Representative ROC curves for false positive rates below 0.1 for the approaches using \edgeR{}, {\tt baySeq},  and \ShrinkBayes{} described in Section \ref{s:alternative}, along with the independence and dependence methods using \RStan{} described in Section \ref{s:gene_heterosis}.}
+\caption{Representative ROC curves for false positive rates below 0.1 for the approaches using \edgeR{} and \baySeq{} described in Section \ref{s:alternative}, the \ShrinkBayes{} approach described in Section \ref{s:shrinkbayes} and our eBayes approach described in Section \ref{s:ebayes}.}
 \label{f:roc}
@@
 For this simulation, we can see that the approaches based on the model in Section \ref{s:model}, i.e.
-\RStan{} and \ShrinkBayes{}, provide the best true positive rate for a given false positive rate.
+eBayes and \ShrinkBayes{}, provide the best true positive rate for a given false positive rate.
 Also, as expected, as the sample size increases, our ability to distinguish genes with heterosis from genes without improves.
@@
 \centerline{\includegraphics[width=\textwidth]{auc-facet-TRUE}}
-\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different sample sizes for the approaches using \edgeR{}, {\tt baySeq},  and \ShrinkBayes{} described in Section \ref{s:alternative} , along with the independence and dependence methods using \RStan{} described in Section \ref{s:gene_heterosis}.
-Each line is a different simulation while the blue box indicates mean AUC and mean AUC plus and minus one standard error across the simulations.}
+\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different replicates per variety for the approaches using \edgeR{} and \baySeq{} described in Section \ref{s:alternative}, the \ShrinkBayes{} approach described in Section \ref{s:shrinkbayes} and our eBayes approach described in Section \ref{s:ebayes}.
+Each line is a different simulation while the blue box indicates mean AUC (plus or minus one standard error).}
 \label{f:auc}
 \end{figure}
-The pattern appears to hold, with the \RStan{} and \ShrinkBayes{} techniques outperforming the other two.
-%While there does not appear to be a significant difference between \ShrinkBayes{} and the independence model using \RStan{}, performance for \RStan{} appears to decrease for the dependence model where an unstructured covariance is estimated, at least for smaller sample sizes.
+The pattern appears to hold, with the eBayes and \ShrinkBayes{} methods outperforming the other two.
+With 4 replicates per variety, there does not appear to be much of a difference between \ShrinkBayes{} and our eBayes approach, but as the number of replicates increases our eBayes approach appears to improve relative to \ShrinkBayes{}.
 
@@
 
-We used our method to analyze a maize data set \citep{paschold2012complementation} on gene expression in parental lines B73 and Mo17 and the hybrid genotype (B73$\times$Mo17).
+We used our method to analyze a maize data set \citep{paschold2012complementation} of \RNAseq{} gene expression in parental lines B73 and Mo17 and the hybrid genotype (B73$\times$Mo17).
 The data are available in the Supplemental Materials.
@@
 As explained in Section~\ref{s:sim_data}, we analyzed a set of $G=27,888$ genes for $V=3$ varieties and $n_v=4$ replicates ($v=1,2,3$).
-
-We found MAP estimates for both the independence and dependence models as shown in Table \ref{t:map}. 
-\input{include/map-laplace}
-The MAP estimates for the hierarchical means, $\eta$, and standard deviations, $\sigma$, are consistent between the independence and dependence models with, perhaps, a slightly larger mean and smaller standard deviation for the logarithm of the overdispersion parameter, $\psi_g$, in the dependence model relative to the independence model.
-In the dependence model, there appears to be high correlation amongst $\mu_{g1}$, $\mu_{g2}$, and $\mu_{g3}$, and negative correlation between $\psi_g$ and each of $\mu_{g1}$, $\mu_{g2}$, and $\mu_{g3}$.
-The normalization factors, and their variance, are consistent between the two models, with the estimates for the dependence model being consistently 0.02 larger than the estimates for the independence model, and both indicating that there is about a 25\% increase in counts between the least and most thoroughly sequenced samples.
-
-Independently for each gene, we ran four Markov chains conditional on the estimated hyperparameters to obtain approximate posterior distributions for all gene-specific parameters.
-Each Markov chain used initial values randomly drawn by {\tt Stan} and was run for a total of 2000 iterations, with 1000 iterations of burn-in and the remaining 1000 draws thinned by a factor of 4 for storage reasons.
-% From Dan "Do we need to be more specific about what method was used to generate the chains?"
-The potential scale reduction factor, $\hat{R}$, was computed for each parameter \citep{Gelm:Rubi:infe:1992, Broo:Gelm:gene:1997}.
-If the maximum $\hat{R}$ was larger than 1.1 for any parameter, the total number of iterations was increased by 500, with the burn-in increased proportionally, and the chain was rerun.
-This process was repeated until the maximum $\hat{R}$ was less than 1.1 for all parameters.
+We estimated the hyperparameters to be $\hat{\eta}_\psi = 3.44$, $\hat{\sigma}_\phi = 2.74$, $\hat{\eta}_\alpha = -0.05$, $\hat{\sigma}_\alpha=0.36$, $\hat{\eta}_\delta = 0.08$, $\hat{\sigma}_\delta=0.33$, $\hat{\eta}_\psi = -2.35$, and $\hat{\sigma}_\psi=0.60$.
 
@@
 0 & \mbox{otherwise},
-\end{array} 
+\end{array}.
 \right. 
@@
 \end{equation}
-where $\hat{\mu}_{gi}$ is the estimated posterior mean of $\mu_{gv}$ for $v=1,2,3$.
 Thus, the estimated effect size is the difference between hybrid mean and the nearest parent with negative values indicating LPH and positive values indicating HPH.
@@
 Figure \ref{f:volcano} provides a volcano plot, in this case a bivariate histogram, to visualize the maximum of the probabilities of LPH and HPH versus estimated effect size. 
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{volcano}}
+\caption{A bivariate histogram of the maximum of the LPH and HPH probabilities versus estimated effect size defined in equation \ref{e:effect_size} for the B73 $\times$ Mo17 maize experiment.}
+\label{f:volcano}
+\end{figure}
 The figure shows a peak at zero for probabilities below 0.5 with estimated effect sizes of zero.
@@
 
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{maize_volcano_dexp}}
-\caption{A bivariaet histogram of the maximum of the LPH and HPH probabilities versus estimated effect size defined in equation \ref{e:effect_size} for the B73 $\times$ Mo17 maize experiment.}
-\label{f:volcano}
-\end{figure}
+\section{Discussion}
+\label{s:discussion}
 
@@
 Figure \ref{f:scatterplot} provides marginal and bivariate histograms for the gene-specific parameters.
-
 \begin{figure}
-\centerline{\includegraphics[width=\textwidth]{scatterplot_dexp}}
+\centerline{\includegraphics[width=\textwidth]{estimates}}
 \caption{Marginal and bivariate histograms for gene-specific parameter posterior meansfor the B73 $\times$ Mo17 maize experiment.
@@
 
-Marginal and bivariate histograms for the model with an estimated covariance look similar although the estimated distribution $N(\hat{\eta},\hat{\Sigma})$ has increased variance for $\mu$ and decreased variance for $\psi$ as was previously seen in Table \ref{t:map} (results not shown).
-
-\section{Discussion}
-\label{s:discussion}
-
 Geneticists speculate that gene expression heterosis is one possible explanation of phenotypic heterosis of traits, such as plant height or grain yield.
@@
 We followed their approach, modified to allow for \RNAseq{} read counts as measures of transcript abundance.
-We developed an empirical Bayes approach based on obtaining MAP estimates for hyperparameters followed by MCMC to estimate gene-specific parameters.
-This methodology can be conveniently implemented in the statistical software, {\tt Stan}.
+We developed an empirical Bayes approach based on obtaining estimates for hyperparameters followed by MCMC to estimate gene-specific parameters.
 The empirical Bayes posteriors can be used to estimate posterior probabilities of high and low parent heterosis.
-
 Through a simulation study, we demonstrated that this method outperformed alternative methods (that are not designed to detect heterosis), and performed comparably well with an implementation of our model in \ShrinkBayes{}, which estimates the full posterior via INLA.
 We then demonstrated the use of the methodology on a maize experiment in which phenotypic heterosis is well known.
+
 % Estimates of heterosis to compare with Ji et. al.

### 2015-07-14 ebayes_rnaseq_heterosis.tex
through discussion

@@
 
-With almost 30,000 genes in our data set, we can evaluate the distribution assumptions for the gene-specific parameters.
-Figure \ref{f:scatterplot} provides marginal and bivariate histograms for the gene-specific parameters.
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{estimates}}
-\caption{Marginal and bivariate histograms for gene-specific parameter posterior meansfor the B73 $\times$ Mo17 maize experiment.
-For reference, the marginal histograms have a dashed line indicating the appropriate marginal probability density functions in the hierarchical model with the MAP estimates.}
-\label{f:scatterplot}
-\end{figure}
-
 Geneticists speculate that gene expression heterosis is one possible explanation of phenotypic heterosis of traits, such as plant height or grain yield.
@@
 
-% Estimates of heterosis to compare with Ji et. al.
-In addition to providing estimates of heterosis probabilities and effect sizes, this real data analysis exposed violations of our normality assumptions, indicating room to improve the model and increase the power to detect gene expression heterosis.
-Some of our ongoing and future research will develop the details of such an extension.
+Although our method appears to improve upon existing methods, we believe our approach can be improved by refining the hierarchical model for the gene-specific parameter distribution.
+Figure \ref{f:scatterplot} shows marginal and bivariate histograms for the empirical Bayes posterior means for the gene-specific parameters.
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{estimates}}
+\caption{Marginal and bivariate histograms for gene-specific parameter posterior meansfor the B73 $\times$ Mo17 maize experiment.
+For reference, the marginal histograms have a dashed line indicating the appropriate marginal probability density functions in the hierarchical model with the MAP estimates.}
+\label{f:scatterplot}
+\end{figure}
+These figures show markedly departure from marginal model assumptions, e.g.
+normality assumptions for $\phi_g$ and $\psi_g$, and independence assumptions, e.g.
+$\phi_g$ versus $\psi_g$ and $\alpha_g$ versus $\delta_g$.
+The plot of $\phi_g$ versus $\psi_g$ shows a pattern where the mean overdispersion decreases as the mean expression level increases.
+The plot of $\alpha_g$ versus $\delta_g$ shows a rotated V pattern where $\delta_g$ appears to be equal to $|\alpha_g|$.
+This pattern is consistent with Mendel's Law of Dominance where the hybrid has mean expression equal to the parent with higher mean expression.
+
+In addition to improving the hierarchical distribution, we believe better estimates of the parameters of this distribution, i.e.
+the hyperparameters, will also improve detection of gene-expression heterosis.
+Our current method, based on moment matching of essentially independently estimated gene-specific parameters provides consistent estimators as the number of replicates per variety increases.
+But typically the number of replicates per variety is quite small, in our data there are only four replicates per variety, and therefore assymptotic justifications appear deficient.
+There are a variety of hyperparameter estimation approaches to explore, e.g.
+expectation-maximization algorithms or fully Bayesian approaches.
+
+Notwithstanding these improvements, we believe our approach is a computationally efficient method that can immediately aid scientists who are interested in identifying candidate genes to explore a genetic mechanism of heterosis.
 

### 2015-07-14 supp/figs.R
create figs for real data

@@ -1,0 +1,51 @@
+library(ggplot2)
+
+d = readRDS("script.rds")
+
+d$prob_HPH = d$prob_HPH - d$prob_LPH # fix until model.stan is updated
+
+
+##########################################################################################
+# Volcano plot
+##########################################################################################
+ggplot(d, aes(effectiveSize, pmax(prob_LPH, prob_HPH))) + 
+  stat_binhex(color="white",bin=100) +
+  theme_bw() + 
+  scale_fill_gradientn(trans='log',breaks=c(1,10,100,1000), colours=c("gray","black")) +
+  labs(x="Effect size", y="Maximum heterosis probabilities")
+
+ggsave("../include/volcano.pdf", width=8, height=8)
+  
+
+##########################################################################################
+# Hyperparameter plot
+##########################################################################################
+parms = c("phi","alpha","delta","psi")
+estimates = d[,parms]
+plots = list()
+
+n = ncol(estimates)
+for (i in 1:n) {
+  for (j in 1:n) {
+    p = (i-1)*n+j # plot number
+    if (i>j) 
+      plots[[p]] = grid.rect(gp=gpar(col="white"))
+    if (i==j) {
+      plots[[p]] = ggplot(estimates, aes_string(parms[i])) + 
+        geom_histogram(fill='gray') +
+        theme_bw() + 
+        labs(x=parse(text=parms[i]))
+    }
+    if (i<j) {   
+      plots[[p]] = ggplot(estimates, aes_string(parms[j], parms[i])) + 
+        stat_binhex(bin=100) +
+        theme_bw() + 
+        scale_fill_gradientn(trans='log',breaks=c(1,10,100,1000), guide=FALSE, colours=c("gray","black")) + 
+        labs(x=parse(text=parms[j]), y=parse(text=parms[i]))
+    }
+  }
+}
+
+do.call("grid.arrange", c(plots, ncol=4))
+
+ggsave("../include/estimates.pdf", do.call("arrangeGrob", c(plots, ncol=4)), width=8, height=8)

### 2015-07-15 ebayes_rnaseq_heterosis.tex
typos

@@
 The  gene-specific parameter estimates were treated as the true parameter values for the simulation study so that our simulated datasets mimicked the existing structure among the gene-variety means of the original maize data.
-If $\hat{\delta}_g<-|\hat{\alpha}_g$ ($\hat{\delta}_g >|\hat{\alpha}_g|$)  the gene was considered to have LPH (HPH).
+If $\hat{\delta}_g<-|\hat{\alpha}_g|$ ($\hat{\delta}_g >|\hat{\alpha}_g|$)  the gene was considered to have LPH (HPH).
 
@@
 \centerline{\includegraphics[width=\textwidth]{volcano}}
-\caption{A bivariate histogram of the maximum of the LPH and HPH probabilities versus estimated effect size defined in equation \ref{e:effect_size} for the B73 $\times$ Mo17 maize experiment.}
+\caption{A bivariate histogram of the maximum of the LPH and HPH probabilities versus estimated effect size defined in equation \eqref{e:effect_size} for the B73 $\times$ Mo17 maize experiment.}
 \label{f:volcano}

### 2015-07-15 ebayes_rnaseq_heterosis.tex
post Dan comments

--- 2015-07-14 ebayes_rnaseq_heterosis.tex
@@
 \newcommand{\ShrinkBayes}{{\tt ShrinkBayes}}
-\newcommand{\RNAseq}{RNAseq}
+\newcommand{\RNAseq}{RNA-seq}
 
@@
 \begin{keywords}
-Empirical Bayes; Gene expression; Heterosis; Hierarchical model; Negative binomial; \RNAseq{}.
+Hierarchical model; Negative binomial; \RNAseq{}; Bayesian LASSO; Parallel computing; \Stan{}.
 \end{keywords}
@@
 Heterosis, or hybrid vigor, occurs when hybrid progeny display phenotypes that are superior to the phenotypes of their parents.
-The heterosis phenomenon was scientifically documented in plants by \cite{darwin1876effects}and has long been used to improve agricultural production.
+The heterosis phenomenon was scientifically documented in plants by \cite{darwin1876effects} and has long been used to improve agricultural production.
 One classic example involves hybrid maize offspring that are taller, faster to mature, and yield considerably more grain than their inbred parents \citep{hallauer1981quantitative, hallauer2010quantitative}.
@@
 Building on the work of \citeauthor{ji2014estimation} with the normal data model, we construct a hierarchical model based on a negative binomial data model.
-We also utilize an empirical Bayes approach to obtain \emph{maximum a posterior} (MAP) estimates of the hyperparameters and the posterior distributions for the gene-specific parameters conditional on those hyperparameters.
+We also utilize an empirical Bayes approach to obtain estimates of the hyperparameters and the posterior distributions for the gene-specific parameters conditional on those hyperparameters.
 
@@
 
-An RNA sequencing (\RNAseq{}) experiment to identify genes demonstrating expression heterosis consists of three genetic varieties: two parental varieties (P1 and P2) and a cross between these two varieties called the hybrid (H).
+We considering an RNA sequencing (\RNAseq{}) experiment that involves at least three genetic varieties: two parental varieties (P1 and P2) and a cross between these two varieties called the hybrid (H).
 For each variety, replicate RNA samples are isolated and assessed for quality.
@@
 
-To use \RNAseq{} counts to identify genes displaying expression heterosis, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the parameters using an empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate empirical Bayes posterior probabilities for heterosis.
+To use \RNAseq{} counts to identify genes displaying expression heterosis, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the parameters using an empirical Bayes procedure implemented in the statistical software \Stan{}, and calculate empirical Bayes posterior probabilities for heterosis.
 
@@
 \begin{equation} 
-Y_{gvi} \stackrel{ind}{\sim} NB\left(e^{\mu_{gv}+\gamma_{vi}},e^{\psi_g}\right), 
+Y_{gvi} \stackrel{ind}{\sim} \mbox{NB}\left(e^{\mu_{gv}+\gamma_{vi}},e^{\psi_g}\right), 
 \label{e:data}
 \end{equation}
-where $NB(\xi,e^\psi)$ indicates a negative binomial distribution with expectation $\xi$ and variance $\xi+e^\psi\xi^2$, and $ind$ indicates the observations are conditionally independent.
+where NB$(\xi,e^\psi)$ indicates a negative binomial distribution with expectation $\xi$ and variance $\xi+e^\psi\xi^2$, and $ind$ indicates the observations are conditionally independent.
 As shown in equation \eqref{e:data}, our data model involves gene-specific overdispersion $\psi_g$ and a mean that depends on the gene-variety combination through $\mu_{gv}$ and on the sample though  $\gamma_{vi}$.
@@
 Following \cite{ji2014estimation}, we reparameterize the gene-variety mean structure into the gene-specific parental average ($\phi_g$), half-parental difference ($\alpha_g$), and hybrid difference ($\delta_g$).
-Specifically, for heterosis $V=3$ and let $v=1,2$ indicate the two parental varieties and $v=3$ indicate the hybrid, then
+Specifically, for our heterosis where $V=3$, we let $v=1,2$ indicate the two parental varieties and $v=3$ indicate the hybrid, then
 \[ \phi_g = \frac{\mu_{g1}+\mu_{g2}}{2}, \quad 
@@
 \]
-where $\alpha\sim La(\eta,\sigma)$ has a probability density function given by $p(\alpha|\eta,\sigma) = \exp(-|\alpha-\eta|/\sigma)/2\sigma$ with expectation $\eta$ and variance $2\sigma^2$.
+where $\alpha\sim La(\eta,\sigma)$ has a probability density function given by $La(\alpha;\eta,\sigma) = \exp(-|\alpha-\eta|/\sigma)/2\sigma$ with expectation $\eta$ and variance $2\sigma^2$.
 We assume independence amongst the parental averages, half parental differences, hybrid differences, and overdispersion parameters.
@@
 
-To perform this Bayesian analysis, we used the statistical software \Stan{} which implements a variant of Markov chain Monte Carlo (MCMC) called Hamiltonion Monte Carlo (HMC) \citep{neal2011mcmc} to obtaining samples from the posterior in equation \eqref{e:condind}.
+To perform this Bayesian analysis, we used the statistical software \Stan{} which implements a variant of Markov chain Monte Carlo (MCMC) called Hamiltonion Monte Carlo (HMC) \citep{neal2011mcmc} to obtain samples from the posterior in equation \eqref{e:condind}.
 We ran 4 simultaneous chains with random initial starting values for 1000 burnin (and tuning) iterations followed by another 1000 iterations thinned by 4 (to reduce storage space) for inference.
@@
 %P(LPH_g|y) \approx 
-P(LPH_g|y,\hat{\pi},\hat{\gamma}) = P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi},\hat{\gamma}\right) = \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right),
+P(\mbox{LPH}_g|y,\hat{\pi},\hat{\gamma}) = P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi},\hat{\gamma}\right) = \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right),
 \label{e:probs}
@@
 % \end{array} \right.
-P_{\ShrinkBayes{}}(LPH_g|y) \equiv P(\delta_g>-\alpha_g|y)\mathrm{I}\left(\hat{\delta}_g > -\hat{\alpha}_g \ge 0\right) + P(\delta_g>\alpha_g|y) \mathrm{I}\left(\hat{\delta}_g > \hat{\alpha}_g \ge 0\right).
+P_{\ShrinkBayes{}}(\mbox{LPH}_g|y) \equiv P(\delta_g<-\alpha_g|y)\mathrm{I}\left(\hat{\delta}_g < -\hat{\alpha}_g \le 0\right) + P(\delta_g<\alpha_g|y) \mathrm{I}\left(\hat{\delta}_g < \hat{\alpha}_g \ge 0\right).
 %P_{\ShrinkBayes{}}(HPH|y) \equiv \left\{ \begin{array}{ll}
@@
 
-For each of the methods above, as well as the method described in Section \ref{s:gene_heterosis}, we sorted genes according to the computed measure of the strength of evidence for heterosis.
+For the methods in Sections \ref{s:ebayes}, \ref{s:shrinkbayes}, and \ref{s:alternative}, we sorted genes according to the computed measure of the strength of evidence for heterosis.
 From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to distinguish genes with heterosis, as defined in Section \ref{s:sim_data},
@@
 \end{figure}
-For this simulation, we can see that the approaches based on the model in Section \ref{s:model}, i.e.
+For this simulation, we can see that the approaches based most closely on the model in Section \ref{s:model}, i.e.
 eBayes and \ShrinkBayes{}, provide the best true positive rate for a given false positive rate.
@@
 As explained in Section~\ref{s:sim_data}, we analyzed a set of $G=27,888$ genes for $V=3$ varieties and $n_v=4$ replicates ($v=1,2,3$).
-We estimated the hyperparameters to be $\hat{\eta}_\psi = 3.44$, $\hat{\sigma}_\phi = 2.74$, $\hat{\eta}_\alpha = -0.05$, $\hat{\sigma}_\alpha=0.36$, $\hat{\eta}_\delta = 0.08$, $\hat{\sigma}_\delta=0.33$, $\hat{\eta}_\psi = -2.35$, and $\hat{\sigma}_\psi=0.60$.
+We estimated the hyperparameters to be $\hat{\eta}_\phi = 3.44$, $\hat{\sigma}_\phi = 2.74$, $\hat{\eta}_\alpha = -0.05$, $\hat{\sigma}_\alpha=0.36$, $\hat{\eta}_\delta = 0.08$, $\hat{\sigma}_\delta=0.33$, $\hat{\eta}_\psi = -2.35$, and $\hat{\sigma}_\psi=0.60$.
 
@@
 The empirical Bayes posteriors can be used to estimate posterior probabilities of high and low parent heterosis.
-Through a simulation study, we demonstrated that this method outperformed alternative methods (that are not designed to detect heterosis), and performed comparably well with an implementation of our model in \ShrinkBayes{}, which estimates the full posterior via INLA.
+Through a simulation study, we demonstrated that this method outperformed alternative methods (that are not designed to detect heterosis), and performed comparably well with a similar model in \ShrinkBayes{}, which estimates the full posterior via INLA.
 We then demonstrated the use of the methodology on a maize experiment in which phenotypic heterosis is well known.
@@
 Our current method, based on moment matching of essentially independently estimated gene-specific parameters provides consistent estimators as the number of replicates per variety increases.
-But typically the number of replicates per variety is quite small, in our data there are only four replicates per variety, and therefore assymptotic justifications appear deficient.
+But typically the number of replicates per variety is quite small, in our data there are only four replicates per variety, and therefore assymptotic justifications are deficient.
 There are a variety of hyperparameter estimation approaches to explore, e.g.
@@
 
-The supplementary materials include three plain text files: the maize experiment data ({\tt data.csv}), a script to run our method ({\tt script.R}) on this data, and the {\tt Stan} model ({\tt model.stan}).
+The supplementary materials include three plain text files: the maize experiment data ({\tt data.csv}), a script to run our method ({\tt script.R}) on this data, and the \Stan{} model ({\tt model.stan}).
 %the all-genes-independent (\verb@model_ind.txt@) and covariance (\verb@model_cov.txt@) models used in obtaining MAP estimates, along with the single-gene-independent (\verb@sg_model_ind.txt@) and covariance (\verb@sg_model_cov.txt@) models used in performing the MCMC on single genes.

### 2015-07-15 supp/figs.R
fixed HPH probability calculation

@@ -1,9 +1,7 @@
 library(ggplot2)
+library(gridExtra)
 
 d = readRDS("script.rds")
-
-d$prob_HPH = d$prob_HPH - d$prob_LPH # fix until model.stan is updated
-
 
 ##########################################################################################
 # Volcano plot

### 2015-07-15 supp/results/combine.R

@@ -7,7 +7,7 @@ combined = ddply(expand.grid(r = c(4,8), i = 1:10), .(r,i), function(x) {
   data.frame(geneid = rownames(d),
              sim = x$i, 
              sample.size = x$r, 
-             phph = d$prob_HPH - d$prob_LPH, # temporary fix until model.stan gets fixed
+             phph = d$prob_HPH, 
              plph = d$prob_LPH)
 })
 

### 2015-07-20 ebayes_rnaseq_heterosis.tex
more Dan edits
blinded version

@@
 
-\newcommand{\blind}{0} % 0 = not blind, 1 = blind
+\newcommand{\blind}{1} % 0 = not blind, 1 = blind
 
@@
 
-\date{{\it Received December} 2014}
+\date{{\it Received July} 2015}
 
@@
 We demonstrate improvements over alternative methods via a simulation study based on a maize experiment and then analyze that maize experiment with our newly proposed methodology.
-This article has supplementary material online.
+% This article has supplementary material online.
 \end{abstract}
@@
 \phi_g \stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2), \quad \mbox{and}\quad 
-\psi_g \stackrel{ind}{\sim} N(0,\sigma_\psi^2). 
+\psi_g \stackrel{ind}{\sim} N(\eta_\psi,\sigma_\psi^2). 
 \]
@@
 %&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right.\right) \right] p(\theta_g|\hat{\eta},\hat{\sigma}) \\
-\propto \prod_{g=1}^G & \left[ \,\prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi};e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right)  \right. \\
+\propto \prod_{g=1}^G & \left[ \,\prod_{v=1}^V \prod_{i=1}^{n_v} \mbox{NB}\left(y_{gvi};e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right)  \right. \\
 &\times 
@@
 
+In addition to the approach above, we utilized \ShrinkBayes{} to estimate heterosis probabilities with two modifications described here.
 \ShrinkBayes{} utilizes integrated nested Laplace approximation (INLA) \citep{rue2009approximate} in combination with empirical Bayes ideas \citep{van2014shrinkbayes}.
@@
 
-\section*{Supplementary Materials}
+%\section*{Supplementary Materials}
 
-The supplementary materials include three plain text files: the maize experiment data ({\tt data.csv}), a script to run our method ({\tt script.R}) on this data, and the \Stan{} model ({\tt model.stan}).
-%the all-genes-independent (\verb@model_ind.txt@) and covariance (\verb@model_cov.txt@) models used in obtaining MAP estimates, along with the single-gene-independent (\verb@sg_model_ind.txt@) and covariance (\verb@sg_model_cov.txt@) models used in performing the MCMC on single genes.
+%The supplementary materials include three plain text files: the maize experiment data ({\tt data.csv}), a script to run our method ({\tt script.R}) on this data, and the \Stan{} model ({\tt model.stan}).
 

### 2015-08-11 ebayes_rnaseq_heterosis.tex
blinded; updated date

@@
 We demonstrate improvements over alternative methods via a simulation study based on a maize experiment and then analyze that maize experiment with our newly proposed methodology.
-% This article has supplementary material online.
+This article has supplementary material online.
 \end{abstract}
@@
 
-Notwithstanding these improvements, we believe our approach is a computationally efficient method that can immediately aid scientists who are interested in identifying candidate genes to explore a genetic mechanism of heterosis.
+Notwithstanding these improvements, we believe our approach is a computationally efficient method that can immediately aid scientists who are interested in identifying candidate genes for a genetic mechanism of heterosis.
 
@@
 
-%\section*{Supplementary Materials}
+\section*{Supplementary Materials}
 
-%The supplementary materials include three plain text files: the maize experiment data ({\tt data.csv}), a script to run our method ({\tt script.R}) on this data, and the \Stan{} model ({\tt model.stan}).
+The supplementary materials include these plain text files: the maize experiment data ({\tt data.csv}), a script to run our method ({\tt script.R}) on these data, a script to create the figures ({\tt figs.R}), two functions ({\tt get\_hyperparameters.R} and {\tt single\_gene\_analysis.R}), and the \Stan{} model ({\tt model.stan}).
 

### 2015-08-11 supp/make_Makefile.R
add zip target

@@ -10,9 +10,12 @@ screen_width = 60
 cat(rep('#', screen_width), '\n', sep='', file=makefile, append=FALSE)
 cat('#     DO NOT TOUCH \n', file=makefile, append=TRUE)
 cat(rep('#', screen_width), '\n', sep='', file=makefile, append=TRUE)
-cat('# Automatically created using make_makefile.R\n\n', 
+cat('# Automatically created using make_Makefile.R\n\n', 
     file=makefile, 
     append=TRUE)
+
+cat('files = README.txt script.R figs.R get_hyperparameters.R single_gene_analysis.R model.stan\n\nzip: $(files); zip supp.zip $(files)
+', file=makefile, append=TRUE)
 
 for (r in n_reps) {
   cat(rep("#",screen_width), "\n", sep='', file=makefile, append=TRUE)
@@ -20,7 +23,7 @@ for (r in n_reps) {
   cat(rep("#",screen_width), "\n", sep='', file=makefile, append=TRUE)  
   
   # Vectors of files
-  result_file = paste0("results/results-", r, "-", 1:n_sims,'.rds')
+  result_file = paste0("results-2/results-", r, "-", 1:n_sims,'.rds')
   data_file   = paste0("sim-data/sim-", r, "-", 1:n_sims,'.rds')
   Rout_file   = paste0("sim-", r, "-", 1:n_sims,'.Rout')
   
@@ -30,8 +33,8 @@ for (r in n_reps) {
     append=TRUE)
   
   for (i in 1:n_sims) {
-    cat(result_file[i], ": sim-script.R ", data_file[i], "\n\t", 
-        r_command, " '--args r=", r, " i=", i, "' sim-script.R ",
+    cat(result_file[i], ": sim-script2.R ", data_file[i], "\n\t", 
+        r_command, " '--args r=", r, " i=", i, "' sim-script2.R ",
         Rout_file[i], " \n\n",
         sep='',
         file=makefile,

### 2015-08-11 supp/make_Makefile.R
revert emittman changes

@@ -23,7 +23,7 @@ for (r in n_reps) {
   cat(rep("#",screen_width), "\n", sep='', file=makefile, append=TRUE)  
   
   # Vectors of files
-  result_file = paste0("results-2/results-", r, "-", 1:n_sims,'.rds')
+  result_file = paste0("results/results-", r, "-", 1:n_sims,'.rds')
   data_file   = paste0("sim-data/sim-", r, "-", 1:n_sims,'.rds')
   Rout_file   = paste0("sim-", r, "-", 1:n_sims,'.Rout')
   
@@ -33,7 +33,7 @@ for (r in n_reps) {
     append=TRUE)
   
   for (i in 1:n_sims) {
-    cat(result_file[i], ": sim-script2.R ", data_file[i], "\n\t", 
+    cat(result_file[i], ": sim-script.R ", data_file[i], "\n\t", 
         r_command, " '--args r=", r, " i=", i, "' sim-script2.R ",
         Rout_file[i], " \n\n",
         sep='',

### 2015-08-11 supp/make_Makefile.R
now reverted all emittman changes

@@ -34,7 +34,7 @@ for (r in n_reps) {
   
   for (i in 1:n_sims) {
     cat(result_file[i], ": sim-script.R ", data_file[i], "\n\t", 
-        r_command, " '--args r=", r, " i=", i, "' sim-script2.R ",
+        r_command, " '--args r=", r, " i=", i, "' sim-script.R ",
         Rout_file[i], " \n\n",
         sep='',
         file=makefile,

### 2015-08-11 supp/make_Makefile.R
added catf function; added normal sub targets

@@ -7,44 +7,46 @@ r_command = 'R CMD BATCH --vanilla'
 
 screen_width = 60
 
-cat(rep('#', screen_width), '\n', sep='', file=makefile, append=FALSE)
-cat('#     DO NOT TOUCH \n', file=makefile, append=TRUE)
-cat(rep('#', screen_width), '\n', sep='', file=makefile, append=TRUE)
-cat('# Automatically created using make_Makefile.R\n\n', 
-    file=makefile, 
-    append=TRUE)
+catf = function(...) cat(..., sep='', file=makefile, append=TRUE)
 
-cat('files = README.txt script.R figs.R get_hyperparameters.R single_gene_analysis.R model.stan\n\nzip: $(files); zip supp.zip $(files)
-', file=makefile, append=TRUE)
+cat(rep('#', screen_width), '\n', file=makefile, append=FALSE)
+catf('#     DO NOT TOUCH \n')
+catf(rep('#', screen_width), '\n')
+catf('# Automatically created using make_Makefile.R\n\n')
 
+# Main targets
+methods = c('laplace')
+
+catf('all: ', paste(methods, collapse=' '), '\n\n')
+
+catf('laplace: ', paste('laplace',n_reps, sep='', collapse=' '), '\n\n')
+
+catf('files = README.txt script.R figs.R get_hyperparameters.R single_gene_analysis.R model.stan\n\nzip: $(files); zip supp.zip $(files)\n\n')
+
+
+# Sub targets
+
+for (m in methods) {
 for (r in n_reps) {
-  cat(rep("#",screen_width), "\n", sep='', file=makefile, append=TRUE)
-  cat("# Targets for ", r, "reps per variety\n", file=makefile, append=TRUE)
-  cat(rep("#",screen_width), "\n", sep='', file=makefile, append=TRUE)  
+  catf(rep('#',screen_width), '\n')
+  catf('# Targets for ', r, 'reps per variety\n')
+  catf(rep('#',screen_width), '\n')  
   
   # Vectors of files
-  result_file = paste0("results/results-", r, "-", 1:n_sims,'.rds')
-  data_file   = paste0("sim-data/sim-", r, "-", 1:n_sims,'.rds')
-  Rout_file   = paste0("sim-", r, "-", 1:n_sims,'.Rout')
+  result_file = paste0('results/',m,'/results-', r, '-', 1:n_sims,'.rds')
[…]

### 2015-08-11 supp/make_Makefile.R
indented

@@ -15,7 +15,7 @@ catf(rep('#', screen_width), '\n')
 catf('# Automatically created using make_Makefile.R\n\n')
 
 # Main targets
-methods = c('laplace')
+methods = c('laplace','normal')
 
 catf('all: ', paste(methods, collapse=' '), '\n\n')
 
@@ -27,26 +27,26 @@ catf('files = README.txt script.R figs.R get_hyperparameters.R single_gene_analy
 # Sub targets
 
 for (m in methods) {
-for (r in n_reps) {
-  catf(rep('#',screen_width), '\n')
-  catf('# Targets for ', r, 'reps per variety\n')
-  catf(rep('#',screen_width), '\n')  
-  
-  # Vectors of files
-  result_file = paste0('results/',m,'/results-', r, '-', 1:n_sims,'.rds')
-  data_file   = paste0('sim-data/sim-',          r, '-', 1:n_sims,'.rds')
-  Rout_file   = paste0('Rout/',m,'/sim-',        r, '-', 1:n_sims,'.Rout')
-  
-  catf(m,r, ': ', paste(result_file, collapse=' '),'\n\n')
-  
-  for (i in 1:n_sims) {
-    catf(result_file[i], ": sim-script.R ", data_file[i], "\n\t", 
-        r_command, " '--args r=", r, " i=", i, "' sim-script.R ",
-        Rout_file[i], " \n\n")
+  for (r in n_reps) {
+    catf(rep('#',screen_width), '\n')
+    catf('# Targets for ', r, 'reps per variety\n')
+    catf(rep('#',screen_width), '\n')  
+    
+    # Vectors of files
+    result_file = paste0('results/',m,'/results-', r, '-', 1:n_sims,'.rds')
+    data_file   = paste0('sim-data/sim-',          r, '-', 1:n_sims,'.rds')
+    Rout_file   = paste0('Rout/',m,'/sim-',        r, '-', 1:n_sims,'.Rout')
+    
+    catf(m,r, ': ', paste(result_file, collapse=' '),'\n\n')
+    
+    for (i in 1:n_sims) {
+      catf(result_file[i], ": sim-script.R ", data_file[i], "\n\t", 
+           r_command, " '--args r=", r, " i=", i, "' sim-script.R ",
+           Rout_file[i], " \n\n")
+    }
+    
+    catf("clean",r,":\n\trm -fv ", 
+         paste(result_file, collapse=' '),"\n\n")
[…]

### 2015-08-11 supp/make_Makefile.R
add normal main target

@@ -19,7 +19,9 @@ methods = c('laplace','normal')
 
 catf('all: ', paste(methods, collapse=' '), '\n\n')
 
-catf('laplace: ', paste('laplace',n_reps, sep='', collapse=' '), '\n\n')
+for (m in methods) {
+  catf(m, ': ', paste(m, n_reps, sep='', collapse=' '), '\n\n')
+}
 
 catf('files = README.txt script.R figs.R get_hyperparameters.R single_gene_analysis.R model.stan\n\nzip: $(files); zip supp.zip $(files)\n\n')
 
@@ -37,7 +39,7 @@ for (m in methods) {
     data_file   = paste0('sim-data/sim-',          r, '-', 1:n_sims,'.rds')
     Rout_file   = paste0('Rout/',m,'/sim-',        r, '-', 1:n_sims,'.Rout')
     
-    catf(m,r, ': ', paste(result_file, collapse=' '),'\n\n')
+    catf(m, r, ': ', paste(result_file, collapse=' '),'\n\n')
     
     for (i in 1:n_sims) {
       catf(result_file[i], ": sim-script.R ", data_file[i], "\n\t", 
@@ -45,7 +47,7 @@ for (m in methods) {
            Rout_file[i], " \n\n")
     }
     
-    catf("clean",r,":\n\trm -fv ", 
+    catf("clean-", m, r,":\n\trm -fv ", 
          paste(result_file, collapse=' '),"\n\n")
   }
 }

### 2015-08-11 supp/get_hyperparameters.R
I believe normal model will run now

@@ -1,4 +1,4 @@
-get_hyperparameters = function(d, variety) {
+get_hyperparameters = function(d, variety, method) {
   # phi, alpha, delta parameterization
   design = cbind(1,
                  c(1,-1,0)[as.numeric(variety)],
@@ -24,13 +24,19 @@ get_hyperparameters = function(d, variety) {
     group_by(parameter) %>%
     summarize(mean=mean(value), sd=sd(value))
   
+  # Fix sd for Laplace priors
+  if (method=='laplace') {
+    ss$sd[ss$parameter=='alpha'] = ss$sd[ss$parameter=='alpha']/sqrt(2)
+    ss$sd[ss$parameter=='delta'] = ss$sd[ss$parameter=='delta']/sqrt(2)
+  }
+  
   list(eta_phi   = ss$mean[ss$parameter=='phi'],
        eta_alpha = ss$mean[ss$parameter=='alpha'],
        eta_delta = ss$mean[ss$parameter=='delta'],
        eta_psi   = ss$mean[ss$parameter=='psi'],
        sigma_phi   = ss$sd[ss$parameter=='phi'],
-       sigma_alpha = ss$sd[ss$parameter=='alpha']/sqrt(2), # Fix sd for Laplace priors
-       sigma_delta = ss$sd[ss$parameter=='delta']/sqrt(2), # Fix sd for Laplace priors
+       sigma_alpha = ss$sd[ss$parameter=='alpha'], 
+       sigma_delta = ss$sd[ss$parameter=='delta'], 
        sigma_psi   = ss$sd[ss$parameter=='psi'],
        c = fit$offset[1,] - mean(fit$offset[1,]))
 }

### 2015-08-11 supp/make_Makefile.R

@@ -43,7 +43,7 @@ for (m in methods) {
     
     for (i in 1:n_sims) {
       catf(result_file[i], ": sim-script.R ", data_file[i], "\n\t", 
-           r_command, " '--args r=", r, " i=", i, "' sim-script.R ",
+           r_command, " '--args r=", r, " i=", i, ," m=", m, "' sim-script.R ",
            Rout_file[i], " \n\n")
     }
     

### 2015-08-11 supp/sim-script.R
I believe normal model will run now

@@ -4,11 +4,18 @@ library(dplyr)
 library(reshape2)
 library(edgeR)
 
+# Need to pass as arguments (or assign)
+# r: number of replicates per variety (4, 8, 16)
+# i: simulation number (1-10)
+# m: statistical method ('laplace' or 'normal')
+
+
 source("get_hyperparameters.R")
 source("single_gene_analysis.R")
 
 # Compile stan model
-model = stan_model("model.stan")
+if (m == 'laplace') model = stan_model("laplace.stan")
+if (m == 'normal' ) model = stan_model("normal.stan")
 
 # Use command line arguments
 # R CMD BATCH --vanilla '--args i=1' sim-script.R 
@@ -32,7 +39,7 @@ variety = factor(gsub("_[0-9]{1,2}", "", names(d)), levels=c("B73","Mo17","B73xM
 
 # trim genes for average count greater than 1
 # d = d[which(rowMeans(d)>1),] 
-hyperparameters = get_hyperparameters(d, variety)
+hyperparameters = get_hyperparameters(d, variety, method=m)
 
 
 

### 2015-08-11 supp/make_Makefile.R
clean working

@@ -9,7 +9,7 @@ screen_width = 60
 
 catf = function(...) cat(..., sep='', file=makefile, append=TRUE)
 
-cat(rep('#', screen_width), '\n', file=makefile, append=FALSE)
+cat(rep('#', screen_width), '\n', sep='', file=makefile, append=FALSE)
 catf('#     DO NOT TOUCH \n')
 catf(rep('#', screen_width), '\n')
 catf('# Automatically created using make_Makefile.R\n\n')
@@ -17,13 +17,21 @@ catf('# Automatically created using make_Makefile.R\n\n')
 # Main targets
 methods = c('laplace','normal')
 
-catf('all: ', paste(methods, collapse=' '), '\n\n')
+catf('all: ', paste(methods, collapse=' '), '\n')
 
 for (m in methods) {
-  catf(m, ': ', paste(m, n_reps, sep='', collapse=' '), '\n\n')
+  catf(m, ': ', paste(m, n_reps, sep='', collapse=' '), '\n')
 }
 
-catf('files = README.txt script.R figs.R get_hyperparameters.R single_gene_analysis.R model.stan\n\nzip: $(files); zip supp.zip $(files)\n\n')
+
+catf('\nclean: ', paste('clean-', methods, sep='', collapse=' '), '\n')
+
+for (m in methods) {
+  catf('clean-', m, ': ', paste('clean-', m, n_reps, sep='', collapse=' '), '\n')
+}
+
+
+catf('\nfiles = README.txt script.R figs.R get_hyperparameters.R single_gene_analysis.R model.stan\n\nzip: $(files); zip supp.zip $(files)\n\n')
 
 
 # Sub targets
@@ -31,7 +39,7 @@ catf('files = README.txt script.R figs.R get_hyperparameters.R single_gene_analy
 for (m in methods) {
   for (r in n_reps) {
     catf(rep('#',screen_width), '\n')
-    catf('# Targets for ', r, 'reps per variety\n')
+    catf('# Targets for ', r, ' reps per variety using method ', m, '\n')
     catf(rep('#',screen_width), '\n')  
     
     # Vectors of files
@@ -43,12 +51,12 @@ for (m in methods) {
     
     for (i in 1:n_sims) {
       catf(result_file[i], ": sim-script.R ", data_file[i], "\n\t", 
-           r_command, " '--args r=", r, " i=", i, ," m=", m, "' sim-script.R ",
[…]

### 2015-08-11 supp/results/combine.R
added 16 reps to combine.R

@@ -2,7 +2,7 @@
 
 library(plyr)
 
-combined = ddply(expand.grid(r = c(4,8), i = 1:10), .(r,i), function(x) {  
+combined = ddply(expand.grid(r = c(4,8,16), i = 1:10), .(r,i), function(x) {  
   d = readRDS(paste("results-",x$r,"-",x$i,".rds", sep=''))
   data.frame(geneid = rownames(d),
              sim = x$i, 

### 2015-08-11 ebayes_rnaseq_heterosis.tex

--- 2015-07-20 ebayes_rnaseq_heterosis.tex
@@
 %\section*{Supplementary Materials}
-
 %The supplementary materials include three plain text files: the maize experiment data ({\tt data.csv}), a script to run our method ({\tt script.R}) on this data, and the \Stan{} model ({\tt model.stan}).

### 2015-08-11 supp/sim-script.R
extract command line args first

@@ -9,16 +9,7 @@ library(edgeR)
 # i: simulation number (1-10)
 # m: statistical method ('laplace' or 'normal')
 
-
-source("get_hyperparameters.R")
-source("single_gene_analysis.R")
-
-# Compile stan model
-if (m == 'laplace') model = stan_model("laplace.stan")
-if (m == 'normal' ) model = stan_model("normal.stan")
-
-# Use command line arguments
-# R CMD BATCH --vanilla '--args i=1' sim-script.R 
+# R CMD BATCH --vanilla '--args r=4 i=1 m ="laplace"' sim-script.R 
 args=(commandArgs(TRUE))
 
 if(length(args)==0){
@@ -30,6 +21,17 @@ if(length(args)==0){
     eval(parse(text=args[[i]]))
   }
 }
+
+
+
+
+source("get_hyperparameters.R")
+source("single_gene_analysis.R")
+
+# Compile stan model
+if (m == 'laplace') model = stan_model("laplace.stan")
+if (m == 'normal' ) model = stan_model("normal.stan")
+
 
 data_file = paste0("sim-data/sim-", r, "-", i,'.rds')
 

### 2015-08-17 supp/sim-script.R
changed iterator

@@ -17,8 +17,8 @@ if(length(args)==0){
   ##supply default values
   i = 1
 }else{
-  for(i in 1:length(args)){
-    eval(parse(text=args[[i]]))
+  for(ii in 1:length(args)){
+    eval(parse(text=args[[ii]]))
   }
 }
 

### 2015-08-18 response.tex
added editor/reviewer comments

\documentclass{article}

\newcommand{\comment}[1]{\textit{#1}}
\newcommand{\response}[1]{#1}

\begin{document}

Editor comments:

\comment{The proposed model appears to be an extension of Ji et al. (2014) albeit with a differently specified likelihood function (negative binomial or Poisson-gamma) as perceived to be more appropriate for count data relative to the Gaussian likelihood function pursued in Ji et al. (2014). Both reviewers indicate that they are little underwhelmed by the performance of your proposed procedure relative to some of the existing methods that you compare against (i.e., ShrinkBayes, edgeR and baySeq).  Furthermore, they seem somewhat concerned about an ‘apples vs. oranges’ types of comparisons since the proposed hypothesis testing strategy in the proposed method is really not an option in the other methods.
Perhaps a reference to the promise of allele specific expression analyses (Discussion section) for inferring upon gene-expression heterosis might be useful to point readers into other directions with this work (Bell et al., 2013; Wei and Wang, 2013).}

\comment{I realize that Ji et al.’s (2014) work was intended for microarray data, or more specifically Gaussian likelihoods, but there has been some work that suggests that Gaussian likelihood specifications in MAANOVA may, oddly enough, have better properties for the analysis of RNA-seq data than more popular methods specifically intended for count data (edgeR and DESeq).  See Reeb and Steibel (2013) and also comments from one reviewer along these lines.  Since Ji et al.’s work also derives from ISU, have you assessed its performance relative to your proposed method on RNA-seq data?}

\comment{One seemingly glaring omission from this work is any comparative assessment of the proposed versus other methods (e.g. ShrinkBayes) on the actual maize dataset in Section 4.  Why is that missing?
The comments made by one reviewer regarding a relative assessment of computing time relative to the other methods deserves attention.}

\comment{Page 1.  I was instructed that heterosis is merely characterized by the hybrid performance not being equal to the average of the two parental lines.  So in your very first sentence, I might write “Heterosis, or hybrid vigor, occurs when hybrid progeny display phenotypes that are superior to the average phenotypes of their parents.” i.e., I might be only interested in the $\delta_g$’s. However, your later characterization of heterosis is really something called overdominance to me…where the hybrid performance exceeds the performance of either parent.  I wonder if you could then be more specific in your definitions in your first sentence…and also in your third sentence in this particular paragraph.}

\comment{Page 1 lines 48-59.  Perhaps this was already captured well in the Ji et al. (2014) paper but why not simply test heterosis as the difference between the hybrid and the average parental performance (i.e. $\mu_3 – 0.5 \mu_1 -0.5 \mu_2$).  Again, I think your issue pertains really to an overdominance test of some sort.}

\comment{Page 2 line 35.  “We CONSIDER an…”}

\comment{Page 3 Line 7.  I don’t it is sufficient to mention the statistical software “Stan”.  Please provide a suitable description and citation.}

\comment{Page 3 Line 26.  There are several ways to construct a Poisson-gamma distribution, all leading to different specifications of the relationship between the mean and variance of a negative binomial.  I apologize for the admittedly vain self-citation but there is one specification, for example, that specifies the variance to be a linear function of the mean of a negative binomial…see Equation 2.12b in  Tempelman and Gianola (1996).  I guess I’m not sure why most RNA-seq data analysts gravitate to a specification where the variance is specified to be linear and quadratic function of the mean as you do here.}

\comment{Page 3. Bottom.  Shouldn’t you at least provide a justification when one would use a normal prior and when one would or should use a Laplace prior?}

\comment{Page 4, bottom half.  The strategy described here, also somewhat attested to by a reviewer, is a little disconcerting with respect to the “double estimation” of gene-specific parameters…first to estimate they hyperparameters and then again conditioned on the estimates of these hyperparameters.  Perhaps you should characterize that these second-stage gene-specific estimates are shrunk relative to those estimates used to estimate the hyperparameters and that they then have lead to predictions with better properties?  Ideally, I would have liked to see these hyperparameters be also formally estimated together with the gene specific parameters within one MCMC analyses…authors should address why they didn’t do this.}

\comment{Page 5, lines 37-47.  Since you’re doing MCMC, why not simply tabulate the number of times samples of $\mu_3$ exceeds either $\mu_1$ or $\mu_2$….and/or the number of times that $\mu_3$ is less than $\mu_1$ or $\mu_2$.  I suspect that what your test is really doing….in which case it might be useful to explicitly state this equivalence as such.}

\comment{Page 6, top.  Is it possible then to assess your proposed method using Laplace versus normal priors to assess if that seems to be contributing to any differences between ShrinkBayes and your proposed method?  Maybe that is the real reason for the small improvements?...., and not the differences in sophistication of the constructed statistical tests?}

\comment{Page 7, line 15.  Why delete the low counts?  Data quality issues?}

\comment{Page 7, lines 17-22.  Can you provide us any information on how well the hyperparameters were estimated based on your empirical Bayes strategy?}

\comment{Page 8, line 37.  It is not obvious to me why $H_g2*$ (are the two parents the same?) is a necessary component of a heterosis (overdominance) test!?}

\comment{Page 9, lines 23-25.  In tandem with another reviewer’s comments, the real question is why?  Is it because of the suboptimality of the test that you contrived for ShrinkBayesA?}

\comment{Page 10, Line 29.  Don’t you mean “nadir” rather than peak?}

\comment{Page 11 line 39.  Is it really “independence” or nearly non-identifiability that is the issue?}

\comment{Page 11 last paragraph.  So this then begs an assessment of how good/poor these hyperparameter estimates were.}

\comment{Page 11 last line “ASYMPTOTIC”}}

Reviewer \#1

\comment{The authors develop a hierarchical negative binomial model for drawing inferences from a composite null hypothesis, that tests gene expression heterosis. Simulations are performed to describe the performance of this new method along with a data analysis demonstrating an application to real data. This manuscript represents a well-conceived and executed methodological development in the statistical analysis of RNA-Seq gene expression data. However, the performance of the new method is underwhelming compared to existing methods, tempering my enthusiasm for the likely impact of this paper.}

\comment{(1) In the simulations, the authors should also compare the eBayes method to the method of Ji et al. (2014). Although developed for continuous microarray data, the authors could apply a simple transformation to the count data in order to make them continuous. I would be curious to know whether the eBayes method out-performs the method of Ji et al. (2014) on the transformed data.}

\comment{(2) The authors should provide more details on the computational price of the eBayes method? How long (in comparison to edgeR, baySeq, ShrinkBayes) did the eBayes method take to run? Are these methods comparable in terms of computational cost?}

\comment{(3) In the description of the simulation results, the eBayes and ShrinkBayes methods display very similar performance. The distinction between the results on the Figures is quite minimal. The authors should attempt to quantify the difference in performance in a single metric. For instance, in Figure 1, take the average percent difference in TPR across the FPRs that were used to generate the plot.}

Reviewer \#4

\comment{1. The authors propose a method to assess "gene expression heterosis."   While it is mentioned that this type of heterosis is a possible explanation for phenotypic heterosis, I'd like more discussion regarding this point.  The two references seem a bit outdated (for bioinformatics) so I'm wondering if this explanation is still being considered.  If it is, than I'd like to see more discussion about how the analytic results would be used to investigate this explanation.   Wouldn't scientists have specific phenotypic traits in mind and know of genes associated with these traits?  I'm a little confused why the search through the whole genome for this type of heterosis if they're looking to draw inference on the association.}

\comment{2.      The model appears reasonable and is adapted from on used with microarrays.  To support their approach, they simulate data using their model (ideal case) and find that their approach does slightly better than ShrinkBayes.
a.      How robust do you feel your method is?  There is no discussion along these lines.
b.      Why is there more of an improvement over ShrinkBayes with more replicates?  Is it because of the method used to calculate the posterior probabilities?  I'd like to see some discussion on this
c.      The ROC curves are not particularly impressive.  Even with 16 reps, the true positive rate is below 50\% (controlling for the false positive rate).  The real data example included only 4 reps, which is far more likely a scenario.  Very little discussion is devoted to this.
3.      Minor edits or comments
a.      Start of Section 2 on page 2  :  Replace "We considering" with just "Consider"
b.      Page 3: "The parental averages and overdispersion parameters are assumed to follow normal distributions, i.e.,"
c.      Right before Section 2.2…it is assumed apriori independence, correct?
d.      Page 4: You have a two-step approach where you estimate the parameters of interest to get estimates of the hyperparameters and then restimate.  How much change is there is these estimates?  Why aren't the methods of McCarthy used as a competitor?
e.      Page 11 "These figures show marked departures from ……"
f.      Page 11 : Does it really matter that the posterior is different from the prior?  One often assume aprior independence but expects there to be posterior correlation.  Perhaps I'm just missing something in this discussion.}

\end{document}

### 2015-08-18 ebayes_rnaseq_heterosis.tex
added some responses

@@
 
-We considering an RNA sequencing (\RNAseq{}) experiment that involves at least three genetic varieties: two parental varieties (P1 and P2) and a cross between these two varieties called the hybrid (H).
+We consider an RNA sequencing (\RNAseq{}) experiment that involves at least three genetic varieties: two parental varieties (P1 and P2) and a cross between these two varieties called the hybrid (H).
 For each variety, replicate RNA samples are isolated and assessed for quality.
@@
 
-We assume a hierarchical model for the gene-specific mean parameters and overdispersion parameters.
-The parental averages and overdispersion parameters are assumed to follow a normal distribution, i.e.
+We assume a hierarchical model for the gene-specific mean parameters and overdispersion parameters. The parental averages and overdispersion parameters are assumed to follow normal distributions, i.e.
 \[ 
@@
 where $\alpha\sim La(\eta,\sigma)$ has a probability density function given by $La(\alpha;\eta,\sigma) = \exp(-|\alpha-\eta|/\sigma)/2\sigma$ with expectation $\eta$ and variance $2\sigma^2$.
-We assume independence amongst the parental averages, half parental differences, hybrid differences, and overdispersion parameters.
+We assume \emph{a priori} independence amongst the parental averages, half parental differences, hybrid differences, and overdispersion parameters.
 
@@
 \end{figure}
-These figures show markedly departure from marginal model assumptions, e.g.
+These figures show markedly departures from marginal model assumptions, e.g.
 normality assumptions for $\phi_g$ and $\psi_g$, and independence assumptions, e.g.
@@
 Our current method, based on moment matching of essentially independently estimated gene-specific parameters provides consistent estimators as the number of replicates per variety increases.
-But typically the number of replicates per variety is quite small, in our data there are only four replicates per variety, and therefore assymptotic justifications are deficient.
+But typically the number of replicates per variety is quite small, in our data there are only four replicates per variety, and therefore asymptotic justifications are deficient.
 There are a variety of hyperparameter estimation approaches to explore, e.g.

### 2015-08-18 response.tex
added some responses

@@
 
+\response{Fixed.}
+
 \comment{Page 3 Line 7.  I don’t it is sufficient to mention the statistical software “Stan”.  Please provide a suitable description and citation.}
@@
 
+\response{Fixed.}
+
 Reviewer \#1
@@
 The model appears reasonable and is adapted from on used with microarrays.
-To support their approach, they simulate data using their model (ideal case) and find that their approach does slightly better than ShrinkBayes.
-a.      How robust do you feel your method is?  There is no discussion along these lines.
-b.      Why is there more of an improvement over ShrinkBayes with more replicates?  Is it because of the method used to calculate the posterior probabilities?  I'd like to see some discussion on this
-c.
+To support their approach, they simulate data using their model (ideal case) and find that their approach does slightly better than ShrinkBayes.}
+
+\comment{a. How robust do you feel your method is?  There is no discussion along these lines.}
+
+\comment{b.
+Why is there more of an improvement over ShrinkBayes with more replicates?
+Is it because of the method used to calculate the posterior probabilities?
+I'd like to see some discussion on this.}
+
+\comment{c.
 The ROC curves are not particularly impressive.
@@
 The real data example included only 4 reps, which is far more likely a scenario.
-Very little discussion is devoted to this.
-3.      Minor edits or comments
-a.      Start of Section 2 on page 2  :  Replace "We considering" with just "Consider"
-b.      Page 3: "The parental averages and overdispersion parameters are assumed to follow normal distributions, i.e.,"
-c.      Right before Section 2.2…it is assumed apriori independence, correct?
-d.
+Very little discussion is devoted to this.}
+
+\comment{3.      Minor edits or comments}
+
+\comment{a.      Start of Section 2 on page 2  :  Replace "We considering" with just "Consider"}
+
+\response{Fixed.}
+
+\comment{b.      Page 3: "The parental averages and overdispersion parameters are assumed to follow normal distributions, i.e.,"}
+
+\response{Fixed.}
+
+\comment{c.      Right before Section 2.2…it is assumed apriori independence, correct?}
+
+\response{Yes, added \emph{a priori}.}
+
+\comment{d.
 Page 4: You have a two-step approach where you estimate the parameters of interest to get estimates of the hyperparameters and then restimate.
 How much change is there is these estimates?
-Why aren't the methods of McCarthy used as a competitor?
-e.      Page 11 "These figures show marked departures from ……"
-f.
+Why aren't the methods of McCarthy used as a competitor?}
+
+\comment{e.      Page 11 "These figures show marked departures from ……"}
+
+\response{Fixed.}
+
+\comment{f.
 Page 11 : Does it really matter that the posterior is different from the prior?

### 2015-08-18 response.tex
responses to all comments, I htink

@@
 \documentclass{article}
+
+\usepackage{color}
+
+\usepackage[parfill]{parskip}
 
@@
 \newcommand{\response}[1]{#1}
+\newcommand{\todo}[1]{{\color{red} #1}}
 
@@
 Furthermore, they seem somewhat concerned about an ‘apples vs.
-oranges’ types of comparisons since the proposed hypothesis testing strategy in the proposed method is really not an option in the other methods.
-Perhaps a reference to the promise of allele specific expression analyses (Discussion section) for inferring upon gene-expression heterosis might be useful to point readers into other directions with this work (Bell et al., 2013; Wei and Wang, 2013).}
+oranges’ types of comparisons since the proposed hypothesis testing strategy in the proposed method is really not an option in the other methods.}
+
+\comment{Perhaps a reference to the promise of allele specific expression analyses (Discussion section) for inferring upon gene-expression heterosis might be useful to point readers into other directions with this work (Bell et al., 2013; Wei and Wang, 2013).}
 
@@
 
+% Email sent to Tieming Ji to see if she has the script.
+
 \comment{One seemingly glaring omission from this work is any comparative assessment of the proposed versus other methods (e.g.
 ShrinkBayes) on the actual maize dataset in Section 4.
-Why is that missing?
-The comments made by one reviewer regarding a relative assessment of computing time relative to the other methods deserves attention.}
+Why is that missing?}
+
+\comment{The comments made by one reviewer regarding a relative assessment of computing time relative to the other methods deserves attention.}
 
@@
 However, your later characterization of heterosis is really something called overdominance to me…where the hybrid performance exceeds the performance of either parent.
-I wonder if you could then be more specific in your definitions in your first sentence…and also in your third sentence in this particular paragraph.}
-
-\comment{Page 1 lines 48-59.
+I wonder if you could then be more specific in your definitions in your first sentence…and also in your third sentence in this particular paragraph.
+lines 48-59.
 Perhaps this was already captured well in the Ji et al.
 (2014) paper but why not simply test heterosis as the difference between the hybrid and the average parental performance (i.e.
-$\mu_3 – 0.5 \mu_1 -0.5 \mu_2$).
+$\mu_3 - 0.5 \mu_1 - 0.5 \mu_2$).
 Again, I think your issue pertains really to an overdominance test of some sort.}
+
+\response{Yes, overdominance is the scenario were are interested in. We have added the term in the introduction as well as in Section 2.3 to make this clear.}
 
@@
 \comment{Page 3 Line 7.  I don’t it is sufficient to mention the statistical software “Stan”.  Please provide a suitable description and citation.}
+
+\response{Added citation.}
 
@@
 
+\todo{currently have normal versions running}
+
 \comment{Page 4, bottom half.
@@
 
+\response{We have added a paragraph to the beginning of Section 2.2 to share our experience with a fully MCMC analysis which took too long and had poor mixing.}
+
 \comment{Page 5, lines 37-47.
@@
 I suspect that what your test is really doing….in which case it might be useful to explicitly state this equivalence as such.}
+
+\response{Yes, that is what the test is doing. We have now stated the case for LPH, i.e. ``$\delta_g < -|\alpha_g|$ (equivalently $\mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}$).''
 
@@
 
+\todo{currently have normal versions running}
+
 \comment{Page 7, line 15.  Why delete the low counts?  Data quality issues?}
@@
 
+\todo{No, because we did not simulate from our model since we wanted to maintain the inherent structure of the gene-specific parameters.
+Perhaps include a sentence based on the other simulation study.
+Perhaps say that we do not believe they are estimated very well.}
+
 \comment{Page 8, line 37.  It is not obvious to me why $H_g2*$ (are the two parents the same?) is a necessary component of a heterosis (overdominance) test!?}
+
+\response{It is not necessary, but it is sufficient. If the parents are the same, then the hybrid must have overdominance.}
 
@@
 
+\todo{Why does ShrinkBayes perform worse? When the normal results are done, we can speak more directly to this.}
+
 \comment{Page 10, Line 29.  Don’t you mean “nadir” rather than peak?}
+
+\response{No, although we do mean ridge rather than peak, i.e. the count is high for effect size of zero and probabilities below 0.5.}
 
@@
 
+\todo{??}
+
 \comment{Page 11 last paragraph.  So this then begs an assessment of how good/poor these hyperparameter estimates were.}
+
+\todo{Add discussion of quality of hyperparameter estimates.}
 
@@
 
+\todo{Will run when we get the code. If we don't get the code, will figure out how to run it. Ideally it would be nice to see the eBayes version as well as the approximation in Ji et al.}
+
 \comment{(2) The authors should provide more details on the computational price of the eBayes method?
@@
 Are these methods comparable in terms of computational cost?}
+
+\todo{We will do this.}
 
@@
 For instance, in Figure 1, take the average percent difference in TPR across the FPRs that were used to generate the plot.}
+
+\todo{I think the main point is that they aren't very different. Somehow I am not able to communicate that the model is novel and that here are a couple of different estimation approaches.}
 
@@
 
+\todo{Dan? Even though some genes may be good candidates. An exploratory analysis to suggest other genes seems relevant.}
+
 \comment{2.
@@
 
+\todo{Good question. Although we suggest the gene-specific parameter distributions do not fit very well, how much better do we really think we can get if we model it well?}
+
 \comment{b.
@@
 I'd like to see some discussion on this.}
+
+\todo{Once normal runs are completed, we can comment on this more.}
 
@@
 Very little discussion is devoted to this.}
+
+\todo{Yes, but we haven't created artificially large signals.}
 
@@
 
+\todo{Comment on the change in gene-specific parameter estimates, e.g. shrinkage.}
+
 \comment{e.      Page 11 "These figures show marked departures from ……"}
@@
 
+\todo{Yes, because we are only estimating the parameters of assumed marginal distributions that are assumed independent.
+If we were performing a fully Bayesian analysis, then we could learn about the joint distribution of the gene-specific parameters.}
+
 \end{document}

### 2015-08-18 ebayes_rnaseq_heterosis.tex
responses to all comments, I htink

@@
 
-Heterosis, or hybrid vigor, occurs when hybrid progeny display phenotypes that are superior to the phenotypes of their parents.
+Heterosis, or hybrid vigor, occurs when hybrid progeny display phenotypes that are superior to the average phenotypes of their parents.
 The heterosis phenomenon was scientifically documented in plants by \cite{darwin1876effects} and has long been used to improve agricultural production.
+Overdominance is a particularly extreme case of heterosis where the hybrid phenotype lies outside the range of the parental phenotypes.
 One classic example involves hybrid maize offspring that are taller, faster to mature, and yield considerably more grain than their inbred parents \citep{hallauer1981quantitative, hallauer2010quantitative}.
@@
 The remainder of the paper proceeds as follows.
-Section \ref{s:method} presents the proposed hierarchical model, an empirical Bayes method of estimating the parameters, and the calculation of posterior probabilities of gene expression heterosis.
+Section \ref{s:method} presents the proposed hierarchical model, an empirical Bayes method of estimating the parameters, and the calculation of posterior probabilities of overdominance.
 Section \ref{s:simulation} presents a simulation study based on a maize experiment and compares our approach to alternative methods.
@@
 
-%We assume improper uniform priors for $\eta_\phi$, $\eta_\alpha$, and $\eta_\delta$ and independent half-cauchy priors for all standard deviations \citep{gelman2006prior}.
-
 \subsection{Empirical Bayes}
@@
 
-We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$, normalization factors $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, and hyperparameters $\pi = (\eta, \sigma)$ where $\eta=(\eta_\phi,\eta_\alpha,\eta_\delta, \eta_\psi)$ and $\sigma=(\sigma_\phi,\sigma_\alpha,\sigma_\delta,\sigma_\psi)$.
+Initial attempts to estimate the models above via Markov chain Monte Carlo (MCMC) failed due to high computational costs and poor mixing in the resulting chains.
+For example, we implemented the model in the statistical software \Stan{} (discussed at the end of this section), ran the MCMC on a simulated data set with 10,000 genes for 2 months on a state-of-the-art linux server, and obtained potential scale reduction factors that suggested we would need to run at least ten times as long to obtain convergence.
+Although there are certainly improvements that could be made to decrease computational costs and improve mixing, we opted for an empirical Bayes approach.
+This approach may be a reasonable approximation to a fully Bayesian approach when estimating models with large numbers of genes as the posterior distributions for the hyperparameters may be tightly peaked.
+
+We categorize the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$, normalization factors $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, and hyperparameters $\pi = (\eta, \sigma)$ where $\eta=(\eta_\phi,\eta_\alpha,\eta_\delta, \eta_\psi)$ and $\sigma=(\sigma_\phi,\sigma_\alpha,\sigma_\delta,\sigma_\psi)$.
 We obtain estimates for the hyperparameters and then base gene-specific inference on the posterior conditional on these estimates.
@@
 
-To perform this Bayesian analysis, we used the statistical software \Stan{} which implements a variant of Markov chain Monte Carlo (MCMC) called Hamiltonion Monte Carlo (HMC) \citep{neal2011mcmc} to obtain samples from the posterior in equation \eqref{e:condind}.
+To perform this Bayesian analysis, we used the statistical software \Stan{} \citep{stan-software:2014} executed through the \RStan{} interface \citep{rstan-software:2014} in R \citep{R2014}.
+\Stan{} implements a variant of MCMC called Hamiltonion Monte Carlo \citep{neal2011mcmc} to obtain samples from the posterior in equation \eqref{e:condind}.
 We ran 4 simultaneous chains with random initial starting values for 1000 burnin (and tuning) iterations followed by another 1000 iterations thinned by 4 (to reduce storage space) for inference.
@@
 
-For a specific gene $g$, low-parent heterosis (LPH) of gene expression occurs when average expression in the hybrid is less than the average expression of either parent, i.e.
-$\delta_g < -|\alpha_g|$, and high-parent heterosis (HPH) occurs when average expression in the hybrid is greater than the average expression of either parent, i.e.
+In the maize context that motivates this work, we are particularly interested in overdominance, either low-parent heterosis (LPH) or high-parent heterosis (HPH), in gene expression. 
+For a specific gene $g$, LPH occurs when average expression in the hybrid is less than the average expression of either parent, i.e.
+$\delta_g < -|\alpha_g|$ (equivalently $\mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}$), and HPH occurs when average expression in the hybrid is greater than the average expression of either parent, i.e.
 $\delta_g > |\alpha_g|$.
@@
 %P(LPH_g|y) \approx 
-P(\mbox{LPH}_g|y,\hat{\pi},\hat{\gamma}) = P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi},\hat{\gamma}\right) = \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right),
+P(\mbox{LPH}_g|y,\hat{\pi},\hat{\gamma}) = 
+%P\left(\left. \mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}\right|y,\hat{\pi},\hat{\gamma}\right) = 
+P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi},\hat{\gamma}\right) = \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right), 
 \label{e:probs}
@@
 Geneticists can use this list to prioritize future experiments to understand the molecular genetic mechanisms for heterosis.
-
-% \subsection{Implementation in \Stan{}}
-
-%To perform the optimization required for the MAP estimate $\hat{\pi}$, we use the statistical software, {\tt Stan} \citep{stan-software:2014}, executed through the \RStan{} interface \citep{rstan-software:2014} in R \citep{R2014}.
-The Stan software allows the construction of general Bayesian models with the primary restriction that all unknown parameters are continuous.
-With a fully constructed model, the software can obtain MAP estimates \cite[Section 50.3,][]{stan-manual:2014} using the {\tt optimizing} function in \RStan{}.
-Presumably due to the large dimension of the optimization, it was necessary to use the L-BFGS algorithm \cite[Section 55,][]{stan-manual:2014} and reasonable starting values obtained via \edgeR{} \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor}.
 
@@
 We set this measure to zero if the maximum likelihood estimate, calculated using \edgeR{}, of $\mu_{g3}$ is between the maximum likelihood estimates of $\mu_{g1}$ and $\mu_{g2}$.
-Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the two hypotheses that allow for extreme heterosis.
+Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the two hypotheses that allow for overdominance.
 
@@
 \end{figure}
-The figure shows a peak at zero for probabilities below 0.5 with estimated effect sizes of zero.
+The figure shows a ridge at an effect size of zero for probabilities below 0.5.
 Above a probability of 0.5, we see a prototypical volcano pattern, with increased heterosis probability corresponding to larger estimated effect sizes and no estimated effect sizes near zero for genes with high heterosis probability.

### 2015-08-18 supp/fig_compare_gene_specific_parameter_estimates.R
script to construct figure to compare gene specific parameter estimates

@@ -1,0 +1,55 @@
+library(rstan)
+library(plyr)
+library(dplyr)
+library(reshape2)
+library(edgeR)
+library(ggplot2)
+
+##################################################
+# Get hyperparameters
+##################################################
+d = read.csv("data.csv", row.names=1)
+variety = factor(gsub("_[0-9]{1,2}", "", names(d)), levels=c("B73","Mo17","B73xMo17"))
+
+# phi, alpha, delta parameterization
+design = cbind(1,
+               c(1,-1,0)[as.numeric(variety)],
+               variety == 'B73xMo17')
+
+# GLM fit using edgeR
+fit = d %>% 
+  DGEList() %>%
+  calcNormFactors %>%
+  estimateCommonDisp %>%
+  estimateGLMTagwiseDisp(design) %>%
+  glmFit(design)
+
+# Calculate gene-specific estimates for phi, alpha, delta, and psi
+hat = data.frame(phi   = fit$coefficients[,1] + mean(fit$offset[1,]),
+                 alpha = -fit$coefficients[,2],
+                 delta = fit$coefficients[,3],
+                 psi   = log(fit$dispersion),
+                 method="edgeR")
+hat$gene = rownames(hat)
+
+
+# Read eBayes parameter estimates
+tmp = readRDS("script.rds")[,c("phi","alpha","delta","psi")]
+tmp$gene = rownames(tmp)
+tmp$method = 'eBayes'
+
+# Combine estimates
+combined = rbind(melt(hat, id.vars=c("method","gene")),
+                 melt(tmp, id.vars=c("method","gene")))
+casted = dcast(combined, gene+variable~method)
+
+casted$variable = revalue(casted$variable, c("phi"   = "parental average",
+                                             "alpha" = "half-parental difference",
+                                             "delta" = "hybrid effect",
+                                             "psi"   = "overdispersion"))
+
+# Plot estimates
+ggplot(casted, aes(edgeR, eBayes)) + 
+  stat_binhex() + 
+  geom_abline(intercept=0,slope=1) +
+  facet_wrap(~variable, scales='free')

### 2015-08-18 supp/get_hyperparameters.R
updated comment

@@ -12,7 +12,7 @@ get_hyperparameters = function(d, variety, method) {
     estimateGLMTagwiseDisp(design) %>%
     glmFit(design)
   
-  # Calculate gene-specific estimates for phi, alpha, and delta
+  # Calculate gene-specific estimates for phi, alpha, delta, and psi
   hat = data.frame(gene = 1:length(fit$dispersion),
                    phi   = fit$coefficients[,1] + mean(fit$offset[1,]),
                    alpha = fit$coefficients[,2],

### 2015-08-18 supp/fig_compare_gene_specific_parameter_estimates.R
fixed design

@@ -26,7 +26,7 @@ fit = d %>%
 
 # Calculate gene-specific estimates for phi, alpha, delta, and psi
 hat = data.frame(phi   = fit$coefficients[,1] + mean(fit$offset[1,]),
-                 alpha = -fit$coefficients[,2],
+                 alpha = fit$coefficients[,2], 
                  delta = fit$coefficients[,3],
                  psi   = log(fit$dispersion),
                  method="edgeR")
@@ -37,6 +37,7 @@ hat$gene = rownames(hat)
 tmp = readRDS("script.rds")[,c("phi","alpha","delta","psi")]
 tmp$gene = rownames(tmp)
 tmp$method = 'eBayes'
+tmp$alpha = -tmp$alpha # Fixed in the stan model files, but the rds file has swapped parents
 
 # Combine estimates
 combined = rbind(melt(hat, id.vars=c("method","gene")),

### 2015-08-18 supp/fig_compare_gene_specific_parameter_estimates.R
code for fig incorporated into figs.R

@@ -1,56 +1,0 @@
-library(rstan)
-library(plyr)
-library(dplyr)
-library(reshape2)
-library(edgeR)
-library(ggplot2)
-
-##################################################
-# Get hyperparameters
-##################################################
-d = read.csv("data.csv", row.names=1)
-variety = factor(gsub("_[0-9]{1,2}", "", names(d)), levels=c("B73","Mo17","B73xMo17"))
-
-# phi, alpha, delta parameterization
-design = cbind(1,
-               c(1,-1,0)[as.numeric(variety)],
-               variety == 'B73xMo17')
-
-# GLM fit using edgeR
-fit = d %>% 
-  DGEList() %>%
-  calcNormFactors %>%
-  estimateCommonDisp %>%
-  estimateGLMTagwiseDisp(design) %>%
-  glmFit(design)
-
-# Calculate gene-specific estimates for phi, alpha, delta, and psi
-hat = data.frame(phi   = fit$coefficients[,1] + mean(fit$offset[1,]),
-                 alpha = fit$coefficients[,2], 
-                 delta = fit$coefficients[,3],
-                 psi   = log(fit$dispersion),
-                 method="edgeR")
-hat$gene = rownames(hat)
-
-
-# Read eBayes parameter estimates
-tmp = readRDS("script.rds")[,c("phi","alpha","delta","psi")]
-tmp$gene = rownames(tmp)
-tmp$method = 'eBayes'
-tmp$alpha = -tmp$alpha # Fixed in the stan model files, but the rds file has swapped parents
-
-# Combine estimates
-combined = rbind(melt(hat, id.vars=c("method","gene")),
-                 melt(tmp, id.vars=c("method","gene")))
-casted = dcast(combined, gene+variable~method)
-
-casted$variable = revalue(casted$variable, c("phi"   = "parental average",
-                                             "alpha" = "half-parental difference",
-                                             "delta" = "hybrid effect",
-                                             "psi"   = "overdispersion"))
-
-# Plot estimates
-ggplot(casted, aes(edgeR, eBayes)) + 
-  stat_binhex() + 
-  geom_abline(intercept=0,slope=1) +
-  facet_wrap(~variable, scales='free')

### 2015-08-18 supp/figs.R
code for fig incorporated into figs.R

@@ -47,3 +47,66 @@ for (i in 1:n) {
 do.call("grid.arrange", c(plots, ncol=4))
 
 ggsave("../include/estimates.pdf", do.call("arrangeGrob", c(plots, ncol=4)), width=8, height=8)
+
+
+
+##########################################################################################
+# Comparison of gene specific parameter estimates
+##########################################################################################
+
+library(dplyr)
+library(reshape2)
+library(edgeR)
+
+d = read.csv("data.csv", row.names=1)
+variety = factor(gsub("_[0-9]{1,2}", "", names(d)), levels=c("B73","Mo17","B73xMo17"))
+
+# phi, alpha, delta parameterization
+design = cbind(1,
+               c(1,-1,0)[as.numeric(variety)],
+               variety == 'B73xMo17')
+
+# GLM fit using edgeR
+fit = d %>% 
+  DGEList() %>%
+  calcNormFactors %>%
+  estimateCommonDisp %>%
+  estimateGLMTagwiseDisp(design) %>%
+  glmFit(design)
+
+# Calculate gene-specific estimates for phi, alpha, delta, and psi
+hat = data.frame(phi   = fit$coefficients[,1] + mean(fit$offset[1,]),
+                 alpha = fit$coefficients[,2], 
+                 delta = fit$coefficients[,3],
+                 psi   = log(fit$dispersion),
+                 method="edgeR")
+hat$gene = rownames(hat)
+
+
+# Read eBayes parameter estimates
+tmp = readRDS("script.rds")[,c("phi","alpha","delta","psi")]
+tmp$gene = rownames(tmp)
+tmp$method = 'eBayes'
+tmp$alpha = -tmp$alpha # Fixed in the stan model files, but the rds file has swapped parents
+
+# Combine estimates
+combined = rbind(melt(hat, id.vars=c("method","gene")),
+                 melt(tmp, id.vars=c("method","gene")))
+casted = dcast(combined, gene+variable~method)
+
+casted$variable = revalue(casted$variable, c("phi"   = "parental average",
+                                             "alpha" = "half-parental difference",
+                                             "delta" = "hybrid effect",
+                                             "psi"   = "overdispersion"))
+
+# Plot estimates
+ggplot(casted, aes(edgeR, eBayes)) + 
+  stat_binhex() + 
+  geom_abline(intercept=0,slope=1) +
+  facet_wrap(~variable, scales='free')
+
+
+ggsave(file="../include/gene_specific_estimates.pdf", width=8, height=8)
+
+
+

### 2015-08-18 ebayes_rnaseq_heterosis.tex
include gene-specific parameter comparison

@@
 
+Figure \ref{f:gene_specific_estimates} provides point estimates from the initial \edgeR{} estimation step and after the eBayes procedure described in Section \ref{s:ebayes}. 
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{gene_specific_estimates.pdf}}
+\caption{Two-dimensional histogram of point estimates from \edgeR{} and eBayes (posterior means) along with the $y=x$ diagonal.}
+\label{f:gene_specific_estimates}
+\end{figure}
+This figure shows shrinkage for large absolute values of $\alpha$ and $\delta$ from the \edgeR{} estimation toward $\hat{\eta}_\alpha$ and $\hat{\eta}_\delta$ which are both approximately zero.
+The figure also shows decreased values for the overdispersion parameter with larger decreases for high and low values of overdispersion.
+Finally, very little, if any, shrinkage is observed for $\phi$.
+
 With posteriors for all parameters, we can calculate empirical Bayes posterior probabilities for LPH and HPH.

### 2015-08-18 response.tex
now have all responses
add shrinkage discussion and a prior independence response

@@
 
+\todo{We don't have artificially large signals. I didn't perceive the ``apples vs. oranges'' comparison even though I agree.}
+
 \comment{Perhaps a reference to the promise of allele specific expression analyses (Discussion section) for inferring upon gene-expression heterosis might be useful to point readers into other directions with this work (Bell et al., 2013; Wei and Wang, 2013).}
+
+\todo{Add references to allele specific expression.}
 
@@
 
-% Email sent to Tieming Ji to see if she has the script.
+\todo{Run Ji approach}
 
@@
 
+\todo{How would we compare this?}
+
 \comment{The comments made by one reviewer regarding a relative assessment of computing time relative to the other methods deserves attention.}
+
+\todo{Add computing time.}
 
@@
 \comment{Page 7, line 15.  Why delete the low counts?  Data quality issues?}
+
+\todo{Why did we do this? The other methods fail?}
 
@@
 
-\todo{Comment on the change in gene-specific parameter estimates, e.g. shrinkage.}
+\response{We have added a figure (Fig 3) to compare gene-specific parameter estimates and a paragraphing describing the comparison in Section 4.
+We aren't exactly sure which McCarthy methods you are referring to.
+We already use an approach based on edgeR and the point estimates themselves do not naturally lead to a statistic for determining the plausibility of heterosis.}
 
@@
 
-\todo{Yes, because we are only estimating the parameters of assumed marginal distributions that are assumed independent.
-If we were performing a fully Bayesian analysis, then we could learn about the joint distribution of the gene-specific parameters.}
+\todo{Due to the nature of an empirical Bayes analysis, we are learning nothing about the distribution of the gene-specific parameter estimates other than through the hyperparameters.
+Yes, we can learn something about the parameters for a particular gene, but the information content in the data is very poor, i.e.
+only 12 total observations to infer about 4 parameters.
+Had we performed a fully Bayesian analysis, we could have learned more about the distribution of the gene-specific parameter estimates via the joint distributions of the hyperparameters (we believe this is what you are referring to).
+We believe that improved estimation of the gene-specific parameter distribution should lead to improved ability to detect heterosis.}
 

### 2015-08-24 ebayes_rnaseq_heterosis.tex
updated fig caption

@@
 \centerline{\includegraphics[width=\textwidth]{exampleROC0_1}}
-\caption{Representative ROC curves for false positive rates below 0.1 for the approaches using \edgeR{} and \baySeq{} described in Section \ref{s:alternative}, the \ShrinkBayes{} approach described in Section \ref{s:shrinkbayes} and our eBayes approach described in Section \ref{s:ebayes}.}
+\caption{Representative ROC curves for false positive rates below 0.1 for the approaches using \edgeR{} and \baySeq{} described in Section \ref{s:alternative}, the approach of \cite{} using log-counts, the \ShrinkBayes{} approach described in Section \ref{s:shrinkbayes} and the eBayes approach described in Section \ref{s:ebayes} using both normal and laplace distributions for the half-parental difference and hybrid effect.}
 \label{f:roc}

### 2015-08-24 supp/ji-sim-script.R
incorporate normalization factors into Ji for sims

@@ -14,10 +14,15 @@ names(d) = c(paste("B73_",1:n, sep=''), paste('Mo17_',1:n, sep=''), paste('B73xM
 variety = factor(gsub("_[0-9]{1,2}", "", names(d)), levels=c("B73","Mo17","B73xMo17"))
 d$geneID = 1:nrow(d)
 
+fit = d[,-1] %>% # remove geneID
+  DGEList() %>%
+  calcNormFactors 
+lnorm_factors = log(fit$samples$norm.factors)
+
 ld = d %>%
   melt(id.vars="geneID", variable.name="sampleID", value.name="count") %>%
   mutate(variety = factor(gsub("_[0-9]{1,2}", "", sampleID), levels=c("B73","Mo17","B73xMo17")),
-         y = log(count+1)) 
+         y = log(count+1)-lnorm_factors) 
   
 # Summary statistics
 tmp = ld %>% 

### 2015-08-24 supp/ji-sim-script.R
add edgeR library

@@ -1,5 +1,6 @@
 library(plyr)
 library(dplyr)
+library(edgeR)
 library(reshape2)
 
 source("args.R")

### 2015-08-24 ebayes_rnaseq_heterosis.tex
update alternative methods to include Ji

@@
 
-Existing \RNAseq{} data analysis methods are not specifically designed to identify gene expression heterosis, but we adapted the approaches in \edgeR{} and {\tt baySeq} to provide a measure of the strength of heterosis for each gene such that large values of this measure indicate support for heterosis.
+We compared our method to that of \cite{ji2014estimation}, which assumes normality in the response, by modeling the logarithm of the RNAseq count plus one divided by the normalization factor, i.e.
+$\log([Y_{gvi}+1]/\gamma_{vi})$.
+Use of the normalization factor here provides two advantages: 1) counts are properly scaled by the thoroughness of the sequencing of the sample and 2) for genes with no variability within variety (which actually occurs in our maize data set), use of the normalization factors allows the approach of \citeauthor{ji2014estimation} to still execute.
+
+In addition to the approach of \citeauthor{ji2014estimation}, we modified two existing \RNAseq{} approaches, \edgeR{} and \baqSeq{}, for use in the overdominance context.
+For each method, we attempted to provide a measure of the strength of overdominance for each gene such that large values of this measure indicate support for heterosis.
 
@@
 
-{\tt baySeq} allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct.
+\baqSeq{} allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct.
 In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene.

### 2015-08-24 ebayes_rnaseq_heterosis.tex
update fig captions

@@
 \centerline{\includegraphics[width=\textwidth]{auc-facet-TRUE}}
-\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different replicates per variety for the approaches using \edgeR{} and \baySeq{} described in Section \ref{s:alternative}, the \ShrinkBayes{} approach described in Section \ref{s:shrinkbayes} and our eBayes approach described in Section \ref{s:ebayes}.
+\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different replicates per variety for the approaches using \citeauthor{ji2014estimation}, \edgeR{}, and \baySeq{} as described in Section \ref{s:alternative}, the \ShrinkBayes{} approach described in Section \ref{s:shrinkbayes} and the eBayes approach described in Section \ref{s:ebayes} using both normal and laplace distributions for the half-parental difference and hybrid effect.
 Each line is a different simulation while the blue box indicates mean AUC (plus or minus one standard error).}

### 2015-08-24 ebayes_rnaseq_heterosis.tex
updated methods to discuss normal distributions for half-parental differences and hybrid effects

@@
 
-We assume a hierarchical model for the gene-specific mean parameters and overdispersion parameters. The parental averages and overdispersion parameters are assumed to follow normal distributions, i.e.
+We assume a hierarchical model for the gene-specific mean parameters and overdispersion parameters.
+Initially, we assume the parental averages, half-parently differences, hybrid effect, and overdispersion parameters follow normal distributions, i.e.
 \[ 
 %\gamma_{vi} \stackrel{ind}{\sim} N(0,\sigma_\gamma^2), \quad 
-\phi_g \stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2), \quad \mbox{and}\quad 
+\phi_g \stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2), 
+\alpha_g \stackrel{ind}{\sim} N(\eta_\alpha, \sigma_\alpha^2),
+\delta_g \stackrel{ind}{\sim} N(\eta_\delta, \sigma_\delta^2), \mbox{ and }
 \psi_g \stackrel{ind}{\sim} N(\eta_\psi,\sigma_\psi^2). 
 \]
-To allow for more shrinkage with the half-parental and hybrid differences, we use a Laplace (or double exponential prior) and thus implement a Bayesian LASSO \citep{park2008bayesian}, i.e. 
+Empirical plots of estimated values of $\alpha_g$ and $\delta_g$ for our maize data set (described in Section \ref{s:maize}) suggest that the distribution of these parameters are more peaked near zero and have heavier tails than a normal distribution allows.
+For differential expression between the two parental phenotypes, we would expect many genes to have small effects and only a few genes to have relatively large effects.
+For these many genes with small effects, we might expect the hybrid to act similar to its parents and therefore also have many genes where the hybrid effect is small and only a few genes where this hybrid effect is large.
+For these reasons, we also assessed Laplace (or double exponential) distributions for the half-parental difference and hybrid effect and thus implement a Bayesian LASSO \citep{park2008bayesian}, i.e.
 \[ 
@@
 where $\alpha\sim La(\eta,\sigma)$ has a probability density function given by $La(\alpha;\eta,\sigma) = \exp(-|\alpha-\eta|/\sigma)/2\sigma$ with expectation $\eta$ and variance $2\sigma^2$.
-We assume \emph{a priori} independence amongst the parental averages, half parental differences, hybrid differences, and overdispersion parameters.
+Whether using normal or Laplace distributions, we assume \emph{a priori} independence amongst the parental averages, half parental differences, hybrid effects, and overdispersion parameters.
 
@@
 \centerline{\includegraphics[width=\textwidth]{exampleROC0_1}}
-\caption{Representative ROC curves for false positive rates below 0.1 for the approaches using \edgeR{} and \baySeq{} described in Section \ref{s:alternative}, the approach of \cite{} using log-counts, the \ShrinkBayes{} approach described in Section \ref{s:shrinkbayes} and the eBayes approach described in Section \ref{s:ebayes} using both normal and laplace distributions for the half-parental difference and hybrid effect.}
+\caption{Representative ROC curves for false positive rates below 0.1 for the approaches using \citeauthor{ji2014estimation}, \edgeR{}, and \baySeq{} described in Section \ref{s:alternative}, the \ShrinkBayes{} approach described in Section \ref{s:shrinkbayes} and the eBayes approach described in Section \ref{s:ebayes} using both normal and laplace distributions for the half-parental difference and hybrid effect.}
 \label{f:roc}

### 2015-08-24 ebayes_rnaseq_heterosis.tex
typo

@@
 
-In addition to the approach of \citeauthor{ji2014estimation}, we modified two existing \RNAseq{} approaches, \edgeR{} and \baqSeq{}, for use in the overdominance context.
+In addition to the approach of \citeauthor{ji2014estimation}, we modified two existing \RNAseq{} approaches, \edgeR{} and \baySeq{}, for use in the overdominance context.
 For each method, we attempted to provide a measure of the strength of overdominance for each gene such that large values of this measure indicate support for heterosis.
@@
 
-\baqSeq{} allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct.
+\baySeq{} allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct.
 In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene.

### 2015-08-24 response.tex
updated responses

@@
 
-Editor comments:
+\response{We thank the AE and two reviewers for their comments which dramatically improved the manuscript.}
+
+Associate Editor comments:
 
@@
 
-\todo{We don't have artificially large signals. I didn't perceive the ``apples vs. oranges'' comparison even though I agree.}
+\todo{We don't have artificially large signals. I didn't perceive the ``apples vs. oranges'' comments from the reviews. Nonetheless, I agree.}
 
@@
 
-\todo{Run Ji approach}
+\response{We have included Ji et al.'s work as another alternative method, see Sections 3.2 and 3.3.}
 
@@
 
+\todo{??}
+
 \comment{Page 3. Bottom.  Shouldn’t you at least provide a justification when one would use a normal prior and when one would or should use a Laplace prior?}
 
-\todo{currently have normal versions running}
+\response{We have assessed the use of normal distributions see Section 3.3 and also provided rationale for why Laplace distributions would be expected to perform better in this scenario at the end of Section 2.1.}
 
@@
 
-\todo{currently have normal versions running}
+\response{We have assessed the use of normal distributions, see Section 3.3.}
 
@@
 
-\todo{Why does ShrinkBayes perform worse? When the normal results are done, we can speak more directly to this.}
+\todo{Still need to figure out how to respond.}
 
@@
 
-\todo{Will run when we get the code. If we don't get the code, will figure out how to run it. Ideally it would be nice to see the eBayes version as well as the approximation in Ji et al.}
+\response{We have assessed the approach of Ji et al. on transformed data, see Sections 3.2 and 3.3.}
 
@@
 
-\todo{Once normal runs are completed, we can comment on this more.}
+\todo{Figure out how to respond.}
 

### 2015-08-24 ebayes_rnaseq_heterosis.tex
prose for figs

@@
 The  gene-specific parameter estimates were treated as the true parameter values for the simulation study so that our simulated datasets mimicked the existing structure among the gene-variety means of the original maize data.
-If $\hat{\delta}_g<-|\hat{\alpha}_g|$ ($\hat{\delta}_g >|\hat{\alpha}_g|$)  the gene was considered to have LPH (HPH).
 
@@
 We repeated this simulation process 10 times for each of 4, 8, and 16 replicates per variety, reusing normalization factors when necessary.
-% From Dan "Did we consider changing the number of zeros allowed when the number of reps is higher?
-I like the idea of simplifying all this by just picking an average count threshold that his high enough to avoid problems.
-I think that may be what others have done in the past."
+
+For a particular gene, the truth was determined via the estimated values for $\alpha_g$ and $\delta_g$.
+Specifically, if $|\hat{\delta}_g| >|\hat{\alpha}_g|$ the gene was considered to have overdominance.
+For many overdominance genes, the value of $|\hat{\delta}_g|$ was only slightly larger than $|\hat{\alpha}_g|$.
+Thus there are many genes in these simulated data sets that are considered to have overdominance, but whose signal in the simulations will be extremely small.
+Conversely, there are many non-overdominance genes whose value of $|\hat{\delta}_g|$ was only slightly smaller than $|\hat{\alpha}_g|$, but whose simulated data will look similar to many overdominance genes.
+Therefore, we expect it will be difficult to accurately identify overdominance genes, but believe this is indicative of real applications as well.
 
@@
 In addition to the approach of \citeauthor{ji2014estimation}, we modified two existing \RNAseq{} approaches, \edgeR{} and \baySeq{}, for use in the overdominance context.
-For each method, we attempted to provide a measure of the strength of overdominance for each gene such that large values of this measure indicate support for heterosis.
+For each method, we attempted to provide a measure of the strength of overdominance for each gene such that large values of this measure indicate support for overdominance.
 
 \edgeR{} can be used to test for differential expression between any two varieties based on the fit of a negative binomial log-linear model \citep{robinson2007moderated, robinson2010edgeR}.
-To construct a measure of heterosis, we computed the maximum likelihood estimates of the $\mu_{gv}$ parameters for all genes using \edgeR{}'s built-in Fisher scoring algorithm, and then used likelihood ratio tests to calculate two $p$-values for each gene: $p_{g1}$  for testing $H_{g01}:\mu_{g1}=\mu_{g3}$ and $p_{g2}$ for testing $H_{g02}:\mu_{g2}=\mu_{g3}$.
+To construct a measure of overdominance, we computed the maximum likelihood estimates of the $\mu_{gv}$ parameters for all genes using \edgeR{}'s built-in Fisher scoring algorithm, and then used likelihood ratio tests to calculate two $p$-values for each gene: $p_{g1}$  for testing $H_{g01}:\mu_{g1}=\mu_{g3}$ and $p_{g2}$ for testing $H_{g02}:\mu_{g2}=\mu_{g3}$.
 Then, for each gene, we define a new $p$-value denoted as $p_g$ and set to $p_g=1$ if the estimate of $\mu_{g3}$ falls between the estimates of $\mu_{g1}$ and $\mu_{g2}$ and $p_g=\max\{p_{g1},p_{g2}\}/2$ otherwise.
-(For all relevant significance thresholds $\omega$ near 0, it can be shown that rejecting the null hypothesis of no heterosis of expression for gene $g$ whenever $p_g\leq \omega$ results in a test that is asymptotically size $\omega$.)  We then use $1-p_g$ as a final measure positively associated with strength of evidence for gene expression heterosis.
+(For all relevant significance thresholds $\omega$ near 0, it can be shown that rejecting the null hypothesis of no overdominance of expression for gene $g$ whenever $p_g\leq \omega$ results in a test that is asymptotically size $\omega$.)  We then use $1-p_g$ as a final measure positively associated with strength of evidence for overdominance.
 
 \baySeq{} allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct.
-In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene.
+In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of overdominance for each gene.
 We set this measure to zero if the maximum likelihood estimate, calculated using \edgeR{}, of $\mu_{g3}$ is between the maximum likelihood estimates of $\mu_{g1}$ and $\mu_{g2}$.
@@
 
-For the methods in Sections \ref{s:ebayes}, \ref{s:shrinkbayes}, and \ref{s:alternative}, we sorted genes according to the computed measure of the strength of evidence for heterosis.
-From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to distinguish genes with heterosis, as defined in Section \ref{s:sim_data},
+For the methods in Sections \ref{s:ebayes}, \ref{s:shrinkbayes}, and \ref{s:alternative}, we sorted genes according to the computed measure of the strength of evidence for overdominance.
+From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to distinguish genes with overdominance, as defined in Section \ref{s:sim_data},
 % From Dan, "why this citation?"
-from those without heterosis \citep{landau2013dispersion}. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
+from those without overdominance \citep{landau2013dispersion}. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
 \begin{figure}[htbp]
 \centerline{\includegraphics[width=\textwidth]{exampleROC0_1}}
-\caption{Representative ROC curves for false positive rates below 0.1 for the approaches using \citeauthor{ji2014estimation}, \edgeR{}, and \baySeq{} described in Section \ref{s:alternative}, the \ShrinkBayes{} approach described in Section \ref{s:shrinkbayes} and the eBayes approach described in Section \ref{s:ebayes} using both normal and laplace distributions for the half-parental difference and hybrid effect.}
+\caption{Example ROC curves for false positive rates below 0.1 for the approaches using \citeauthor{ji2014estimation}, \edgeR{}, and \baySeq{} described in Section \ref{s:alternative}, the \ShrinkBayes{} approach described in Section \ref{s:shrinkbayes} and the eBayes approach described in Section \ref{s:ebayes} using both normal and laplace distributions for the half-parental difference and hybrid effect.}
 \label{f:roc}
 \end{figure}
-For this simulation, we can see that the approaches based most closely on the model in Section \ref{s:model}, i.e.
+The ROC curves are modest, e.g.
+for a false positive rate of 5\%, the best performing methods only achieved a true positive rate of just over 15\%.
+This is consistent with our expectation discussed in Section \ref{s:sim_data} due to the high signal-to-noise ratio in these simulated data.
+
+For this simulation, we can see that the approaches based on the model in Section \ref{s:model}, i.e.
 eBayes and \ShrinkBayes{}, provide the best true positive rate for a given false positive rate.
-Also, as expected, as the sample size increases, our ability to distinguish genes with heterosis from genes without improves.
+Also, as expected, as the sample size increases, our ability to distinguish genes with overdominance from genes without improves.
 
-Figure \ref{f:auc} provides the area under the ROC curve below a false positive rate of 0.1 across the 10 simulations for each of the 3 different sample sizes. 
+Figure \ref{f:auc} provides the area under the ROC curve (AUC) below a false positive rate of 0.1 across the 10 simulations for each of the 3 different sample sizes. 
 \begin{figure}
 \centerline{\includegraphics[width=\textwidth]{auc-facet-TRUE}}
-\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different replicates per variety for the approaches using \citeauthor{ji2014estimation}, \edgeR{}, and \baySeq{} as described in Section \ref{s:alternative}, the \ShrinkBayes{} approach described in Section \ref{s:shrinkbayes} and the eBayes approach described in Section \ref{s:ebayes} using both normal and laplace distributions for the half-parental difference and hybrid effect.
+\caption{Area under the ROC curves (AUC) below a false positive rate of 0.1 for 3 different replicates per variety for the approaches using \citeauthor{ji2014estimation}, \edgeR{}, and \baySeq{} as described in Section \ref{s:alternative}, the \ShrinkBayes{} approach described in Section \ref{s:shrinkbayes} and the eBayes approach described in Section \ref{s:ebayes} using both normal and laplace distributions for the half-parental difference and hybrid effect.
 Each line is a different simulation while the blue box indicates mean AUC (plus or minus one standard error).}
@@
 \end{figure}
-The pattern appears to hold, with the eBayes and \ShrinkBayes{} methods outperforming the other two.
-With 4 replicates per variety, there does not appear to be much of a difference between \ShrinkBayes{} and our eBayes approach, but as the number of replicates increases our eBayes approach appears to improve relative to \ShrinkBayes{}.
+Similar to the single ROC curve, the eBayes and \ShrinkBayes{} methods appear to outperform the other method.
+With 4 replicates per variety, there does not appear to be much of a difference between \ShrinkBayes{} and the eBayes approaches, but as the number of replicates increases the eBayes approaches appear to improve relative to \ShrinkBayes{}.
+Two difference exist between the \ShrinkBayes{} and eBayes approaches that could explain the difference in AUC: 1) \ShrinkBayes{} utilizes a different empirical Bayes approach for estimating both the hyperparameters and the gene-specific parameters and 2) the measure of overdominance is slightly different due to INLA not providing a full posterior.
+
+There also appears to be a pattern of the eBayes approach with Laplace distributions for the half-parental differences and hybrid effect systematically performing better than the approach that uses normal distributions.
+We assume this is because the Laplace distributions are better approximations to the true underlying distribution for these parameters.
 

### 2015-08-24 ebayes_rnaseq_heterosis.tex
percentage improvement

@@
 \end{figure}
-Similar to the single ROC curve, the eBayes and \ShrinkBayes{} methods appear to outperform the other method.
-With 4 replicates per variety, there does not appear to be much of a difference between \ShrinkBayes{} and the eBayes approaches, but as the number of replicates increases the eBayes approaches appear to improve relative to \ShrinkBayes{}.
+Similar to the single ROC curve, the eBayes and \ShrinkBayes{} methods appear to outperform the other methods in terms of AUC.
+This improvement ranges from about a 20\% improvement over Ji et al.
+to about a 100\% improvement over \edgeR{} (which was not designed for overdominance testing).
+
+With 4 replicates per variety, there does not appear to be much of a difference between \ShrinkBayes{} and the eBayes approaches, but as the number of replicates increases, the eBayes approaches appear to improve relative to \ShrinkBayes{}.
 Two difference exist between the \ShrinkBayes{} and eBayes approaches that could explain the difference in AUC: 1) \ShrinkBayes{} utilizes a different empirical Bayes approach for estimating both the hyperparameters and the gene-specific parameters and 2) the measure of overdominance is slightly different due to INLA not providing a full posterior.

### 2015-08-24 response.tex
more responses

@@
 
-\todo{Why did we do this? The other methods fail?}
+\todo{During the course of our research, we were initially estimating hyperparameters based on moment matching of gene-specific parameters estimated using negative binomial regression independently for all genes.
+This procedure had issues whenever a gene-variety combination had counts that were all zeros, thus leading to an unidentified regression parameter, or too little variability, leading to an unidentified overdispersion parameter.
+This cut off appeared to fix the problem.
+None of the current methods suffer from this problem, in particular the \edgeR{} based parameter estimations have a built shrinkage to estimate parameters.
+We did not change the simulations we were using and thus report the deletion of low counts for reproducible accuracy.
+We do not believe removing this deletion would change any of our results.}
 
@@
 
-\todo{Still need to figure out how to respond.}
+\response{In Section 3.3, we highlighted the differences between ShrinkBayes and the eBayes approaches, but do not try to specifically state which difference is the cause.}
 
@@
 
-\todo{??}
+\response{We're sorry, but we don't understand this comment. We believe all the parameters in the model are identified and therefore are not sure what non-identifiability is being referenced.}
 
@@
 
+\response{We assume that you mean the improvement compared to existing methods is underwhelming.
+To more accurately address the improvement, we calculated percentage improvement which ranges from 20\% to 100\% and discussed these in Section 3.3.}
+
 \comment{(1) In the simulations, the authors should also compare the eBayes method to the method of Ji et al.
@@
 
-\response{We have assessed the approach of Ji et al. on transformed data, see Sections 3.2 and 3.3.}
+\response{We have assessed the approach of Ji et al.
+on transformed data, see Sections 3.2 and 3.3.
+The bottom line is that it doesn't perform as well as the methods based on our hierarchical negative-binomial model.}
 
@@
 
-\todo{I think the main point is that they aren't very different. Somehow I am not able to communicate that the model is novel and that here are a couple of different estimation approaches.}
+\response{The area under the curve attempts to do exactly this, i.e.
+a single metric, to compare the methods.
+We do not try to over-emphasize the difference between the ShrinkBayes and eBayes methods, which are both based on our novel model, as we do not believe there is much of a difference.}
 
@@
 
+\response{We simulate data from the negative-binomial portion of the model, but do not simulate from the hierarchical portion of the model.
+Instead, we use \edgeR{} estimates for gene-specific parameters which provide some shrinkage for the overdispersion parameters but none for the other gene-specific parameters.
+We have tried to make this clear in Section 3.1.
+In addition, we have performed the analysis by simulating entirely from our model and achieve similar results.}
+
+\response{We also tried to emphasize in Section 2 that ``our'' approach includes the implementation in ShrinkBayes as the model itself is novel and the eBayes and ShrinkBayes are just two different approaches to estimation.}
+
 \comment{a. How robust do you feel your method is?  There is no discussion along these lines.}
@@
 
-\todo{Figure out how to respond.}
+\response{We have added discussion Section 3.3.
+This discussion points out the two main differences between ShrinkBayes and the eBayes approaches, but does not try to state which is the cause because, frankly, we don't know.}
 

### 2015-08-24 supp/sim/ji-sim-script.R
reorganized supp into directories

@@ -5,10 +5,10 @@ library(reshape2)
 
 source("args.R")
 
-source("empirical_estimate_function.R")
+source("../common/empirical_estimate_function.R")
 
 # Data set
-data_file = paste0("sim-data/sim-", r, "-", i,'.rds')
+data_file = paste0("data/sim-", r, "-", i,'.rds')
 
 d = as.data.frame(readRDS(data_file)); n = ncol(d)/3
 names(d) = c(paste("B73_",1:n, sep=''), paste('Mo17_',1:n, sep=''), paste('B73xMo17_',1:n, sep=''))

### 2015-08-24 supp/sim/create_Makefile.R
reorganized supp into directories

@@ -43,14 +43,14 @@ for (m in methods) {
     
     # Vectors of files
     result_file = paste0('results/',m,'/results-', r, '-', 1:n_sims,'.rds')
-    data_file   = paste0('sim-data/sim-',          r, '-', 1:n_sims,'.rds')
-    Rout_file   = paste0('Rout/',m,'/sim-',        r, '-', 1:n_sims,'.Rout')
+    data_file   = paste0('data/sim-',          r, '-', 1:n_sims,'.rds')
+    Rout_file   = paste0('Rout/',m,'/',        r, '-', 1:n_sims,'.Rout')
     
     catf(m, r, ': ', paste(result_file, collapse=' '),'\n')
     
     for (i in 1:n_sims) {
       catf(result_file[i], 
-           ifelse(m=='ji', ": ji-sim-script.R ", ": sim-script.R "), # a hack
+           ifelse(m=='ji', ": ji-sim-script.R ", ": script.R "), # a hack
            data_file[i], 
            "\n\t", 
            "$(RCMD) '--args r=", 
@@ -59,7 +59,7 @@ for (m in methods) {
            i,
            ' m=\"', 
            m, 
-           ifelse(m=='ji', "\"' ji-sim-script.R ", "\"' sim-script.R "), # hack
+           ifelse(m=='ji', "\"' ji-sim-script.R ", "\"' script.R "), # hack
            Rout_file[i], 
            " \n")
     }

### 2015-08-24 supp/sim/sim-script.R

@@ -6,15 +6,15 @@ library(edgeR)
 
 source("args.R")
 
-source("get_hyperparameters.R")
-source("single_gene_analysis.R")
+source("../common/get_hyperparameters.R")
+source("../common/single_gene_analysis.R")
 
 # Compile stan model
-if (m == 'laplace') model = stan_model("laplace.stan")
-if (m == 'normal' ) model = stan_model("normal.stan")
+if (m == 'laplace') model = stan_model("../common/laplace.stan")
+if (m == 'normal' ) model = stan_model("../common/normal.stan")
 
 
-data_file = paste0("sim-data/sim-", r, "-", i,'.rds')
+data_file = paste0("data/sim-", r, "-", i,'.rds')
 
 d = as.data.frame(readRDS(data_file)); n = ncol(d)/3
 names(d) = c(paste("B73_",1:n, sep=''), paste('Mo17_',1:n, sep=''), paste('B73xMo17_',1:n, sep=''))

### 2015-08-24 response.tex
one more comment
fixed edgeR

@@
 
-\todo{We don't have artificially large signals. I didn't perceive the ``apples vs. oranges'' comments from the reviews. Nonetheless, I agree.}
+\response{We have assessed the Ji et al. approach and included it in Sections 3.2 and 3.3.}
 
@@
 This cut off appeared to fix the problem.
-None of the current methods suffer from this problem, in particular the \edgeR{} based parameter estimations have a built shrinkage to estimate parameters.
+None of the current methods suffer from this problem, in particular the edgeR based parameter estimations have a built shrinkage to estimate parameters.
 We did not change the simulations we were using and thus report the deletion of low counts for reproducible accuracy.
@@
 \response{We simulate data from the negative-binomial portion of the model, but do not simulate from the hierarchical portion of the model.
-Instead, we use \edgeR{} estimates for gene-specific parameters which provide some shrinkage for the overdispersion parameters but none for the other gene-specific parameters.
+Instead, we use edgeR estimates for gene-specific parameters which provide some shrinkage for the overdispersion parameters but none for the other gene-specific parameters.
 We have tried to make this clear in Section 3.1.

### 2015-08-24 supp/real/normal-parallel.R
distinct script

@@ -1,0 +1,48 @@
+library(rstan)
+library(plyr)
+library(dplyr)
+library(reshape2)
+library(edgeR)
+
+source("../common/get_hyperparameters.R")
+source("../common/single_gene_analysis.R")
+
+##################################################
+# Get hyperparameters
+##################################################
+d = read.csv("data/data.csv", row.names=1)
+variety = factor(gsub("_[0-9]{1,2}", "", names(d)), levels=c("B73","Mo17","B73xMo17"))
+
+# trim genes for average count greater than 1
+#d = d[which(rowMeans(d)>1),] 
+hyperparameters = get_hyperparameters(d, variety, method="normal")
+
+
+
+##################################################
+# Perform individual gene analsysis (in parallel)
+##################################################
+
+# Compile stan model
+model = stan_model("../common/normal.stan")
+
+# This will take a long time, so parallelize if possible
+require(doMC)
+registerDoMC(5)
+
+sink("script.temp")
+
+analysis = adply(d,
+                 1,
+                 function(x) single_gene_analysis(x),
+                 .id = 'gene',
+                 .parallel = TRUE,
+                 .paropts = list(.export=c('single_gene_analysis','model','hyperparameters'), .packages='rstan'))
+
+sink()
+unlink("script.temp")
+
+rownames(analysis) = rownames(d)
+saveRDS(analysis, file="results/normal-parallel.rds")
+
+q(ifelse(interactive(), "ask", "no"))

### 2015-08-24 supp/real/normal.R

@@ -1,0 +1,44 @@
+library(rstan)
+library(plyr)
+library(dplyr)
+library(reshape2)
+library(edgeR)
+
+source("../common/get_hyperparameters.R")
+source("../common/single_gene_analysis.R")
+
+##################################################
+# Get hyperparameters
+##################################################
+d = read.csv("data/data.csv", row.names=1)
+variety = factor(gsub("_[0-9]{1,2}", "", names(d)), levels=c("B73","Mo17","B73xMo17"))
+
+# trim genes for average count greater than 1
+#d = d[which(rowMeans(d)>1),] 
+hyperparameters = get_hyperparameters(d, variety, method="normal")
+
+
+
+##################################################
+# Perform individual gene analsysis (in parallel)
+##################################################
+
+# Compile stan model
+model = stan_model("../common/normal.stan")
+
+sink("script.temp")
+
+analysis = adply(d,
+                 1,
+                 function(x) single_gene_analysis(x),
+                 .id = 'gene',
+                 .parallel = FALSE,
+                 .paropts = list(.export=c('single_gene_analysis','model','hyperparameters'), .packages='rstan'))
+
+sink()
+unlink("script.temp")
+
+rownames(analysis) = rownames(d)
+saveRDS(analysis, file="results/normal.rds")
+
+q(ifelse(interactive(), "ask", "no"))

### 2015-08-24 supp/real/laplace-parallel.R
distinct script

@@ -4,18 +4,18 @@ library(dplyr)
 library(reshape2)
 library(edgeR)
 
-source("get_hyperparameters.R")
-source("single_gene_analysis.R")
+source("../common/get_hyperparameters.R")
+source("../common/single_gene_analysis.R")
 
 ##################################################
 # Get hyperparameters
 ##################################################
-d = read.csv("data.csv", row.names=1)
+d = read.csv("data/data.csv", row.names=1)
 variety = factor(gsub("_[0-9]{1,2}", "", names(d)), levels=c("B73","Mo17","B73xMo17"))
 
 # trim genes for average count greater than 1
 #d = d[which(rowMeans(d)>1),] 
-hyperparameters = get_hyperparameters(d, variety)
+hyperparameters = get_hyperparameters(d, variety, method='laplace')
 
 
 
@@ -24,15 +24,11 @@ hyperparameters = get_hyperparameters(d, variety)
 ##################################################
 
 # Compile stan model
-model = stan_model("model.stan")
+model = stan_model("../common/laplace.stan")
 
 # This will take a long time, so parallelize if possible
-if (parallel <- require(doMC)) {
-  registerDoMC()
-} else if (parallel <- require(doParallel)) {
-  cl = makeCluster(detectCores(), type='PSOCK')
-  registerDoParallel(cl)
-}
+require(doMC)
+registerDoMC(5)
 
 sink("script.temp")
 
@@ -40,13 +36,13 @@ analysis = adply(d,
                  1,
                  function(x) single_gene_analysis(x),
                  .id = 'gene',
-                 .parallel = parallel,
+                 .parallel = TRUE,
                  .paropts = list(.export=c('single_gene_analysis','model','hyperparameters'), .packages='rstan'))
 
 sink()
 unlink("script.temp")
 
 rownames(analysis) = rownames(d)
-saveRDS(analysis, file="script.rds")
+saveRDS(analysis, file="results/laplace-parallel.rds")
 
 q(ifelse(interactive(), "ask", "no"))

### 2015-08-24 supp/real/laplace.R

@@ -1,0 +1,44 @@
+library(rstan)
+library(plyr)
+library(dplyr)
+library(reshape2)
+library(edgeR)
+
+source("../common/get_hyperparameters.R")
+source("../common/single_gene_analysis.R")
+
+##################################################
+# Get hyperparameters
+##################################################
+d = read.csv("data/data.csv", row.names=1)
+variety = factor(gsub("_[0-9]{1,2}", "", names(d)), levels=c("B73","Mo17","B73xMo17"))
+
+# trim genes for average count greater than 1
+#d = d[which(rowMeans(d)>1),] 
+hyperparameters = get_hyperparameters(d, variety, method='laplace')
+
+
+
+##################################################
+# Perform individual gene analsysis (in parallel)
+##################################################
+
+# Compile stan model
+model = stan_model("../common/laplace.stan")
+
+sink("script.temp")
+
+analysis = adply(d,
+                 1,
+                 function(x) single_gene_analysis(x),
+                 .id = 'gene',
+                 .parallel = FALSE,
+                 .paropts = list(.export=c('single_gene_analysis','model','hyperparameters'), .packages='rstan'))
+
+sink()
+unlink("script.temp")
+
+rownames(analysis) = rownames(d)
+saveRDS(analysis, file="results/laplace.rds")
+
+q(ifelse(interactive(), "ask", "no"))

### 2015-08-24 supp/common/get_hyperparameters.R
add dependent libraries

@@ -1,4 +1,8 @@
 get_hyperparameters = function(d, variety, method) {
+  require(reshape2)
+  require(dplyr)
+  require(edgeR)
+  
   # phi, alpha, delta parameterization
   design = cbind(1,
                  c(1,-1,0)[as.numeric(variety)],

### 2015-08-24 supp/real/ji.R
fixed dirs

@@ -3,10 +3,10 @@ library(dplyr)
 library(reshape2)
 library(edgeR)
 
-source("empirical_estimate_function.R")
+source("../common/empirical_estimate_function.R")
 
 # Data set
-d = read.csv("data.csv")
+d = read.csv("data/data.csv")
 variety = factor(gsub("_[0-9]{1,2}", "", names(d)[-1]), levels=c("B73","Mo17","B73xMo17"))
 
 # GLM fit using edgeR
@@ -45,3 +45,7 @@ est = estimate(simd = NULL,              # Was used for ROC calculations
                I = tmp$n[1],             # number of reps per variety
                J = nrow(d),              # number of genes 
                mcn = 1000)               # Monte Carlo reps?
+
+q(ifelse(interactive(), "ask","no"))
+
+

### 2015-08-24 supp/real/ji.R
save Ji results

@@ -46,6 +46,8 @@ est = estimate(simd = NULL,              # Was used for ROC calculations
                J = nrow(d),              # number of genes 
                mcn = 1000)               # Monte Carlo reps?
 
+writeRDS(est, file='results/ji.rds')
+
 q(ifelse(interactive(), "ask","no"))
 
 

### 2015-08-25 supp/real/laplace-parallel.R
combine eBayes approaches into single file

@@ -1,48 +1,4 @@
-library(rstan)
-library(plyr)
-library(dplyr)
-library(reshape2)
-library(edgeR)
+parallel = TRUE
+method = 'laplace'
 
-source("../common/get_hyperparameters.R")
-source("../common/single_gene_analysis.R")
-
-##################################################
-# Get hyperparameters
-##################################################
-d = read.csv("data/data.csv", row.names=1)
-variety = factor(gsub("_[0-9]{1,2}", "", names(d)), levels=c("B73","Mo17","B73xMo17"))
-
-# trim genes for average count greater than 1
-#d = d[which(rowMeans(d)>1),] 
-hyperparameters = get_hyperparameters(d, variety, method='laplace')
-
-
-
-##################################################
-# Perform individual gene analsysis (in parallel)
-##################################################
-
-# Compile stan model
-model = stan_model("../common/laplace.stan")
-
-# This will take a long time, so parallelize if possible
-require(doMC)
-registerDoMC(5)
-
-sink("script.temp")
-
-analysis = adply(d,
-                 1,
-                 function(x) single_gene_analysis(x),
-                 .id = 'gene',
-                 .parallel = TRUE,
-                 .paropts = list(.export=c('single_gene_analysis','model','hyperparameters'), .packages='rstan'))
-
-sink()
-unlink("script.temp")
-
-rownames(analysis) = rownames(d)
-saveRDS(analysis, file="results/laplace-parallel.rds")
-
-q(ifelse(interactive(), "ask", "no"))
+source("eBayes.R")

### 2015-08-25 supp/real/laplace.R

@@ -1,44 +1,4 @@
-library(rstan)
-library(plyr)
-library(dplyr)
-library(reshape2)
-library(edgeR)
+parallel = FALSE
+method = 'laplace'
 
-source("../common/get_hyperparameters.R")
-source("../common/single_gene_analysis.R")
-
-##################################################
-# Get hyperparameters
-##################################################
-d = read.csv("data/data.csv", row.names=1)
-variety = factor(gsub("_[0-9]{1,2}", "", names(d)), levels=c("B73","Mo17","B73xMo17"))
-
-# trim genes for average count greater than 1
-#d = d[which(rowMeans(d)>1),] 
-hyperparameters = get_hyperparameters(d, variety, method='laplace')
-
-
-
-##################################################
-# Perform individual gene analsysis (in parallel)
-##################################################
-
-# Compile stan model
-model = stan_model("../common/laplace.stan")
-
-sink("script.temp")
-
-analysis = adply(d,
-                 1,
-                 function(x) single_gene_analysis(x),
-                 .id = 'gene',
-                 .parallel = FALSE,
-                 .paropts = list(.export=c('single_gene_analysis','model','hyperparameters'), .packages='rstan'))
-
-sink()
-unlink("script.temp")
-
-rownames(analysis) = rownames(d)
-saveRDS(analysis, file="results/laplace.rds")
-
-q(ifelse(interactive(), "ask", "no"))
+source("eBayes.R")

### 2015-08-25 supp/real/normal-parallel.R

@@ -1,48 +1,4 @@
-library(rstan)
-library(plyr)
-library(dplyr)
-library(reshape2)
-library(edgeR)
+parallel = TRUE
+method = 'normal'
 
-source("../common/get_hyperparameters.R")
-source("../common/single_gene_analysis.R")
-
-##################################################
-# Get hyperparameters
-##################################################
-d = read.csv("data/data.csv", row.names=1)
-variety = factor(gsub("_[0-9]{1,2}", "", names(d)), levels=c("B73","Mo17","B73xMo17"))
-
-# trim genes for average count greater than 1
-#d = d[which(rowMeans(d)>1),] 
-hyperparameters = get_hyperparameters(d, variety, method="normal")
-
-
-
-##################################################
-# Perform individual gene analsysis (in parallel)
-##################################################
-
-# Compile stan model
-model = stan_model("../common/normal.stan")
-
-# This will take a long time, so parallelize if possible
-require(doMC)
-registerDoMC(5)
-
-sink("script.temp")
-
-analysis = adply(d,
-                 1,
-                 function(x) single_gene_analysis(x),
-                 .id = 'gene',
-                 .parallel = TRUE,
-                 .paropts = list(.export=c('single_gene_analysis','model','hyperparameters'), .packages='rstan'))
-
-sink()
-unlink("script.temp")
-
-rownames(analysis) = rownames(d)
-saveRDS(analysis, file="results/normal-parallel.rds")
-
-q(ifelse(interactive(), "ask", "no"))
+source("eBayes.R")

### 2015-08-25 supp/real/parallel.R

@@ -1,0 +1,7 @@
+# Sets up parallelization if possible
+
+if (require(doMC)) {
+  registerDoMC(5)
+} else {
+  parallel=FALSE
+}

### 2015-08-25 supp/real/eBayes.R
combine eBayes approaches into single file

@@ -1,0 +1,52 @@
+library(rstan)
+library(plyr)
+library(dplyr)
+library(reshape2)
+library(edgeR)
+
+source("../common/get_hyperparameters.R")
+source("../common/single_gene_analysis.R")
+
+if (parallel) source("parallel.R")
+
+##################################################
+# Get hyperparameters
+##################################################
+d = read.csv("data/data.csv", row.names=1)
+variety = factor(gsub("_[0-9]{1,2}", "", names(d)), levels=c("B73","Mo17","B73xMo17"))
+
+# trim genes for average count greater than 1
+#d = d[which(rowMeans(d)>1),] 
+hyperparameters = get_hyperparameters(d, variety, method=method)
+
+
+
+##################################################
+# Perform individual gene analsysis (in parallel)
+##################################################
+
+# Compile stan model
+model = stan_model(paste0("../common/",method,".stan"))
+
+con = 'script.temp'
+sink(con)
+sink(con, type='message')
+
+analysis = adply(d,
+                 1,
+                 function(x) single_gene_analysis(x),
+                 .id = 'gene',
+                 .parallel = parallel,
+                 .paropts = list(.export=c('single_gene_analysis','model','hyperparameters'), .packages='rstan'))
+
+sink()
+sink(type='message')
+unlink(con)
+
+rownames(analysis) = rownames(d)
+saveRDS(analysis, file=paste0("results/",
+                              method,
+                              ifelse(parallel, '-parallel', NULL),
+                              ".rds"))
+
+q(ifelse(interactive(), "ask", "no"))

### 2015-08-25 supp/real/normal.R

@@ -1,44 +1,4 @@
-library(rstan)
-library(plyr)
-library(dplyr)
-library(reshape2)
-library(edgeR)
+parallel = FALSE
+method = 'normal'
 
-source("../common/get_hyperparameters.R")
-source("../common/single_gene_analysis.R")
-
-##################################################
-# Get hyperparameters
-##################################################
-d = read.csv("data/data.csv", row.names=1)
-variety = factor(gsub("_[0-9]{1,2}", "", names(d)), levels=c("B73","Mo17","B73xMo17"))
-
-# trim genes for average count greater than 1
-#d = d[which(rowMeans(d)>1),] 
-hyperparameters = get_hyperparameters(d, variety, method="normal")
-
-
-
-##################################################
-# Perform individual gene analsysis (in parallel)
-##################################################
-
-# Compile stan model
-model = stan_model("../common/normal.stan")
-
-sink("script.temp")
-
-analysis = adply(d,
-                 1,
-                 function(x) single_gene_analysis(x),
-                 .id = 'gene',
-                 .parallel = FALSE,
-                 .paropts = list(.export=c('single_gene_analysis','model','hyperparameters'), .packages='rstan'))
-
-sink()
-unlink("script.temp")
-
-rownames(analysis) = rownames(d)
-saveRDS(analysis, file="results/normal.rds")
-
-q(ifelse(interactive(), "ask", "no"))
+source("eBayes.R")

### 2015-08-25 supp/real/ji.R
typo

@@ -46,7 +46,7 @@ est = estimate(simd = NULL,              # Was used for ROC calculations
                J = nrow(d),              # number of genes 
                mcn = 1000)               # Monte Carlo reps?
 
-writeRDS(est, file='results/ji.rds')
+saveRDS(est, file='results/ji.rds')
 
 q(ifelse(interactive(), "ask","no"))
 

### 2015-08-25 supp/real/utils.R
add Will's files

@@ -1,0 +1,321 @@
+trim_genes <- function(counts, group, geneid, z.allow.tr = 3, mean.thr = exp(1)){
+  #counts is a num_gene x num_sample matrix
+  #group is a num_sample length factor mapping columns to groups
+  #geneid is a num_gene length vector rows to genes, i.e. 1:G or descriptive names
+  #z.allow.tr is maximum zeros allowable for any treatment(gene)
+  #mean.thr is a minimum average count for a gene
+  G = length(geneid)
+  if(G != nrow(counts)) stop("length(geneid) must match nrow(counts)")
+  if(length(group) != ncol(counts)) stop("length(group) must match ncol(counts)")
+  key_for_id = cbind(geneid,1:G)
+  counts = as.matrix(counts)
+  
+  #Which genes have a treatment with excessive zeros?
+  flag_zeros <- NULL
+  for(i in group){
+    flags  = apply(counts, MARGIN=1,FUN= function(x){
+        
+        sum(x[which(group==i)]==0) > z.allow.tr       #indicator for group i (for all genes) 
+        })
+        
+    flag_zeros = c(flag_zeros, which(flags))
+  }                  
+    
+  #Which genes have too low average expression?
+  flags2 = apply(counts, MARGIN=1, FUN = function(x) mean(x) < mean.thr)
+  flag_low = which(flags2)
+  
+  allflags = sort(unique(c(flag_zeros,flag_low)))
+  
+  
+  return(list(counts[-allflags,],geneid[-allflags]))
+}
+
+edgeR.fit = function(counts, group){
+  dge <- DGEList(counts = counts, group = group)
+  design <- model.matrix(~ -1 + group)
+  dge <- calcNormFactors(dge)
+  dge <- estimateGLMCommonDisp(dge, design)
+  dge <- estimateGLMTagwiseDisp(dge, design)
+
+  glmFit(dge, design)
+}
+
+edgeR.ord = function(fit){
+  apply(fit$coef, 1, function(x){
+    if(x["grouphybrid"] < min(x[c("groupparent1", "groupparent2")]))
+      return("hybrid < min")
+    else if(x["grouphybrid"] > max(x[c("groupparent1", "groupparent2")]))
[…]

### 2015-08-25 supp/real/real_data_pvals.R
add Will's files

@@ -1,0 +1,22 @@
+library(edgeR)
+library(baySeq)
+library(ShrinkBayes)
+
+source("andyspost.R")
+source("utils.R")
+
+real_data_pvals = function(ncores = 1){
+  group = as.factor(rep(c("parent1", "parent2", "hybrid"), each = 4))
+  cts = as.matrix(readRDS("paschold_counts.rds"))
+  cts = trim_genes(cts, group, rownames(cts))[[1]]
+  for(mtd in c("edgeR", "baySeq", "ShrinkBayes")){
+    for(cp in ncores){
+      print(mtd)
+      print(cp)
+      t = proc.time()
+      r = ranks1dataset(mtd, 4, 1, cts, group, ncpus = ncores)
+      saveRDS(proc.time() - t, paste0("time_", mtd, "_", cp, "_cores", ".rds"))
+      saveRDS(r, paste0("pvals_", mtd, "_", cp, "_cores", ".rds"))
+    }
+  }
+}

### 2015-08-25 supp/real/real_data_pvals.R
do not use ShrinkBayes

@@ -1,6 +1,6 @@
 library(edgeR)
 library(baySeq)
-library(ShrinkBayes)
+#library(ShrinkBayes)
 
 source("andyspost.R")
 source("utils.R")
@@ -9,7 +9,7 @@ real_data_pvals = function(ncores = 1){
   group = as.factor(rep(c("parent1", "parent2", "hybrid"), each = 4))
   cts = as.matrix(readRDS("paschold_counts.rds"))
   cts = trim_genes(cts, group, rownames(cts))[[1]]
-  for(mtd in c("edgeR", "baySeq", "ShrinkBayes")){
+  for(mtd in c("edgeR", "baySeq")){ # add back "ShrinkBayes"
     for(cp in ncores){
       print(mtd)
       print(cp)

### 2015-08-25 supp/real/real_data_pvals.R
don't trim genes

@@ -8,7 +8,7 @@ source("utils.R")
 real_data_pvals = function(ncores = 1){
   group = as.factor(rep(c("parent1", "parent2", "hybrid"), each = 4))
   cts = as.matrix(readRDS("paschold_counts.rds"))
-  cts = trim_genes(cts, group, rownames(cts))[[1]]
+#  cts = trim_genes(cts, group, rownames(cts))[[1]]
   for(mtd in c("edgeR", "baySeq")){ # add back "ShrinkBayes"
     for(cp in ncores){
       print(mtd)

### 2015-08-25 supp/real/real_data_pvals.R
run the function

@@ -20,3 +20,5 @@ real_data_pvals = function(ncores = 1){
     }
   }
 }
+
+real_data_pvals()

### 2015-08-25 supp/real/laplace.R
echo=T

@@ -1,4 +1,4 @@
 parallel = FALSE
 method = 'laplace'
 
-source("eBayes.R")
+source("eBayes.R", echo=TRUE)

### 2015-08-25 supp/real/normal.R

@@ -1,4 +1,4 @@
 parallel = FALSE
 method = 'normal'
 
-source("eBayes.R")
+source("eBayes.R", echo=TRUE)

### 2015-08-25 supp/real/laplace-parallel.R
echo=T

@@ -1,4 +1,4 @@
 parallel = TRUE
 method = 'laplace'
 
-source("eBayes.R")
+source("eBayes.R", echo=TRUE)

### 2015-08-25 supp/real/normal-parallel.R

@@ -1,4 +1,4 @@
 parallel = TRUE
 method = 'normal'
 
-source("eBayes.R")
+source("eBayes.R", echo=TRUE)

### 2015-08-25 supp/real/eBayes.R
sink output and messages

@@ -28,9 +28,9 @@ hyperparameters = get_hyperparameters(d, variety, method=method)
 # Compile stan model
 model = stan_model(paste0("../common/",method,".stan"))
 
-con = 'script.temp'
-sink(con)
-sink(con, type='message')
+con = file('script.temp')
+sink(con, append=TRUE)
+sink(con, append=TRUE, type='message')
 
 analysis = adply(d,
                  1,
@@ -41,7 +41,8 @@ analysis = adply(d,
 
 sink()
 sink(type='message')
-unlink(con)
+close(con)
+unlink(con, force=TRUE)
 
 rownames(analysis) = rownames(d)
 saveRDS(analysis, file=paste0("results/",

### 2015-08-26 ebayes_rnaseq_heterosis.tex
added computation time

@@
 
+We analyzed the data using all the methods compared in the previous section.
+The computation time on a Dell Precision T3600 with a 3.6GHz intel Xeon processor was 14 seconds for \edgeR{}, 1.3 minutes for \citeauthor{ji2014estimation}, 10 hours for the eBayes approaches, and 17 hours for \baySeq{}.
+In the eBayes approach, the vast majority of the time is spent on independent MCMC analysis for each gene.
+Thus we parallelized this step using {\tt doMC} \citep{revolution2014domc} and {\tt plyr} \citep{hadley2011split}.
+When parallelized across 5 cores, we the eBayes approaches took about 2.5 hours.
+
 Figure \ref{f:gene_specific_estimates} provides point estimates from the initial \edgeR{} estimation step and after the eBayes procedure described in Section \ref{s:ebayes}. 
@@
 \centerline{\includegraphics[width=\textwidth]{gene_specific_estimates.pdf}}
-\caption{Two-dimensional histogram of point estimates from \edgeR{} and eBayes (posterior means) along with the $y=x$ diagonal.}
+\caption{Two-dimensional histogram of point estimates from \edgeR{} and posterior means from eBayes using Laplace distributions along with the $y=x$ diagonal.}
 \label{f:gene_specific_estimates}
@@
 \centerline{\includegraphics[width=\textwidth]{estimates}}
-\caption{Marginal and bivariate histograms for gene-specific parameter posterior meansfor the B73 $\times$ Mo17 maize experiment.
-For reference, the marginal histograms have a dashed line indicating the appropriate marginal probability density functions in the hierarchical model with the MAP estimates.}
+\caption{Marginal and bivariate histograms of posterior means for gene-specific parameters for the B73 $\times$ Mo17 maize experiment.}
 \label{f:scatterplot}

### 2015-08-26 ebayes_rnaseq_heterosis.tex
fixed up real data results

@@
 We used our method to analyze a maize data set \citep{paschold2012complementation} of \RNAseq{} gene expression in parental lines B73 and Mo17 and the hybrid genotype (B73$\times$Mo17).
-The data are available in the Supplemental Materials.
 Each variety had four biological replicates measured with Illumina methodology and equipment.
@@
 For more specifics, please see \cite{paschold2012complementation}.
-
-As explained in Section~\ref{s:sim_data}, we analyzed a set of $G=27,888$ genes for $V=3$ varieties and $n_v=4$ replicates ($v=1,2,3$).
-We estimated the hyperparameters to be $\hat{\eta}_\phi = 3.44$, $\hat{\sigma}_\phi = 2.74$, $\hat{\eta}_\alpha = -0.05$, $\hat{\sigma}_\alpha=0.36$, $\hat{\eta}_\delta = 0.08$, $\hat{\sigma}_\delta=0.33$, $\hat{\eta}_\psi = -2.35$, and $\hat{\sigma}_\psi=0.60$.
 
@@
 
-Figure \ref{f:gene_specific_estimates} provides point estimates from the initial \edgeR{} estimation step and after the eBayes procedure described in Section \ref{s:ebayes}. 
+For the eBayes approach with the Laplace distributions, we estimated the hyperparameters to be $\hat{\eta}_\phi = 3.44$, $\hat{\sigma}_\phi = 2.74$, $\hat{\eta}_\alpha = -0.05$, $\hat{\sigma}_\alpha=0.36$, $\hat{\eta}_\delta = 0.08$, $\hat{\sigma}_\delta=0.33$, $\hat{\eta}_\psi = -2.35$, and $\hat{\sigma}_\psi=0.60$.
+We then performed independent MCMC analysis for each gene conditional on these hyperparameters.
+As with the simulation study, we ran 4 chains simultaneously and doubled the number of MCMC iterations until each gene-specific parameter had an effective sample size above 1,000.
+
+Figure \ref{f:gene_specific_estimates} provides point estimates of the gene-specific parameters from the initial \edgeR{} estimation step and after the eBayes with Laplace distribution procedure described in Section \ref{s:ebayes}.
 \begin{figure}

### 2015-08-26 ebayes_rnaseq_heterosis.tex
added allele-specific expression

@@
 The empirical Bayes posteriors can be used to estimate posterior probabilities of high and low parent heterosis.
-Through a simulation study, we demonstrated that this method outperformed alternative methods (that are not designed to detect heterosis), and performed comparably well with a similar model in \ShrinkBayes{}, which estimates the full posterior via INLA.
+Through a simulation study, we demonstrated that this method outperformed alternative methods, and performed comparably well with a similar model in \ShrinkBayes{}, which estimates the posterior via INLA.
 We then demonstrated the use of the methodology on a maize experiment in which phenotypic heterosis is well known.
@@
 Notwithstanding these improvements, we believe our approach is a computationally efficient method that can immediately aid scientists who are interested in identifying candidate genes for a genetic mechanism of heterosis.
+
+An alternative explanation for phenotypic heterosis is preferrential gene expression of alleles from certain parents for certain genes.
+Analyses of allele-specific expression \citep{bell2013rna,wei2013computational} could cmoplement the search for gene expression heterosis in the quest for the mechanism behind phenotypic heterosis.
 

### 2015-08-26 response.tex
all todo's removed except robust

@@
 
+\usepackage{fullpage}
 \usepackage{color}
@@
 
-\todo{Add references to allele specific expression.}
+\response{Added at the end of the Discussion section}
 
@@
 
-\todo{How would we compare this?}
+\response{We have provided a comparison of the computational time on the real data set using these methods.}
 
@@
 
-\todo{Add computing time.}
+\response{Added. See section 4.}
 
@@
 
-\todo{??}
+\response{We cannot answer why most RNAseq data analysts gravitate to this specification. We appreciate the reference and will consider the alternative specification in the future.}
 
@@
 
-\todo{During the course of our research, we were initially estimating hyperparameters based on moment matching of gene-specific parameters estimated using negative binomial regression independently for all genes.
+\response{ShrinkBayes failed for us when too many zeros were present. We do not believe the comparison of other methods would change if we did not remove the low counts.}
+
+%\todo{During the course of our research, we were initially estimating hyperparameters based on moment matching of gene-specific parameters estimated using negative binomial regression independently for all genes.
 This procedure had issues whenever a gene-variety combination had counts that were all zeros, thus leading to an unidentified regression parameter, or too little variability, leading to an unidentified overdispersion parameter.
@@
 
-\todo{No, because we did not simulate from our model since we wanted to maintain the inherent structure of the gene-specific parameters.
-Perhaps include a sentence based on the other simulation study.
-Perhaps say that we do not believe they are estimated very well.}
+\response{Since we did not simulate data from our model, there are no true values of the hyperparameters and therefore we cannot directly compare our estimates to a truth.
+We did try to point out in the Discussion that we believe these estimates could be improved.}
 
@@
 
-\todo{Add discussion of quality of hyperparameter estimates.}
+\response{As mentioned above, as we have no true values of the hyperparameters, we cannot quantitatively assess the quality of the estimates..}
 
@@
 
-\todo{We will do this.}
+\response{Added. See section 4.}
 
@@
 \response{The area under the curve attempts to do exactly this, i.e.
-a single metric, to compare the methods.
+a single metric to compare the methods.
 We do not try to over-emphasize the difference between the ShrinkBayes and eBayes methods, which are both based on our novel model, as we do not believe there is much of a difference.}
@@
 
-\todo{Dan? Even though some genes may be good candidates. An exploratory analysis to suggest other genes seems relevant.}
+\response{Even though scientists may have specific candidates. An exploratory analysis to suggest other genes seems useful.}
 
@@
 
-\todo{Good question. Although we suggest the gene-specific parameter distributions do not fit very well, how much better do we really think we can get if we model it well?}
+\todo{Add robustness discussion?}
 
@@
 
-\response{We have added discussion Section 3.3.
+\response{We have added discussion in Section 3.3.
 This discussion points out the two main differences between ShrinkBayes and the eBayes approaches, but does not try to state which is the cause because, frankly, we don't know.}
@@
 
-\todo{Yes, but we haven't created artificially large signals.}
+\response{We have added discussion at the end of Section 3.1.
+The bottom line is that the ROC curves are not particularly impressive because we did not use artificially large signals, but we feel this is likely representative of real data.}
 
@@
 
-\todo{Due to the nature of an empirical Bayes analysis, we are learning nothing about the distribution of the gene-specific parameter estimates other than through the hyperparameters.
-Yes, we can learn something about the parameters for a particular gene, but the information content in the data is very poor, i.e.
-only 12 total observations to infer about 4 parameters.
-Had we performed a fully Bayesian analysis, we could have learned more about the distribution of the gene-specific parameter estimates via the joint distributions of the hyperparameters (we believe this is what you are referring to).
-We believe that improved estimation of the gene-specific parameter distribution should lead to improved ability to detect heterosis.}
+\response{Yes, we believe it does matter.
+Due to the nature of an empirical Bayes analysis, we only learn about the joint distribution of the gene-specific parameters through the hyperparameters, i.e.
+the gene-specific parameters will still be assumed to be independent and from the marginal distribution that was assumed (normal or Laplace).
+As an example, if the expression level ($\phi$) and overdispersion ($\psi$) parameters were assumed to have a bivariate normal distribution, then when inferring about $\psi$ we would use information about $\phi$, i.e.
+larger values of $\phi$ lead, on average, to smaller values for $\psi$.
+But this does not occur in the current model.}
 

### 2015-08-26 response.tex
updated some responses

@@
 
-\response{Added at the end of the Discussion section}
+\response{Added at the end of the Discussion section.}
 
@@
 
-\response{We have provided a comparison of the computational time on the real data set using these methods.}
+\response{We have provided a comparison of the computational time on the real data set using these methods.} \todo{What else should we do?}
 
@@
 
-\response{Yes, overdominance is the scenario were are interested in. We have added the term in the introduction as well as in Section 2.3 to make this clear.}
+\response{Yes, overdominance is the scenario were are interested in. We have added the term in the introduction as well as throughout the manuscript to make this clear.}
 
@@
 
-\response{It is not necessary, but it is sufficient. If the parents are the same, then the hybrid must have overdominance.}
+\response{It is not necessary, but it is sufficient. If parental expressions are the same and the hybrid is not, then the hybrid must be exhibiting overdominance.}
 
@@
 
-\response{As mentioned above, as we have no true values of the hyperparameters, we cannot quantitatively assess the quality of the estimates..}
+\response{As mentioned above, as we have no true values of the hyperparameters, we cannot quantitatively assess the quality of the estimates.}
 
@@
 \response{We assume that you mean the improvement compared to existing methods is underwhelming.
-To more accurately address the improvement, we calculated percentage improvement which ranges from 20\% to 100\% and discussed these in Section 3.3.}
+To more accurately address the improvement, we calculated percentage improvement in AUC which ranges from 20\% to 100\% and discussed these in Section 3.3.}
 
@@
 a single metric to compare the methods.
-We do not try to over-emphasize the difference between the ShrinkBayes and eBayes methods, which are both based on our novel model, as we do not believe there is much of a difference.}
+We do not try to over-emphasize the difference between the ShrinkBayes and eBayes methods, which are both based on our model, as we do not believe there is much of a difference.}
 
@@
 
-\response{We also tried to emphasize in Section 2 that ``our'' approach includes the implementation in ShrinkBayes as the model itself is novel and the eBayes and ShrinkBayes are just two different approaches to estimation.}
+\response{We also tried to emphasize in Section 2 that ``our'' approach includes the implementation in ShrinkBayes as the model itself is novel and eBayes and ShrinkBayes are just two different approaches to estimation.}
 

### 2015-08-27 supp/real/eBayes.R
fixed RDS file name

@@ -47,7 +47,7 @@ unlink(con, force=TRUE)
 rownames(analysis) = rownames(d)
 saveRDS(analysis, file=paste0("results/",
                               method,
-                              ifelse(parallel, '-parallel', NULL),
+                              ifelse(parallel, '-parallel', ''),
                               ".rds"))
 
 q(ifelse(interactive(), "ask", "no"))

### 2015-08-31 ebayes_rnaseq_heterosis.tex
Dan changes other than overdominance

@@
 
-Heterosis, or hybrid vigor, occurs when hybrid progeny display phenotypes that are superior to the average phenotypes of their parents.
+Heterosis exists when the expected value of a hybrid phenotype differs from the average of the expected phenotypic values of the hybrid's parents.
 The heterosis phenomenon was scientifically documented in plants by \cite{darwin1876effects} and has long been used to improve agricultural production.
@@
 
-To use \RNAseq{} counts to identify genes displaying expression heterosis, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the parameters using an empirical Bayes procedure implemented in the statistical software \Stan{}, and calculate empirical Bayes posterior probabilities for heterosis.
+To use \RNAseq{} counts to identify genes displaying expression heterosis, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the parameters using an empirical Bayes procedure, and calculate empirical Bayes posterior probabilities for heterosis.
 
@@
 Following \cite{ji2014estimation}, we reparameterize the gene-variety mean structure into the gene-specific parental average ($\phi_g$), half-parental difference ($\alpha_g$), and hybrid difference ($\delta_g$).
-Specifically, for our heterosis where $V=3$, we let $v=1,2$ indicate the two parental varieties and $v=3$ indicate the hybrid, then
+For our heterosis study where $V=3$, we let $v=1,2$ indicate the two parental varieties and $v=3$ indicate the hybrid.
+The reparameterization is
 \[ \phi_g = \frac{\mu_{g1}+\mu_{g2}}{2}, \quad 
@@
 We assume a hierarchical model for the gene-specific mean parameters and overdispersion parameters.
-Initially, we assume the parental averages, half-parently differences, hybrid effect, and overdispersion parameters follow normal distributions, i.e.
+Initially, we assume the parental averages, half-parental differences, hybrid effect, and overdispersion parameters follow normal distributions, i.e.
 \[ 
@@
 To obtain normalization factors $\hat{\gamma}$, we use the weighted trimmed mean of $M$ values (TMM) method of \cite{robinson2010scaling}.
-We then use the methods of \cite{mccarthy2012differential}, to obtain genewise dispersion estimates, $\hat{\psi}_g$, and their generalized linear model methods to obtain estimates for the remaining gene-specific parameters, i.e.
+We then use the methods of \cite{robinson2010edgeR} version 3.8.6, to obtain genewise dispersion estimates, $\hat{\psi}_g$, and their generalized linear model methods to obtain estimates for the remaining gene-specific parameters, i.e.
 $\hat{\phi}_g$, $\hat{\alpha}_g$, and $\hat{\delta}_g$.
-Using $\hat{\theta}_g = (\hat{\phi}_g,\hat{\alpha}_g,\hat{\delta}_g,\hat{\psi}_g)$, we estimate hyperparameters for the location and scale parameters in the hierarchical model using a central method of moments approach, e.g.
-$\hat{\eta}_\phi = \sum_{g=1}^G \hat{\phi}_g/G$ and $\hat{\sigma}_\phi^2 = \sum_{g=1}^G (\hat{\phi}_g - \hat{\eta}_\phi)^2/(G-1)$ (and similarly for $\hat{\eta}_\psi$ and $\hat{\sigma}_\psi$) and $\hat{\eta}_\alpha = \sum_{g=1}^G \hat{\alpha}_g/G$ and $2\hat{\sigma}_\alpha^2 = \sum_{g=1}^G (\hat{\alpha}_g - \hat{\eta}_\alpha)^2/(G-1)$ (and similarly for $\hat{\eta}_\delta$ and $\hat{\sigma}_\delta$).
+Using $\hat{\theta}_g = (\hat{\phi}_g,\hat{\alpha}_g,\hat{\delta}_g,\hat{\psi}_g)$, we estimate hyperparameters for the location and scale parameters in the hierarchical model using a central method of moments approach.
+For example, $\hat{\eta}_\phi = \sum_{g=1}^G \hat{\phi}_g/G$ and $\hat{\sigma}_\phi^2 = \sum_{g=1}^G (\hat{\phi}_g - \hat{\eta}_\phi)^2/(G-1)$ (and similarly for $\hat{\eta}_\psi$ and $\hat{\sigma}_\psi$) and $\hat{\eta}_\alpha = \sum_{g=1}^G \hat{\alpha}_g/G$ and $2\hat{\sigma}_\alpha^2 = \sum_{g=1}^G (\hat{\alpha}_g - \hat{\eta}_\alpha)^2/(G-1)$ (and similarly for $\hat{\eta}_\delta$ and $\hat{\sigma}_\delta$) for the model assuming Laplace distributions.
 
@@
 \Stan{} implements a variant of MCMC called Hamiltonion Monte Carlo \citep{neal2011mcmc} to obtain samples from the posterior in equation \eqref{e:condind}.
-We ran 4 simultaneous chains with random initial starting values for 1000 burnin (and tuning) iterations followed by another 1000 iterations thinned by 4 (to reduce storage space) for inference.
+We ran 4 simultaneous chains with random initial starting values for 1000 burn-in (and tuning) iterations followed by another 1000 iterations thinned by 4 (to reduce storage space) for inference.
 We monitored convergence using the potential scale reduction factor and effective sample size (ESS) for $\phi_g$, $\alpha_g$, $\delta_g$, and $\psi_g$ \citep{Gelm:Rubi:infe:1992}.
-If the minimum ESS was less than 1000, we reran the chains with double the iterations for both burnin and inference.
+If the minimum ESS was less than 1000, we reran the chains with double the iterations for both burn-in and inference.
 We continued this restarting and doubling until we obtained minimum ESS greater than 1000 for all parameters.
@@
 
-In the maize context that motivates this work, we are particularly interested in overdominance, either low-parent heterosis (LPH) or high-parent heterosis (HPH), in gene expression. 
+In the maize context that motivates this work, we are interested in overdominance, either low-parent heterosis (LPH) or high-parent heterosis (HPH), in gene expression. 
 For a specific gene $g$, LPH occurs when average expression in the hybrid is less than the average expression of either parent, i.e.
-$\delta_g < -|\alpha_g|$ (equivalently $\mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}$), and HPH occurs when average expression in the hybrid is greater than the average expression of either parent, i.e.
-$\delta_g > |\alpha_g|$.
+$\delta_g < -|\alpha_g|$ or, equivalently, $\mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}$), and HPH occurs when average expression in the hybrid is greater than the average expression of either parent, i.e.
+$\delta_g > |\alpha_g|$ or, equivalently, $\mu_{g3} > \max\{\mu_{g1},\mu_{g2}\}$).
 % and high-parent  heterosis (HPH) of gene expression occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$.
@@
 
+We will use the term eBayes to refer to the approaches defined in Sections \ref{s:model}-\ref{s:gene_heterosis} and add parenthetical labels ``normal'' and ``Laplace'' to distinguish between the two eBayes approaches.
+
 \subsection{Implementation in \ShrinkBayes{}} \label{s:shrinkbayes}
@@
 % \end{array} \right.
-P_{\ShrinkBayes{}}(\mbox{LPH}_g|y) \equiv P(\delta_g<-\alpha_g|y)\mathrm{I}\left(\hat{\delta}_g < -\hat{\alpha}_g \le 0\right) + P(\delta_g<\alpha_g|y) \mathrm{I}\left(\hat{\delta}_g < \hat{\alpha}_g \ge 0\right).
+P_{\ShrinkBayes{}}(\mbox{LPH}_g|y) \equiv P(\delta_g<-\alpha_g|y)\mathrm{I}\left(\hat{\delta}_g < -\hat{\alpha}_g \le 0\right) + P(\delta_g<\alpha_g|y) \mathrm{I}\left(\hat{\delta}_g < \hat{\alpha}_g \le 0\right).
 %P_{\ShrinkBayes{}}(HPH|y) \equiv \left\{ \begin{array}{ll}
@@
 Conversely, there are many non-overdominance genes whose value of $|\hat{\delta}_g|$ was only slightly smaller than $|\hat{\alpha}_g|$, but whose simulated data will look similar to many overdominance genes.
-Therefore, we expect it will be difficult to accurately identify overdominance genes, but believe this is indicative of real applications as well.
+Therefore, we expect it will be difficult to accurately identify overdominance genes, but believe this level of difficulty is representative of real applications.
 
@@
 
-We compared our method to that of \cite{ji2014estimation}, which assumes normality in the response, by modeling the logarithm of the RNAseq count plus one divided by the normalization factor, i.e.
-$\log([Y_{gvi}+1]/\gamma_{vi})$.
-Use of the normalization factor here provides two advantages: 1) counts are properly scaled by the thoroughness of the sequencing of the sample and 2) for genes with no variability within variety (which actually occurs in our maize data set), use of the normalization factors allows the approach of \citeauthor{ji2014estimation} to still execute.
+We compared our method to that of \cite{ji2014estimation}, which assumes normality in the response, by modeling the logarithm of the RNAseq count plus one scaled by the normalization factor, i.e.
+$\log([Y_{gvi}+1])-\gamma_{vi}$.
+Use of the normalization factor here provides two advantages: 1) counts are properly adjusted for the thoroughness of the sequencing of the sample and 2) for genes with no count variation within variety (which actually occurs in our maize data set), use of the normalization factors allows the approach of \citeauthor{ji2014estimation} to still execute.
 
@@
 To construct a measure of overdominance, we computed the maximum likelihood estimates of the $\mu_{gv}$ parameters for all genes using \edgeR{}'s built-in Fisher scoring algorithm, and then used likelihood ratio tests to calculate two $p$-values for each gene: $p_{g1}$  for testing $H_{g01}:\mu_{g1}=\mu_{g3}$ and $p_{g2}$ for testing $H_{g02}:\mu_{g2}=\mu_{g3}$.
-Then, for each gene, we define a new $p$-value denoted as $p_g$ and set to $p_g=1$ if the estimate of $\mu_{g3}$ falls between the estimates of $\mu_{g1}$ and $\mu_{g2}$ and $p_g=\max\{p_{g1},p_{g2}\}/2$ otherwise.
+Then, for each gene, we defined a new $p$-value denoted as $p_g$ and set to $p_g=1$ if the estimate of $\mu_{g3}$ falls between the estimates of $\mu_{g1}$ and $\mu_{g2}$ and $p_g=\max\{p_{g1},p_{g2}\}/2$ otherwise.
 (For all relevant significance thresholds $\omega$ near 0, it can be shown that rejecting the null hypothesis of no overdominance of expression for gene $g$ whenever $p_g\leq \omega$ results in a test that is asymptotically size $\omega$.)  We then use $1-p_g$ as a final measure positively associated with strength of evidence for overdominance.
@@
 
-For the methods in Sections \ref{s:ebayes}, \ref{s:shrinkbayes}, and \ref{s:alternative}, we sorted genes according to the computed measure of the strength of evidence for overdominance.
+For the methods in Sections \ref{s:method} and \ref{s:alternative}, we sorted genes according to the computed measure of the strength of evidence for overdominance.
 From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to distinguish genes with overdominance, as defined in Section \ref{s:sim_data},
 % From Dan, "why this citation?"
-from those without overdominance \citep{landau2013dispersion}. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
+from those without overdominance. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
 \begin{figure}[htbp]
@@
 \end{figure}
-The ROC curves are modest, e.g.
+The ROC curves indicate modest performance, e.g.
 for a false positive rate of 5\%, the best performing methods only achieved a true positive rate of just over 15\%.
-This is consistent with our expectation discussed in Section \ref{s:sim_data} due to the high signal-to-noise ratio in these simulated data.
+This is consistent with our expectation discussed in Section \ref{s:sim_data} due to the low signal-to-noise ratio in these simulated data.
 
@@
 Thus we parallelized this step using {\tt doMC} \citep{revolution2014domc} and {\tt plyr} \citep{hadley2011split}.
-When parallelized across 5 cores, we the eBayes approaches took about 2.5 hours.
+When parallelized across 5 cores, the eBayes approaches took about 2.5 hours.
 
@@
 
-Although our method appears to improve upon existing methods, we believe our approach can be improved by refining the hierarchical model for the gene-specific parameter distribution.
+Although our method appears to hold some advantage over existing methods, we believe our approach can be improved by refining the hierarchical model for the gene-specific parameter distribution.
 Figure \ref{f:scatterplot} shows marginal and bivariate histograms for the empirical Bayes posterior means for the gene-specific parameters.
@@
 the hyperparameters, will also improve detection of gene-expression heterosis.
-Our current method, based on moment matching of essentially independently estimated gene-specific parameters provides consistent estimators as the number of replicates per variety increases.
-But typically the number of replicates per variety is quite small, in our data there are only four replicates per variety, and therefore asymptotic justifications are deficient.
+Our current method, based on moment matching of essentially independently estimated gene-specific parameters, provides consistent estimators as the number of replicates per variety increases.
+But typically the number of replicates per variety is quite small.
+In our data there are only four replicates per variety, and therefore asymptotic justifications are deficient.
 There are a variety of hyperparameter estimation approaches to explore, e.g.
@@
 
-An alternative explanation for phenotypic heterosis is preferrential gene expression of alleles from certain parents for certain genes.
-Analyses of allele-specific expression \citep{bell2013rna,wei2013computational} could cmoplement the search for gene expression heterosis in the quest for the mechanism behind phenotypic heterosis.
+An alternative explanation for phenotypic heterosis is preferential gene expression of alleles from certain parents for certain genes.
+Analyses of allele-specific expression \citep{bell2013rna,wei2013computational} could complement the search for gene expression heterosis in the quest for understanding the molecular genetic mechanisms behind phenotypic heterosis.
 

### 2015-08-31 supp/real/figs.R
volcano plot for all approaches

@@ -15,6 +15,11 @@ ggplot(d, aes(effectiveSize, pmax(prob_LPH, prob_HPH))) +
 ggsave("../include/volcano.pdf", width=8, height=8)
   
 
+
+
+
+
+
 ##########################################################################################
 # Hyperparameter plot
 ##########################################################################################
@@ -58,7 +63,7 @@ library(dplyr)
 library(reshape2)
 library(edgeR)
 
-d = read.csv("data.csv", row.names=1)
+d = read.csv("data/data.csv", row.names=1)
 variety = factor(gsub("_[0-9]{1,2}", "", names(d)), levels=c("B73","Mo17","B73xMo17"))
 
 # phi, alpha, delta parameterization
@@ -113,3 +118,43 @@ ggsave(file="../include/gene_specific_estimates.pdf", width=8, height=8)
 
 
 
+
+
+####################################################################################
+# All volcano plots
+####################################################################################
+
+d = readRDS("results/laplace.rds")
+laplace = data.frame(p = pmax(d$prob_LPH, d$prob_HPH),
+                     e = d$effectiveSize,
+                     method = 'laplace')
+
+d = readRDS("results/normal.rds")
+normal = data.frame(p = pmax(d$prob_LPH, d$prob_HPH),
+                     e = d$effectiveSize,
+                     method = 'normal')
+
+d = readRDS("results/ji.rds")
+ji = data.frame(p = with(d, pmax(lph.p, hph.p)),
+                e = with(d,(delta.pos - abs.alpha.pos) * (delta.pos >  abs.alpha.pos) + 
+                           (delta.pos + abs.alpha.pos) * (delta.pos < -abs.alpha.pos)),
+                method = 'ji')
+
+# Need hat from above
+d = readRDS("pvals_edgeR_1_cores.rds")
+edgeR = data.frame(p = 1-d[names(d) %in% rownames(hat)],
+                   e = with(hat, (delta-abs(alpha)) *(delta > abs(alpha)) + 
+                                 (delta+abs(alpha)) * (delta < -abs(alpha))),
+                   method = 'edgeR')
+
+d = readRDS("pvals_baySeq_1_cores.rds")
+baySeq = data.frame(p = d[names(d) %in% rownames(hat)],
+                   e = edgeR$e),
+                   method = 'baySeq')
+
+ggplot(rbind(laplace,normal,ji,edgeR,baySeq), aes(e,p)) +
[…]

### 2015-08-31 response.tex
added todo
added comparison

@@
 
-\response{We have added a paragraph to the beginning of Section 2.2 to share our experience with a fully MCMC analysis which took too long and had poor mixing.}
+\response{We now compare the estimated gene-specific parameters to the eBayes estimates in Figure 3.
+We have added a paragraph to the beginning of Section 2.2 to share our experience with a fully MCMC analysis which took too long and had poor mixing.}
 
@@
 
-\response{Even though scientists may have specific candidates. An exploratory analysis to suggest other genes seems useful.}
+\response{Even though scientists may have specific candidates. An exploratory analysis to suggest other genes seems useful.} \todo{Updated references.}
 

### 2015-08-31 supp/real/figs.R
volcano plot for all approaches

@@ -141,20 +141,33 @@ ji = data.frame(p = with(d, pmax(lph.p, hph.p)),
                 method = 'ji')
 
 # Need hat from above
-d = readRDS("pvals_edgeR_1_cores.rds")
-edgeR = data.frame(p = 1-d[names(d) %in% rownames(hat)],
-                   e = with(hat, (delta-abs(alpha)) *(delta > abs(alpha)) + 
-                                 (delta+abs(alpha)) * (delta < -abs(alpha))),
-                   method = 'edgeR')
+tmp = readRDS("pvals_edgeR_1_cores.rds")
+d = data.frame(p = 2*(1-tmp)-1+1*(tmp==1),
+               gene = names(tmp),
+               method = 'edgeR')
+edgeR = merge(d, hat[,c('alpha','delta','gene')])
+edgeR$e = with(edgeR, 
+                (delta-abs(alpha)) * (delta >  abs(alpha)) + 
+                  (delta+abs(alpha)) * (delta < -abs(alpha)))
 
-d = readRDS("pvals_baySeq_1_cores.rds")
-baySeq = data.frame(p = d[names(d) %in% rownames(hat)],
-                   e = edgeR$e),
-                   method = 'baySeq')
 
-ggplot(rbind(laplace,normal,ji,edgeR,baySeq), aes(e,p)) +
+tmp = readRDS("pvals_baySeq_1_cores.rds")
+d = data.frame(p = 1-tmp,
+               gene = names(tmp),
+               method = 'baySeq')
+baySeq = merge(d, hat[,c('alpha','delta','gene')])
+baySeq$e = with(baySeq, 
+                (delta-abs(alpha)) * (delta >  abs(alpha)) + 
+                (delta+abs(alpha)) * (delta < -abs(alpha)))
+
+ggplot(rbind(laplace,
+             normal,
+             ji,
+             edgeR[,c('p','e','method')],
+             baySeq[,c('p','e','method')]), 
+       aes(e,p)) +
   stat_binhex() +
   facet_wrap(~method) + 
   theme_bw() + 
   scale_fill_gradientn(trans='log',breaks=c(1,10,100,1000), colours=c("gray","black")) +
-  labs(x="Effect size", y="Maximum heterosis probabilities")
+  labs(x="Effect size", y="Heterosis measure")

### 2015-08-31 ebayes_rnaseq_heterosis.tex
some changes

@@
 Although the heterosis phenomenon is extensively utilized in agriculture, the molecular basis of heterosis is still largely unknown.
-In an effort to understand phenotypic heterosis at the molecular level, researchers have begun to measure transcript abundance levels of thousands of genes in parental inbred lines and their hybrid offspring using RNA sequencing (\RNAseq{}) technology.
+In an effort to understand phenotypic heterosis at the molecular level, researchers are measuring transcript abundance levels of thousands of genes in parental inbred lines and their hybrid offspring using RNA sequencing (\RNAseq{}) technology.
 The resulting data allow researchers to search for evidence of gene expression heterosis as one potential molecular mechanism underlying heterosis of agriculturally important traits.
@@
 The heterosis phenomenon was scientifically documented in plants by \cite{darwin1876effects} and has long been used to improve agricultural production.
-Overdominance is a particularly extreme case of heterosis where the hybrid phenotype lies outside the range of the parental phenotypes.
+Overdominance is a particularly extreme case of heterosis where the expected value of the hybrid phenotype lies outside the range of the expected values of the parental phenotypes.
 One classic example involves hybrid maize offspring that are taller, faster to mature, and yield considerably more grain than their inbred parents \citep{hallauer1981quantitative, hallauer2010quantitative}.
@@
 Section \ref{s:simulation} presents a simulation study based on a maize experiment and compares our approach to alternative methods.
-Section \ref{s:maize} analyzes a maize experiment where phenotypic heterosis is well established and identifies genes demonstrating expression heterosis.
+Section \ref{s:maize} analyzes a maize experiment where phenotypic heterosis is well established and identifies genes demonstrating expression overdominance.
 The data and code necessary to perform this analysis are available in the Supplemental Material.
@@
 
-Following \cite{ji2014estimation}, we reparameterize the gene-variety mean structure into the gene-specific parental average ($\phi_g$), half-parental difference ($\alpha_g$), and hybrid difference ($\delta_g$).
+Following \cite{ji2014estimation}, we reparameterize the gene-variety mean structure into the gene-specific parental average ($\phi_g$), half-parental difference ($\alpha_g$), and hybrid effect ($\delta_g$).
 For our heterosis study where $V=3$, we let $v=1,2$ indicate the two parental varieties and $v=3$ indicate the hybrid.
@@
 %\gamma_{vi} \stackrel{ind}{\sim} N(0,\sigma_\gamma^2), \quad 
-\phi_g \stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2), 
-\alpha_g \stackrel{ind}{\sim} N(\eta_\alpha, \sigma_\alpha^2),
-\delta_g \stackrel{ind}{\sim} N(\eta_\delta, \sigma_\delta^2), \mbox{ and }
+\phi_g \stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2), \,
+\alpha_g \stackrel{ind}{\sim} N(\eta_\alpha, \sigma_\alpha^2), \,
+\delta_g \stackrel{ind}{\sim} N(\eta_\delta, \sigma_\delta^2), \, \mbox{ and } \,
 \psi_g \stackrel{ind}{\sim} N(\eta_\psi,\sigma_\psi^2). 
@@
 For these many genes with small effects, we might expect the hybrid to act similar to its parents and therefore also have many genes where the hybrid effect is small and only a few genes where this hybrid effect is large.
-For these reasons, we also assessed Laplace (or double exponential) distributions for the half-parental difference and hybrid effect and thus implement a Bayesian LASSO \citep{park2008bayesian}, i.e.
+For these reasons, we also assessed Laplace (or double exponential) distributions for the half-parental difference and hybrid effect and thus implement a Bayesian LASSO \citep{park2008bayesian,hans2009bayesian}, i.e.
 \[ 
@@
 
-Initial attempts to estimate the models above via Markov chain Monte Carlo (MCMC) failed due to high computational costs and poor mixing in the resulting chains.
+Initial attempts to perform a fully Bayesian analysis via Markov chain Monte Carlo (MCMC) failed due to high computational costs and poor mixing in the resulting chains.
 For example, we implemented the model in the statistical software \Stan{} (discussed at the end of this section), ran the MCMC on a simulated data set with 10,000 genes for 2 months on a state-of-the-art linux server, and obtained potential scale reduction factors that suggested we would need to run at least ten times as long to obtain convergence.
@@
 To obtain normalization factors $\hat{\gamma}$, we use the weighted trimmed mean of $M$ values (TMM) method of \cite{robinson2010scaling}.
-We then use the methods of \cite{robinson2010edgeR} version 3.8.6, to obtain genewise dispersion estimates, $\hat{\psi}_g$, and their generalized linear model methods to obtain estimates for the remaining gene-specific parameters, i.e.
-$\hat{\phi}_g$, $\hat{\alpha}_g$, and $\hat{\delta}_g$.
+We use \edgeR{} to obtain genewise dispersion estimates, $\hat{\psi}_g$, and the generalized linear model methods to obtain estimates for the remaining gene-specific parameters, i.e.
+$\hat{\phi}_g$, $\hat{\alpha}_g$, and $\hat{\delta}_g$ \citep{robinson2010edgeR}.
 Using $\hat{\theta}_g = (\hat{\phi}_g,\hat{\alpha}_g,\hat{\delta}_g,\hat{\psi}_g)$, we estimate hyperparameters for the location and scale parameters in the hierarchical model using a central method of moments approach.
@@
 For a specific gene $g$, LPH occurs when average expression in the hybrid is less than the average expression of either parent, i.e.
-$\delta_g < -|\alpha_g|$ or, equivalently, $\mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}$), and HPH occurs when average expression in the hybrid is greater than the average expression of either parent, i.e.
-$\delta_g > |\alpha_g|$ or, equivalently, $\mu_{g3} > \max\{\mu_{g1},\mu_{g2}\}$).
+$\delta_g < -|\alpha_g|$ or, equivalently, $\mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}$, and HPH occurs when average expression in the hybrid is greater than the average expression of either parent, i.e.
+$\delta_g > |\alpha_g|$ or, equivalently, $\mu_{g3} > \max\{\mu_{g1},\mu_{g2}\}$.
 % and high-parent  heterosis (HPH) of gene expression occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$.
@@
 One limitation with inferential methods based on INLA is that all distributions, except for the data distribution, must have tails as light or lighter than the normal density.
-Thus, we cannot implement the Laplace priors for the half-parental difference ($\alpha_g$) and the hybrid difference ($\delta_g$) and instead use normal priors in this situation.
+Thus, we cannot implement the Laplace priors for the half-parental difference ($\alpha_g$) and the hybrid effect ($\delta_g$) and instead use normal priors in this situation.
 An additional limitation is that INLA provides approximations to marginal posteriors for parameters or linear combinations of parameters, but not an approximation to the full posterior.
 Since we are interested in non-linear quantities such as $P(\delta_g > |\alpha_g||y)$, we cannot compute these directly using \ShrinkBayes{}.
-Instead, for \ShrinkBayes{}, we calculate heterosis probabilities conditional on posterior means for $\hat{\delta}_g$ and $\hat{\alpha}_g$, e.g.
+Instead, for \ShrinkBayes{}, we calculate heterosis probabilities conditional on posterior means for the half-parental difference and hybrid effect, i.e.
+$\hat{\delta}_g$ and $\hat{\alpha}_g$.
+For example,
 \[ 
@@
 Section \ref{s:maize} describes the maize dataset in detail.
-We compared our method to approaches using the R packages \edgeR{}, \baySeq{} \citep{hardcastle2010bayseq, hardcastle2012baySeq}, and \ShrinkBayes{} \citep{van2014shrinkbayes}.
+We compared our method to approaches using the R packages \edgeR{} , \baySeq{} \citep{hardcastle2010bayseq, hardcastle2012baySeq}, and \ShrinkBayes{} \citep{van2014shrinkbayes}.
 

### 2015-09-01 ebayes_rnaseq_heterosis.tex
overdominance -> heterosis

@@
 \begin{abstract}
-Heterosis, or hybrid vigor, refers to the enhancements in the phenotype of hybrid progeny relative to their inbred parents.
+An important type of heterosis, known as hybrid vigor, refers to the enhancements in the phenotype of hybrid progeny relative to their inbred parents.
 Although the heterosis phenomenon is extensively utilized in agriculture, the molecular basis of heterosis is still largely unknown.
@@
 Heterosis exists when the expected value of a hybrid phenotype differs from the average of the expected phenotypic values of the hybrid's parents.
-The heterosis phenomenon was scientifically documented in plants by \cite{darwin1876effects} and has long been used to improve agricultural production.
-Overdominance is a particularly extreme case of heterosis where the expected value of the hybrid phenotype lies outside the range of the expected values of the parental phenotypes.
+The most interesting and useful form of heterosis, known as hybrid vigor, occurs when hybrid progeny display a mean phenotype that is superior to  both parental phenotypic means.
+This heterosis phenomenon was scientifically documented in plants by \cite{darwin1876effects} and has long been used to improve agricultural production.
 One classic example involves hybrid maize offspring that are taller, faster to mature, and yield considerably more grain than their inbred parents \citep{hallauer1981quantitative, hallauer2010quantitative}.
-Despite intensive study and successful use of heterosis in agriculture, the basic molecular genetic mechanisms remain largely unknown \citep{coors1999genetics, lippman2007heterosis}.
-One potential explanation, known as gene expression heterosis, is enhanced expression of one or more genes in hybrids compared to their inbred parents \citep{swanson2006all, springer2007allelic}.
+
+Depending on whether large or small values of a phenotype are favorable, hybrid vigor can occur if the mean phenotype of a hybrid is greater than both parental means or less than both parental means.
+We refer to the former as high-parent heterosis (HPH) and the latter as low-parent heterosis (LPH).
+Note that heterosis, HPH, and LPH are similar to the quantitative genetics concepts of dominance, heterosis, and underdominance.
+However, the various forms of dominance are usually reserved for describing the association of mean phenotype with homozygous and heterozygous genotypes at a single genetic locus.
+Heterosis involves a comparison of inbreds (simultaneously homozygous at many loci) and hybrids (simultaneously heterozygous at many loci).
+For simplicity, throughout the remainder of the article, we will use the term heterosis to describe the situation where the hybrid mean is more extreme than the parental means, i.e.
+we say there is heterosis if and only if either HPH or LPH holds.
+
+Despite intensive study and successful use of heterosis in agriculture, the basic molecular genetic mechanisms remain poorly understood \citep{chen2013genomic}.
+One potential explanation, known as gene expression heterosis, is enhanced expression of one or more genes in hybrids compared to their inbred parents.
+Gene expression heterosis has been investigated in maize by \cite{swanson2006all} and \cite{springer2007allelic} and is conceptually depicted in the right column of Figure 1b in  \cite{chen2013genomic}.
 
@@
 The remainder of the paper proceeds as follows.
-Section \ref{s:method} presents the proposed hierarchical model, an empirical Bayes method of estimating the parameters, and the calculation of posterior probabilities of overdominance.
+Section \ref{s:method} presents the proposed hierarchical model, an empirical Bayes method of estimating the parameters, and the calculation of posterior probabilities of heterosis.
 Section \ref{s:simulation} presents a simulation study based on a maize experiment and compares our approach to alternative methods.
-Section \ref{s:maize} analyzes a maize experiment where phenotypic heterosis is well established and identifies genes demonstrating expression overdominance.
+Section \ref{s:maize} analyzes a maize experiment where phenotypic heterosis is well established and identifies genes demonstrating expression heterosis.
 The data and code necessary to perform this analysis are available in the Supplemental Material.
@@
 
-In the maize context that motivates this work, we are interested in overdominance, either low-parent heterosis (LPH) or high-parent heterosis (HPH), in gene expression. 
+In the maize context that motivates this work, we are interested in heterosis, either low-parent heterosis (LPH) or high-parent heterosis (HPH), in gene expression. 
 For a specific gene $g$, LPH occurs when average expression in the hybrid is less than the average expression of either parent, i.e.
@@
 For a particular gene, the truth was determined via the estimated values for $\alpha_g$ and $\delta_g$.
-Specifically, if $|\hat{\delta}_g| >|\hat{\alpha}_g|$ the gene was considered to have overdominance.
-For many overdominance genes, the value of $|\hat{\delta}_g|$ was only slightly larger than $|\hat{\alpha}_g|$.
-Thus there are many genes in these simulated data sets that are considered to have overdominance, but whose signal in the simulations will be extremely small.
-Conversely, there are many non-overdominance genes whose value of $|\hat{\delta}_g|$ was only slightly smaller than $|\hat{\alpha}_g|$, but whose simulated data will look similar to many overdominance genes.
-Therefore, we expect it will be difficult to accurately identify overdominance genes, but believe this level of difficulty is representative of real applications.
+Specifically, if $|\hat{\delta}_g| >|\hat{\alpha}_g|$ the gene was considered to have heterosis.
+For many heterosis genes, the value of $|\hat{\delta}_g|$ was only slightly larger than $|\hat{\alpha}_g|$.
+Thus there are many genes in these simulated data sets that are considered to have heterosis, but whose signal in the simulations will be extremely small.
+Conversely, there are many non-heterosis genes whose value of $|\hat{\delta}_g|$ was only slightly smaller than $|\hat{\alpha}_g|$, but whose simulated data will look similar to many heterosis genes.
+Therefore, we expect it will be difficult to accurately identify heterosis genes, but believe this level of difficulty is representative of real applications.
 
@@
 
-In addition to the approach of \citeauthor{ji2014estimation}, we modified two existing \RNAseq{} approaches, \edgeR{} and \baySeq{}, for use in the overdominance context.
-For each method, we attempted to provide a measure of the strength of overdominance for each gene such that large values of this measure indicate support for overdominance.
+In addition to the approach of \citeauthor{ji2014estimation}, we modified two existing \RNAseq{} approaches, \edgeR{} and \baySeq{}, for use in the heterosis context.
+For each method, we attempted to provide a measure of the strength of heterosis for each gene such that large values of this measure indicate support for heterosis.
 
 \edgeR{} can be used to test for differential expression between any two varieties based on the fit of a negative binomial log-linear model \citep{robinson2007moderated, robinson2010edgeR}.
-To construct a measure of overdominance, we computed the maximum likelihood estimates of the $\mu_{gv}$ parameters for all genes using \edgeR{}'s built-in Fisher scoring algorithm, and then used likelihood ratio tests to calculate two $p$-values for each gene: $p_{g1}$  for testing $H_{g01}:\mu_{g1}=\mu_{g3}$ and $p_{g2}$ for testing $H_{g02}:\mu_{g2}=\mu_{g3}$.
+To construct a measure of heterosis, we computed the maximum likelihood estimates of the $\mu_{gv}$ parameters for all genes using \edgeR{}'s built-in Fisher scoring algorithm, and then used likelihood ratio tests to calculate two $p$-values for each gene: $p_{g1}$  for testing $H_{g01}:\mu_{g1}=\mu_{g3}$ and $p_{g2}$ for testing $H_{g02}:\mu_{g2}=\mu_{g3}$.
 Then, for each gene, we defined a new $p$-value denoted as $p_g$ and set to $p_g=1$ if the estimate of $\mu_{g3}$ falls between the estimates of $\mu_{g1}$ and $\mu_{g2}$ and $p_g=\max\{p_{g1},p_{g2}\}/2$ otherwise.
-(For all relevant significance thresholds $\omega$ near 0, it can be shown that rejecting the null hypothesis of no overdominance of expression for gene $g$ whenever $p_g\leq \omega$ results in a test that is asymptotically size $\omega$.)  We then use $1-p_g$ as a final measure positively associated with strength of evidence for overdominance.
+(For all relevant significance thresholds $\omega$ near 0, it can be shown that rejecting the null hypothesis of no heterosis of expression for gene $g$ whenever $p_g\leq \omega$ results in a test that is asymptotically size $\omega$.)  We then use $1-p_g$ as a final measure positively associated with strength of evidence for heterosis.
 
 \baySeq{} allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct.
-In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of overdominance for each gene.
+In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene.
 We set this measure to zero if the maximum likelihood estimate, calculated using \edgeR{}, of $\mu_{g3}$ is between the maximum likelihood estimates of $\mu_{g1}$ and $\mu_{g2}$.
-Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the two hypotheses that allow for overdominance.
+Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the two hypotheses that allow for heterosis.
 
@@
 
-For the methods in Sections \ref{s:method} and \ref{s:alternative}, we sorted genes according to the computed measure of the strength of evidence for overdominance.
-From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to distinguish genes with overdominance, as defined in Section \ref{s:sim_data},
+For the methods in Sections \ref{s:method} and \ref{s:alternative}, we sorted genes according to the computed measure of the strength of evidence for heterosis.
+From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to distinguish genes with heterosis, as defined in Section \ref{s:sim_data},
 % From Dan, "why this citation?"
-from those without overdominance. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
+from those without heterosis. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
 \begin{figure}[htbp]
@@
 eBayes and \ShrinkBayes{}, provide the best true positive rate for a given false positive rate.
-Also, as expected, as the sample size increases, our ability to distinguish genes with overdominance from genes without improves.
+Also, as expected, as the sample size increases, our ability to distinguish genes with heterosis from genes without improves.
 
@@
 This improvement ranges from about a 20\% improvement over Ji et al.
-to about a 100\% improvement over \edgeR{} (which was not designed for overdominance testing).
+to about a 100\% improvement over \edgeR{} (which was not designed for heterosis testing).
 
 With 4 replicates per variety, there does not appear to be much of a difference between \ShrinkBayes{} and the eBayes approaches, but as the number of replicates increases, the eBayes approaches appear to improve relative to \ShrinkBayes{}.
-Two difference exist between the \ShrinkBayes{} and eBayes approaches that could explain the difference in AUC: 1) \ShrinkBayes{} utilizes a different empirical Bayes approach for estimating both the hyperparameters and the gene-specific parameters and 2) the measure of overdominance is slightly different due to INLA not providing a full posterior.
+Two difference exist between the \ShrinkBayes{} and eBayes approaches that could explain the difference in AUC: 1) \ShrinkBayes{} utilizes a different empirical Bayes approach for estimating both the hyperparameters and the gene-specific parameters and 2) the measure of heterosis is slightly different due to INLA not providing a full posterior.
 

### 2015-09-01 ebayes_rnaseq_heterosis.tex
use EH

@@
 Heterosis involves a comparison of inbreds (simultaneously homozygous at many loci) and hybrids (simultaneously heterozygous at many loci).
-For simplicity, throughout the remainder of the article, we will use the term heterosis to describe the situation where the hybrid mean is more extreme than the parental means, i.e.
+For simplicity, throughout the remainder of the article, we will use the term \emph{extreme heterosis} to describe the situation where the hybrid mean is more extreme than the parental means, i.e.
 we say there is heterosis if and only if either HPH or LPH holds.
@@
 
-To use \RNAseq{} counts to identify genes displaying expression heterosis, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the parameters using an empirical Bayes procedure, and calculate empirical Bayes posterior probabilities for heterosis.
+To use \RNAseq{} counts to identify genes displaying expression heterosis, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the parameters using an empirical Bayes procedure, and calculate empirical Bayes posterior probabilities for extreme heterosis.
 
@@
 
-In the maize context that motivates this work, we are interested in heterosis, either low-parent heterosis (LPH) or high-parent heterosis (HPH), in gene expression. 
+In the maize context that motivates this work, we are interested in extreme heterosis (EH), either low-parent heterosis (LPH) or high-parent heterosis (HPH), in gene expression. 
 For a specific gene $g$, LPH occurs when average expression in the hybrid is less than the average expression of either parent, i.e.
@@
 
-In addition to the approach above, we utilized \ShrinkBayes{} to estimate heterosis probabilities with two modifications described here.
+In addition to the approach above, we utilized \ShrinkBayes{} to estimate EH probabilities with two modifications described here.
 \ShrinkBayes{} utilizes integrated nested Laplace approximation (INLA) \citep{rue2009approximate} in combination with empirical Bayes ideas \citep{van2014shrinkbayes}.
@@
 Since we are interested in non-linear quantities such as $P(\delta_g > |\alpha_g||y)$, we cannot compute these directly using \ShrinkBayes{}.
-Instead, for \ShrinkBayes{}, we calculate heterosis probabilities conditional on posterior means for the half-parental difference and hybrid effect, i.e.
+Instead, for \ShrinkBayes{}, we calculate EH probabilities conditional on posterior means for the half-parental difference and hybrid effect, i.e.
 $\hat{\delta}_g$ and $\hat{\alpha}_g$.
@@
 
-To assess the efficacy of our method to identify genes demonstrating heterosis, we used a maize data set with parental varieties B73 and Mo17 and the hybrid variety (B73 $\times$ Mo17) \citep{paschold2012complementation} to determine realistic parameter values for a simulation study.
+To assess the efficacy of our method to identify genes demonstrating EH, we used a maize data set with parental varieties B73 and Mo17 and the hybrid variety (B73 $\times$ Mo17) \citep{paschold2012complementation} to determine realistic parameter values for a simulation study.
 Section \ref{s:maize} describes the maize dataset in detail.
@@
 For a particular gene, the truth was determined via the estimated values for $\alpha_g$ and $\delta_g$.
-Specifically, if $|\hat{\delta}_g| >|\hat{\alpha}_g|$ the gene was considered to have heterosis.
+Specifically, if $|\hat{\delta}_g| >|\hat{\alpha}_g|$ the gene was considered to have EH.
 For many heterosis genes, the value of $|\hat{\delta}_g|$ was only slightly larger than $|\hat{\alpha}_g|$.
-Thus there are many genes in these simulated data sets that are considered to have heterosis, but whose signal in the simulations will be extremely small.
-Conversely, there are many non-heterosis genes whose value of $|\hat{\delta}_g|$ was only slightly smaller than $|\hat{\alpha}_g|$, but whose simulated data will look similar to many heterosis genes.
-Therefore, we expect it will be difficult to accurately identify heterosis genes, but believe this level of difficulty is representative of real applications.
+Thus there are many genes in these simulated data sets that are considered to have EH, but whose signal in the simulations will be extremely small.
+Conversely, there are many non-heterosis genes whose value of $|\hat{\delta}_g|$ was only slightly smaller than $|\hat{\alpha}_g|$, but whose simulated data will look similar to many EH genes.
+Therefore, we expect it will be difficult to accurately identify EH genes, but believe this level of difficulty is representative of real applications.
 
@@
 In addition to the approach of \citeauthor{ji2014estimation}, we modified two existing \RNAseq{} approaches, \edgeR{} and \baySeq{}, for use in the heterosis context.
-For each method, we attempted to provide a measure of the strength of heterosis for each gene such that large values of this measure indicate support for heterosis.
+For each method, we attempted to provide a measure of the strength of EH for each gene such that large values of this measure indicate support for EH.
 
 \edgeR{} can be used to test for differential expression between any two varieties based on the fit of a negative binomial log-linear model \citep{robinson2007moderated, robinson2010edgeR}.
-To construct a measure of heterosis, we computed the maximum likelihood estimates of the $\mu_{gv}$ parameters for all genes using \edgeR{}'s built-in Fisher scoring algorithm, and then used likelihood ratio tests to calculate two $p$-values for each gene: $p_{g1}$  for testing $H_{g01}:\mu_{g1}=\mu_{g3}$ and $p_{g2}$ for testing $H_{g02}:\mu_{g2}=\mu_{g3}$.
+To construct a measure of EH, we computed the maximum likelihood estimates of the $\mu_{gv}$ parameters for all genes using \edgeR{}'s built-in Fisher scoring algorithm, and then used likelihood ratio tests to calculate two $p$-values for each gene: $p_{g1}$  for testing $H_{g01}:\mu_{g1}=\mu_{g3}$ and $p_{g2}$ for testing $H_{g02}:\mu_{g2}=\mu_{g3}$.
 Then, for each gene, we defined a new $p$-value denoted as $p_g$ and set to $p_g=1$ if the estimate of $\mu_{g3}$ falls between the estimates of $\mu_{g1}$ and $\mu_{g2}$ and $p_g=\max\{p_{g1},p_{g2}\}/2$ otherwise.
-(For all relevant significance thresholds $\omega$ near 0, it can be shown that rejecting the null hypothesis of no heterosis of expression for gene $g$ whenever $p_g\leq \omega$ results in a test that is asymptotically size $\omega$.)  We then use $1-p_g$ as a final measure positively associated with strength of evidence for heterosis.
+(For all relevant significance thresholds $\omega$ near 0, it can be shown that rejecting the null hypothesis of no EH for gene $g$ whenever $p_g\leq \omega$ results in a test that is asymptotically size $\omega$.)  We then use $1-p_g$ as a final measure positively associated with strength of evidence for EH.
 
 \baySeq{} allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct.
-In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene.
+In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of EH for each gene.
 We set this measure to zero if the maximum likelihood estimate, calculated using \edgeR{}, of $\mu_{g3}$ is between the maximum likelihood estimates of $\mu_{g1}$ and $\mu_{g2}$.
-Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the two hypotheses that allow for heterosis.
+Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the two hypotheses that allow for EH.
 
@@
 
-For the methods in Sections \ref{s:method} and \ref{s:alternative}, we sorted genes according to the computed measure of the strength of evidence for heterosis.
-From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to distinguish genes with heterosis, as defined in Section \ref{s:sim_data},
+For the methods in Sections \ref{s:method} and \ref{s:alternative}, we sorted genes according to the computed measure of the strength of evidence for EH.
+From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to distinguish genes with EH, as defined in Section \ref{s:sim_data},
 % From Dan, "why this citation?"
-from those without heterosis. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
+from those without EH. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
 \begin{figure}[htbp]
@@
 eBayes and \ShrinkBayes{}, provide the best true positive rate for a given false positive rate.
-Also, as expected, as the sample size increases, our ability to distinguish genes with heterosis from genes without improves.
+Also, as expected, as the sample size increases, our ability to distinguish genes with EH from genes without improves.
 
@@
 With 4 replicates per variety, there does not appear to be much of a difference between \ShrinkBayes{} and the eBayes approaches, but as the number of replicates increases, the eBayes approaches appear to improve relative to \ShrinkBayes{}.
-Two difference exist between the \ShrinkBayes{} and eBayes approaches that could explain the difference in AUC: 1) \ShrinkBayes{} utilizes a different empirical Bayes approach for estimating both the hyperparameters and the gene-specific parameters and 2) the measure of heterosis is slightly different due to INLA not providing a full posterior.
+Two difference exist between the \ShrinkBayes{} and eBayes approaches that could explain the difference in AUC: 1) \ShrinkBayes{} utilizes a different empirical Bayes approach for estimating both the hyperparameters and the gene-specific parameters and 2) the measure of EH is slightly different due to INLA not providing a full posterior.
 
@@
 The figure shows a ridge at an effect size of zero for probabilities below 0.5.
-Above a probability of 0.5, we see a prototypical volcano pattern, with increased heterosis probability corresponding to larger estimated effect sizes and no estimated effect sizes near zero for genes with high heterosis probability.
+Above a probability of 0.5, we see a prototypical volcano pattern, with increased EH probability corresponding to larger estimated effect sizes and no estimated effect sizes near zero for genes with high EH probability.
 We also see asymmetry, with larger negative effect sizes than positive effect sizes due to genes with hybrid counts near zero and relatively high parental counts.
-Genes with high estimated posterior probabilities of heterosis and large estimated effect sizes are candidates for further investigation.
+Genes with high estimated posterior probabilities of EH and large estimated effect sizes are candidates for further investigation.
 

### 2015-09-01 response.tex
overdominance response

@@
 
-\response{Yes, overdominance is the scenario were are interested in. We have added the term in the introduction as well as throughout the manuscript to make this clear.}
+\response{Many publications mentioning heterosis do not define it clearly.
+For example, the 2013 {\em Nature Reviews Genetics} paper by Chen that we now cite in our revised manuscript states the following definition: ``Heterosis (Also known as hybrid vigour).
+When hybrids display increased levels of growth, survival or fitness relative to their parents.''  Like Chen, some other author present heterosis as synonymous with hybrid vigor, and then go on to describe hybrid vigor as the phenomenon in which the hybrid outperforms the parents.
+It may not be so clear what ``outperforms the parents'' means, but the idea behind hybrid vigor and the reason it is of such interest (particularly in maize) and so useful is that the hybrid is better than {\bf both parents}, not simply better than the average.
+That said, we do agree with your definition of heterosis.
+We originally wrote that first sentence so that it would be true regardless of which definition of heterosis one takes.
+We were saying that heterosis occurs when hybrid vigor occurs.
+The reverse is not necessarily so, and we did not claim it to be in that first sentence.
+However, we acknowledge that our first sentence was misleading because most readers would naturally take the sentence for a definition.
+Furthermore, for simplicity, we did start using ``heterosis'' as a synonym for the situation where the hybrid mean expression is more extreme than both parents.
+That is really what we are interested in because the phenotypic traits we see in maize have this pattern, so it is only natural to seek genes whose expression has this same pattern.
+In the revised manuscript, we have been more careful with terminology.
+We have defined heterosis to be a difference between the hybrid mean the average of the parental means.
+We have also introduced a new term (extreme heterosis, EH) to specify the situation where the hybrid mean is more extreme than both parental means.
+We continue to use low-parent heterosis (LPH) and high-parent heterosis (HPH) to distinguish the two types of EH.
+We also explain in the manuscript why we prefer the LPH/HPH terminology to underdominance/overdominance.
+In our experience, these latter terms are used to describe the ``gene action'' at a single locus with respect to some quantitative trait.
+We can't asses the action of any locus on any of our expression traits in data from inbred and hybrid crosses.}
 

### 2015-09-01 ebayes_rnaseq_heterosis.tex
last paragraph

@@
 
-Notwithstanding these improvements, we believe our approach is a computationally efficient method that can immediately aid scientists who are interested in identifying candidate genes for a genetic mechanism of heterosis.
+Notwithstanding these improvements, we believe our approach is a computationally efficient method that can immediately aid scientists who are interested in identifying candidate genes involved in a genetic mechanism of heterosis.
 
-An alternative explanation for phenotypic heterosis is preferential gene expression of alleles from certain parents for certain genes.
-Analyses of allele-specific expression \citep{bell2013rna,wei2013computational} could complement the search for gene expression heterosis in the quest for understanding the molecular genetic mechanisms behind phenotypic heterosis.
+This paper has focused on statistical methods for detecting EH in gene expression using RNA-seq data.
+EH at the transcript level is only one of multiple molecular genetic mechanisms that may play roles in establishing hybrid vigor.
+Complementation \citep{paschold2012complementation}, allele-specific expression \citep{bell2013rna,wei2013computational}, and other complex forms of genomic and epigenetic interaction \citep{chen2013genomic} are all plausible as mechanisms partially responsible for phenotypic heterosis.
+Further study of these phenomena using modern genomic technologies and appropriate statistical methods should enhance our understanding of hybrid vigor.
 

### 2015-09-01 response.tex
updated responses

@@
 ShrinkBayes) on the actual maize dataset in Section 4.
-Why is that missing?}
+Why is that missing?
+The comments made by one reviewer regarding a relative assessment of computing time relative to the other methods deserves attention.}
 
-\response{We have provided a comparison of the computational time on the real data set using these methods.} \todo{What else should we do?}
-
-\comment{The comments made by one reviewer regarding a relative assessment of computing time relative to the other methods deserves attention.}
+\response{We have provided a comparison of the computational time on the real data set using these methods. We have compared point estimates from edgeR to those using eBayes (laplace). }
 
@@
 
-\response{Even though scientists may have specific candidates. An exploratory analysis to suggest other genes seems useful.} \todo{Updated references.}
+\response{We now include a reference to Chen (2013), a Nature Reviews Genetics article entitled ``Genomic and epigenetic insights into the molecular bases of heterosis''.
+Even though scientists may have specific candidates.
+An exploratory analysis to suggest other genes seems useful.}
 

### 2015-09-01 ebayes_rnaseq_heterosis.tex
computation time

@@
 We analyzed the data using all the methods compared in the previous section.
-The computation time on a Dell Precision T3600 with a 3.6GHz intel Xeon processor was 14 seconds for \edgeR{}, 1.3 minutes for \citeauthor{ji2014estimation}, 10 hours for the eBayes approaches, and 17 hours for \baySeq{}.
+The computation time on a Dell Precision T3600 with a 3.6GHz Intel Xeon processor was 14 seconds for \edgeR{}, 1.3 minutes for \citeauthor{ji2014estimation}, 10 hours for the eBayes approaches, and 17 hours for \baySeq{}.
 In the eBayes approach, the vast majority of the time is spent on independent MCMC analysis for each gene.
@@
 \centerline{\includegraphics[width=\textwidth]{gene_specific_estimates.pdf}}
-\caption{Two-dimensional histogram of point estimates from \edgeR{} and posterior means from eBayes using Laplace distributions along with the $y=x$ diagonal.}
+\caption{Two-dimensional histogram of point estimates from \edgeR{} and posterior means from eBayes (Laplace) along with the $y=x$ diagonal.}
 \label{f:gene_specific_estimates}
@@
 \centerline{\includegraphics[width=\textwidth]{volcano}}
-\caption{A bivariate histogram of the maximum of the LPH and HPH probabilities versus estimated effect size defined in equation \eqref{e:effect_size} for the B73 $\times$ Mo17 maize experiment.}
+\caption{A bivariate histogram of the maximum of the LPH and HPH probabilities versus estimated effect size defined in equation \eqref{e:effect_size} for the B73 $\times$ Mo17 maize experiment using eBayes (Laplace).}
 \label{f:volcano}

### 2015-09-01 response.tex
include Dan edits

@@
 
-\response{We have assessed the Ji et al. approach and included it in Sections 3.2 and 3.3.}
+\response{We have assessed the Ji et al.
+approach and included it in Sections 3.2 and 3.3.
+Although the Ji et al.\ approach performed better than edgeR and baySeq, its performance was considerably worse than the performance of the methods we proposed.
+Regarding the `apples vs.\ oranges comparison', we agree that the proposed hypothesis testing strategy is not available in other methods.
+This was part of the motivation for developing our methods.
+Our simulations show that doing the best we know how to do with existing methods is worse than doing what we proposed.
+We think the comparison is relevant and answers a question that some readers would have raised had we not included methods like edgeR and baySeq in our simulation.}
 
@@
 
-\response{Added at the end of the Discussion section.}
+\response{Thanks for pointing out these papers.  We have added a new concluding paragraph to the paper that cites the papers you mentioned as well as two others.}
 
@@
 
-\response{We have provided a comparison of the computational time on the real data set using these methods. We have compared point estimates from edgeR to those using eBayes (laplace). }
+\response{We have added a comparison of the computing time required for each method in Section 4 of the revised manuscript.}
 
@@
 
-\response{Fixed.}
+\response{We fixed the error.  Thanks for catching this.}}
 
@@
 
-\response{Added citation.}
+\response{We agree and have added the appropriate citation.}}
 
@@
 
-\response{We cannot answer why most RNAseq data analysts gravitate to this specification. We appreciate the reference and will consider the alternative specification in the future.}
+\response{You have raised in interesting point.
+If $y_i|\lambda_i$ is Poisson with mean $\lambda_i$, and $\lambda_i$ is ${\rm gamma}(\alpha_i,\beta)$ for all $i=1,\ldots, n$, then (for all $i$) the marginal distribution of $y_i$ is negative binomial with mean $\mu_i$ and variance $\phi\mu_i$, where $\mu_i=\alpha_i\beta$ and $\phi=1+\beta$.
+Thus, if we assume the $\beta$ parameter is constant across observations, we have a linear mean-variance relationship.
+On the other hand, if $y_i|\lambda_i$ is Poisson with mean $\lambda_i$, and $\lambda_i$ is ${\rm gamma}(\alpha,\beta_i)$ for all $i=1,\ldots, n$, then (for all $i$) the marginal distribution of $y_i$ is negative binomial with mean $\mu_i$ and variance $\mu_i+\phi\mu_i^2$, where $\mu_i=\alpha\beta_i$ and $\phi=1/\alpha$.
+Thus, if we assume the $\alpha$ parameter is constant across observations, we have a quadratic mean-variance relationship.
+Past research suggests that RNA-seq data that includes biological (rather than only technical) variation tends to exhibit a quadratic mean-variance relationship.
+Thus, we have focused on the latter type of negative binomial model specification.
+It is conceivable that the linear mean-variance relationship could be appropriate for some subset of genes in an RNA-seq dataset.
+That is an interesting model selection question, but one that we think should be left for future work focused specifically on this issue.
+}
 
@@
 
-\response{We have assessed the use of normal distributions see Section 3.3 and also provided rationale for why Laplace distributions would be expected to perform better in this scenario at the end of Section 2.1.}
+\response{We have assessed the use of normal distributions see Section 3.3 and also provided rationale for why Laplace distributions would be expected to perform better in this scenario at the end of Section 2.1.
+Our simulation results in Figure 2 show that our Laplace model specification is preferred over the normal model specification.}
 
@@
 
-\response{We now compare the estimated gene-specific parameters to the eBayes estimates in Figure 3.
-We have added a paragraph to the beginning of Section 2.2 to share our experience with a fully MCMC analysis which took too long and had poor mixing.}
+\response{We have added a paragraph to the beginning of Section 2.2 to share our experience with a fully MCMC analysis which took too long and had poor mixing.
+We have added a new Figure 3, which shows that the second-stage gene-specific estimates are shrunk relative to the original edgeR estimates.}
 
@@
 
-\response{We have assessed the use of normal distributions, see Section 3.3.}
+\response{We have now assessed the use of normal distributions.
+The results in Section 3.3 and Figure 2 show that there is an advantage to using Laplace distributions instead of normals.
+Our eBayes implementation of our model performs quite similarly to our ShrinkBayes implementation of our model.}
 
@@
 
-\response{ShrinkBayes failed for us when too many zeros were present. We do not believe the comparison of other methods would change if we did not remove the low counts.}
+\response{ShrinkBayes failed for us when too many zeros were present.
+We do not believe the comparison of other methods would change if we did not remove genes with low counts.
+In practice, it is common to remove low count genes from RNA-seq datasets prior to analysis because these genes contain little information about differential expression and lead to computational difficulties for some methods.
+With that being said, we did not remove any genes from the real data analysis.}
 
@@
 
-\response{It is not necessary, but it is sufficient. If parental expressions are the same and the hybrid is not, then the hybrid must be exhibiting overdominance.}
+\response{It is not necessary, but it is sufficient. If parental expressions are the same and the hybrid is not, then the hybrid must be exhibiting either LPH or HPH.}
 
@@
 
-\response{No, although we do mean ridge rather than peak, i.e. the count is high for effect size of zero and probabilities below 0.5.}
+\response{No, although we do mean ridge rather than peak, i.e. there are many genes with effect size of zero and probabilities below 0.5.}
 

### 2015-09-01 response.tex
compiles

@@
 
-\response{We fixed the error.  Thanks for catching this.}}
+\response{We fixed the error.  Thanks for catching this.}
 
@@
 
-\response{We agree and have added the appropriate citation.}}
+\response{We agree and have added the appropriate citation.}
 
@@
 
-\response{Yes, that is what the test is doing. We have now stated the case for LPH, i.e. ``$\delta_g < -|\alpha_g|$ (equivalently $\mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}$).''
+\response{Yes, that is what the test is doing. We have now stated the equivalence for both LPH and HPH, e.g. for LPH: $\delta_g < -|\alpha_g|$ is equivalent to $\mu_{g3} < \min(\mu_{g1},\mu_{g2})$.}
 
@@
 
-\response{It is not necessary, but it is sufficient. If parental expressions are the same and the hybrid is not, then the hybrid must be exhibiting either LPH or HPH.}
+\response{Yes, that is the hypothesis that the two parents are the same, but the hybrid is different.
+It is not necessary, but it is sufficient.
+If parental expressions are the same and the hybrid is not, then the hybrid must be exhibiting either LPH or HPH.}
 
@@
 
-\comment{Page 11 last line “ASYMPTOTIC”}}
+\comment{Page 11 last line “ASYMPTOTIC”}
 
@@
 
-\response{We now include a reference to Chen (2013), a Nature Reviews Genetics article entitled ``Genomic and epigenetic insights into the molecular bases of heterosis''.
+\response{We now include a reference to Chen (2013), a Nature Reviews Genetics article entitled ``Genomic and epigenetic insights into the molecular bases of heterosis'' as well as three other articles published in 2012 or later.
 Even though scientists may have specific candidates.
@@
 
-\todo{Add robustness discussion?}
+\response{In the discussion, we mention the departures from model assumptions observed.
+Despite these violations, the results seems reasonable and thus suggest some level of robustness.
+We believe that further analysis of robustness is better left to future work specifically devoted to addressing this question.}
 

### 2015-09-02 ebayes_rnaseq_heterosis.tex
small typos

@@
 
-To use \RNAseq{} counts to identify genes displaying expression heterosis, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the parameters using an empirical Bayes procedure, and calculate empirical Bayes posterior probabilities for extreme heterosis.
+To use \RNAseq{} counts to identify genes displaying expression heterosis, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the hyperparameters using an empirical Bayes procedure, and calculate empirical Bayes posterior probabilities for extreme heterosis.
 
@@
 
-In the maize context that motivates this work, we are interested in extreme heterosis (EH), either low-parent heterosis (LPH) or high-parent heterosis (HPH), in gene expression. 
-For a specific gene $g$, LPH occurs when average expression in the hybrid is less than the average expression of either parent, i.e.
-$\delta_g < -|\alpha_g|$ or, equivalently, $\mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}$, and HPH occurs when average expression in the hybrid is greater than the average expression of either parent, i.e.
-$\delta_g > |\alpha_g|$ or, equivalently, $\mu_{g3} > \max\{\mu_{g1},\mu_{g2}\}$.
+In the maize context that motivates this work, we are interested in extreme heterosis (EH), i.e. either low-parent heterosis (LPH) or high-parent heterosis (HPH), in gene expression. 
+For a specific gene $g$, LPH occurs when expected expression in the hybrid is less than the expected expression of either parent, i.e.
+$\mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}$ or, equivalently, $\delta_g < -|\alpha_g|$, and HPH occurs when expected expression in the hybrid is greater than the exptected expression of either parent, i.e.
+$\mu_{g3} > \max\{\mu_{g1},\mu_{g2}\}$ or, equivalently, $\delta_g > |\alpha_g|$.
 % and high-parent  heterosis (HPH) of gene expression occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$.
@@
 %P\left(\left. \mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}\right|y,\hat{\pi},\hat{\gamma}\right) = 
-P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi},\hat{\gamma}\right) = \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right), 
+P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi},\hat{\gamma}\right) \approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right), 
 \label{e:probs}
@@
 
-Using these parameters and normalization factors, we simulated data according to the negative binomial model in \eqref{e:data} independently for each gene.
+Using these parameters and normalization factors, we simulated data according to the negative binomial model in equation \eqref{e:data} independently for each gene.
 For each simulation, we analyzed a subset of 25,000 genes selected randomly from genes with simulated counts at least one on average and with at most two zeros for each variety across replications.
@@
 We compared our method to that of \cite{ji2014estimation}, which assumes normality in the response, by modeling the logarithm of the RNAseq count plus one scaled by the normalization factor, i.e.
-$\log([Y_{gvi}+1])-\gamma_{vi}$.
+$\log(Y_{gvi}+1)-\gamma_{vi}$.
 Use of the normalization factor here provides two advantages: 1) counts are properly adjusted for the thoroughness of the sequencing of the sample and 2) for genes with no count variation within variety (which actually occurs in our maize data set), use of the normalization factors allows the approach of \citeauthor{ji2014estimation} to still execute.
@@
 With 4 replicates per variety, there does not appear to be much of a difference between \ShrinkBayes{} and the eBayes approaches, but as the number of replicates increases, the eBayes approaches appear to improve relative to \ShrinkBayes{}.
-Two difference exist between the \ShrinkBayes{} and eBayes approaches that could explain the difference in AUC: 1) \ShrinkBayes{} utilizes a different empirical Bayes approach for estimating both the hyperparameters and the gene-specific parameters and 2) the measure of EH is slightly different due to INLA not providing a full posterior.
+Two differences exist between the \ShrinkBayes{} and eBayes approaches that could explain the difference in AUC: 1) \ShrinkBayes{} utilizes a different empirical Bayes approach for estimating both the hyperparameters and the gene-specific parameters and 2) the measure of EH is slightly different due to INLA not providing a full posterior.
 
-There also appears to be a pattern of the eBayes approach with Laplace distributions for the half-parental differences and hybrid effect systematically performing better than the approach that uses normal distributions.
-We assume this is because the Laplace distributions are better approximations to the true underlying distribution for these parameters.
+There also appears to be a pattern of the eBayes (Laplace) systematically performing better than eBayes (normal).
+We suspect this is because the Laplace distributions are better approximations to the true underlying distribution for these parameters and we discuss this in Section \ref{s:discussion}.
 
@@
 
-We used our method to analyze a maize data set \citep{paschold2012complementation} of \RNAseq{} gene expression in parental lines B73 and Mo17 and the hybrid genotype (B73$\times$Mo17).
+We used our method to analyze a maize data set \citep{paschold2012complementation} of \RNAseq{} gene expression in parental lines B73 and Mo17 and the hybrid genotype (B73$\times$Mo17) with a total of 39,656 genes.
 Each variety had four biological replicates measured with Illumina methodology and equipment.
@@
 We analyzed the data using all the methods compared in the previous section.
-The computation time on a Dell Precision T3600 with a 3.6GHz Intel Xeon processor was 14 seconds for \edgeR{}, 1.3 minutes for \citeauthor{ji2014estimation}, 10 hours for the eBayes approaches, and 17 hours for \baySeq{}.
+The computation time on a desktop with two 4-core 3.6GHz Intel Xeon processor was 14 seconds for \edgeR{}, 1.3 minutes for \citeauthor{ji2014estimation}, 10 hours for the eBayes approaches, and 17 hours for \baySeq{}.
 In the eBayes approach, the vast majority of the time is spent on independent MCMC analysis for each gene.
@@
 When parallelized across 5 cores, the eBayes approaches took about 2.5 hours.
+ShrinkBayes took 12 hours on a cluster node with two 8-core 2.6GHz Intel Haswell E5-2640 v3 processors where the code was also parallelized across 5 cores.
 
-For the eBayes approach with the Laplace distributions, we estimated the hyperparameters to be $\hat{\eta}_\phi = 3.44$, $\hat{\sigma}_\phi = 2.74$, $\hat{\eta}_\alpha = -0.05$, $\hat{\sigma}_\alpha=0.36$, $\hat{\eta}_\delta = 0.08$, $\hat{\sigma}_\delta=0.33$, $\hat{\eta}_\psi = -2.35$, and $\hat{\sigma}_\psi=0.60$.
+For the eBayes (Laplace), we estimated the hyperparameters to be $\hat{\eta}_\phi = 3.44$, $\hat{\sigma}_\phi = 2.74$, $\hat{\eta}_\alpha = -0.05$, $\hat{\sigma}_\alpha=0.36$, $\hat{\eta}_\delta = 0.08$, $\hat{\sigma}_\delta=0.33$, $\hat{\eta}_\psi = -2.35$, and $\hat{\sigma}_\psi=0.60$.
 We then performed independent MCMC analysis for each gene conditional on these hyperparameters.
@@
 
-Figure \ref{f:gene_specific_estimates} provides point estimates of the gene-specific parameters from the initial \edgeR{} estimation step and after the eBayes with Laplace distribution procedure described in Section \ref{s:ebayes}.
+Figure \ref{f:gene_specific_estimates} provides point estimates of the gene-specific parameters from the initial \edgeR{} estimation step and after the eBayes (Laplace) procedure described in Section \ref{s:ebayes}.
 \begin{figure}
@@
 For each gene, the quantity of interest is likely to be the maximum of these two probabilities.
-For each gene with a high probability of either HPH or LPH, the magnitude of the effect is of interest.
-We calculate
+For each gene with a high probability of either HPH or LPH, the magnitude of the effect is of interest, thus we calculate
 \begin{equation}
@@
 \end{equation}
-Thus, the estimated effect size is the difference between hybrid mean and the nearest parent with negative values indicating LPH and positive values indicating HPH.
+This estimated effect size is the difference between hybrid mean and the nearest parent with negative values indicating LPH and positive values indicating HPH.
 If the hybrid mean is estimated to be between the parents, this effect is defined to be zero.
@@
 Although our method appears to hold some advantage over existing methods, we believe our approach can be improved by refining the hierarchical model for the gene-specific parameter distribution.
-Figure \ref{f:scatterplot} shows marginal and bivariate histograms for the empirical Bayes posterior means for the gene-specific parameters.
+Figure \ref{f:scatterplot} shows marginal and bivariate histograms for eBayes (Laplace) posterior means for the gene-specific parameters.
 \begin{figure}
@@
 These figures show markedly departures from marginal model assumptions, e.g.
-normality assumptions for $\phi_g$ and $\psi_g$, and independence assumptions, e.g.
-$\phi_g$ versus $\psi_g$ and $\alpha_g$ versus $\delta_g$.
+normality assumptions for $\phi_g$ and $\psi_g$, and independence assumptions for $(\phi_g,\psi_g)$ and $(\alpha_g,\delta_g)$.
 The plot of $\phi_g$ versus $\psi_g$ shows a pattern where the mean overdispersion decreases as the mean expression level increases.
 The plot of $\alpha_g$ versus $\delta_g$ shows a rotated V pattern where $\delta_g$ appears to be equal to $|\alpha_g|$.
-This pattern is consistent with Mendel's Law of Dominance where the hybrid has mean expression equal to the parent with higher mean expression.
+This V pattern is consistent with Mendel's Law of Dominance where the hybrid has mean expression equal to the parent with higher mean expression.
 
@@
 In our data there are only four replicates per variety, and therefore asymptotic justifications are deficient.
-There are a variety of hyperparameter estimation approaches to explore, e.g.
+There are a variety of alternative estimation approaches to explore, e.g.
 expectation-maximization algorithms or fully Bayesian approaches.

### 2015-09-02 response.tex
hyperparameter estimates

@@
 
-\response{We have added a comparison of the computing time required for each method in Section 4 of the revised manuscript.}
+\response{We have added a comparison of the computing time required for each method in Section 4 of the revised manuscript.
+We have also added a comparison of edgeR estimates to eBayes (Laplace) in Figure 3 which show shrinkage for half-parental differences and hybrid effects.}
 
-\response{Added. See section 4.}
+\response{Added in Section 4. The bottom line is edgeR takes seconds, Ji et al. takes minutes, the other methods take hours.}
 
@@
 
-%\todo{During the course of our research, we were initially estimating hyperparameters based on moment matching of gene-specific parameters estimated using negative binomial regression independently for all genes.
-This procedure had issues whenever a gene-variety combination had counts that were all zeros, thus leading to an unidentified regression parameter, or too little variability, leading to an unidentified overdispersion parameter.
-This cut off appeared to fix the problem.
-None of the current methods suffer from this problem, in particular the edgeR based parameter estimations have a built shrinkage to estimate parameters.
-We did not change the simulations we were using and thus report the deletion of low counts for reproducible accuracy.
-We do not believe removing this deletion would change any of our results.}
-
 \comment{Page 7, lines 17-22.  Can you provide us any information on how well the hyperparameters were estimated based on your empirical Bayes strategy?}
 
-\response{Since we did not simulate data from our model, there are no true values of the hyperparameters and therefore we cannot directly compare our estimates to a truth.
-We did try to point out in the Discussion that we believe these estimates could be improved.}
+\response{We have assessed the hyperparameter estimates in Section 3.4 and Table 1 where the estimates seem reasonable and improve with increased number of replicates per variety.}
 
@@
 
-\response{No, although we do mean ridge rather than peak, i.e. there are many genes with effect size of zero and probabilities below 0.5.}
+\response{No, although we do mean ridge rather than peak, i.e. there are many genes with effect size of zero and probabilities below 0.5, and have updated the manuscript accordingly.}
 
@@
 
-\response{As mentioned above, as we have no true values of the hyperparameters, we cannot quantitatively assess the quality of the estimates.}
+\response{We have assessed the hyperparameter estimates in Section 3.4 and Table 1 where the estimates seem reasonable and improve with increased number of replicates per variety.}
 
@@
 \response{We assume that you mean the improvement compared to existing methods is underwhelming.
-To more accurately address the improvement, we calculated percentage improvement in AUC which ranges from 20\% to 100\% and discussed these in Section 3.3.}
+To more accurately address the improvement, we calculated percentage improvement in AUC which ranges from 20\% to 100\% and discuss these in Section 3.3.}
 
@@
 
-\response{Added. See section 4.}
+\response{Added in Section 4. The bottom line is edgeR takes seconds, Ji et al. takes minutes, the other methods take hours.}
 

### 2015-09-02 ebayes_rnaseq_heterosis.tex
hyperparameter estimates

@@
 
+\subsection{Hyperparameter estimation}
+
+As described in Section \ref{s:simulation}, we simulated data only from the negative-binomial portion of the model, but not from the hierarchical distributions.
+Thus, there are no true values for the hyperparameters to compare against.
+But, we can compute means and standard deviations of the gene-specific parameters that were used in the simulation study and compare these to the estimated hyperparameter values.
+\begin{table}
+\caption{Empirical values calculated from the gene-specific parameters used in the simulation and mean of estimated values across the 10 simulations within replicate number.
+Maximum standard deviation in the estimates across the 10 simulations was 0.004.}
+\label{t:hyperparameter}
+\begin{tabular}{l|r|rrr|}
+& \multicolumn{1}{c|}{Empirical} & \multicolumn{3}{c}{\# of replicates}\\
+Parameter & value & 4 & 8 & 16 \\
+\hline
+% with (SD)
+% $\eta_\phi$ & 4.675 & 4.681 (0.004) & 4.741 (0.002) & 4.817 (0.003) \\
+% $\eta_\alpha$ & -0.017 & -0.023 (0.002) & -0.018 (0.003) & -0.014 (0.002)  \\
+% $\eta_\delta$ & -0.046 & -0.046 (0.002) & -0.031 (0.002) & -0.022 (0.003)  \\
+% $\eta_\phi$ & -2.713 & -2.290 (0.004) &  -2.335 (0.003) & -2.387 (0.002)\\
+% \hline
+% $\sigma_\phi$ & 1.815 & 1.821 (0.002) & 1.779 (0.002) & 1.726 (0.003)\\
+% $\sigma_\alpha$ & 0.364 & 0.459 (0.002) & 0.399 (0.002) & 0.357 (0.002) \\
+% $\sigma_\delta$ & 0.145 & 0.389 (0.002) & 0.317 (0.002) & 0.268 (0.001) \\
+% $\sigma_\psi$ & 0.737 & 0.401 (0.001) & 0.493 (0.001) & 0.542 (0.001) \\
+% 3 decimal places
+% $\eta_\phi$ & 4.675 & 4.681 & 4.741  & 4.817  \\
+% $\eta_\alpha$ & -0.017 & -0.023  & -0.018 & -0.014  \\
+% $\eta_\delta$ & -0.046 & -0.046 & -0.031  & -0.022   \\
+% $\eta_\phi$ & -2.713 & -2.290  &  -2.335  & -2.387\\
+% \hline
+% $\sigma_\phi$ & 1.815 & 1.821  & 1.779  & 1.726 \\
+% $\sigma_\alpha$ & 0.364 & 0.459  & 0.399  & 0.357  \\
+% $\sigma_\delta$ & 0.145 & 0.389  & 0.317  & 0.268  \\
+% $\sigma_\psi$ & 0.737 & 0.401 & 0.493 & 0.542 \\
+$\eta_\phi$ & 4.7\phantom{00} & 4.7\phantom{00} & 4.7\phantom{00}  & 4.8\phantom{00}  \\
+$\eta_\alpha$ & -0.017 & -0.023  & -0.018 & -0.014  \\
+$\eta_\delta$ & -0.046 & -0.046 & -0.031  & -0.022   \\
+$\eta_\phi$ & -2.7\phantom{00} & -2.3\phantom{00}  &  -2.3\phantom{00}  & -2.4\phantom{00} \\
+\hline
+$\sigma_\phi$ & 1.8\phantom{0} & 1.8\phantom{0}  & 1.8\phantom{0}  & 1.7\phantom{0} \\
+$\sigma_\alpha$ & 0.36 & 0.46  & 0.40  & 0.36  \\
+$\sigma_\delta$ & 0.15 & 0.39  & 0.32  & 0.27  \\
+$\sigma_\psi$ & 0.74 & 0.40 & 0.49 & 0.54 \\
+\hline
+\end{tabular}
+\end{table}
+Table \ref{t:hyperparameter} provides the computed means and standard deviations from the gene-specific parameter values, called empirical values, and compares these values against the mean of estimated hyperparameters across the 10 simulations within replicate number.
+For the mean parameters, there is reasonable agreement amongst the different numbers of replicates with a somewhat high estimate for the mean of the overdispersion parameter $(\eta_\phi)$ and, perhaps, a trend for the mean of the hybrid effect ($\eta_\delta$), although we would not have expected this.
+For the standard deviation parameters, $\sigma_\phi$ is consistent across the replicate numbers, $\sigma_\alpha$ and $\sigma_\delta$ mean estimates are decreasing toward the empirical values as the number of replicates increases, and $\sigma_\phi$ is increasing toward the empirical value as the number of replicates increases.
+%We expect this pattern for $\sigma_\alpha$ and $\sigma_\delta$ since these are based on moment matching of independent gene-specific parameter estimates.
+For low numbers of replicates, these independent estimates will have relatively large sampling variability that will cause a bias toward larger values for the moment-matched standard deviation estimates.
+
 \section{Searching for gene expression heterosis in the maize experiment}

### 2015-09-02 response.tex
fixed for Dan's comments

@@
 
-\response{We have assessed the hyperparameter estimates in Section 3.4 and Table 1 where the estimates seem reasonable and improve with increased number of replicates per variety.}
+\response{We have assessed the hyperparameter estimates in Section 3.4 and Table 1 where the estimates seem reasonable and generally improve with increased number of replicates per variety.}
 

### 2015-09-02 ebayes_rnaseq_heterosis.tex
fixed for Dan's comments

@@
 An important type of heterosis, known as hybrid vigor, refers to the enhancements in the phenotype of hybrid progeny relative to their inbred parents.
-Although the heterosis phenomenon is extensively utilized in agriculture, the molecular basis of heterosis is still largely unknown.
+Although hybrid vigor is extensively utilized in agriculture, its molecular basis  is still largely unknown.
 In an effort to understand phenotypic heterosis at the molecular level, researchers are measuring transcript abundance levels of thousands of genes in parental inbred lines and their hybrid offspring using RNA sequencing (\RNAseq{}) technology.
@@
 We refer to the former as high-parent heterosis (HPH) and the latter as low-parent heterosis (LPH).
-Note that heterosis, HPH, and LPH are similar to the quantitative genetics concepts of dominance, heterosis, and underdominance.
+Note that heterosis, HPH, and LPH are similar to the quantitative genetics concepts of dominance, overdominance, and underdominance.
 However, the various forms of dominance are usually reserved for describing the association of mean phenotype with homozygous and heterozygous genotypes at a single genetic locus.
 Heterosis involves a comparison of inbreds (simultaneously homozygous at many loci) and hybrids (simultaneously heterozygous at many loci).
-For simplicity, throughout the remainder of the article, we will use the term \emph{extreme heterosis} to describe the situation where the hybrid mean is more extreme than the parental means, i.e.
-we say there is heterosis if and only if either HPH or LPH holds.
+For simplicity, throughout the remainder of the article, we will use the term \emph{extreme heterosis} (EH) to describe the situation where the hybrid mean is more extreme than the parental means, i.e.
+we say there is EH if and only if either HPH or LPH holds.
 
 Despite intensive study and successful use of heterosis in agriculture, the basic molecular genetic mechanisms remain poorly understood \citep{chen2013genomic}.
-One potential explanation, known as gene expression heterosis, is enhanced expression of one or more genes in hybrids compared to their inbred parents.
-Gene expression heterosis has been investigated in maize by \cite{swanson2006all} and \cite{springer2007allelic} and is conceptually depicted in the right column of Figure 1b in  \cite{chen2013genomic}.
+One potential explanation, is EH of gene expression, i.e.
+enhanced (or suppressed) expression of one or more genes in hybrids compared to their inbred parents.
+Gene expression heterosis has been investigated in maize by \cite{swanson2006all} and \cite{springer2007allelic} and EH of gene expression is conceptually depicted in the right column of Figure 1b in \cite{chen2013genomic}.
 
@@
 The remainder of the paper proceeds as follows.
-Section \ref{s:method} presents the proposed hierarchical model, an empirical Bayes method of estimating the parameters, and the calculation of posterior probabilities of heterosis.
+Section \ref{s:method} presents the proposed hierarchical model, an empirical Bayes method of estimating the parameters, and the calculation of posterior probabilities of EH.
 Section \ref{s:simulation} presents a simulation study based on a maize experiment and compares our approach to alternative methods.
-Section \ref{s:maize} analyzes a maize experiment where phenotypic heterosis is well established and identifies genes demonstrating expression heterosis.
-The data and code necessary to perform this analysis are available in the Supplemental Material.
+Section \ref{s:maize} analyzes a maize experiment where hybrid vigor is well established and identifies genes demonstrating EH of expression.
 Section \ref{s:discussion} summarizes the work and suggests directions for future research.
@@
 
-We consider an RNA sequencing (\RNAseq{}) experiment that involves at least three genetic varieties: two parental varieties (P1 and P2) and a cross between these two varieties called the hybrid (H).
+We consider an RNA sequencing (\RNAseq{}) experiment that involves at least three genetic varieties: two parental varieties and a cross between these two varieties called the hybrid.
 For each variety, replicate RNA samples are isolated and assessed for quality.
@@
 
-To use \RNAseq{} counts to identify genes displaying expression heterosis, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the hyperparameters using an empirical Bayes procedure, and calculate empirical Bayes posterior probabilities for extreme heterosis.
+To use \RNAseq{} counts to identify genes displaying EH of expression, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the hyperparameters using an empirical Bayes procedure, and calculate empirical Bayes posterior probabilities for EH.
 
@@
 
-We will use the term eBayes to refer to the approaches defined in Sections \ref{s:model}-\ref{s:gene_heterosis} and add parenthetical labels ``normal'' and ``Laplace'' to distinguish between the two eBayes approaches.
+We will use the term eBayes to refer to the approaches defined in Sections \ref{s:model}--\ref{s:gene_heterosis} and add parenthetical labels ``normal'' and ``Laplace'' to distinguish between the two eBayes approaches.
 
@@
 
-We compared our method to that of \cite{ji2014estimation}, which assumes normality in the response, by modeling the logarithm of the RNAseq count plus one scaled by the normalization factor, i.e.
+We compared our method to that of \cite{ji2014estimation}, which assumes normality in the response, by modeling the logarithm of the RNAseq count plus one adjusted by the normalization factor, i.e.
 $\log(Y_{gvi}+1)-\gamma_{vi}$.
@@
 There also appears to be a pattern of the eBayes (Laplace) systematically performing better than eBayes (normal).
-We suspect this is because the Laplace distributions are better approximations to the true underlying distribution for these parameters and we discuss this in Section \ref{s:discussion}.
+We suspect this is because the Laplace distributions are better approximations to the true underlying distribution for these parameters, and we discuss this in Section \ref{s:discussion}.
 
@@
 \label{t:hyperparameter}
-\begin{tabular}{l|r|rrr|}
-& \multicolumn{1}{c|}{Empirical} & \multicolumn{3}{c}{\# of replicates}\\
+\begin{tabular}{l|c|ccc|}
+& \multicolumn{1}{c|}{Empirical} & \multicolumn{3}{c}{Number of replicates}\\
 Parameter & value & 4 & 8 & 16 \\
@@
 $\eta_\delta$ & -0.046 & -0.046 & -0.031  & -0.022   \\
-$\eta_\phi$ & -2.7\phantom{00} & -2.3\phantom{00}  &  -2.3\phantom{00}  & -2.4\phantom{00} \\
+$\eta_\psi$ & -2.7\phantom{00} & -2.3\phantom{00}  &  -2.3\phantom{00}  & -2.4\phantom{00} \\
 \hline
-$\sigma_\phi$ & 1.8\phantom{0} & 1.8\phantom{0}  & 1.8\phantom{0}  & 1.7\phantom{0} \\
-$\sigma_\alpha$ & 0.36 & 0.46  & 0.40  & 0.36  \\
-$\sigma_\delta$ & 0.15 & 0.39  & 0.32  & 0.27  \\
-$\sigma_\psi$ & 0.74 & 0.40 & 0.49 & 0.54 \\
+$\sigma_\phi$ & 1.8\phantom{00} & 1.8\phantom{00}  & 1.8\phantom{00}  & 1.7\phantom{00} \\
+$\sigma_\alpha$ & 0.36\phantom{0} & 0.46\phantom{0}  & 0.40\phantom{0}  & 0.36\phantom{0}  \\
+$\sigma_\delta$ & 0.15\phantom{0} & 0.39\phantom{0}  & 0.32\phantom{0}  & 0.27\phantom{0}  \\
+$\sigma_\psi$ & 0.74\phantom{0} & 0.40\phantom{0} & 0.49\phantom{0} & 0.54\phantom{0} \\
 \hline
@@
 Table \ref{t:hyperparameter} provides the computed means and standard deviations from the gene-specific parameter values, called empirical values, and compares these values against the mean of estimated hyperparameters across the 10 simulations within replicate number.
-For the mean parameters, there is reasonable agreement amongst the different numbers of replicates with a somewhat high estimate for the mean of the overdispersion parameter $(\eta_\phi)$ and, perhaps, a trend for the mean of the hybrid effect ($\eta_\delta$), although we would not have expected this.
+For the mean parameters, there is reasonable agreement amongst the different numbers of replicates with a somewhat high estimate for the mean of the overdispersion parameter $(\eta_\psi)$ and, perhaps, a trend for the mean of the hybrid effect ($\eta_\delta$), although we would not have expected this.
 For the standard deviation parameters, $\sigma_\phi$ is consistent across the replicate numbers, $\sigma_\alpha$ and $\sigma_\delta$ mean estimates are decreasing toward the empirical values as the number of replicates increases, and $\sigma_\phi$ is increasing toward the empirical value as the number of replicates increases.
@@
 \end{figure}
-This figure shows shrinkage for large absolute values of $\alpha$ and $\delta$ from the \edgeR{} estimation toward $\hat{\eta}_\alpha$ and $\hat{\eta}_\delta$ which are both approximately zero.
+This figure shows shrinkage for large absolute estimates of $\alpha_g$ and $\delta_g$ from the \edgeR{} estimation toward $\hat{\eta}_\alpha$ and $\hat{\eta}_\delta$ which are both approximately zero.
 The figure also shows decreased values for the overdispersion parameter with larger decreases for high and low values of overdispersion.
-Finally, very little, if any, shrinkage is observed for $\phi$.
+Finally, very little, if any, shrinkage is observed for $\phi_g$ estimates.
 
 With posteriors for all parameters, we can calculate empirical Bayes posterior probabilities for LPH and HPH.
-For each gene, the quantity of interest is likely to be the maximum of these two probabilities.
+For each gene, the quantity of interest is the maximum of these two probabilities.
 For each gene with a high probability of either HPH or LPH, the magnitude of the effect is of interest, thus we calculate
@@
 
-Geneticists speculate that gene expression heterosis is one possible explanation of phenotypic heterosis of traits, such as plant height or grain yield.
-Existing methods for identifying differential gene expression based on \RNAseq{} data are not directly applicable for detecting heterosis genes.
+Geneticists speculate that gene expression EH is one possible explanation of hybrid vigor of traits, such as plant height or grain yield.
+Existing methods for identifying differential gene expression based on \RNAseq{} data are not directly applicable for detecting EH genes.
 \cite{ji2014estimation} introduced an empirical Bayes approach based on a hierarchical model for microarray data.
@@
 expectation-maximization algorithms or fully Bayesian approaches.
-
 Notwithstanding these improvements, we believe our approach is a computationally efficient method that can immediately aid scientists who are interested in identifying candidate genes involved in a genetic mechanism of heterosis.

### 2015-09-04 response.tex
response done

--- 2015-09-02 response.tex
@@
 In the revised manuscript, we have been more careful with terminology.
-We have defined heterosis to be a difference between the hybrid mean the average of the parental means.
+We have defined heterosis to be a difference between the hybrid mean and the average of the parental means.
 We have also introduced a new term (extreme heterosis, EH) to specify the situation where the hybrid mean is more extreme than both parental means.
@@
 
-\response{We have assessed the hyperparameter estimates in Section 3.4 and Table 1 where the estimates seem reasonable and improve with increased number of replicates per variety.}
+\response{Since we did not simulate from the hierarchical portion of the model, there are no true values of the hyperparameters to compare against.
+We tried to get at this question by calculating means and standard deviations of the true values for the gene-specific parameters in each simulation.
+Table \ref{t:hyperparameter} shows root mean squared error (RMSE) of the estimated hyperparameters.
+Because each simulation has a different value of the hyperparameters (due to different genes randomly being included), the RMSEs are hard to interpret.
+For this reason, we did not include this table in the manuscript because we did not believe it added much value.
+}
 
+\input{include/rmse}
+
+\newpage
 \comment{Page 8, line 37.  It is not obvious to me why $H_g2*$ (are the two parents the same?) is a necessary component of a heterosis (overdominance) test!?}
@@
 
-\response{We have assessed the hyperparameter estimates in Section 3.4 and Table 1 where the estimates seem reasonable and improve with increased number of replicates per variety.}
+\response{See discussion above and Table \ref{t:hyperparameter}.}
 

### 2015-09-04 ebayes_rnaseq_heterosis.tex
response done

@@
 \alpha_g = \frac{\mu_{g1}-\mu_{g2}}{2}, \quad \mbox{and} \quad
-\delta_g = \phi_g-\mu_{g3}. \]
+\delta_g = \mu_{g3}-\phi_g. \]
 
@@
 
-\subsection{Hyperparameter estimation}
-
-As described in Section \ref{s:simulation}, we simulated data only from the negative-binomial portion of the model, but not from the hierarchical distributions.
+% \subsection{Hyperparameter estimation}
+% 
+% As described in Section \ref{s:simulation}, we simulated data only from the negative-binomial portion of the model, but not from the hierarchical distributions.
 Thus, there are no true values for the hyperparameters to compare against.
 But, we can compute means and standard deviations of the gene-specific parameters that were used in the simulation study and compare these to the estimated hyperparameter values.
-\begin{table}
-\caption{Empirical values calculated from the gene-specific parameters used in the simulation and mean of estimated values across the 10 simulations within replicate number.
+% \include{include/rmse}
+% \include{include/rae}
+% % \begin{table}
+% % \caption{Empirical values calculated from the gene-specific parameters used in the simulation and mean of estimated values across the 10 simulations within replicate number.
 Maximum standard deviation in the estimates across the 10 simulations was 0.004.}
-\label{t:hyperparameter}
-\begin{tabular}{l|c|ccc|}
-& \multicolumn{1}{c|}{Empirical} & \multicolumn{3}{c}{Number of replicates}\\
-Parameter & value & 4 & 8 & 16 \\
-\hline
-% with (SD)
-% $\eta_\phi$ & 4.675 & 4.681 (0.004) & 4.741 (0.002) & 4.817 (0.003) \\
-% $\eta_\alpha$ & -0.017 & -0.023 (0.002) & -0.018 (0.003) & -0.014 (0.002)  \\
-% $\eta_\delta$ & -0.046 & -0.046 (0.002) & -0.031 (0.002) & -0.022 (0.003)  \\
-% $\eta_\phi$ & -2.713 & -2.290 (0.004) &  -2.335 (0.003) & -2.387 (0.002)\\
-% \hline
-% $\sigma_\phi$ & 1.815 & 1.821 (0.002) & 1.779 (0.002) & 1.726 (0.003)\\
-% $\sigma_\alpha$ & 0.364 & 0.459 (0.002) & 0.399 (0.002) & 0.357 (0.002) \\
-% $\sigma_\delta$ & 0.145 & 0.389 (0.002) & 0.317 (0.002) & 0.268 (0.001) \\
-% $\sigma_\psi$ & 0.737 & 0.401 (0.001) & 0.493 (0.001) & 0.542 (0.001) \\
-% 3 decimal places
-% $\eta_\phi$ & 4.675 & 4.681 & 4.741  & 4.817  \\
-% $\eta_\alpha$ & -0.017 & -0.023  & -0.018 & -0.014  \\
-% $\eta_\delta$ & -0.046 & -0.046 & -0.031  & -0.022   \\
-% $\eta_\phi$ & -2.713 & -2.290  &  -2.335  & -2.387\\
-% \hline
-% $\sigma_\phi$ & 1.815 & 1.821  & 1.779  & 1.726 \\
-% $\sigma_\alpha$ & 0.364 & 0.459  & 0.399  & 0.357  \\
-% $\sigma_\delta$ & 0.145 & 0.389  & 0.317  & 0.268  \\
-% $\sigma_\psi$ & 0.737 & 0.401 & 0.493 & 0.542 \\
-$\eta_\phi$ & 4.7\phantom{00} & 4.7\phantom{00} & 4.7\phantom{00}  & 4.8\phantom{00}  \\
-$\eta_\alpha$ & -0.017 & -0.023  & -0.018 & -0.014  \\
-$\eta_\delta$ & -0.046 & -0.046 & -0.031  & -0.022   \\
-$\eta_\psi$ & -2.7\phantom{00} & -2.3\phantom{00}  &  -2.3\phantom{00}  & -2.4\phantom{00} \\
-\hline
-$\sigma_\phi$ & 1.8\phantom{00} & 1.8\phantom{00}  & 1.8\phantom{00}  & 1.7\phantom{00} \\
-$\sigma_\alpha$ & 0.36\phantom{0} & 0.46\phantom{0}  & 0.40\phantom{0}  & 0.36\phantom{0}  \\
-$\sigma_\delta$ & 0.15\phantom{0} & 0.39\phantom{0}  & 0.32\phantom{0}  & 0.27\phantom{0}  \\
-$\sigma_\psi$ & 0.74\phantom{0} & 0.40\phantom{0} & 0.49\phantom{0} & 0.54\phantom{0} \\
-\hline
-\end{tabular}
-\end{table}
-Table \ref{t:hyperparameter} provides the computed means and standard deviations from the gene-specific parameter values, called empirical values, and compares these values against the mean of estimated hyperparameters across the 10 simulations within replicate number.
+% % \label{t:hyperparameter}
+% % \begin{tabular}{c|c|ccc|}
+% % & \multicolumn{1}{c|}{Empirical} & \multicolumn{3}{c}{Number of replicates}\\
+% % Parameter & value & 4 & 8 & 16 \\
+% % \hline
+% % $\eta_\phi$ & \phantom{-}4.6\phantom{00} & \phantom{-}4.7\phantom{00} & \phantom{-}4.7\phantom{00}  & \phantom{-}4.8\phantom{00}  \\
+% % $\eta_\alpha$ & -0.028 & -0.023  & -0.018 & -0.014  \\
+% % $\eta_\delta$ & \phantom{-}0.070 & \phantom{-}0.046 & \phantom{-}0.031  & \phantom{-}0.022   \\
+% % $\eta_\psi$ & -2.4\phantom{00} & -2.3\phantom{00}  &  -2.3\phantom{00}  & -2.4\phantom{00} \\
+% % \hline
+% % $\sigma_\phi$ & \phantom{-}1.8\phantom{00} & \phantom{-}1.8\phantom{00}  & \phantom{-}1.8\phantom{00}  & \phantom{-}1.7\phantom{00} \\
+% % $\sigma_\alpha$ & \phantom{-}0.45\phantom{0} & \phantom{-}0.46\phantom{0}  & \phantom{-}0.40\phantom{0}  & \phantom{-}0.36\phantom{0}  \\
+% % $\sigma_\delta$ & \phantom{-}0.31\phantom{0} & \phantom{-}0.39\phantom{0}  & \phantom{-}0.32\phantom{0}  & \phantom{-}0.27\phantom{0}  \\
+% % $\sigma_\psi$ & \phantom{-}0.59\phantom{0} & \phantom{-}0.40\phantom{0} & \phantom{-}0.49\phantom{0} & \phantom{-}0.54\phantom{0} \\
+% % \hline
+% % \end{tabular}
+% % \end{table}
+% Table \ref{t:hyperparameter} provides the computed means and standard deviations from the gene-specific parameter values, called empirical values, and compares these values against the mean of estimated hyperparameters across the 10 simulations within replicate number.
 For the mean parameters, there is reasonable agreement amongst the different numbers of replicates with a somewhat high estimate for the mean of the overdispersion parameter $(\eta_\psi)$ and, perhaps, a trend for the mean of the hybrid effect ($\eta_\delta$), although we would not have expected this.

### 2015-09-16 ebayes_rnaseq_heterosis.tex
final version

@@
 
-\newcommand{\blind}{1} % 0 = not blind, 1 = blind
+\newcommand{\blind}{0} % 0 = not blind, 1 = blind
 
-%\newcommand{\Stan}{{\tt Stan}}
-\newcommand{\Stan}{Stan}
+\newcommand{\Stan}{{\tt Stan}}
 \newcommand{\RStan}{{\tt RStan}}
@@
 \begin{keywords}
-Hierarchical model; Negative binomial; \RNAseq{}; Bayesian LASSO; Parallel computing; \Stan{}.
+Hierarchical model; Negative binomial; \RNAseq{}; Bayesian LASSO; Parallel computing; hybrid vigor.
 \end{keywords}
@@
 Initial attempts to perform a fully Bayesian analysis via Markov chain Monte Carlo (MCMC) failed due to high computational costs and poor mixing in the resulting chains.
-For example, we implemented the model in the statistical software \Stan{} (discussed at the end of this section), ran the MCMC on a simulated data set with 10,000 genes for 2 months on a state-of-the-art linux server, and obtained potential scale reduction factors that suggested we would need to run at least ten times as long to obtain convergence.
+For example, we implemented the model in the statistical software \Stan{} (discussed at the end of this section), ran the MCMC on a simulated data set with 10,000 genes for 2 months on a state-of-the-art linux server, and obtained potential scale reduction factors \citep{Gelm:Rubi:infe:1992} that suggested we would need to run at least ten times as long to obtain convergence.
 Although there are certainly improvements that could be made to decrease computational costs and improve mixing, we opted for an empirical Bayes approach.
@@
 Conditional on the estimated normalization factors $\hat{\gamma}$ and hyperparameters $\hat{\pi}$, we perform a Bayesian analysis to re-estimate the gene-specific parameters and describe their uncertainty.
-Conditional on $\hat{\gamma}$ and $\hat{\pi}$ the gene-specific parameters are independent, i.e.
+Equation \ref{e:condind} shows that conditional on $\hat{\gamma}$ and $\hat{\pi}$ the gene-specific parameters are independent and therefore conditional posterior inference across the genes can be parallelized.
+In this equation, the densities for $\alpha_g$ and $\delta_g$ will depend on whether we are assuming normal or Laplace distributions.
 \begin{equation}
@@
 N(\phi_g;\hat{\eta}_\phi, \hat{\sigma}_\phi^2)
-La(\alpha_g;\hat{\eta}_\alpha, \hat{\sigma}_\alpha)
-La(\delta_g;\hat{\eta}_\delta, \hat{\sigma}_\delta)
+p(\alpha_g;\hat{\eta}_\alpha, \hat{\sigma}_\alpha)
+p(\delta_g;\hat{\eta}_\delta, \hat{\sigma}_\delta)
 N(\psi_g;\hat{\eta}_\psi, \hat{\sigma}_\psi^2) 
@@
 \end{equation}
-Thus, conditional posterior inference across the genes can be obtained independently and in parallel.
 
-To perform this Bayesian analysis, we used the statistical software \Stan{} \citep{stan-software:2014} executed through the \RStan{} interface \citep{rstan-software:2014} in R \citep{R2014}.
+To perform the conditional posterior inference on the gene-specific parameters, we used the statistical software \Stan{} \citep{stan-software:2014} executed through the \RStan{} interface \citep{rstan-software:2014} in R \citep{R2014}.
 \Stan{} implements a variant of MCMC called Hamiltonion Monte Carlo \citep{neal2011mcmc} to obtain samples from the posterior in equation \eqref{e:condind}.
-We ran 4 simultaneous chains with random initial starting values for 1000 burn-in (and tuning) iterations followed by another 1000 iterations thinned by 4 (to reduce storage space) for inference.
+We ran 4 simultaneous chains with random initial starting values for 1000 burn-in (and tuning) iterations followed by another 1000 iterations retaining every fourth sample (to reduce storage space) for inference.
 We monitored convergence using the potential scale reduction factor and effective sample size (ESS) for $\phi_g$, $\alpha_g$, $\delta_g$, and $\psi_g$ \citep{Gelm:Rubi:infe:1992}.
@@
 
-We will use the term eBayes to refer to the approaches defined in Sections \ref{s:model}--\ref{s:gene_heterosis} and add parenthetical labels ``normal'' and ``Laplace'' to distinguish between the two eBayes approaches.
-
 \subsection{Implementation in \ShrinkBayes{}} \label{s:shrinkbayes}
@@
 %the heterosis probability for gene $g$ is zero if $|\hat{\delta}_g| \le |\hat{\alpha}_g|$, $P(\delta_g>\alpha_g|y)$ if $\hat{\delta}_g > -\hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g > 0 |y)$ if $\hat{\delta}_g > \hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g < 0 |y)$ if $\hat{\delta}_g < \hat{\alpha}_g \le 0$, and $P(\delta_g + \alpha_g < 0 |y)$ if $\hat{\delta}_g < -\hat{\alpha}_g \le 0$.
+
+We will use the term eBayes to refer to the approach defined in Sections \ref{s:model}--\ref{s:gene_heterosis} and add parenthetical labels ``normal'' and ``Laplace'' to specify whether we are assuming normal or Laplace distributions for half-parental differences and hybrid effects.
 
@@
 Then, for each gene, we defined a new $p$-value denoted as $p_g$ and set to $p_g=1$ if the estimate of $\mu_{g3}$ falls between the estimates of $\mu_{g1}$ and $\mu_{g2}$ and $p_g=\max\{p_{g1},p_{g2}\}/2$ otherwise.
-(For all relevant significance thresholds $\omega$ near 0, it can be shown that rejecting the null hypothesis of no EH for gene $g$ whenever $p_g\leq \omega$ results in a test that is asymptotically size $\omega$.)  We then use $1-p_g$ as a final measure positively associated with strength of evidence for EH.
+For all relevant significance thresholds $\omega$ near 0, it can be shown that rejecting the null hypothesis of no EH for gene $g$ whenever $p_g\leq \omega$ results in a test that is asymptotically size $\omega$.
+We then use $1-p_g$ as a final measure positively associated with strength of evidence for EH.
 
@@
 
-%\section*{Supplementary Materials}
-%The supplementary materials include three plain text files: the maize experiment data ({\tt data.csv}), a script to run our method ({\tt script.R}) on this data, and the \Stan{} model ({\tt model.stan}).
+\section*{Supplementary Materials}
 
+The supplementary materials include links to the data as well as code to reproduce the analysis in this manuscript.
 \bibliography{jarad}

### 2015-09-16 supp.tex
add supplementary material

\documentclass[10pt]{article}

\usepackage{fullpage, hyperref}

\usepackage{amsmath}
\usepackage[figuresright]{rotating}

\def\bSig\mathbf{\Sigma}
\newcommand{\VS}{V\&S}
\newcommand{\tr}{\mbox{tr}}

\newcommand{\Stan}{{\tt Stan}}
\newcommand{\RStan}{{\tt RStan}}
\newcommand{\edgeR}{{\tt edgeR}}
\newcommand{\baySeq}{{\tt baySeq}}
\newcommand{\ShrinkBayes}{{\tt ShrinkBayes}}
\newcommand{\RNAseq}{RNA-seq}

%\renewcommand{\gamma}{c}

\title{Supplmentary material: Empirical Bayes analysis of \RNAseq{} data for detection of gene expression heterosis}

\author{Jarad Niemi, 
Eric Mittman, 
Will Landau, and 
Dan Nettleton \\
Department of Statistics, Iowa State University, Ames, Iowa, U.S.A.}

\begin{document}

\maketitle

The data and code necessary to reproduce the analyses in the manuscript are available at \url{https://github.com/jarad/niemi2015empirical}. 
The repository contains a {\tt README.md} file that describes the directory structure and files within those directories.
In particular, the data used in the real data analysis are available at \url{https://github.com/jarad/niemi2015empirical/tree/master/real/data}. 
If these data or code are used in a future publication, please cite the manuscript associated with this supplementary material.

\end{document}

### 2015-09-16 ebayes_rnaseq_heterosis.tex
typo
removed comments

@@
 \newcommand{\RNAseq}{RNA-seq}
-
-%\renewcommand{\gamma}{c}
 
@@
 \begin{keywords}
-Hierarchical model; Negative binomial; \RNAseq{}; Bayesian LASSO; Parallel computing; hybrid vigor.
+Hierarchical model; Negative binomial; \RNAseq{}; Bayesian LASSO; Parallel computing; Hybrid vigor.
 \end{keywords}
@@
 \[ 
-%\gamma_{vi} \stackrel{ind}{\sim} N(0,\sigma_\gamma^2), \quad 
 \phi_g \stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2), \,
@@
 p\left(\theta\left|y,\hat{\pi},\hat{\gamma}\right.\right) 
-%&= \prod_{g=1}^G p\left(\theta_g\left|y_g,\hat{\pi}\right.\right) = \prod_{g=1}^G p\left(y_g\left|\theta_g,\hat{\gamma}\right.\right) p\left(\theta_g\left|\hat{\pi}\right.\right) \nonumber \\
-%&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right.\right) \right] p(\theta_g|\hat{\eta},\hat{\sigma}) \\
 \propto \prod_{g=1}^G & \left[ \,\prod_{v=1}^V \prod_{i=1}^{n_v} \mbox{NB}\left(y_{gvi};e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right)  \right. \\
@@
 $\mu_{g3} > \max\{\mu_{g1},\mu_{g2}\}$ or, equivalently, $\delta_g > |\alpha_g|$.
-% and high-parent  heterosis (HPH) of gene expression occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$.
-For simplicity, we define \emph{heterosis} of gene expression to be the event of LPH or HPH.
-The most important hypotheses are
-% \begin{align}
-% H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
-% H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3},
-% \label{e:hypotheses}
-% \end{align}
-% where $H_{g0}$ indicates no heterosis for gene $g$ and $H_{g1}$ indicates  heterosis for gene $g$. 
-% We evaluate these hypotheses based on empirical Bayes estimates of their posterior probabilities, e.g., 
-% \begin{align}
-% P\left(H_{g1}|y, \hat{\pi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\pi}_{MAP}\right) \nonumber \\
-% &\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right), \label{e:probs}
-% \end{align}
-% where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise.
 We evaluate these probabilities based on empirical Bayes estimates of their posterior probabilities, e.g., 
 \begin{equation}
-%P(LPH_g|y) \approx 
 P(\mbox{LPH}_g|y,\hat{\pi},\hat{\gamma}) = 
-%P\left(\left. \mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}\right|y,\hat{\pi},\hat{\gamma}\right) = 
 P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi},\hat{\gamma}\right) \approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right), 
@@
 \[ 
-% P_{\ShrinkBayes{}}(LPH_g|y) \equiv \left\{ \begin{array}{ll}
-% P(\delta_g>-\alpha_g|y) & \mbox{if } \hat{\delta}_g > -\hat{\alpha}_g \ge 0 \\
-% P(\delta_g>\phantom{-}\alpha_g|y) & \mbox{if } \hat{\delta}_g > \phantom{-}\hat{\alpha}_g \ge 0 \\
-% 0 & \mbox{otherwise}
-% \end{array} \right.
 P_{\ShrinkBayes{}}(\mbox{LPH}_g|y) \equiv P(\delta_g<-\alpha_g|y)\mathrm{I}\left(\hat{\delta}_g < -\hat{\alpha}_g \le 0\right) + P(\delta_g<\alpha_g|y) \mathrm{I}\left(\hat{\delta}_g < \hat{\alpha}_g \le 0\right).
-%P_{\ShrinkBayes{}}(HPH|y) \equiv \left\{ \begin{array}{ll}
-%P(\delta_g<-\alpha_g|y) & \mbox{if } \hat{\delta}_g > -\hat{\alpha}_g \ge 0 \\
-%(\delta_g<\phantom{-}\alpha_g|y) & \mbox{if } \hat{\delta}_g > \phantom{-}\hat{\alpha}_g \ge 0 \\
-%0 & \mbox{otherwise}
-%\end{array} \right.
 \]
 HPH probability is defined similarly with all inequalities reversed. As before, we construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities.
-%the heterosis probability for gene $g$ is zero if $|\hat{\delta}_g| \le |\hat{\alpha}_g|$, $P(\delta_g>\alpha_g|y)$ if $\hat{\delta}_g > -\hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g > 0 |y)$ if $\hat{\delta}_g > \hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g < 0 |y)$ if $\hat{\delta}_g < \hat{\alpha}_g \le 0$, and $P(\delta_g + \alpha_g < 0 |y)$ if $\hat{\delta}_g < -\hat{\alpha}_g \le 0$.
 
@@
 \label{s:simulation}
-
-% \subsection{Coverage for our model}
-% \input{../temp/analyze-heterosis/FIGURES/coverage}
 
@@
 From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to distinguish genes with EH, as defined in Section \ref{s:sim_data},
-% From Dan, "why this citation?"
 from those without EH. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
@@
 We suspect this is because the Laplace distributions are better approximations to the true underlying distribution for these parameters, and we discuss this in Section \ref{s:discussion}.
-
-% \subsection{Hyperparameter estimation}
-% 
-% As described in Section \ref{s:simulation}, we simulated data only from the negative-binomial portion of the model, but not from the hierarchical distributions.
-Thus, there are no true values for the hyperparameters to compare against.
-But, we can compute means and standard deviations of the gene-specific parameters that were used in the simulation study and compare these to the estimated hyperparameter values.
-% \include{include/rmse}
-% \include{include/rae}
-% % \begin{table}
-% % \caption{Empirical values calculated from the gene-specific parameters used in the simulation and mean of estimated values across the 10 simulations within replicate number.
-Maximum standard deviation in the estimates across the 10 simulations was 0.004.}
-% % \label{t:hyperparameter}
-% % \begin{tabular}{c|c|ccc|}
-% % & \multicolumn{1}{c|}{Empirical} & \multicolumn{3}{c}{Number of replicates}\\
-% % Parameter & value & 4 & 8 & 16 \\
-% % \hline
-% % $\eta_\phi$ & \phantom{-}4.6\phantom{00} & \phantom{-}4.7\phantom{00} & \phantom{-}4.7\phantom{00}  & \phantom{-}4.8\phantom{00}  \\
-% % $\eta_\alpha$ & -0.028 & -0.023  & -0.018 & -0.014  \\
-% % $\eta_\delta$ & \phantom{-}0.070 & \phantom{-}0.046 & \phantom{-}0.031  & \phantom{-}0.022   \\
-% % $\eta_\psi$ & -2.4\phantom{00} & -2.3\phantom{00}  &  -2.3\phantom{00}  & -2.4\phantom{00} \\
-% % \hline
-% % $\sigma_\phi$ & \phantom{-}1.8\phantom{00} & \phantom{-}1.8\phantom{00}  & \phantom{-}1.8\phantom{00}  & \phantom{-}1.7\phantom{00} \\
-% % $\sigma_\alpha$ & \phantom{-}0.45\phantom{0} & \phantom{-}0.46\phantom{0}  & \phantom{-}0.40\phantom{0}  & \phantom{-}0.36\phantom{0}  \\
-% % $\sigma_\delta$ & \phantom{-}0.31\phantom{0} & \phantom{-}0.39\phantom{0}  & \phantom{-}0.32\phantom{0}  & \phantom{-}0.27\phantom{0}  \\
-% % $\sigma_\psi$ & \phantom{-}0.59\phantom{0} & \phantom{-}0.40\phantom{0} & \phantom{-}0.49\phantom{0} & \phantom{-}0.54\phantom{0} \\
-% % \hline
-% % \end{tabular}
-% % \end{table}
-% Table \ref{t:hyperparameter} provides the computed means and standard deviations from the gene-specific parameter values, called empirical values, and compares these values against the mean of estimated hyperparameters across the 10 simulations within replicate number.
-For the mean parameters, there is reasonable agreement amongst the different numbers of replicates with a somewhat high estimate for the mean of the overdispersion parameter $(\eta_\psi)$ and, perhaps, a trend for the mean of the hybrid effect ($\eta_\delta$), although we would not have expected this.
-For the standard deviation parameters, $\sigma_\phi$ is consistent across the replicate numbers, $\sigma_\alpha$ and $\sigma_\delta$ mean estimates are decreasing toward the empirical values as the number of replicates increases, and $\sigma_\phi$ is increasing toward the empirical value as the number of replicates increases.
-%We expect this pattern for $\sigma_\alpha$ and $\sigma_\delta$ since these are based on moment matching of independent gene-specific parameter estimates.
-For low numbers of replicates, these independent estimates will have relatively large sampling variability that will cause a bias toward larger values for the moment-matched standard deviation estimates.
 
@@
 \begin{equation}
-% \mbox{estimated effect size}_g = \left\{ 
-% \begin{array}{ll}
-% \hat{\mu}_{g3} - \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} < \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
-% \hat{\mu}_{g3} - \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} > \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
-% 0 & \mbox{otherwise},
-% \end{array} 
-% \right. 
 \mbox{estimated effect size}_g \equiv \left\{ 

### 2015-09-16 response2.tex
revision 2

\documentclass{article}

\usepackage{amsmath}
\usepackage{fullpage}
\usepackage{color}

\usepackage[parfill]{parskip}

\newcommand{\comment}[1]{\textit{#1}}
\newcommand{\response}[1]{#1}
\newcommand{\todo}[1]{{\color{red} #1}}

\begin{document}

\response{We thank the AE for additional editorial feedback. In addition to addressing all feedback, we have added Supplementary Material that provides URLs to code and data used in the manuscript. Our changes due to feedback are shown below.}

\comment{Are you sure you want to include a software product as a keyword?}

\response{Replaced Stan with ``Hybrid vigor''.}

\comment{Your call but wouldn't you like to put software names / functions, etc. in a different font for emphasis}

\response{Changed Stan to {\tt Stan}.}

\comment{You probably should cite Gelmand and Rubin here, I think.}

\response{Added a citation to Gelman and Rubin.}

\comment{Ok...but you're contradicting yourself then on page 15 where you seem to cast doubt on your EB approach.}

\response{Here we state ``This approach may be a reasonable approximation'', but in the Discussion we openly express our doubts about whether it is a reasonable approximation.}

\comment{Does this mean you're saving every 4 iterations?? Maybe this expression is obvious to most Bayesians but I'm not sure.}

\response{Changed ``thinned by 4'' to ``retaining every fourth sample''.}

\comment{I'm not sure that you've adequately characterized what the two approaches are up until now? For example, which of the two are represented in Equation (2)?}

\response{Equation (2) has been updated to 
\begin{equation}
\begin{split}
p\left(\theta\left|y,\hat{\pi},\hat{\gamma}\right.\right) 
\propto \prod_{g=1}^G & \left[ \,\prod_{v=1}^V \prod_{i=1}^{n_v} \mbox{NB}\left(y_{gvi};e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right)  \right. \\
&\times 
\left. \vphantom{\prod_{v=1}^V}
N(\phi_g;\hat{\eta}_\phi, \hat{\sigma}_\phi^2)
p(\alpha_g;\hat{\eta}_\alpha, \hat{\sigma}_\alpha)
p(\delta_g;\hat{\eta}_\delta, \hat{\sigma}_\delta)
N(\psi_g;\hat{\eta}_\psi, \hat{\sigma}_\psi^2) 
\right]
\label{e:condind}
\end{split}
\end{equation}
with the accompanying sentence ``In this equation, the densities for $\alpha_g$ and $\delta_g$ will depend on whether we are assuming normal or Laplace distributions.'' And the last sentence of section 2.3 has been changed to ``We will use the term eBayes to refer to the approach defined in Sections 2.1--2.3 and add parenthetical labels ``normal'' and ``Laplace'' to specify whether we are assuming normal or Laplace distributions for half-parental differences and hybrid effects.''}

\comment{Is it really necessary to include the parentheses?}

\response{Removed the parentheses.}

\comment{Delete ``markedly''.}

\response{Deleted ``markedly''.}

\end{document}

### 2015-09-16 title_page.tex
revision 2

--- 2015-09-16 ebayes_rnaseq_heterosis.tex
@@
 
-\date{{\it Received July} 2015}
-
 \pagerange{\pageref{firstpage}--\pageref{lastpage}} 
@@
 
-%  A maximum of six (6) tables or figures combined is often required.''
-
-\section{Introduction}
-\label{s:intro}
-
-Heterosis exists when the expected value of a hybrid phenotype differs from the average of the expected phenotypic values of the hybrid's parents.
-The most interesting and useful form of heterosis, known as hybrid vigor, occurs when hybrid progeny display a mean phenotype that is superior to  both parental phenotypic means.
-This heterosis phenomenon was scientifically documented in plants by \cite{darwin1876effects} and has long been used to improve agricultural production.
-One classic example involves hybrid maize offspring that are taller, faster to mature, and yield considerably more grain than their inbred parents \citep{hallauer1981quantitative, hallauer2010quantitative}.
-
-Depending on whether large or small values of a phenotype are favorable, hybrid vigor can occur if the mean phenotype of a hybrid is greater than both parental means or less than both parental means.
-We refer to the former as high-parent heterosis (HPH) and the latter as low-parent heterosis (LPH).
-Note that heterosis, HPH, and LPH are similar to the quantitative genetics concepts of dominance, overdominance, and underdominance.
-However, the various forms of dominance are usually reserved for describing the association of mean phenotype with homozygous and heterozygous genotypes at a single genetic locus.
-Heterosis involves a comparison of inbreds (simultaneously homozygous at many loci) and hybrids (simultaneously heterozygous at many loci).
-For simplicity, throughout the remainder of the article, we will use the term \emph{extreme heterosis} (EH) to describe the situation where the hybrid mean is more extreme than the parental means, i.e.
-we say there is EH if and only if either HPH or LPH holds.
-
-Despite intensive study and successful use of heterosis in agriculture, the basic molecular genetic mechanisms remain poorly understood \citep{chen2013genomic}.
-One potential explanation, is EH of gene expression, i.e.
-enhanced (or suppressed) expression of one or more genes in hybrids compared to their inbred parents.
-Gene expression heterosis has been investigated in maize by \cite{swanson2006all} and \cite{springer2007allelic} and EH of gene expression is conceptually depicted in the right column of Figure 1b in \cite{chen2013genomic}.
-
-Recently, \cite{ji2014estimation} introduced an approach to assess gene expression heterosis using microarray data under the assumption that these data are continuous.
-They built a normal hierarchical model for microarray measurements of transcript abundance that allows borrowing of information across genes to estimate means and variances.
-They introduced an empirical Bayes framework that first estimates model hyperparameters, then estimates the posterior distribution for gene-specific parameters conditional on those hyperparameters, and finally computes heterosis probabilities based on integrals of regions under this posterior.
-This development was necessary due to the composite null hypotheses in tests for heterosis.
-These hypotheses, which many available methods do not fully accommodate, remain a challenge in the transition from continuous measurements of transcript abundance to count-based measurements that arise from RNA sequencing (\RNAseq{}) technology.
-Building on the work of \citeauthor{ji2014estimation} with the normal data model, we construct a hierarchical model based on a negative binomial data model.
-We also utilize an empirical Bayes approach to obtain estimates of the hyperparameters and the posterior distributions for the gene-specific parameters conditional on those hyperparameters.
-
-The remainder of the paper proceeds as follows.
-Section \ref{s:method} presents the proposed hierarchical model, an empirical Bayes method of estimating the parameters, and the calculation of posterior probabilities of EH.
-Section \ref{s:simulation} presents a simulation study based on a maize experiment and compares our approach to alternative methods.
-Section \ref{s:maize} analyzes a maize experiment where hybrid vigor is well established and identifies genes demonstrating EH of expression.
-Section \ref{s:discussion} summarizes the work and suggests directions for future research.
-
-\section{Empirical Bayes identification of gene expression heterosis from \RNAseq{} read counts}
-\label{s:method}
-
-We consider an RNA sequencing (\RNAseq{}) experiment that involves at least three genetic varieties: two parental varieties and a cross between these two varieties called the hybrid.
-For each variety, replicate RNA samples are isolated and assessed for quality.
-Complementary DNA (cDNA) libraries, consisting of short cDNA fragments derived from RNA, are constructed.
-Then, next generation sequencing technology is used to determine the \emph{reads}, or nucleotide sequences, in the cDNA libraries.
-These reads are processed using bioinformatic algorithms to match the reads to genes, or specific gene transcripts, exons, microRNAs, etc.
-The results of read processing are summarized by a gene $\times$ sample matrix of counts.
-See \cite{datta2014statistical} for more details on \RNAseq{} experiments and data from a statistical perspective, and see \cite{paschold2012complementation} for the biological background behind the use of \RNAseq{} to study gene expression heterosis.
-
-To use \RNAseq{} counts to identify genes displaying EH of expression, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the hyperparameters using an empirical Bayes procedure, and calculate empirical Bayes posterior probabilities for EH.
-
-\subsection{Hierarchical model for \RNAseq{} counts}
-\label{s:model}
-
-Let $Y_{gvi}$ be the count for gene $g=1,\ldots,G$, variety $v=1,\ldots,V$, and replicate $i=1,\ldots,n_v$.  We assume
-\begin{equation} 
-Y_{gvi} \stackrel{ind}{\sim} \mbox{NB}\left(e^{\mu_{gv}+\gamma_{vi}},e^{\psi_g}\right), 
-\label{e:data}
-\end{equation}
-where NB$(\xi,e^\psi)$ indicates a negative binomial distribution with expectation $\xi$ and variance $\xi+e^\psi\xi^2$, and $ind$ indicates the observations are conditionally independent.
-As shown in equation \eqref{e:data}, our data model involves gene-specific overdispersion $\psi_g$ and a mean that depends on the gene-variety combination through $\mu_{gv}$ and on the sample though  $\gamma_{vi}$.
-The $\mu_{gv}$ terms are of primary scientific interest; the $\gamma_{vi}$ terms are normalization factors that account for differences in the thoroughness of sequencing from sample to sample.
-
-Following \cite{ji2014estimation}, we reparameterize the gene-variety mean structure into the gene-specific parental average ($\phi_g$), half-parental difference ($\alpha_g$), and hybrid effect ($\delta_g$).
-For our heterosis study where $V=3$, we let $v=1,2$ indicate the two parental varieties and $v=3$ indicate the hybrid.
-The reparameterization is
-\[ \phi_g = \frac{\mu_{g1}+\mu_{g2}}{2}, \quad 
-\alpha_g = \frac{\mu_{g1}-\mu_{g2}}{2}, \quad \mbox{and} \quad
-\delta_g = \mu_{g3}-\phi_g. \]
-
-We assume a hierarchical model for the gene-specific mean parameters and overdispersion parameters.
-Initially, we assume the parental averages, half-parental differences, hybrid effect, and overdispersion parameters follow normal distributions, i.e.
-\[ 
-\phi_g \stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2), \,
-\alpha_g \stackrel{ind}{\sim} N(\eta_\alpha, \sigma_\alpha^2), \,
-\delta_g \stackrel{ind}{\sim} N(\eta_\delta, \sigma_\delta^2), \, \mbox{ and } \,
-\psi_g \stackrel{ind}{\sim} N(\eta_\psi,\sigma_\psi^2). 
-\]
-Empirical plots of estimated values of $\alpha_g$ and $\delta_g$ for our maize data set (described in Section \ref{s:maize}) suggest that the distribution of these parameters are more peaked near zero and have heavier tails than a normal distribution allows.
-For differential expression between the two parental phenotypes, we would expect many genes to have small effects and only a few genes to have relatively large effects.
-For these many genes with small effects, we might expect the hybrid to act similar to its parents and therefore also have many genes where the hybrid effect is small and only a few genes where this hybrid effect is large.
-For these reasons, we also assessed Laplace (or double exponential) distributions for the half-parental difference and hybrid effect and thus implement a Bayesian LASSO \citep{park2008bayesian,hans2009bayesian}, i.e.
-\[ 
-\alpha_g \stackrel{ind}{\sim} La(\eta_\alpha, \sigma_\alpha) \qquad \mbox{and} \qquad
-\delta_g \stackrel{ind}{\sim} La(\eta_\delta, \sigma_\delta) 
-\]
-where $\alpha\sim La(\eta,\sigma)$ has a probability density function given by $La(\alpha;\eta,\sigma) = \exp(-|\alpha-\eta|/\sigma)/2\sigma$ with expectation $\eta$ and variance $2\sigma^2$.
-Whether using normal or Laplace distributions, we assume \emph{a priori} independence amongst the parental averages, half parental differences, hybrid effects, and overdispersion parameters.
-
-\subsection{Empirical Bayes}
-\label{s:ebayes}
-
-Initial attempts to perform a fully Bayesian analysis via Markov chain Monte Carlo (MCMC) failed due to high computational costs and poor mixing in the resulting chains.
-For example, we implemented the model in the statistical software \Stan{} (discussed at the end of this section), ran the MCMC on a simulated data set with 10,000 genes for 2 months on a state-of-the-art linux server, and obtained potential scale reduction factors \citep{Gelm:Rubi:infe:1992} that suggested we would need to run at least ten times as long to obtain convergence.
-Although there are certainly improvements that could be made to decrease computational costs and improve mixing, we opted for an empirical Bayes approach.
-This approach may be a reasonable approximation to a fully Bayesian approach when estimating models with large numbers of genes as the posterior distributions for the hyperparameters may be tightly peaked.
-
-We categorize the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$, normalization factors $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, and hyperparameters $\pi = (\eta, \sigma)$ where $\eta=(\eta_\phi,\eta_\alpha,\eta_\delta, \eta_\psi)$ and $\sigma=(\sigma_\phi,\sigma_\alpha,\sigma_\delta,\sigma_\psi)$.
-We obtain estimates for the hyperparameters and then base gene-specific inference on the posterior conditional on these estimates.
-
-To obtain normalization factors $\hat{\gamma}$, we use the weighted trimmed mean of $M$ values (TMM) method of \cite{robinson2010scaling}.
-We use \edgeR{} to obtain genewise dispersion estimates, $\hat{\psi}_g$, and the generalized linear model methods to obtain estimates for the remaining gene-specific parameters, i.e.
-$\hat{\phi}_g$, $\hat{\alpha}_g$, and $\hat{\delta}_g$ \citep{robinson2010edgeR}.
-Using $\hat{\theta}_g = (\hat{\phi}_g,\hat{\alpha}_g,\hat{\delta}_g,\hat{\psi}_g)$, we estimate hyperparameters for the location and scale parameters in the hierarchical model using a central method of moments approach.
-For example, $\hat{\eta}_\phi = \sum_{g=1}^G \hat{\phi}_g/G$ and $\hat{\sigma}_\phi^2 = \sum_{g=1}^G (\hat{\phi}_g - \hat{\eta}_\phi)^2/(G-1)$ (and similarly for $\hat{\eta}_\psi$ and $\hat{\sigma}_\psi$) and $\hat{\eta}_\alpha = \sum_{g=1}^G \hat{\alpha}_g/G$ and $2\hat{\sigma}_\alpha^2 = \sum_{g=1}^G (\hat{\alpha}_g - \hat{\eta}_\alpha)^2/(G-1)$ (and similarly for $\hat{\eta}_\delta$ and $\hat{\sigma}_\delta$) for the model assuming Laplace distributions.
-
-Conditional on the estimated normalization factors $\hat{\gamma}$ and hyperparameters $\hat{\pi}$, we perform a Bayesian analysis to re-estimate the gene-specific parameters and describe their uncertainty.
-Equation \ref{e:condind} shows that conditional on $\hat{\gamma}$ and $\hat{\pi}$ the gene-specific parameters are independent and therefore conditional posterior inference across the genes can be parallelized.
-In this equation, the densities for $\alpha_g$ and $\delta_g$ will depend on whether we are assuming normal or Laplace distributions.
-\begin{equation}
-\begin{split}
-p\left(\theta\left|y,\hat{\pi},\hat{\gamma}\right.\right) 
-\propto \prod_{g=1}^G & \left[ \,\prod_{v=1}^V \prod_{i=1}^{n_v} \mbox{NB}\left(y_{gvi};e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right)  \right. \\
-&\times 
-\left. \vphantom{\prod_{v=1}^V}
-N(\phi_g;\hat{\eta}_\phi, \hat{\sigma}_\phi^2)
-p(\alpha_g;\hat{\eta}_\alpha, \hat{\sigma}_\alpha)
-p(\delta_g;\hat{\eta}_\delta, \hat{\sigma}_\delta)
-N(\psi_g;\hat{\eta}_\psi, \hat{\sigma}_\psi^2) 
-\right]
-\label{e:condind}
-\end{split}
-\end{equation}
-
-To perform the conditional posterior inference on the gene-specific parameters, we used the statistical software \Stan{} \citep{stan-software:2014} executed through the \RStan{} interface \citep{rstan-software:2014} in R \citep{R2014}.
-\Stan{} implements a variant of MCMC called Hamiltonion Monte Carlo \citep{neal2011mcmc} to obtain samples from the posterior in equation \eqref{e:condind}.
-We ran 4 simultaneous chains with random initial starting values for 1000 burn-in (and tuning) iterations followed by another 1000 iterations retaining every fourth sample (to reduce storage space) for inference.
-We monitored convergence using the potential scale reduction factor and effective sample size (ESS) for $\phi_g$, $\alpha_g$, $\delta_g$, and $\psi_g$ \citep{Gelm:Rubi:infe:1992}.
-If the minimum ESS was less than 1000, we reran the chains with double the iterations for both burn-in and inference.
-We continued this restarting and doubling until we obtained minimum ESS greater than 1000 for all parameters.
-
-\subsection{Gene expression heterosis}
-\label{s:gene_heterosis}
-
-In the maize context that motivates this work, we are interested in extreme heterosis (EH), i.e. either low-parent heterosis (LPH) or high-parent heterosis (HPH), in gene expression. 
-For a specific gene $g$, LPH occurs when expected expression in the hybrid is less than the expected expression of either parent, i.e.
-$\mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}$ or, equivalently, $\delta_g < -|\alpha_g|$, and HPH occurs when expected expression in the hybrid is greater than the exptected expression of either parent, i.e.
-$\mu_{g3} > \max\{\mu_{g1},\mu_{g2}\}$ or, equivalently, $\delta_g > |\alpha_g|$.
-We evaluate these probabilities based on empirical Bayes estimates of their posterior probabilities, e.g., 
-\begin{equation}
-P(\mbox{LPH}_g|y,\hat{\pi},\hat{\gamma}) = 
-P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi},\hat{\gamma}\right) \approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right), 
-\label{e:probs}
-\end{equation}
-where $\left(\delta_g^{(m)},\alpha_g^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise.
-HPH probability is defined similarly with the inequality reversed and without the negative sign.
-We construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities.
-Geneticists can use this list to prioritize future experiments to understand the molecular genetic mechanisms for heterosis.
-
-\subsection{Implementation in \ShrinkBayes{}} \label{s:shrinkbayes}
-
-In addition to the approach above, we utilized \ShrinkBayes{} to estimate EH probabilities with two modifications described here.
-\ShrinkBayes{} utilizes integrated nested Laplace approximation (INLA) \citep{rue2009approximate} in combination with empirical Bayes ideas \citep{van2014shrinkbayes}.
-One limitation with inferential methods based on INLA is that all distributions, except for the data distribution, must have tails as light or lighter than the normal density.
-Thus, we cannot implement the Laplace priors for the half-parental difference ($\alpha_g$) and the hybrid effect ($\delta_g$) and instead use normal priors in this situation.
-An additional limitation is that INLA provides approximations to marginal posteriors for parameters or linear combinations of parameters, but not an approximation to the full posterior.
-Since we are interested in non-linear quantities such as $P(\delta_g > |\alpha_g||y)$, we cannot compute these directly using \ShrinkBayes{}.
-Instead, for \ShrinkBayes{}, we calculate EH probabilities conditional on posterior means for the half-parental difference and hybrid effect, i.e.
-$\hat{\delta}_g$ and $\hat{\alpha}_g$.
-For example,
-\[ 
-P_{\ShrinkBayes{}}(\mbox{LPH}_g|y) \equiv P(\delta_g<-\alpha_g|y)\mathrm{I}\left(\hat{\delta}_g < -\hat{\alpha}_g \le 0\right) + P(\delta_g<\alpha_g|y) \mathrm{I}\left(\hat{\delta}_g < \hat{\alpha}_g \le 0\right).
-\]
-HPH probability is defined similarly with all inequalities reversed. As before, we construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities.
-
-We will use the term eBayes to refer to the approach defined in Sections \ref{s:model}--\ref{s:gene_heterosis} and add parenthetical labels ``normal'' and ``Laplace'' to specify whether we are assuming normal or Laplace distributions for half-parental differences and hybrid effects.
-
-\section{Simulation study based on a maize experiment}
-\label{s:simulation}
-
-To assess the efficacy of our method to identify genes demonstrating EH, we used a maize data set with parental varieties B73 and Mo17 and the hybrid variety (B73 $\times$ Mo17) \citep{paschold2012complementation} to determine realistic parameter values for a simulation study.
-Section \ref{s:maize} describes the maize dataset in detail.
-We compared our method to approaches using the R packages \edgeR{} , \baySeq{} \citep{hardcastle2010bayseq, hardcastle2012baySeq}, and \ShrinkBayes{} \citep{van2014shrinkbayes}.
-
-\subsection{Constructing simulated data}
-\label{s:sim_data}
-
-We used the methods described at the beginning of Section \ref{s:ebayes} to obtain normalization factors $\hat{\gamma}$ and gene-specific parameter estimates $\hat{\theta}_g$ for all genes using the \edgeR{} package \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor} applied to the maize data on 27,888 genes with average count at least one and at most two zero read counts for each variety across the four replicates.
-This analysis produced sample-specific normalization factors of $\hat{\gamma}=
-(0.074, -0.059, -0.074, -0.014, -0.014, -0.124,  0.093,  0.063,  0.021, -0.037,  0.049,  0.021)$.
-The  gene-specific parameter estimates were treated as the true parameter values for the simulation study so that our simulated datasets mimicked the existing structure among the gene-variety means of the original maize data.
-
-Using these parameters and normalization factors, we simulated data according to the negative binomial model in equation \eqref{e:data} independently for each gene.
-For each simulation, we analyzed a subset of 25,000 genes selected randomly from genes with simulated counts at least one on average and with at most two zeros for each variety across replications.
-We repeated this simulation process 10 times for each of 4, 8, and 16 replicates per variety, reusing normalization factors when necessary.
-
-For a particular gene, the truth was determined via the estimated values for $\alpha_g$ and $\delta_g$.
-Specifically, if $|\hat{\delta}_g| >|\hat{\alpha}_g|$ the gene was considered to have EH.
-For many heterosis genes, the value of $|\hat{\delta}_g|$ was only slightly larger than $|\hat{\alpha}_g|$.
-Thus there are many genes in these simulated data sets that are considered to have EH, but whose signal in the simulations will be extremely small.
-Conversely, there are many non-heterosis genes whose value of $|\hat{\delta}_g|$ was only slightly smaller than $|\hat{\alpha}_g|$, but whose simulated data will look similar to many EH genes.
-Therefore, we expect it will be difficult to accurately identify EH genes, but believe this level of difficulty is representative of real applications.
-
-\subsection{Alternative methods}
-\label{s:alternative}
-
-We compared our method to that of \cite{ji2014estimation}, which assumes normality in the response, by modeling the logarithm of the RNAseq count plus one adjusted by the normalization factor, i.e.
-$\log(Y_{gvi}+1)-\gamma_{vi}$.
-Use of the normalization factor here provides two advantages: 1) counts are properly adjusted for the thoroughness of the sequencing of the sample and 2) for genes with no count variation within variety (which actually occurs in our maize data set), use of the normalization factors allows the approach of \citeauthor{ji2014estimation} to still execute.
-
-In addition to the approach of \citeauthor{ji2014estimation}, we modified two existing \RNAseq{} approaches, \edgeR{} and \baySeq{}, for use in the heterosis context.
-For each method, we attempted to provide a measure of the strength of EH for each gene such that large values of this measure indicate support for EH.
-
-\edgeR{} can be used to test for differential expression between any two varieties based on the fit of a negative binomial log-linear model \citep{robinson2007moderated, robinson2010edgeR}.
-To construct a measure of EH, we computed the maximum likelihood estimates of the $\mu_{gv}$ parameters for all genes using \edgeR{}'s built-in Fisher scoring algorithm, and then used likelihood ratio tests to calculate two $p$-values for each gene: $p_{g1}$  for testing $H_{g01}:\mu_{g1}=\mu_{g3}$ and $p_{g2}$ for testing $H_{g02}:\mu_{g2}=\mu_{g3}$.
-Then, for each gene, we defined a new $p$-value denoted as $p_g$ and set to $p_g=1$ if the estimate of $\mu_{g3}$ falls between the estimates of $\mu_{g1}$ and $\mu_{g2}$ and $p_g=\max\{p_{g1},p_{g2}\}/2$ otherwise.
-For all relevant significance thresholds $\omega$ near 0, it can be shown that rejecting the null hypothesis of no EH for gene $g$ whenever $p_g\leq \omega$ results in a test that is asymptotically size $\omega$.
-We then use $1-p_g$ as a final measure positively associated with strength of evidence for EH.
-
-\baySeq{} allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct.
-In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of EH for each gene.
-We set this measure to zero if the maximum likelihood estimate, calculated using \edgeR{}, of $\mu_{g3}$ is between the maximum likelihood estimates of $\mu_{g1}$ and $\mu_{g2}$.
-Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the two hypotheses that allow for EH.
-
-\subsection{Results}
-
-For the methods in Sections \ref{s:method} and \ref{s:alternative}, we sorted genes according to the computed measure of the strength of evidence for EH.
-From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to distinguish genes with EH, as defined in Section \ref{s:sim_data},
-from those without EH. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
-\begin{figure}[htbp]
-\centerline{\includegraphics[width=\textwidth]{exampleROC0_1}}
-\caption{Example ROC curves for false positive rates below 0.1 for the approaches using \citeauthor{ji2014estimation}, \edgeR{}, and \baySeq{} described in Section \ref{s:alternative}, the \ShrinkBayes{} approach described in Section \ref{s:shrinkbayes} and the eBayes approach described in Section \ref{s:ebayes} using both normal and laplace distributions for the half-parental difference and hybrid effect.}
-\label{f:roc}
-\end{figure}
-The ROC curves indicate modest performance, e.g.
-for a false positive rate of 5\%, the best performing methods only achieved a true positive rate of just over 15\%.
-This is consistent with our expectation discussed in Section \ref{s:sim_data} due to the low signal-to-noise ratio in these simulated data.
-
-For this simulation, we can see that the approaches based on the model in Section \ref{s:model}, i.e.
-eBayes and \ShrinkBayes{}, provide the best true positive rate for a given false positive rate.
-Also, as expected, as the sample size increases, our ability to distinguish genes with EH from genes without improves.
-
-Figure \ref{f:auc} provides the area under the ROC curve (AUC) below a false positive rate of 0.1 across the 10 simulations for each of the 3 different sample sizes. 
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{auc-facet-TRUE}}
-\caption{Area under the ROC curves (AUC) below a false positive rate of 0.1 for 3 different replicates per variety for the approaches using \citeauthor{ji2014estimation}, \edgeR{}, and \baySeq{} as described in Section \ref{s:alternative}, the \ShrinkBayes{} approach described in Section \ref{s:shrinkbayes} and the eBayes approach described in Section \ref{s:ebayes} using both normal and laplace distributions for the half-parental difference and hybrid effect.
-Each line is a different simulation while the blue box indicates mean AUC (plus or minus one standard error).}
-\label{f:auc}
-\end{figure}
-Similar to the single ROC curve, the eBayes and \ShrinkBayes{} methods appear to outperform the other methods in terms of AUC.
-This improvement ranges from about a 20\% improvement over Ji et al.
-to about a 100\% improvement over \edgeR{} (which was not designed for heterosis testing).
-
-With 4 replicates per variety, there does not appear to be much of a difference between \ShrinkBayes{} and the eBayes approaches, but as the number of replicates increases, the eBayes approaches appear to improve relative to \ShrinkBayes{}.
-Two differences exist between the \ShrinkBayes{} and eBayes approaches that could explain the difference in AUC: 1) \ShrinkBayes{} utilizes a different empirical Bayes approach for estimating both the hyperparameters and the gene-specific parameters and 2) the measure of EH is slightly different due to INLA not providing a full posterior.
-
-There also appears to be a pattern of the eBayes (Laplace) systematically performing better than eBayes (normal).
-We suspect this is because the Laplace distributions are better approximations to the true underlying distribution for these parameters, and we discuss this in Section \ref{s:discussion}.
-
-\section{Searching for gene expression heterosis in the maize experiment}
-\label{s:maize}
-
-We used our method to analyze a maize data set \citep{paschold2012complementation} of \RNAseq{} gene expression in parental lines B73 and Mo17 and the hybrid genotype (B73$\times$Mo17) with a total of 39,656 genes.
-Each variety had four biological replicates measured with Illumina methodology and equipment.
-Reads were mapped to the whole reference genome using the short reads aligner, NOVOALIGN.
-For more specifics, please see \cite{paschold2012complementation}.
-
-We analyzed the data using all the methods compared in the previous section.
-The computation time on a desktop with two 4-core 3.6GHz Intel Xeon processor was 14 seconds for \edgeR{}, 1.3 minutes for \citeauthor{ji2014estimation}, 10 hours for the eBayes approaches, and 17 hours for \baySeq{}.
-In the eBayes approach, the vast majority of the time is spent on independent MCMC analysis for each gene.
-Thus we parallelized this step using {\tt doMC} \citep{revolution2014domc} and {\tt plyr} \citep{hadley2011split}.
-When parallelized across 5 cores, the eBayes approaches took about 2.5 hours.
-ShrinkBayes took 12 hours on a cluster node with two 8-core 2.6GHz Intel Haswell E5-2640 v3 processors where the code was also parallelized across 5 cores.
-
-For the eBayes (Laplace), we estimated the hyperparameters to be $\hat{\eta}_\phi = 3.44$, $\hat{\sigma}_\phi = 2.74$, $\hat{\eta}_\alpha = -0.05$, $\hat{\sigma}_\alpha=0.36$, $\hat{\eta}_\delta = 0.08$, $\hat{\sigma}_\delta=0.33$, $\hat{\eta}_\psi = -2.35$, and $\hat{\sigma}_\psi=0.60$.
-We then performed independent MCMC analysis for each gene conditional on these hyperparameters.
-As with the simulation study, we ran 4 chains simultaneously and doubled the number of MCMC iterations until each gene-specific parameter had an effective sample size above 1,000.
-
-Figure \ref{f:gene_specific_estimates} provides point estimates of the gene-specific parameters from the initial \edgeR{} estimation step and after the eBayes (Laplace) procedure described in Section \ref{s:ebayes}.
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{gene_specific_estimates.pdf}}
-\caption{Two-dimensional histogram of point estimates from \edgeR{} and posterior means from eBayes (Laplace) along with the $y=x$ diagonal.}
-\label{f:gene_specific_estimates}
-\end{figure}
-This figure shows shrinkage for large absolute estimates of $\alpha_g$ and $\delta_g$ from the \edgeR{} estimation toward $\hat{\eta}_\alpha$ and $\hat{\eta}_\delta$ which are both approximately zero.
-The figure also shows decreased values for the overdispersion parameter with larger decreases for high and low values of overdispersion.
-Finally, very little, if any, shrinkage is observed for $\phi_g$ estimates.
-
-With posteriors for all parameters, we can calculate empirical Bayes posterior probabilities for LPH and HPH.
-For each gene, the quantity of interest is the maximum of these two probabilities.
-For each gene with a high probability of either HPH or LPH, the magnitude of the effect is of interest, thus we calculate
-\begin{equation}
-\mbox{estimated effect size}_g \equiv \left\{ 
-\begin{array}{ll}
-\hat{\delta}_g - |\hat{\alpha}_g| & \mbox{if } \hat{\delta}_g > \phantom{-}|\alpha_g| \\
-\hat{\delta}_g + |\hat{\alpha}_g| & \mbox{if } \hat{\delta}_g < -|\alpha_g| \\
-0 & \mbox{otherwise},
-\end{array}.
-\right. 
-\label{e:effect_size}
-\end{equation}
-This estimated effect size is the difference between hybrid mean and the nearest parent with negative values indicating LPH and positive values indicating HPH.
-If the hybrid mean is estimated to be between the parents, this effect is defined to be zero.
-
-Figure \ref{f:volcano} provides a volcano plot, in this case a bivariate histogram, to visualize the maximum of the probabilities of LPH and HPH versus estimated effect size. 
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{volcano}}
-\caption{A bivariate histogram of the maximum of the LPH and HPH probabilities versus estimated effect size defined in equation \eqref{e:effect_size} for the B73 $\times$ Mo17 maize experiment using eBayes (Laplace).}
-\label{f:volcano}
-\end{figure}
-The figure shows a ridge at an effect size of zero for probabilities below 0.5.
-Above a probability of 0.5, we see a prototypical volcano pattern, with increased EH probability corresponding to larger estimated effect sizes and no estimated effect sizes near zero for genes with high EH probability.
-We also see asymmetry, with larger negative effect sizes than positive effect sizes due to genes with hybrid counts near zero and relatively high parental counts.
-Genes with high estimated posterior probabilities of EH and large estimated effect sizes are candidates for further investigation.
-
-\section{Discussion}
-\label{s:discussion}
-
-Geneticists speculate that gene expression EH is one possible explanation of hybrid vigor of traits, such as plant height or grain yield.
-Existing methods for identifying differential gene expression based on \RNAseq{} data are not directly applicable for detecting EH genes.
-\cite{ji2014estimation} introduced an empirical Bayes approach based on a hierarchical model for microarray data.
-We followed their approach, modified to allow for \RNAseq{} read counts as measures of transcript abundance.
-We developed an empirical Bayes approach based on obtaining estimates for hyperparameters followed by MCMC to estimate gene-specific parameters.
-The empirical Bayes posteriors can be used to estimate posterior probabilities of high and low parent heterosis.
-Through a simulation study, we demonstrated that this method outperformed alternative methods, and performed comparably well with a similar model in \ShrinkBayes{}, which estimates the posterior via INLA.
-We then demonstrated the use of the methodology on a maize experiment in which phenotypic heterosis is well known.
-
-Although our method appears to hold some advantage over existing methods, we believe our approach can be improved by refining the hierarchical model for the gene-specific parameter distribution.
-Figure \ref{f:scatterplot} shows marginal and bivariate histograms for eBayes (Laplace) posterior means for the gene-specific parameters.
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{estimates}}
-\caption{Marginal and bivariate histograms of posterior means for gene-specific parameters for the B73 $\times$ Mo17 maize experiment.}
-\label{f:scatterplot}
-\end{figure}
-These figures show markedly departures from marginal model assumptions, e.g.
-normality assumptions for $\phi_g$ and $\psi_g$, and independence assumptions for $(\phi_g,\psi_g)$ and $(\alpha_g,\delta_g)$.
-The plot of $\phi_g$ versus $\psi_g$ shows a pattern where the mean overdispersion decreases as the mean expression level increases.
-The plot of $\alpha_g$ versus $\delta_g$ shows a rotated V pattern where $\delta_g$ appears to be equal to $|\alpha_g|$.
-This V pattern is consistent with Mendel's Law of Dominance where the hybrid has mean expression equal to the parent with higher mean expression.
-
-In addition to improving the hierarchical distribution, we believe better estimates of the parameters of this distribution, i.e.
-the hyperparameters, will also improve detection of gene-expression heterosis.
-Our current method, based on moment matching of essentially independently estimated gene-specific parameters, provides consistent estimators as the number of replicates per variety increases.
-But typically the number of replicates per variety is quite small.
-In our data there are only four replicates per variety, and therefore asymptotic justifications are deficient.
-There are a variety of alternative estimation approaches to explore, e.g.
-expectation-maximization algorithms or fully Bayesian approaches.
-Notwithstanding these improvements, we believe our approach is a computationally efficient method that can immediately aid scientists who are interested in identifying candidate genes involved in a genetic mechanism of heterosis.
-
-This paper has focused on statistical methods for detecting EH in gene expression using RNA-seq data.
-EH at the transcript level is only one of multiple molecular genetic mechanisms that may play roles in establishing hybrid vigor.
-Complementation \citep{paschold2012complementation}, allele-specific expression \citep{bell2013rna,wei2013computational}, and other complex forms of genomic and epigenetic interaction \citep{chen2013genomic} are all plausible as mechanisms partially responsible for phenotypic heterosis.
-Further study of these phenomena using modern genomic technologies and appropriate statistical methods should enhance our understanding of hybrid vigor.
-
-\backmatter %  Please keep this command in your document in this position
-
-\if0\blind{
-\section*{Acknowledgements}
-
-The authors thank Andrew Lithio for help in implementing our model in \ShrinkBayes{}.
-Research reported in this publication was supported by National Institute of General Medical Sciences of the National Institutes of Health under award number R01GM109458.
-The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health.
-} \fi
-
-%  If your paper refers to supplementary web material, then you MUST
-%  include this section!!  See Instructions for Authors at the journal
-%  website http://www.biometrics.tibs.org
-
-\section*{Supplementary Materials}
-
-The supplementary materials include links to the data as well as code to reproduce the analysis in this manuscript.
-\bibliography{jarad}
-\bibliographystyle{biom}
-
-% \appendix
-% %  To get the journal style of heading for an appendix, mimic the following.
-% \section{}
-% \subsection{Material here}
-
-\label{lastpage}
-
 \end{document}

### 2015-12-11 slides/CU/JaradNiemi_CSU.Rnw
CSU talk

\documentclass[handout]{beamer}

\usecolortheme[RGB={0,0,144}]{structure}
\usetheme{AnnArbor}\usecolortheme{beaver}

\usepackage{verbatim,xmpmulti,color,multicol,multirow}
\setlength{\unitlength}{\textwidth}  % measure in textwidths
\usepackage[normalem]{ulem}

\graphicspath{{../../include/}{../figs/}{../KState/}}

%\usepackage{beamerthemesplit}
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{enumerate items}[default]
\setbeamertemplate{enumerate subitem}{\alph{enumii}.}
\setbeamertemplate{enumerate subsubitem}{\roman{enumiii}.}
\setkeys{Gin}{width=0.6\textwidth}

\providecommand{\e}{\varepsilon}
\providecommand{\nv}{{}^{-1}}
\providecommand{\ov}[1]{\overline{#1}}
\providecommand{\q}{$\quad$ \newline}
\providecommand{\rt}{\rightarrow}
\providecommand{\vc}[1]{\boldsymbol{#1}}
\providecommand{\wh}[1]{\widehat{#1}}

\title[Bayesian RNAseq heterosis ]{Bayesian analysis for heterosis detection in RNAseq data}
\author[Jarad Niemi]{Dr. Jarad Niemi \\ joint with Eric Mittman, Will Landau, and Dr. Dan Nettleton}
\institute[Iowa State]{Iowa State University}
\date{\today}

\begin{document}

%\section{Temp??} \begin{comment}

<<chunk_options, echo=FALSE, message=FALSE>>=
opts_chunk$set(fig.width=6, 
               fig.height=5, 
               out.width='.8\\linewidth', 
               fig.align='center', 
               size='tiny',
               echo=FALSE)
library(ggplot2)
@

\frame{\titlepage  
{\tiny
This research was supported by National Institute of General Medical Sciences (NIGMS) of the National Institutes of Health and the joint National Science Foundation / NIGMS Mathematical Biology Program under award number R01GM109458. The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health or the National Science Foundation.
}}

\begin{frame}
\frametitle{Outline}

\begin{itemize}
\item Data
  \begin{itemize}
  \item Phenotypic heterosis
  \item RNAseq
  \end{itemize}
\item Methodology
  \begin{itemize}
  \item Model
  \item Empirical Bayes
  \item Stan
  \end{itemize}
\item Results
  \begin{itemize}
  \item Simulation study vs {\tt edgeR}, {\tt baySeq}, Ji et. al., {\tt ShrinkBayes}
  \item Real data analysis
  \end{itemize}
\item Fully Bayesian analysis
  \begin{itemize}
  \item GPUs
  \item Minimizing data transfer
  \item Slice sampling
  \end{itemize}
\end{itemize}

\end{frame}

\section{Data}
\subsection{Phenotypic heterosis}

\begin{frame}
\frametitle{Phenotypic heterosis in maize}
\setkeys{Gin}{width=0.8\textwidth}
\begin{center}
\includegraphics{heterosis_corn}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Phenotypic heterosis in cattle}
\setkeys{Gin}{width=0.8\textwidth}

\vspace{-0.1in}

\begin{center}
\includegraphics{heterosis_cattle1}

\includegraphics{heterosis_cattle2}
\end{center}
\end{frame}

\subsection{RNAseq experiment}
\begin{frame}
\frametitle{RNAseq experiment}
\setkeys{Gin}{width=0.55\textwidth}

\begin{center}
\includegraphics{rnaseq_experiment}
\end{center}

\end{frame}

\begin{frame}
\frametitle{Phenotypic heterosis in RNAseq}
\setkeys{Gin}{width=0.8\textwidth}
\begin{center}
\includegraphics{heterosis_rnaseq}
\end{center}
\end{frame}

\section{Method}
\subsection{Model}

\begin{frame}
\frametitle{RNAseq model}

Let $Y_{gvi}$ be the count for 
\begin{itemize}
\item gene $g=1,\ldots,G$, 
\item variety $v=1,\ldots,V$, and 
\item replicate $i=1,\ldots,n_v$. 
\end{itemize}

\vspace{0.2in} \pause

We assume
\[
Y_{gvi} \stackrel{ind}{\sim} NB\left(e^{\mu_{gv}+c_{vi}},e^{\psi_g}\right) 
\]
\pause
where 
\begin{itemize}
\item $E[Y_{gvi}] = \zeta_{gvi} = e^{\mu_{gv}+c_{vi}}$, \pause 
\item $V[Y_{gvi}] = \zeta_{gvi}+e^{\psi_g}\zeta_{gvi}^2$, \pause and 
\item $c_{vi}$ controls sequencing depth (normalization factors).
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Reparameterization}
\setkeys{Gin}{width=\textwidth}

Convert the variety specific parameters for parent 1 ($\mu_{g1}$), parent 2 ($\mu_{g2}$), and the hybrid ($\mu_{g3}$) \pause to

\begin{itemize}[<+->]
\item Parental average: $\phi_g = (\mu_{g1}+\mu_{g2})/2$
\item Half-parental difference: $\alpha_g = (\mu_{g2}-\mu_{g1})/2$
\item Hybrid effect: $\delta_g = \mu_{g3}-\phi_g$
\end{itemize}

\vspace{0.2in} \pause

and thus 
\[ \begin{array}{rl}
\mu_{g1} &= \phi_g + \alpha_g \\
\mu_{g2} &= \phi_g - \alpha_g \\
\mu_{g3} &= \phi_g + \delta_g
\end{array}\]

\pause
\begin{center}
\includegraphics{pad}
\end{center}

\end{frame}

\begin{frame}
\frametitle{Gene expression heterosis}
\setkeys{Gin}{width=\textwidth}

\begin{center}
\includegraphics{pad}
\end{center}
\pause
We define \alert{gene expression heterosis}:

\[ \begin{array}{rll}
H_{g,HPH}: & \delta_g > \phantom{-}|\alpha_g|  & \mbox{high parent heterosis} \pause \\
H_{g,LPH}: & \delta_g < -|\alpha_g|  & \mbox{low parent heterosis} \pause \\
H_{g0}: & |\delta_g| \le  |\alpha_g| 
\end{array} \]

\pause

We are interested in identifying genes with high posterior probabilities for LPH and HPH, i.e. 
\begin{itemize}
\item $P(H_{g,HPH}|y) = P(\delta_g > \phantom{-}|\alpha_g|\, |y)$ and 
\item $P(H_{g,LPH}|y) = P(\delta_g < -|\alpha_g|\, |y)$. 
\end{itemize}

\end{frame}

\begin{frame}[fragile]
\frametitle{Hierarchical model}
\setkeys{Gin}{width=\textwidth}
\begin{itemize}
\item Data model:
\[ Y_{gvi} \stackrel{ind}{\sim} NB\left(e^{\mu_{gv}+c_{vi}},e^{\psi_g}\right) \]
\[ \mu_{g1} = \phi_g + \alpha_g, \mu_{g2} = \phi_g-\alpha_g, \mu_{g3} = \phi_g + \delta_g \]

\pause

\item Hierarchical structure:
\begin{columns}
\begin{column}{0.5\linewidth}
\[ \begin{array}{rl}
\phi_g &\stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2) \\
\alpha_g &\stackrel{ind}{\sim} La(\eta_\alpha, \sigma_\alpha) \\
\delta_g &\stackrel{ind}{\sim} La(\eta_\delta, \sigma_\delta) \\
\psi_g &\stackrel{ind}{\sim} N(\eta_\psi, \sigma_\psi^2) 
\end{array} \]
\end{column} \pause
\begin{column}{0.5\linewidth}
<<>>=
source("../../sandbox/Laplace.R")
x = seq(-4,4,by=0.1)
d = data.frame(rbind(data.frame(density="normal", x=x, y=dnorm(x)),
                     data.frame(density="Laplace", x=x, y=dlaplace(x,0,1/sqrt(2)))))
ggplot(d, aes(x=x,y=y,color=density)) + 
  geom_line(size=2) + 
  labs(y="f(x)")
@
\end{column}
\end{columns}
\end{itemize}
\end{frame}

% \begin{frame}[fragile]
% \frametitle{Normal vs Laplace}
% <<>>=
% source("../../sandbox/Laplace.R")
% x = seq(-4,4,by=0.1)
% d = data.frame(rbind(data.frame(density="normal", x=x, y=dnorm(x)),
%                      data.frame(density="Laplace", x=x, y=dlaplace(x,0,sqrt(2)))))
% ggplot(d, aes(x=x,y=y,color=density)) + 
%   geom_line() + 
%   labs(y="f(x)")
% @
% \end{frame}

\subsection{Empirical Bayes}
\begin{frame}
\frametitle{Empirical Bayes}

\begin{enumerate}
\item Estimate hyperparameter, $\pi=(\eta,\sigma,c)$:

  \begin{enumerate}
  \item Use {\tt edgeR} to obtain $\hat{c}$ and $\hat{\theta}_g = (\hat{\phi}_g,\hat{\alpha}_g,\hat{\delta}_g,\hat{\psi}_g)$. \pause
  \item Use moment matching, to obtain $\hat{\eta}$ and $\hat{\sigma}$.
  \end{enumerate}

\pause
\item Empirical Bayes posterior

\[ p(\theta|y, \hat{\pi}) = \prod_{g=1}^G p(\theta_g|y_g,\hat{\pi}) \]
where $y_g$ is all observations from gene $g$ \pause and 
{\footnotesize
\[ p(\theta_g|y_g,\hat{\pi}) \propto \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi};e^{\mu_{gv}+\hat{c}_{vi}}, e^{\psi_g}\right) \right] p\left(\theta_g;\hat{\eta},\hat{\sigma}\right) \]
}
\end{enumerate}
\end{frame}

\subsection{Stan}
\begin{frame}
\frametitle{Stan}
\setkeys{Gin}{width=1\textwidth}

\begin{center}
\includegraphics{stan}
\end{center}

For more information, go to \url{http://mc-stan.org/}.

\end{frame}

\begin{frame}[containsverbatim]
\frametitle{}

{\tiny
\begin{verbatim}
data {
  int<lower=1> S;
  int<lower=0> count[S]; 
  int<lower=1,upper=3> variety[S];
  real   eta_phi; real   eta_alpha; real   eta_delta; real   eta_psi; 
  real sigma_phi; real sigma_alpha; real sigma_delta; real sigma_psi;
  vector[S] c;                     // sample sequencing depth
}
transformed data {
  matrix[S,3] X;
  for (s in 1:S) {
    if (variety[s] == 1) { X[s,1] <-  1; X[s,2] <- -1; X[s,3] <-  0; }
    if (variety[s] == 2) { X[s,1] <-  1; X[s,2] <-  1; X[s,3] <-  0; }
    if (variety[s] == 3) { X[s,1] <-  1; X[s,2] <-  0; X[s,3] <-  1; }
  }
}
parameters {
  real phi; real alpha; real delta; real psi;          
}
transformed parameters {
  vector[3] pad;

  pad[1] <- phi;
  pad[2] <- alpha;
  pad[3] <- delta;
}
model {
  phi   ~ normal(            eta_phi,   sigma_phi);
  alpha ~ double_exponential(eta_alpha, sigma_alpha); // Laplace
  delta ~ double_exponential(eta_delta, sigma_delta); // Laplace
  psi   ~ normal(            eta_psi,   sigma_psi);

  count ~ neg_binomial_2_log(X*pad+c, 1/exp(psi));
}
\end{verbatim}
}

\end{frame}

\begin{frame}[fragile]
\frametitle{RStan}

<<echo=TRUE, eval=FALSE>>=
# set up data
# ...
# ...

# get hyperparameter estimates
hyperparameter = get_indep_est(counts.t,group,geneid.t)

# change data to long format
data_stan = w.to.l(counts.t,group)

# compile Stan model (this can take some time, but only needs to be done once)
sg_model = stan_model("sg_model.txt")

# use initial values to obtain hyperparameter estimates
# ...
# ...

# loop over all genes
library(doMC)
registerDoMC()

posterior = dlply(stan_data, .(gene), function(x) {
  dat = list(x, hyperparameter)
  sampling(sg_model, dat, chains=4, iter=2000)
}, .parallel=TRUE)
@

\end{frame}

\section{Results}
\subsection{Simulation study}
\begin{frame}
\frametitle{Simulation study}

Simulation study:
\begin{enumerate}[<+->]
\item Used {\tt edgeR} to estimate $\theta_g$ and $c$ (about 30\% have heterosis)
\item Simulated data from a negative binomial model using the estimated $\theta_g$ and $c$
\item Used each method below to rank genes according to plausibility of heterosis
\item Compared methods based on receiving operating characteristic (ROC) curves
\end{enumerate}

\vspace{0.2in} \pause

Compared our method to 
\begin{itemize}
\item {\tt edgeR}, 
\item {\tt baySeq}, 
\item Ji et. al. log(count+1)
\item {\tt ShrinkBayes} - INLA
\end{itemize}

\end{frame}

\subsection{ROC curves}
\begin{frame}
\frametitle{Example ROC curve}
\setkeys{Gin}{width=\textwidth}
\begin{center}
\includegraphics{exampleROC0_1}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Area under ROC curve}
\setkeys{Gin}{width=\textwidth}
\begin{center}
\includegraphics{auc-facet-TRUE}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Evaluation of empirical Bayes estimates}

In this simulation study, we have no true hyperparameters. \pause But we can evaluate how good our hyperparameters are in estimating means and standard deviations of the gene-specific parameters, \pause e.g. 
\[ 
\eta_\alpha^{\mbox{truth}} \equiv \frac{1}{G} \sum_{g=1}^G \alpha_g^{\mbox{edgeR}} \pause \qquad 
\sigma_\alpha^{2\mbox{truth}} \equiv \frac{1}{G} \sum_{g=1}^G \left( \alpha_g^{\mbox{edgeR}} - \eta_\alpha^{\mbox{truth}}\right)^{2}
\]

\pause

{\tiny
\input{rmse_with_truth}
}
\end{frame}

\subsection{Real data analysis}
\begin{frame}
\frametitle{Real data analysis}

We used our method to analyze a maize data set of RNA-seq gene expression in parental lines B73 and Mo17 and the hybrid genotype (B73$\times$Mo17) with a total of 39,656 genes.

\vspace{0.2in} \pause

Analysis proceeds by 
\begin{itemize}
\item Estimate hyperparameters $\hat{\pi}$
\item In parallel, gene-specific analyses for certain genes\pause, e.g. 
\[ 
\left| \delta_g^{edgeR}\right| > \left| \alpha_g^{edgeR}\right| 
\]
which was 30\% of the original genes.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Shrinkage}
\begin{center}
\includegraphics{gene_specific_estimates}
\end{center}
\end{frame}

% \begin{frame}
% \frametitle{Estimates}
% \begin{center}
% \includegraphics{estimates}
% \end{center}
% \end{frame}

\begin{frame}
\frametitle{Volcano plot for real data}
\setkeys{Gin}{width=0.6\textwidth}
\begin{center}
\includegraphics{volcano}
\end{center}
\end{frame}

\begin{frame}
\frametitle{eBayes heterosis for RNAseq}

Jarad Niemi , Eric Mittman, Will Landau, Dan Nettleton. (2015) Empirical Bayes Analysis of RNA-seq Data for Detection of Gene Expression Heterosis. \emph{Journal of Agricultural, Biological, and Environmental Statistics}. December 2015, Volume 20, Issue 4, pp 614-628

\end{frame}

\section{Fully Bayesian analysis}
\begin{frame}
\frametitle{Fully Bayesian analysis}

An MCMC implementation:
\begin{enumerate}
\item Independently, sample $\phi_g \sim (\phi_g|\ldots)$ for $g=1,\ldots,G$.
\item Independently, sample $\alpha_g \sim (\alpha_g|\ldots)$ for $g=1,\ldots,G$.
\item Independently, sample $\delta_g \sim (\delta_g|\ldots)$ for $g=1,\ldots,G$.
\item Independently, sample $\psi_g \sim (\psi_g|\ldots)$ for $g=1,\ldots,G$.

\vspace{0.1in} \pause

\item Sample hyperparameters:
\begin{itemize}
\item $\eta_\phi,\sigma_\phi^2$ depend on summary \emph{statistics} of $(\phi_1,\ldots,\phi_G)$.
\item $\eta_\alpha,\sigma_\alpha^2$ depend on summary \emph{statistics} of $(\alpha_1,\ldots,\alpha_G)$.
\item $\eta_\delta,\sigma_\delta^2\,$ depend on summary \emph{statistics} of $(\delta_1,\ldots,\delta_G)$.
\item $\eta_\psi,\sigma_\psi^2$ depend on summary \emph{statistics} of $(\psi_1,\ldots,\psi_G)$.
\end{itemize}
\end{enumerate}

\vspace{0.1in} \pause

The problem is that $G$ is large, i.e. ~40k.

\end{frame}

\begin{frame}
\frametitle{Parallel reduction}

A serial sum of $G$ numbers, e.g. $\sum_{g=1}^G \delta_g$, has a computational complexity of $O(G)$. \pause
A parallel reduction has computational complexity of $O\left(\frac{G}{S} \log_2(G) \right)$ where $S$ is ``speedup'', i.e. the number of parallel processes we have.

\vspace{0.2in} \pause

\begin{center}
\includegraphics{parallel_reduction}
\end{center}

\end{frame}

\begin{frame}
\frametitle{Computational tractibility via parallelism}

Amdahl's Law:
\[
\mbox{speedup} = \frac{1}{1-P+\frac{P}{S}}
\]
\pause
where
\begin{itemize}
\item $P$ is the proportion of the execution time that would benefit from speed up
\item $S$ is the speedup of that proportion
\end{itemize}

\vspace{0.2in} \pause

Parallel hardware
\begin{itemize}[<+->]
\item Multi-CPU/multi-core
\item Cluster
\item Graphical processing units (GPUs)
\begin{itemize}
\item Minimize memory transfer between host (CPU) and device (GPU)
\item Only have uniform and normal samplers
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Minimizing data transfer}

Since there are $4G+8\approx 160,000$ parameters, it is not feasible to return MCMC samples for all parameters. \pause We typically need samples to 
\begin{itemize}
\item Assess convergence
\item Perform inference
\end{itemize}

\vspace{0.2in} \pause

Instead, return relevant summary quantities:
\begin{itemize}
\item Convergence for $\psi$
  \begin{itemize}
  \item $\psi$ sum
  \item $\psi^2$ sum 
  \end{itemize}
\item Inference for $\psi$
  \begin{itemize}
  \item $\psi$ sum
  \item $\psi^2$ sum 
  \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Assessing convergence}
Assessing convergence for a quantity of interest $\psi$ using potential scale reduction factor using $M$ chains each with $N$ samples: \pause
\[ 
\hat{R} = 
%\widehat{\mbox{var}}^+(\psi|y) = 
\frac{N-1}{N}W + \frac{1}{N}B
\]
\pause
where
\[ \begin{array}{rl}
B &= \frac{N}{M-1}\sum_{m=1}^M \left(\overline{\psi}_{\cdot m} - \overline{\psi}_{\cdot \cdot}\right)^2 \\
W &= \frac{1}{M}\sum_{m=1}^M s_m^2
\end{array} \]
\pause
and 
\[ \begin{array}{rl}
\overline{\psi}_{\cdot m} & = \frac{1}{N} \sum_{n=1}^N \psi_{nm} \\
\overline{\psi}_{\cdot \cdot} &= \frac{1}{M} \sum_{m=1}^M \overline{\psi}_{\cdot m} \\
s_m^2 &= \frac{1}{N-1} \sum_{n=1}^N \left(\psi_{nm} -\overline{\psi}_{\cdot m} \right)^2 \pause = \frac{1}{N-1} \sum_{n=1}^N \psi_{nm}^2 - \frac{N}{N-1} \overline{\psi}_{\cdot m}^2
\end{array} \]
\end{frame}

\begin{frame}
\frametitle{Inference}

There are two main types of quantities of interest:
\begin{itemize}
\item Parameters, e.g. $\delta_g$ or $\theta_\delta$
\item Functions of parameters, e.g. $\mathrm{I}(\delta_g > |\alpha_g|)$
\end{itemize}
\pause
For both situations, we keep a running sum, e.g. 
\begin{itemize}
\item $\sum_{n=1}^N \delta_{g,n}$ 
\item $\sum_{n=1}^N \mathrm{I}(\delta_{g,n} > |\alpha_{g,n}|)$
\end{itemize}
\pause
Two caveats:
\begin{itemize}
\item Must pre-specify all quantities of interest.
\item Cannot calculate MCMC SE.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Slice sampling}

GPUs only have normal and uniform samplers. 
\pause 
The only full conditional that is normal or uniform is $\theta_\cdot$. 
\pause 
For all other parameters, we use slice sampling.  
\pause
The identity
\[
p(\psi|\ldots) = \int_0^{p(\psi|\ldots)} du
\]
\pause
implies the joint distribution
\[ 
p(\psi,u|\ldots) \propto \mathrm{I}(0<u<p(\psi|\ldots))
\]
\pause
which has the following full conditional distributions
\begin{itemize}
\item $U|\ldots \sim Unif(0,p(\psi|\ldots))$ \pause
\item $\psi|\ldots \sim Unif\{\psi:u<p(\psi|\ldots)\}$
\end{itemize}
\pause
The difficulty is in finding the set $\{\psi:u<p(\psi|\ldots)\}$.
\end{frame}

\begin{frame}
\frametitle{Stepping-out slice sampler}

To sample from $p(\psi|u,\ldots)$, let 
\begin{itemize}
\item $u^{(n)}$ be the current draw for the auxiliary variable $u$ \pause
\item $\psi^{(n-1)}$ be the current draw for $\psi$, \pause and we know $u^{(n)}<p(\psi^{(n-1)}|\ldots)$ \pause
\item $w$ be a tuning parameter that you choose
\end{itemize}
\pause
Perform the following
\begin{enumerate}
\item Randomly place an interval $(\psi_L(u^{(n)}),\psi_U(u^{(n)}))$ of length $w$ around the current value $\psi^{(i-1)}$. \pause
\item Step the endpoints of this interval out in increments of $w$ until $u^{(n)}>p(\psi_L(u^{(n)})|\ldots)$ and $u^{(n)}>p(\psi_B(u^{(n)})|\ldots)$. \pause
\item Sample $\psi^* \sim Unif(\psi_L(u^{(n)}),\psi_L(u^{(n)}))$. \pause
\item If $u^{(n)}<p(\psi^*|\ldots)$, then set $\psi^{(n)} = \psi^*$, \pause otherwise
  \begin{enumerate}
  \item set $\psi_L(u^{(n)}) = \psi^*$ if $\psi^* <  \psi^{(n-1)}$ or
  \item set $\psi_U(u^{(n)}) = \psi^*$ if $\psi^* > \psi^{(n-1)}$ \pause and
  \item return to Step 3.
  \end{enumerate}
\end{enumerate}
\end{frame}

\begin{frame}
\frametitle{}
<<stepping_out_examples>>=
lim = 5
target = function(x) dnorm(x, -2)/2 + dnorm(x, 2)/2
step_out_plot = function(theta, u, w, lim.=lim) {
  curve(target, -lim., lim., axes=F, frame=TRUE, xlab='', ylab='')
  points(theta, u, col='red', pch=19)
  abline(h=u, lty=3, col='gray')

  # Left endpoint
  L = theta - runif(1,0,.5)
  points(L,u)
  while(u<target(L)) points(L <- L-w, u)
  points(L,u, pch=8)

  # Right endpoint
  R = L + w
  points(R,u)
  while(u<target(R)) points(R <- R+w, u)
  points(R,u, pch=8)

  legend('topleft', c('Current', 'Endpoints'), pch=c(19,8), col=c('red','black'))
}

opar = par(mfrow=c(2,1), mar=rep(0.1,4))
w = 1
step_out_plot(2,.15,w)
step_out_plot(2,.05,w)
par(opar)
@
\end{frame}

\begin{frame}
\frametitle{Fully Bayesian analysis}

With $G=40,000$ genes with $V=3$ varieties and $n_v=4$ replicates per variety, \pause 
after 24 hours of running the GPU version of this fully Bayesian analysis, \pause 
we have apparent convergence for all parameters of interest except a handful of gene-specific parameters ($\hat{R}\approx 1.15$).

\vspace{0.2in} \pause

Comments:
\begin{itemize}[<+->]
\item With so many parameters, there is reasonably high probability that some will show lack of convergence.
\item If we remove these genes, all remaining parameters appear to converge. 
\item These genes will have minimal impact on inference for hyperparameters and all other genes. 
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Bottom line:}

\begin{itemize}[<+->]
\item A fully Bayesian analysis will incorporate more of our uncertainty.
\item It may justify our (or another) empirical Bayes approach. 
\end{itemize}

\vspace{0.5in} \pause

\begin{center}
{\Huge
Thank you!
}
\end{center}

\end{frame}

\end{document}

### 2016-04-14 slides/CU/JaradNiemi_CSU.tex
Add CU slides

--- 2015-12-11 slides/CU/JaradNiemi_CSU.Rnw
@@
-\documentclass[handout]{beamer}
+\documentclass[handout]{beamer}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 
@@
 \date{\today}
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 %\section{Temp??} \begin{comment}
-
-<<chunk_options, echo=FALSE, message=FALSE>>=
-opts_chunk$set(fig.width=6, 
-               fig.height=5, 
-               out.width='.8\\linewidth', 
-               fig.align='center', 
-               size='tiny',
-               echo=FALSE)
-library(ggplot2)
-@
 
@@
 \begin{column}{0.5\linewidth}
-<<>>=
-source("../../sandbox/Laplace.R")
-x = seq(-4,4,by=0.1)
-d = data.frame(rbind(data.frame(density="normal", x=x, y=dnorm(x)),
-                     data.frame(density="Laplace", x=x, y=dlaplace(x,0,1/sqrt(2)))))
-ggplot(d, aes(x=x,y=y,color=density)) + 
-  geom_line(size=2) + 
-  labs(y="f(x)")
-@
+\begin{knitrout}\tiny
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=.8\linewidth]{figure/unnamed-chunk-1-1}
+
+}
+
+\end{knitrout}
 \end{column}
@@
 
-<<echo=TRUE, eval=FALSE>>=
-# set up data
-# ...
-# ...
+\begin{knitrout}\tiny
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# set up data}
+\hlcom{# ...}
+\hlcom{# ...}
 
-# get hyperparameter estimates
-hyperparameter = get_indep_est(counts.t,group,geneid.t)
+\hlcom{# get hyperparameter estimates}
+\hlstd{hyperparameter} \hlkwb{=} \hlkwd{get_indep_est}\hlstd{(counts.t,group,geneid.t)}
 
-# change data to long format
-data_stan = w.to.l(counts.t,group)
+\hlcom{# change data to long format}
+\hlstd{data_stan} \hlkwb{=} \hlkwd{w.to.l}\hlstd{(counts.t,group)}
 
-# compile Stan model (this can take some time, but only needs to be done once)
-sg_model = stan_model("sg_model.txt")
+\hlcom{# compile Stan model (this can take some time, but only needs to be done once)}
+\hlstd{sg_model} \hlkwb{=} \hlkwd{stan_model}\hlstd{(}\hlstr{"sg_model.txt"}\hlstd{)}
 
-# use initial values to obtain hyperparameter estimates
-# ...
-# ...
+\hlcom{# use initial values to obtain hyperparameter estimates}
+\hlcom{# ...}
+\hlcom{# ...}
 
-# loop over all genes
-library(doMC)
-registerDoMC()
+\hlcom{# loop over all genes}
+\hlkwd{library}\hlstd{(doMC)}
+\hlkwd{registerDoMC}\hlstd{()}
 
-posterior = dlply(stan_data, .(gene), function(x) {
-  dat = list(x, hyperparameter)
-  sampling(sg_model, dat, chains=4, iter=2000)
-}, .parallel=TRUE)
-@
+\hlstd{posterior} \hlkwb{=} \hlkwd{dlply}\hlstd{(stan_data,} \hlkwd{.}\hlstd{(gene),} \hlkwa{function}\hlstd{(}\hlkwc{x}\hlstd{) \{}
+  \hlstd{dat} \hlkwb{=} \hlkwd{list}\hlstd{(x, hyperparameter)}
+  \hlkwd{sampling}\hlstd{(sg_model, dat,} \hlkwc{chains}\hlstd{=}\hlnum{4}\hlstd{,} \hlkwc{iter}\hlstd{=}\hlnum{2000}\hlstd{)}
+\hlstd{\},} \hlkwc{.parallel}\hlstd{=}\hlnum{TRUE}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
 
@@
 \frametitle{}
-<<stepping_out_examples>>=
-lim = 5
-target = function(x) dnorm(x, -2)/2 + dnorm(x, 2)/2
-step_out_plot = function(theta, u, w, lim.=lim) {
-  curve(target, -lim., lim., axes=F, frame=TRUE, xlab='', ylab='')
-  points(theta, u, col='red', pch=19)
-  abline(h=u, lty=3, col='gray')
+\begin{knitrout}\tiny
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
 
-  # Left endpoint
-  L = theta - runif(1,0,.5)
-  points(L,u)
-  while(u<target(L)) points(L <- L-w, u)
-  points(L,u, pch=8)
+{\centering \includegraphics[width=.8\linewidth]{figure/stepping_out_examples-1}
 
-  # Right endpoint
-  R = L + w
-  points(R,u)
-  while(u<target(R)) points(R <- R+w, u)
-  points(R,u, pch=8)
-
-  legend('topleft', c('Current', 'Endpoints'), pch=c(19,8), col=c('red','black'))
 }
 
-opar = par(mfrow=c(2,1), mar=rep(0.1,4))
-w = 1
-step_out_plot(2,.15,w)
-step_out_plot(2,.05,w)
-par(opar)
-@
+\end{knitrout}
 \end{frame}

### 2016-04-14 shiny/server.R
Remove shiny app that was never completed

@@ -1,15 +1,0 @@
-library(shiny)
-library(plyr)
-library(ggplot2)
-
-# d = read.table("rna_seq.txt", header=TRUE)
-
-shinyServer(function(input,output) {
-  
-#   d_sub = reactive({
-#     d[d$GeneID == input$geneid,]
-#   })
-#   
-  output$data = renderDataTable({ data.frame(a=1) })
-  
-})

### 2016-04-14 shiny/ui.R

@@ -1,22 +1,0 @@
-library(shiny)
-
-d = read.csv("gene_names.csv", stringsAsFactors=FALSE)
-
-
-shinyUI(fluidPage(
-
-  titlePanel("Heterosis simulations"),
-  
-  sidebarLayout(
-  
-    sidebarPanel(
-      selectizeInput('geneid', 'Gene ID', 1:40000)
-    ),
-    
-    mainPanel(
-      tabsetPanel(
-        tabPanel("Data", dataTableOutput('data') )
-      )
-    )
-  )
-))

### 2016-04-14 sandbox/Laplace.R
Remove sandbox

@@ -1,50 +1,0 @@
-#' Laplace (double exponential distribution)
-#' 
-#' Density, distribution function, quantile function, and random generation for the Laplace 
-#' (double exponential) distribution with mean equal to mean and scale equal to sd.
-#' 
-#' @usage dlaplace(x, mean = 0, scale = 1, rate = 1/scale, log = FALSE)
-#' @usage plaplace(q, mean = 0, scale = 1, rate = 1/scale, lower.tail = TRUE, log.p = FALSE)
-#' @usage qlaplace(p, mean = 0, scale = 1, rate = 1/scale, lower.tail = TRUE, log.p = FALSE)
-#' @usage rlaplace(n, mean = 0, scale = 1, rate = 1/scale)
-#' 
-#' @param x vector of quantiles.
-#' @param n number of observations. If length(x) > 1, the length is taken to be the number required.
-#' @param mean vector of means.
-#' @param scale vector of scales.
-#' 
-#' @details The Laplace distribution has density ...
-#' 
-#' @return dlaplace gives the density, plaplace gives the distribution function, 
-#'   qlaplace gives the quantile function, and rlaplace generates random deviates. 
-#'   
-#'   The length of the result is determined by n for rlaplace, and is the maximum 
-#'   of the lengths of the numerical arguments for the other functions.
-#'   
-#'   The numerical arguments other than n are recycled to the length of the result. 
-#'   Only the first elements of the logical arguments are used.
-#'   
-#' @examples dlaplace(0)
-#' 
-dlaplace = function(x, mean = 0, scale = 1, rate = 1/scale, log = FALSE) {
-  if (log) dexp(x=abs(x-mean), rate=rate, log=TRUE)-log(2)
-  dexp(x=abs(x-mean), rate=rate, log=FALSE)/2
-}
-
-rlaplace = function(n, mean = 0, scale = 1, rate = 1/scale) {
-  # sample(c(-1,1), n, replace=TRUE) * rexp(n, rate=rate) + mean
-  u = runif(n=n)-.5
-  mean - sign(x=u)*log(1-2*abs(x=u)) / rate
-}
-
-plaplace = function(q, mean = 0, scale = 1, rate = 1/scale, lower.tail = TRUE, log.p = FALSE) {
-  x = (q-mean)*rate
-  cdf = ifelse(x<0, exp(x)/2, 1-exp(-x)/2)
[…]

### 2016-04-14 sandbox/RUV.R

@@ -1,32 +1,0 @@
-d = read.table("../shiny/rna_seq.txt", header=TRUE)
-
-y = d[,grep('total',names(d))]
-
-filter = apply(y, 1, function(x) length(x[x>5])>=2)
-
-filtered = y[filter,]
-rownames(filtered) = d$GeneID[filter]
-spikes = rownames(filtered)[grep("AC", rownames(filtered))]
-
-
-
-x <- as.factor(rep(c("B","M","BM","MB"),4))
-set <- newSeqExpressionSet(as.matrix(filtered),
-                           phenoData = data.frame(x, row.names=colnames(filtered)))
-set
-
-library(RColorBrewer)
-colors <- brewer.pal(3, "Set2")
-plotRLE(set, outline=FALSE, ylim=c(-4, 4), col=colors[x])
-plotPCA(set, col=colors[x], cex=1.2)
-
-
-set <- betweenLaneNormalization(set, which="upper")
-plotRLE(set, outline=FALSE, ylim=c(-4, 4), col=colors[x])
-plotPCA(set, col=colors[x], cex=1.2)
-
-
-set1 <- RUVg(set, spikes, k=1)
-pData(set1)
-plotRLE(set1, outline=FALSE, ylim=c(-4, 4), col=colors[x])
-plotPCA(set1, col=colors[x], cex=1.2)

### 2016-04-14 sandbox/edgeR_est.R

@@ -1,31 +1,0 @@
-
-edgeR_est <- function(counts, group, geneid){
-
-  require(edgeR)
-  row.names(counts) <- geneid
-  countsDGE         <- DGEList(counts)
-  countsDGE         <- calcNormFactors(countsDGE)
-  design            <- model.matrix(~ 0 + as.factor(group))
-  countsDGE         <- estimateGLMCommonDisp(countsDGE, design)
-  countsDGE         <- estimateGLMTagwiseDisp(countsDGE, design)
-  fit               <- glmFit(countsDGE, design)
-  offset            = fit$offset                                 # offset = log(lib.size * norm.factors)
-  mean.off          = mean(offset[1,])
-  offset.new        = offset - mean.off                          # offset is relative size in our model
-  mu                = fit$coefficients + mean.off
-  lin.comb          = matrix(c( .5, .5, 0,
-                               -.5, .5, 0,
-                               -.5,-.5, 1),3,3,byrow=T )
-  B   = lin.comb %*% t(mu)
-  psi = log(fit$dispersion)
-
-
-  return(list(parameters = data.frame(phi = B[1,], alpha = B[2,], delta = B[3,], psi = psi),
-              hyperparameters = data.frame(parameter = c("phi","alpha","delta","psi"),
-                                           location  = c(rowMeans(B), mean(psi)),
-                                           scale     = c(sd(B[1,]), sd(B[2,])/sqrt(2), sd(B[3,])/sqrt(2), sd(psi))),
-              c = offset.new[1,],
-              sigma_c = sd(offset.new[1,])
-              ))
-}
-

### 2016-04-14 sandbox/em.R

@@ -1,114 +1,0 @@
-library(rstan)
-library(plyr)
-
-# Compile models
-models = list()
-models$de_poisson  = stan_model('stan_models/de_poisson.stan')
-models$de_negbin   = stan_model('stan_models/de_negbin.stan')
-models$het_poisson = stan_model('stan_models/het_poisson.stan')
-models$het_negbin  = stan_model('stan_models/het_negbin.stan')
-
-
-# calculates hyper parameters for normal distribution
-normal_hyper = function(samples, learn_mean) {
-  eta = ifelse(learn_mean, mean(samples), 0)
-  sigma = sqrt(mean((samples-eta)^2))
-  return(list(eta=eta,sigma=sigma))
-}
-
-# calculates hyper parameters for laplace distribution
-laplace_hyper = function(samples, learn_mean) {
-  eta = ifelse(learn_mean, median(samples), 0)
-  sigma = mean(abs(samples-eta))
-  return(list(eta=eta,sigma=sigma))
-}
-
-
-
-
-# MCEM algorithm for differential expression using Poisson model
-em_de_poisson = function(dat, niter, learn_alpha_mean = FALSE, dist = list(alpha = 'normal'),
-                         .parallel=FALSE, verbose=1) {
-  parameter_names = c("phi","alpha")
-  
-  # Build X matrix
-  one_gene = dat[dat$gene==1,]
-  X = cbind(1, 2*(one_gene$variety-1.5))
-  
-  # Initial values
-  initial = ddply(dat, .(gene), function(x) {
-    m = glm(count~0+X, data=x, family="poisson")
-    data.frame(phi = coef(m)[1], alpha = coef(m)[2])
-  })
-  hyper = list()
-  hyper$iteration = 0
-  hyper$eta_phi = mean(initial$phi)
-  hyper$sigma_phi = sd(initial$phi)
-  if (dist$alpha == 'normal') {
-    hyper$eta_alpha = ifelse(learn_alpha_mean, mean(initial$alpha), 0)
-    hyper$sigma_alpha = sqrt(mean((initial$alpha-hyper$eta_alpha)^2))
-  } else if (dist$alpha == 'laplace') {
-    hyper$eta_alpha = ifelse(learn_alpha_mean, median(initial$alpha), 0)
-    hyper$sigma_alpha = mean(abs(initial$alpha-hyper$eta_alpha))
-  } else {
-    stop(paste('Alpha distribution',dist$alpha,'not implemented.'))
-  }
-  
-  # Save structure
[…]

### 2016-04-14 sandbox/em_full.R

@@ -1,135 +1,0 @@
-library(rstan)
-library(reshape2)
-library(doMC)
-#library(dplyr)
-library(ggplot2)
-library(plyr)
-
-source("Laplace.R")
-source("sim_heterosis_data.R")
-
-registerDoMC(cores=8)
-
-m = stan_model("sg_model_pad_dexp.txt")
-
-set.seed(20150408)
-G = 100
-d = sim_heterosis_data(G=G)
-
-# Construct design
-X = cbind(1, rep(c(-1,1,0), each=4), rep(c(0,0,1), each=4))
-#X = cbind(1, rep(c(0,1,0), each=4), rep(c(0,0,1), each=4))
-
-# Use edgeR
-source("edgeR_est.R")
-d_w = dcast(d$data[,c("gene","parent","sample","y")], gene~parent+sample, value.var='y')
-edgeR = edgeR_est(d_w[,-1], substr(names(d_w)[-1],1,1), rownames(d_w))
-
-# Set initial values for hyperparameters
-#gene_names = c("phi","alpha","delta","psi")
-gene_names = c("phi","alpha","delta")
-hyper = with(edgeR$hyperparameters[1:3,], as.list(c(location, scale)))
-names(hyper) = c(paste("eta_",   gene_names, sep=''),
-                 paste("sigma_", gene_names, sep=''))
-hyper$eta_alpha = 0 
-hyper$eta_delta = 0
-n_hyper = length(hyper)
-
-truth = data.frame(variable=names(hyper), value=with(d$hyperparameter, c(location,scale)))
-
-# 
-n_iter = 10
-hyper_keep = matrix(NA, nrow=n_iter, ncol=n_hyper)
-
-if (!interactive()) {
-  cat("MCEM with", G, "genes.\n", file="em_full.Ro")
-  cat("iteration", names(hyper), "iteration_time\n", sep=", ", file="em_full.Ro", add=TRUE)
-}
-
-
-
-# Run EM
-for (i in 1:n_iter) {
-  start_time = proc.time()
-  
-  # MCMC for gene specific parameters
-  mcmc = dlply(d$data, .(gene), function(x) {
-    sampling(m, c(list(S=nrow(x), 
-                       X = X, 
-                       count=x$y, 
-                       c=rep(0, nrow(x))), 
-                  hyper), 
-             c(gene_names), 
-             iter=10^2+(i+2)^2, 
-             chains=4)
-  }, .parallel=TRUE)
[…]

### 2016-04-14 sandbox/find_map_estimates.R

@@ -1,80 +1,0 @@
-library(Rcpp)
-library(reshape2)
-library(plyr)
-library(ggplot2)
-library(MASS)
-
-source("sim_heterosis_data.R")
-source("Laplace.R")
-
-sourceCpp("monte_carlo_integral.cpp", verbose=TRUE)
-
-# Get data
-set.seed(3)
-all = sim_heterosis_data(G=10)
-R = rbind(c(.5, .5, 0),
-          c(.5,-.5, 0),
-          c(.5, .5,-1))
-X = with(all$data[all$data$gene==1,],
-         cbind(parent==1, parent==2, parent==3)) %*% solve(R)
-
-# Observed Fisher information
-
-data_summary = ddply(mutate(all$data, gene=as.factor(gene), parent=as.factor(parent)),
-          .(gene), 
-          function(x) {
-            m = glm.nb(formula = y~0+X, data=x, control=glm.control(maxit=1000))
-            data.frame(
-              parameter = c('phi', 'alpha', 'delta', 'psi'),
-              estimate = c(coef(m), -log(summary(m)$theta)),
-              sd = sqrt(c(diag(summary(m)$cov.scaled),NA)))
-          }, .inform=TRUE)
-
-# Reshape data (put this in a function?)
-count = dcast(all$data[,c("gene","sample","y")], gene ~ sample, value.var='y')
-variety = all$data$parent[1:ncol(count)-1] # hack
-
-
-hyperparameters = melt(all$hyperparameters)
-pi = hyperparameters$value
-# sourceCpp("monte_carlo_integral.cpp", verbose=TRUE)
-
-
-r = ddply(expand.grid(n_sims = 10^seq(1,5), n_genes=c(1)), 
-          .(n_sims, n_genes), 
-          function(x) {
-            rdply(.n=100,
-                  data.frame(log_integral = monte_carlo_integral(t(as.matrix(count[1:x$n_genes,-1])), 
-                                                             variety-1, 
-                                                             pi[1:4], 
-                                                             exp(pi[5:8]), 
-                                                             with(data_summary, estimate[parameter=='phi']),
-                                                             with(data_summary,       sd[parameter=='phi']),
-                                                             x$n_sims)))
-})
-
-r$integral = exp(r$log_integral-max(r$log_integral))
-r$integral = exp(r$log_integral)
[…]

### 2016-04-14 sandbox/monte_carlo_integral.cpp

@@ -1,106 +1,0 @@
-#include <Rcpp.h>
-using namespace Rcpp;
-
-// Below is a simple example of exporting a C++ function to R. You can
-// source this function into an R session using the Rcpp::sourceCpp 
-// function (or via the Source button on the editor toolbar)
-
-// For more on using Rcpp click the Help button on the editor toolbar
-
-void rpadp(double *padp, NumericVector mean, NumericVector scale) {
-  for (int i=0; i<4; i++) 
-    padp[i] = rnorm(1, mean[i], scale[i])[0];
-}
-
-double dpadp(double *padp, NumericVector mean, NumericVector scale) {
-  double sum=0;
-  
-  for (int i=0; i<4; i++)
-    sum += R::dnorm(padp[i], mean[i], scale[i], 1);
-
-  return sum;
-}
-
-void calculate_exp_mu(NumericVector exp_mu, double padp[], IntegerVector variety) {
-  double tmp_mu[3];
-  tmp_mu[0] = padp[0]+padp[1]; // make sure these agree with definition
-  tmp_mu[1] = padp[0]-padp[1];
-  tmp_mu[2] = padp[0]+padp[2];
-  
-  for (int i=0; i < exp_mu.length(); i++)
-    exp_mu[i] = exp(tmp_mu[variety[i]]);
-}
-
-
-
-double logsumexp(double x[], int n) {
-  double max_val = x[0], sum = 0.0;
-  int i;
-  
-  for (i = 1; i< n; i++)
-    if (x[i] > max_val) 
-      max_val = x[i];
-      
-  for (i=0; i<n; i++)
-    sum += exp(x[i] - max_val);
-    
-  return log(sum) + max_val;
-}
-
-// [[Rcpp::export]]
-double monte_carlo_integral(IntegerMatrix count, 
-                            IntegerVector variety, 
-                            NumericVector mean,
-                            NumericVector scale,
-                            NumericVector data_phi_mean,
-                            NumericVector data_phi_scale,
-                            int n_sims) {
-  // count is the observed count (sample x gene, column major for quick access)
-  // variety is the variety identifier for the columns
-  // mean and scale contain the hyperparameters for the phi-alpha-delta-psi distributions
-  // n_sims number of Monte Carlo simulations to use
-  int G = count.ncol(); // number of genes
-  int n = count.nrow(); // number of samples
-  
-  // argument error handling
[…]

### 2016-04-14 sandbox/run_em.R

@@ -1,18 +1,0 @@
-library(doMC)
-library(reshape2)
-library(ggplot2)
-
-registerDoMC(4)
-
-source("Laplace.R")
-source("sim_heterosis_data.R")
-source("em.R")
-
-truth = sim_heterosis_data(10)
-d = subset(truth$data, variety != 3)
-
-tmp = em_de_poisson(d, 10, .parallel=TRUE)
-
-ggplot(melt(tmp, id.var='iteration', variable.name='parameter'), aes(x=iteration, y=value)) + 
-  geom_point() +
-  facet_wrap(~parameter, scales='free')

### 2016-04-14 sandbox/sg_negbinomial.R

@@ -1,151 +1,0 @@
-library(rstan)
-library(reshape2)
-library(doMC)
-#library(dplyr)
-library(ggplot2)
-library(plyr)
-
-source("Laplace.R")
-source("sim_heterosis_data.R")
-
-registerDoMC(cores=6)
-
-m = stan_model("sg_negbinomial.stan")
-
-# Simulate data
-set.seed(20150409)
-G = 100
-d = sim_heterosis_data(G=G)
-
-
-# Construct design
-X = cbind(1, rep(c(-1,1,0), each=4), rep(c(0,0,1), each=4))
-#X = cbind(1, rep(c(0,1,0), each=4), rep(c(0,0,1), each=4))
-
-# Use edgeR
-source("edgeR_est.R")
-d_w = dcast(d$data[,c("gene","variety","sample","count")], gene~variety+sample, value.var='count')
-edgeR = edgeR_est(d_w[,-1], substr(names(d_w)[-1],1,1), rownames(d_w))
-
-# Set initial values for hyperparameters
-#gene_names = c("phi","alpha","delta","psi")
-gene_names = c("phi","alpha","delta","psi")
-hyper = with(edgeR$hyperparameters, as.list(c(location, scale)))
-names(hyper) = c(paste("eta_",   gene_names, sep=''),
-                 paste("sigma_", gene_names, sep=''))
-hyper$eta_alpha = 0 
-hyper$eta_delta = 0
-n_hyper = length(hyper)
-
-truth = data.frame(variable=names(hyper), value=with(d$hyperparameter, c(location,scale)))
-
-# 
-n_iter = 20
-hyper_keep = matrix(NA, nrow=n_iter, ncol=n_hyper)
-hyper_keep[1,] = unlist(hyper)
-
-if (!interactive()) {
-  cat("MCEM with", G, "genes.\n", file="em_full.Ro")
-  cat("iteration", names(hyper), "iteration_time\n", sep=", ", file="em_full.Ro", add=TRUE)
-}
-
-# Initial inits
-inits = alply(edgeR$parameters, 1, function(x) {
-  list(as.list(x), as.list(x), as.list(x), as.list(x))
-})
-
-# Run EM
-for (i in 2:n_iter) {
-  start_time = proc.time()
-  
-  # MCMC for gene specific parameters
-  mcmc = dlply(d$data, .(gene), function(x) {
-    sampling(m, c(list(S=nrow(x), 
-                       X = X, 
-                       count=x$count, 
-                       c=rep(0, nrow(x))), 
[…]

### 2016-04-14 sandbox/sg_poisson.R

@@ -1,155 +1,0 @@
-library(rstan)
-library(reshape2)
-library(doMC)
-#library(dplyr)
-library(ggplot2)
-library(plyr)
-
-source("Laplace.R")
-
-registerDoMC(cores=8)
-
-m = stan_model("sg_poisson.stan")
-m = stan_model("sg_negbinomial.stan")
-
-# Simulate data
-set.seed(20150409)
-G = 1000
-
-d = list()
-d$hyperparameters = data.frame(parameter = c("phi","alpha","delta"),
-                               location  = c(4.6,0,0),
-                               scale     = c(1.8,.1,.1))
-d$parameters = with(d$hyperparameters, 
-                    data.frame(gene = 1:G, 
-                               phi   = rnorm(   G, location[parameter=='phi'  ], scale[parameter=='phi'  ]),
-                               alpha = rlaplace(G, location[parameter=='alpha'], scale[parameter=='alpha']),
-                               delta = rlaplace(G, location[parameter=='delta'], scale[parameter=='delta'])))
-d$data = ddply(d$parameters, .(gene), function(x) {
-  mu = with(x, phi + c(alpha,-alpha,delta))
-  mutate(data.frame(variety = rep(1:3, each=4)),
-         sample  = 1:length(variety),
-         count   = rpois(length(variety), exp(mu[variety])))
-})
-
-
-
-# Construct design
-X = cbind(1, rep(c(-1,1,0), each=4), rep(c(0,0,1), each=4))
-#X = cbind(1, rep(c(0,1,0), each=4), rep(c(0,0,1), each=4))
-
-# Use edgeR
-source("edgeR_est.R")
-d_w = dcast(d$data[,c("gene","variety","sample","count")], gene~variety+sample, value.var='count')
-edgeR = edgeR_est(d_w[,-1], substr(names(d_w)[-1],1,1), rownames(d_w))
-
-# Set initial values for hyperparameters
-#gene_names = c("phi","alpha","delta","psi")
-gene_names = c("phi","alpha","delta")
-hyper = with(edgeR$hyperparameters[1:3,], as.list(c(location, scale)))
-names(hyper) = c(paste("eta_",   gene_names, sep=''),
-                 paste("sigma_", gene_names, sep=''))
-hyper$eta_alpha = 0 
-hyper$eta_delta = 0
[…]

### 2016-04-14 sandbox/sim_heterosis_data.R

@@ -1,49 +1,0 @@
-#' Simulate heterosis data
-#' 
-#' Simulates heterosis data from either hyperparameters (and thus simulating gene-specific parameters)
-#' or gene-specific parameters.
-#' 
-#' @param G integer. The number of genes to simulate.
-#' @param nv integer (or integer vector). indicating the number of samples per genotype
-#' @param parameters. a list containing either hyperparameters or (gene-specific) parameters
-#' @param distributions. a list containing the distribution to use for the gene-specific parameters
-#' 
-#' @return list containing the hyperparameters, gene-specific parameters, and data
-#' 
-
-sim_heterosis_data = function(G=10, nv=4, parameters=NULL, hyperparameters=NULL, distributions=NULL, verbose=0) {
-  require(plyr)
-  if (length(nv==1)) nv = rep(nv,3)
-  
-  if (is.null(hyperparameters) & is.null(parameters)) 
-    if (verbose) message('Using default hyperparameters.')
-    hyperparameters = data.frame(parameter      = c("phi","alpha","delta","psi"),
-                                 location = c(4.6,0,0,-2),
-                                 scale    = c(1.8,.1,.01,.1))
-  
-  # Simulate gene-specific parameters
-  if (is.null(distributions) & is.null(parameters)) {
-    if (verbose) message('Simulating gene-specific parameters.')
-    parameters = rdply(G, {
-      with(hyperparameters, 
-           data.frame(phi   = rnorm   (1, location[parameter == "phi"  ], scale[parameter == "phi"  ]),
-                      alpha = rlaplace(1, location[parameter == "alpha"], scale[parameter == "alpha"]),
-                      delta = rlaplace(1, location[parameter == "delta"], scale[parameter == "delta"]),
-                      psi   = rnorm   (1, location[parameter == "psi"  ], scale[parameter == "psi"  ])))
-    }, .id="gene")
-  }
-  
-  if (verbose) message('Simulating data.')
-  data = ddply(data.frame(gene=1:G), .(gene), function(x) {
-    g = as.numeric(x$gene)
-    mutate(data.frame(variety = rep(1:3, times = nv)),
-           sample = 1:sum(nv),
-           eta = with(parameters, c(phi[g]+alpha[g], phi[g]-alpha[g], phi[g]+delta[g]))[variety],
-           count = rnbinom(length(eta), size = 1/exp(parameters$psi[g]), mu = exp(eta)))
-  }, .inform=T)
-
-  
[…]

### 2016-04-14 sandbox/simpler.R

@@ -1,48 +1,0 @@
-log_sum_exp = function(x) {
-  mv = max(x)
-  log(sum(exp(x-mv)))+mv
-}
-
-n = 12
-J = 10
-phi = rnorm(J,0,1)
-d = data.frame(group = rep(1:J, each=n)) 
-d$y = rnbinom(n*J, size=1, mu=exp(mu[d$group]))
-           
-
-
-integral = Vectorize(function(mu_mean, n_sims=10^5) {
-  daply(d, .(group), function(x) 
-    log_sum_exp(llike(rnorm(n_sims,mu_mean)))-log(n_sims),
-        n_sims=n_sims, mu_mean=mu_mean)
-  
-})
-
-d = ddply(expand.grid(mu_mean = seq(-2,2,by=.5), n_sims=10^(1:5)),
-          .(mu_mean, n_sims), function(x)
-            data.frame(i=integral(x$mu_mean, x$n_sims)))
-ggplot(d, aes(x=mu_mean, y=i, color=factor(n_sims))) + geom_point() + facet_wrap(~n_sims)
-
-
-ddply(d, .(n_sims), function(x) {
-  m = lm(i~mu_mean+I(mu_mean^2), x)
-  data.frame(opt = -coef(m)[2]/2/coef(m)[3])
-})
-
-
-# Monte Carlo
-d = expand.grid(n_sims=10^(1:5), i=1:5)
-r = ddply(d, .(n_sims,i), function(x) {
-  data.frame(y = log_sum_exp(llike(rnorm(n_sims)))-log(x$n_sims))
-})
-
-ggplot(r, aes(x=n_sims, y=y)) + 
-  geom_point() + 
-  geom_hline(yintercept=log(i$value), color='red') +
-  scale_x_log10()
-
-
-ggplot(r, aes(x=n_sims, y=exp(y))) + 
-  geom_point() + 
-  geom_hline(yintercept=i$value, color='red') +
-  scale_x_log10()
